---
title: "Stage 2: Lipid Annotation"
subtitle: "SRM 1950 Database RT adjustment, Multi-evidence annotation, normalization and quality control"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "my_study"
format:
  html:
    output-file: "Annotation_neg.html"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

# Overview

::: {.callout-note}
## Usage
Set `STUDY_ID` and `POLARITY` in the **configuration** section to match the
values used in `Preprocessing_neg.qmd`. The preprocessed objects must
already exist in the `objects/` folder.
:::

This is the second stage of the Untargeted LC-MS-based lipidomics workflow. It handles
**SRM 1950 Database RT adjustment, Multi-evidence lipid annotation, normalization and quality control (QC)**. It uses the preprocessed objects generated by 
`Preprocessing_pos.qmd`.

## Required Input Files

1. **Preprocessed objects** from `Preprocessing_neg.qmd` in `objects/`
2. **SRM 1950-derived lipid database** (`LipidDatabase_R.xlsx`) in the parent study folder
3. **Lipid Reference Set** (`neg_lipid_reference_set.xlsx`) in this folder

---

# Configuration

Set the analysis parameters here. `STUDY_ID` and `POLARITY` must match what was
used in `Preprocessing_neg.qmd`.

```{r config}
#| label: configuration

# ============================================================================
# USER CONFIGURATION - MODIFY THIS SECTION
# ============================================================================

# Project root: path to the repository root (for shared R/ helper functions)
# Use "../../.." from applications/<study>/<polarity>/
PROJECT_ROOT <- "../.."

# Study identifier: must match the value used in Preprocessing.qmd
STUDY_ID <- "my_study"

# Ionization mode: "pos" or "neg"
POLARITY <- "neg"

# Number of CPU cores for parallel processing
CORES_NB <- 4

# m/z tolerance used during preprocessing (needed for grouping parameters)
PPM <- 10

# Matching parameters
MATCH_PPM <- 20        # ppm for database matching
MATCH_RT_TOL <- 20     # RT tolerance in seconds for matching
ISOPEAK_SIM_THRESHOLD <- 0.78  # minimum isotope pattern similarity

# QC filtering
RSD_THRESHOLD <- 0.3   # 30% RSD threshold for QC filtering

# Expected sample types (for validation and CV summary)
EXPECTED_SAMPLE_TYPES <- c("QC")  # e.g., c("QC", "Treatment", "Control")

# Color palette for sample types (NULL = auto-generated from data)
SAMPLE_TYPE_COLORS <- NULL
# Example: c("QC" = "#999999", "Treatment" = "#377EB8", "Control" = "#E41A1C")
```

# Setup

## Load Packages and Helper Functions

```{r load-helpers}
#| label: load-helpers

source(file.path(PROJECT_ROOT, "R/lipid_helpers.R"))
source(file.path(PROJECT_ROOT, "R/imputers.R"))
load_lipid_packages(verbose = TRUE)
```

## Setup Folder Structure

```{r setup-folders}
#| label: setup-folders

folders <- setup_folders(polarity = POLARITY)

SEQ_FILE <- paste0("seq_", POLARITY, "_", STUDY_ID, ".xlsx")
IS_FILE <- if (POLARITY == "pos") "pos_lipid_reference_set.xlsx" else "neg_lipid_reference_set.xlsx"
DB_SHEET <- if (POLARITY == "pos") 4L else 5L
DATA_SUFFIX <- paste0("_", POLARITY, ".mzML")
RT_COL <- if (POLARITY == "pos") "RT ESI(+) (min)" else "RT ESI(\u2013) (min)"

if (.Platform$OS.type == "unix") {
  register(MulticoreParam(CORES_NB))
} else {
  register(SnowParam(CORES_NB))
}
```

---

# Load Preprocessed Data

Load the preprocessed `MsExperiment` and `SummarizedExperiment` objects
generated by `Preprocessing.qmd`.

```{r load-preprocessed}
#| label: load-preprocessed

mse_save_path <- file.path(folders$objects,
                           paste0(STUDY_ID, "_preprocessed_mse_", POLARITY))
res_save_path <- file.path(folders$objects,
                           paste0(STUDY_ID, "_preprocessed_res_", POLARITY))

mse <- readMsObject(XcmsExperiment(), AlabasterParam(path = mse_save_path))
res <- readObject(res_save_path)

# Reload reference lipids and EIC
intern_standard <- prepare_reference_lipids(
  IS_FILE,
  rt_window_left = 30,# Make sure these settings match those used during preprocessing
  rt_window_right = ifelse(POLARITY == "pos", 15, 20)# Make sure these settings match those used during preprocessing
)
eic_is <- extract_is_eics(mse, intern_standard)
fdata <- fData(eic_is)

# Setup color palette
palette_sample_type <- create_color_palette(
  mse,
  column = "sample_type",
  custom_colors = SAMPLE_TYPE_COLORS
)

message("✓ Loaded preprocessed objects")
message("  - MsExperiment: ", length(mse), " samples")
message("  - SummarizedExperiment: ", nrow(res), " features x ", ncol(res), " samples")
```

---

# SRM 1950 Database RT adjustment

By default, the code fits a SCAM model to the lipid reference set and uses the 
observed retention time shifts to adjust the SRM 1950 database RT.

```{r rt-correction-fit}
#| label: rt-correction-fit
#| fig-cap: "RT adjustment model diagnostic plots"

# Prepare grouping parameter
group <- eic_is$sample_type
group[group == "QC"] <- NA
group <- as.factor(group)
param_group <- PeakDensityParam(
  sampleGroups = group,
  minFraction = 0.5,
  binSize = 0.01, ppm = PPM, bw = 2
)

# Fit RT adjustment model
rt_fit <- fit_rt_correction(
  eic_is, intern_standard, param_group,
  output_dir = folders$ref_lipid
)

# Display model summary
summary(rt_fit$fit)

# Validate the RT adjustment model
validate_rt_correction(rt_fit, min_r_squared = 0.90, max_residual = 15)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: RT Correction Model
Review the diagnostic plots in **`r folders$ref_lipid`**.

- Red points (reference lipids) should align closely with the fitted curve.
- **R² > 0.95** is expected for a good fit; values below 0.90 indicate poor calibration.
- **Residuals** should be randomly distributed around zero with no systematic patterns.
- Typical residual range: ±10 seconds; larger deviations suggest outlier reference compounds.
:::

## Load and Correct Lipid Database

The `prepare_lipid_database()` function handles the following database preprocessing steps:

1. Rank filtering - Removes rank=0 (unconfirmed) entries.
2. RT conversion - Converts RT from minutes to seconds.
3. Unique naming - Appends the RT rank for lipids with multiple isomers.
4. Adduct processing - Calculates theoretical m/z values using `adductFormula()`.
5. Deuterium handling - Converts D↔[2H] notation for formula calculations.
6. NA removal - Removes empty rows resulting from the Excel import.
7. RT adjustment - Applies the SCAM model fitted to the lipid reference set.

```{r load-database}
#| label: load-database

lipid_database <- prepare_lipid_database(
  db_path = "../LipidDatabase_R.xlsx",
  sheet = DB_SHEET,
  polarity = POLARITY,
  rt_col = RT_COL,
  rt_fit = rt_fit
)
```

```{r validate-lrs-coverage}
#| label: validate-lrs-coverage

# Check that the LRS contains at least one internal standard per subclass
# present in the lipid database. Uncovered subclasses are reported as a
# warning — add matching IS to the LRS Excel (or harmonize the subclass
# vocabulary between LRS and database) before trusting the IS normalization.
validate_lrs_coverage(intern_standard, lipid_database)
```

```{r plot-rt-correction}
#| label: plot-rt-correction
#| fig-cap: "Database RT vs. adjusted RT with Lipid reference set overlay"

# Plot RT correction results
plot(y = lipid_database$rt_sd, x = lipid_database$rt_adjusted, type = "l",
     xlab = "Adjusted RT", ylab = "Database RT")
points(intern_standard$RT ~ rt_fit$exp_rt, col = "red", pch = 16, cex = 1)
abline(v = min(rt_fit$fit$fitted.values), col = "blue", lwd = 1, lty = 2)
abline(v = max(rt_fit$fit$fitted.values), col = "blue", lwd = 1, lty = 2)
grid()
abline(0, 1, lty = 2, col = "#00000040")
legend("topleft", legend = c("SRM 1950 Database lipids", "Lipid Reference Set"),
       col = c("black", "red"), lty = c(1, NA), pch = c(NA, 16), bty = "n")
```

---

# Multi-evidence lipid annotation

## Rank 1 Matching

Features are matched against the SRM 1950-derived lipid database using primary adducts (Rank 1).
Matches must satisfy both m/z (ppm tolerance) and RT (time tolerance) constraints.

```{r rank1-matching}
#| label: rank1-matching

match_result <- match_features_to_database(
  res,
  lipid_database,
  ppm = MATCH_PPM,
  rt_tol = MATCH_RT_TOL
)

mtched_data <- match_result$mtched_data
query <- match_result$query
lipid_db_r1 <- match_result$lipid_db_r1

# Validate matching results
validate_matches(mtched_data, min_matches = 10, max_ppm_error = 15)
```

## Isotope Pattern Validation

```{r isotope-validation}
#| label: isotope-validation

# Calculate isotope pattern similarity
iso_result <- calculate_isotope_similarity(
  mse, mtched_data, polarity = POLARITY,
  isopeak_threshold = 2,
  similarity_threshold = ISOPEAK_SIM_THRESHOLD
)

mtched_data <- iso_result$mtched_data

message("Matches after isotope filter: ", nrow(mtched_data))
```

```{r plot-isotope-patterns}
#| label: plot-isotope-patterns

# Save isotope pattern comparison plots
plot_isotope_patterns(
  mtched_data,
  iso_result$iso_spectra,
  iso_result$theoretical_spectra,
  folders$iso_pattern
)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Isotope Pattern Validation
Review the isotope-pattern mirror plots in **`r folders$iso_pattern`**.

- The experimental pattern (top) should match the theoretical pattern (bottom).
- Pay attention to compounds with similarity scores near the threshold. Adjust
 `ISOPEAK_SIM_THRESHOLD` if needed.
- Record any potential false-positive matches where the experimental theoretical patterns show clear discrepancies.
:::

## Adduct Matching (Rank > 1)

Following rank 1 matching, search for secondary adducts that provide additional evidence for the lipid identification.
Finding multiple adducts for the same lipid increases annotation confidence.

```{r adduct-matching}
#| label: adduct-matching

mtched_data <- match_adducts(
  mtched_data,
  lipid_database,
  query,
  ppm = MATCH_PPM,
  rt_tol = 5
)
```

## Resolve Ambiguities

```{r resolve-ambiguities}
#| label: resolve-ambiguities

cols_amb <- c("feature_id", "target_mz", "mzmed", "target_rt_adjusted",
              "rtmed", "target_lipid_name_unique", "target_adduct",
              "ppm_error", "score_rt", "target_rank", "score",
              "isopeak_count", "isopeak_sim", "adduct_ratio", "ntch_idx")

# Resolve lipid -> multiple features
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")

rm_vector <- c()
for (i in key[duplicated(key)]) {
  idxs <- which(key == i)
  count <- mtched_data$adduct_ratio[idxs]
  if (which.max(count) == which.min(count)) next
  best_idx <- idxs[which.max(count)]
  rm_vector <- c(rm_vector, setdiff(idxs, best_idx))
}

if (length(rm_vector) > 0) {
  mtched_data <- mtched_data[-rm_vector, ]
  message("Removed ", length(rm_vector), " rows (lower adduct counts)")
}

# Resolve by RT score
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
dupe_keys <- unique(key[duplicated(key)])
rows_to_keep <- rep(TRUE, nrow(mtched_data))

for (k in dupe_keys) {
  idxs <- which(key == k)
  scores <- abs(mtched_data$score_rt[idxs])
  best_score <- min(scores, na.rm = TRUE)
  to_remove <- idxs[(scores - best_score) > 4]
  if (length(to_remove) > 0) rows_to_keep[to_remove] <- FALSE
}

mtched_data <- mtched_data[rows_to_keep, ]
message("After RT score filter: ", nrow(mtched_data), " matches")
```

## sn1/sn2 Isomer Resolution

```{r sm-resolution}
#| label: sm-resolution

mtched_data <- resolve_sm_isomers(mtched_data)
```

## Export Ambiguity Tables for Manual Curation

Export tables for manual review of ambiguous matches:

- Lipid ambiguities: one lipid matches multiple features
- Feature ambiguities: one feature matches multiple lipids (will be merged if not resolved)

```{r export-ambiguity}
#| label: export-ambiguity

ambiguities <- export_ambiguity_tables(mtched_data)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Manual Ambiguity Resolution

Two Excel files are generated for manual review of ambiguous annotations:

1. **`lipid_ambiguity_resolution.xlsx`**: Cases where a single lipid matches multiple features. By default, all entries have `keep_row = VERDADERO`.
2. **`feature_ambiguity_resolution.xlsx`**: Cases where a single feature matches multiple lipids.If these ambiguities are not resolved, the corresponding entries will be merged.

**Action Required:**
- Open each Excel file and review the ambiguous matches.
- Set `keep_row = FALSE` for entries that should be removed.
- Save the files after completing the review.
- Run the code chunk below to apply your curation decisions.
:::

## Apply Manual Curation

```{r apply-curation}
#| label: apply-curation

# Uncomment after manual curation is complete
lipid_amb_res <- read_excel("lipid_ambiguity_resolution.xlsx")
lipid_amb_res$keep_row <- as.logical(lipid_amb_res$keep_row)
mtch_to_be_rm <- lipid_amb_res$ntch_idx[!lipid_amb_res$keep_row]
mtched_data <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm, ]

feature_amb_res <- read_excel("feature_ambiguity_resolution.xlsx")
feature_amb_res$keep_row <- as.logical(feature_amb_res$keep_row)
mtch_to_be_rm_ft <- feature_amb_res$ntch_idx[!feature_amb_res$keep_row]
mtched_data <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm_ft, ]

# Merge remaining duplicated features
mtched_data$nb_annotations <- 1
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
for (i in unique(fids_ambiguous)) {
  idxs <- which(mtched_data$feature_id == i)
  if (length(idxs) > 1) {
    combined_names <- paste(unique(mtched_data$target_lipid_name_unique[idxs]),
                            collapse = "; ")
    combined_adducts <- paste(unique(mtched_data$target_adduct[idxs]),
                              collapse = "; ")
    mtched_data$target_lipid_name_unique[idxs[1]] <- combined_names
    mtched_data$target_adduct[idxs[1]] <- combined_adducts
    mtched_data <- mtched_data[-idxs[-1], ]
    mtched_data$nb_annotations[idxs[1]] <- length(idxs)
  }
}

message("Final matched lipids: ", nrow(mtched_data))
```

## Annotation Summary

```{r annotation-summary}
#| label: annotation-summary

# Coverage summary
n_unique_lipids <- length(unique(mtched_data$target_lipid_name_unique))
n_db_lipids <- length(unique(lipid_db_r1$lipid_name_unique))
coverage <- n_unique_lipids / n_db_lipids * 100

message("Final annotation: ", n_unique_lipids, " unique lipids (",
        round(coverage, 1), "% database coverage)")

# Lipid category distribution
counts <- table(mtched_data$`target_LIPID.CATEGORY..ABBREV.`)
plot_lipid_donut(counts, title = "Annotated Lipid Distribution")
```

---

# Data Normalization

## Create Annotated Result Object

```{r annotated-res}
#| label: annotated-res

# Subset to annotated features only
annotated_res <- res[rownames(res) %in% mtched_data$feature_id, ]

# Add annotation metadata
rowData(annotated_res) <- cbind(
  rowData(annotated_res),
  mtched_data[match(rownames(annotated_res), mtched_data$feature_id),
              c("target_lipid.name", "target_lipid_name_unique",
                "target_LIPID.CATEGORY..ABBREV.",
                "target_LIPID.SUBCLASS..ABBREV.", "target_Adduct",
                "target_IS_norm")]
)

message("Annotated result object: ", nrow(annotated_res), " features")
```

## Missing Value Imputation

We impute it here to perform PCA plotting.

```{r imputation}
#| label: imputation

assays(annotated_res)$raw_filled_imputed <- impute_block_knn_log(
    as.matrix(assay(annotated_res, "raw_filled")), a = 0.4, b = 0.7, k = 5
)
```

## Internal Standard Normalization

```{r is-normalization}
#| label: is-normalization

# Normalize by internal standards
annotated_res <- normalize_by_is(
  annotated_res,
  input_assay = "raw_filled",
  output_assay = "ISnorm_filled",
  is_col = "target_IS_norm",
  lipid_col = "target_lipid.name"
)

# Impute normalized data
assays(annotated_res)$ISnorm_filled_imputed <- impute_block_knn_log(
    as.matrix(assay(annotated_res, "ISnorm_filled")), a = 0.4, b = 0.7, k = 5
)
```

---

# Quality Control

## Relative Standard Deviation (RSD) Filtering

```{r rsd-filter}
#| label: rsd-filter

# Store unfiltered object
res_unfilt <- annotated_res

# Apply RSD filter
annotated_res <- filter_by_qc_rsd(annotated_res, threshold = RSD_THRESHOLD)
```

## Quality Assessment Plots

```{r pca-plot}
#| label: pca-plot
#| fig-cap: "PCA before and after IS normalization, colored by sample type"

p_before <- plot_pca(assay(annotated_res, "raw_filled_imputed"),
                     sample_type = annotated_res$sample_type,
                     palette = palette_sample_type,
                     title = "PCA - Before Normalization")
p_after  <- plot_pca(assay(annotated_res, "ISnorm_filled_imputed"),
                     sample_type = annotated_res$sample_type,
                     palette = palette_sample_type,
                     title = "PCA - After Normalization")

ggsave(file.path(folders$figures, paste0("PCA_before_normalization_", POLARITY, ".png")),
       p_before, width = 7, height = 5, dpi = 300)
ggsave(file.path(folders$figures, paste0("PCA_after_normalization_", POLARITY, ".png")),
       p_after, width = 7, height = 5, dpi = 300)

p_before
p_after
```

```{r rla-plot}
#| label: rla-plot
#| fig-cap: "Relative Log Abundance before and after normalization"

par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))

col_sample_ann <- get_sample_colors(annotated_res$sample_type, palette_sample_type)

boxplot(rowRla(assay(annotated_res, "raw_filled"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
        col = paste0(col_sample_ann, "80"), outline = FALSE,
        xaxt = "n", main = "Raw Data", boxwex = 1)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3)

boxplot(rowRla(assay(annotated_res, "ISnorm_filled"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
        col = paste0(col_sample_ann, "80"), outline = FALSE,
        xaxt = "n", main = "Normalized Data", boxwex = 1)
axis(side = 1, at = seq_len(ncol(annotated_res)),
     labels = annotated_res$sample_name, las = 2, cex.axis = 0.6)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3)
```

## Coeficient of variation (CV) summary

```{r cv-summary}
#| label: cv-summary
#| tbl-cap: "Coefficient of variation summary"

index_study <- annotated_res$sample_type != "QC"
index_QC <- annotated_res$sample_type == "QC"

sample_res <- cbind(
  QC_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_QC], na.rm = TRUE, mad = TRUE),
  QC_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_QC], na.rm = TRUE, mad = TRUE),
  Study_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_study], na.rm = TRUE, mad = TRUE),
  Study_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_study], na.rm = TRUE, mad = TRUE)
)

res_df <- data.frame(
  QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
  QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
  Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
  Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe", digits = 3)
```

---

# Final Results

```{r remove-qc}
#| label: remove-qc

# Remove QCs for final analysis
res_qc <- annotated_res[, annotated_res$sample_type == "QC"]
res_final <- annotated_res[, annotated_res$sample_type != "QC"]

# Add QC CV as feature metadata
rowData(res_final)$qc_cv <- assay(res_qc, "ISnorm_filled") |> rowRsd()

# save objects
saveObject(res_final, file.path(folders$objects, paste0(STUDY_ID, "_final_res_", POLARITY)))
```

## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```

---

# References

1. SRM 1950-derived Lipid Database source: <https://doi.org/10.1016/j.jlr.2024.100671>
2. Original study (MICROSAMPLING dataset): <https://doi.org/10.1016/j.microc.2025.113760>
3. Metabonaut tutorial: <https://rformassspectrometry.github.io/Metabonaut/articles/a-end-to-end-untargeted-metabolomics.html#retention-time-alignment>
4. Application study (HIIE dataset): <https://pubs.acs.org/doi/10.1021/acs.jproteome.5c00480>
