---
title: "Targeted Lipidomics Workflow"
subtitle: "Vendor-independent LC-MS data processing and annotation"
author: "CEMBIO-EURAC Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```


```{r delete-objects}
# FIXME: this should be handled in the chunk save-preprocessed. This is a quick 
# and dirty solution to get this working.
# delete "objects" folder if it exists
if (dir.exists("objects")) {
  unlink("objects", recursive = TRUE)
}

```


# Project Structure Overview

This workflow is designed for targeted lipidomic profiling using LC-MS data.
It can process both positive and negative ionization mode data.

## Folder Hierarchy

```
project_root/
├── Lipidomics_workflow.qmd     # This main workflow file
├── R/
│   └── lipid_helpers.R         # Helper functions (source this first)
├── POS_data/                   # Positive mode .mzML files
├── NEG_data/                   # Negative mode .mzML files
├── objects/                    # Saved R objects (MsExperiment, SummarizedExperiment)
│   ├── preprocessed_mse_pos/
│   ├── preprocessed_mse_neg/
│   ├── preprocessed_res/
│   └── preprocessed_res_neg/
├── figures/                    # Output figures
│   ├── EIC_internal_standards/
│   ├── iso_pattern_check/
│   ├── iso_pattern_check_neg/
│   ├── ref_lipid_image/
│   └── ref_lipid_image_neg/
├── seq_pos.xlsx                # Sample sequence for positive mode
├── seq_neg.xlsx                # Sample sequence for negative mode
├── splashlipidomix_list.xlsx   # Reference lipids for positive mode
├── neg_splashlipidomix_list.xlsx  # Reference lipids for negative mode
└── LipidDatabase_R.xlsx        # Lipid database (sheets 4=POS, 5=NEG)
```

## Required Input Files

Before running, prepare these files:

1. Sample sequence file (`seq_pos.xlsx` or `seq_neg.xlsx`):
   - Columns: `file_name`, `sample_name`, `sample_type`, `injection_index`
   - `sample_type` should include: QC, M-VAMS, W-DBS, C-qDBS, Plasma

2. Reference lipid list (`splashlipidomix_list.xlsx` or `neg_splashlipidomix_list.xlsx`):
   - Columns: `short_name`, `mz`, `RT` (in seconds)

3. Lipid database (`LipidDatabase_R.xlsx`):
   - Sheet 4: Positive mode database
   - Sheet 5: Negative mode database

4. Raw data files in `.mzML` format in the appropriate data folder

---

# Configuration

The workflow processes both positive and negative ionization modes in a single run.
xcms preprocessing steps (1–7) use `PosNegMsExp`/`PosNegXcmsExp` to apply identical parameters to
both modes simultaneously. Annotation steps (8–10) loop over each mode, using
mode-specific reference files and database sheets. Adjust parameters below as needed.

```{r config}
#| label: configuration

# ============================================================================
# USER CONFIGURATION - MODIFY THIS SECTION
# ============================================================================

# Smoke test mode: if TRUE, slice data to a minimal subset (all QCs + 2 study
# samples per type) to allow a fast run-through of the entire pipeline.
# Snapshots are always saved; filename tags distinguish "smoke" vs "full".
SMOKE <- FALSE

# Directory for regression snapshots.
# "tests/snapshots"     → golden baselines (committed, never overwrite)
# "tests/snapshots_dev" → written during active development / refactor work
SNAPSHOT_DIR <- "tests/snapshots_dev"

# Data source: "local", "sqlite", or "metaboLights"
# - "local": Load from local .mzML files in POS_data/ or NEG_data/
# - "sqlite": Load from pre-built SQLite database (faster for repeated analyses)
# - "metaboLights": Download directly from MetaboLights repository (MTBLS10722)
DATA_SOURCE <- "local"

# MetaboLights study ID (only used if DATA_SOURCE = "metaboLights")
MTBLS_ID <- "MTBLS10722"

# SQLite database paths (only used if DATA_SOURCE = "sqlite")
SQLITE_DB_POS <- NULL  # e.g., "POS_data/pilot_pos.sqlite"
SQLITE_DB_NEG <- NULL  # e.g., "NEG_data/pilot_neg.sqlite"

# Number of CPU cores for parallel processing (use 1-2 below your max)
CORES_NB <- 4

# Retention time filter range (seconds) per ionization mode
RT_FILTER_POS <- c(10, 800)
RT_FILTER_NEG <- c(10, 800)

# Chromatographic peak detection parameters
PEAK_WIDTH <- c(4, 8)  # min and max peak width in seconds
PPM <- 10              # m/z tolerance in ppm
SN_THRESHOLD <- 2      # signal-to-noise threshold

# Matching parameters
MATCH_PPM <- 20        # ppm for database matching
MATCH_RT_TOL <- 20     # RT tolerance in seconds for matching
ISOPEAK_SIM_THRESHOLD <- 0.78  # minimum isotope pattern similarity

# QC filtering
RSD_THRESHOLD <- 0.3   # 30% RSD threshold for QC filtering

# Volume correction factors (specific to this study)
VOLUME_FACTORS <- list(
  "W" = 2.5,    # W-DBS: smaller volume injected
  "P" = 1.08    # Plasma: slight adjustment
)

# --- Per-mode file configuration ---
SEQ_FILE_POS <- "seq_pos.xlsx"
SEQ_FILE_NEG <- "seq_neg.xlsx"
IS_FILE_POS  <- "splashlipidomix_list.xlsx"
IS_FILE_NEG  <- "neg_splashlipidomix_list.xlsx"
DB_SHEET_POS <- 4L
DB_SHEET_NEG <- 5L
RT_COL_POS   <- "RT ESI(+) (min)"
RT_COL_NEG   <- "RT ESI(\u2013) (min)"
```

# Setup

## Load Packages and Helper Functions

```{r load-helpers}
#| label: load-helpers

# Source helper functions
source("R/lipid_helpers.R")
source("R/pos_neg_exp.R")

# Load all required packages
load_lipid_packages(verbose = TRUE)
```

```{r snapshot-helper}
#| label: snapshot-helper
#| include: false

# Save a lightweight regression snapshot to SNAPSHOT_DIR.
# Always runs (both SMOKE and full modes); the filename encodes polarity and
# mode so the two sets can be compared independently.
#   step_name    : short identifier, e.g. "step3_chrompeaks"
#   object       : the R object to serialise (keep it small — data frames, vectors)
#   min_rows     : for data frames, minimum acceptable row count (default 1)
#   min_r2       : for RT correction list, minimum acceptable R² (default 0.5)
#   snap_dir     : output directory; defaults to SNAPSHOT_DIR from config
#
# Validation rules applied per object type:
#   data.frame  → nrow >= min_rows; warn on NA-dominated key columns
#   list        → each named numeric element checked for all-NA
#   numeric/list of vectors (adjustedRtime) → each element length > 0
save_snapshot <- function(object, step_name, polarity, smoke,
                          min_rows = 1L, min_r2 = 0.5,
                          snap_dir = SNAPSHOT_DIR) {
    if (!dir.exists(snap_dir)) dir.create(snap_dir, recursive = TRUE)
    mode_tag <- if (isTRUE(smoke)) "smoke" else "full"
    fname <- file.path(snap_dir,
                       paste0(step_name, "_", polarity, "_", mode_tag, ".rds"))

    issues <- character(0)

    if (is.data.frame(object)) {
        if (nrow(object) < min_rows)
            issues <- c(issues, sprintf("data frame has %d rows (min %d)",
                                        nrow(object), min_rows))
        # Warn on columns where >80% of values are NA
        na_frac <- colMeans(is.na(object))
        bad_cols <- names(na_frac[na_frac > 0.8])
        if (length(bad_cols))
            issues <- c(issues, sprintf(
                "column(s) >80%% NA: %s", paste(bad_cols, collapse = ", ")))

    } else if (is.list(object) && !is.data.frame(object)) {
        # adjustedRtime returns a list of per-sample numeric vectors
        if (all(vapply(object, is.numeric, logical(1)))) {
            empty_els <- sum(vapply(object, function(x) length(x) == 0L, logical(1)))
            if (empty_els > 0)
                issues <- c(issues, sprintf("%d element(s) have length 0", empty_els))
            allna_els <- sum(vapply(object, function(x) all(is.na(x)), logical(1)))
            if (allna_els > 0)
                issues <- c(issues, sprintf("%d element(s) are all-NA", allna_els))
        } else {
            # Structured list (step7, step8)
            if (!is.null(object$r_squared) && object$r_squared < min_r2)
                issues <- c(issues, sprintf("R² = %.3f < %.1f", object$r_squared, min_r2))
            if (!is.null(object$n_features) && object$n_features < 1L)
                issues <- c(issues, "n_features = 0")
            na_nums <- Filter(function(x) is.numeric(x) && all(is.na(x)), object)
            if (length(na_nums))
                issues <- c(issues, sprintf("all-NA element(s): %s",
                                            paste(names(na_nums), collapse = ", ")))
        }
    }

    if (length(issues)) {
        warning("Snapshot '", step_name, "' (", mode_tag, "/", polarity, "): ",
                paste(issues, collapse = "; "),
                " — snapshot saved but may be degenerate for regression use")
    }

    saveRDS(object, fname)
    message(sprintf("Snapshot saved [%s/%s]: %s  (%s)",
                    mode_tag, polarity,
                    basename(fname),
                    if (length(issues)) "WARNINGS — see above" else "OK"))
    invisible(fname)
}
```

## Validate Configuration

```{r validate-config}
#| label: validate-config

# Validate parameters for both modes (same xcms params, different RT filter)
for (.pol in c("pos", "neg")) {
  validate_config(
    polarity = .pol,
    data_source = DATA_SOURCE,
    cores_nb = CORES_NB,
    rt_filter = if (.pol == "pos") RT_FILTER_POS else RT_FILTER_NEG,
    peak_width = PEAK_WIDTH,
    ppm = PPM,
    sn_threshold = SN_THRESHOLD,
    match_ppm = MATCH_PPM,
    match_rt_tol = MATCH_RT_TOL,
    isopeak_sim = ISOPEAK_SIM_THRESHOLD,
    rsd_threshold = RSD_THRESHOLD
  )
}
rm(.pol)
```

## Setup Folder Structure

```{r setup-folders}
#| label: setup-folders

# Create folder structure for both modes
folders_pos <- setup_folders(polarity = "pos")
folders_neg <- setup_folders(polarity = "neg")

# Setup parallel processing
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(CORES_NB))
} else {
  register(SnowParam(CORES_NB))
}
```

---

# Data Import

Load MS data based on the selected `DATA_SOURCE` option.

::: {.panel-tabset}

## Local mzML Files

Load data from local `.mzML` files. This is the default option when you have
the raw data files in your `POS_data/` or `NEG_data/` folder.

## SQLite Database

Load from a pre-built SQLite database. This is faster for repeated analyses
as it avoids re-parsing the mzML files each time.

**To create a SQLite database from local mzML files:**

```r
source("R/create_sqlite_database.R")
```

Edit the configuration section at the top of that script to set your polarity
and file paths before running.

## MetaboLights

Download data directly from the MetaboLights repository. This requires an
internet connection and may take some time for the initial download. Data is
cached locally after the first download.

:::

## Validate Input Files

```{r validate-inputs}
#| label: validate-inputs

for (.pol in c("pos", "neg")) {
  .folders  <- if (.pol == "pos") folders_pos else folders_neg
  .seq_file <- if (.pol == "pos") SEQ_FILE_POS else SEQ_FILE_NEG
  .is_file  <- if (.pol == "pos") IS_FILE_POS  else IS_FILE_NEG
  .db_sheet <- if (.pol == "pos") DB_SHEET_POS else DB_SHEET_NEG
  .rt_col   <- if (.pol == "pos") RT_COL_POS   else RT_COL_NEG

  validate_metadata(
    .seq_file,
    expected_sample_types = c("QC", "M-VAMS", "W-DBS", "C-qDBS", "Plasma"),
    data_dir = if (DATA_SOURCE == "local") .folders$data else NULL
  )
  validate_reference_lipids(.is_file)
  validate_lipid_database(
    db_path = "LipidDatabase_R.xlsx",
    sheet = .db_sheet,
    polarity = .pol,
    rt_col = .rt_col
  )
}
rm(.pol, .folders, .seq_file, .is_file, .db_sheet, .rt_col)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Input Validation
Check the output above for any **errors** or **warnings**. Common issues include:

- Missing or misspelled columns in Excel files
- mzML files listed in the metadata but not found on disk
- Unexpected sample types
- Duplicate entries

Fix any reported problems in your input files **before** continuing.
:::

```{r load-data}
#| label: load-data

.load_mode <- function(polarity, folders, seq_file) {
  seq_data <- read_xlsx(seq_file, col_names = TRUE) |> as.data.frame()
  seq_data$polarity <- polarity

  if (DATA_SOURCE == "local") {
    spectra_files <- file.path(folders$data, seq_data$file_name)
    missing_files <- spectra_files[!file.exists(spectra_files)]
    if (length(missing_files) > 0)
      stop("Missing files (", polarity, "):\n", paste(missing_files, collapse = "\n"))
    readMsExperiment(spectraFiles = spectra_files, sampleData = seq_data)

  } else if (DATA_SOURCE == "sqlite") {
    sqlite_db <- if (polarity == "pos") SQLITE_DB_POS else SQLITE_DB_NEG
    if (is.null(sqlite_db))
      sqlite_db <- file.path(folders$data, paste0("pilot_", polarity, ".sqlite"))
    load_from_sqlite(sqlite_db, seq_data)

  } else if (DATA_SOURCE == "metaboLights") {
    assay_name <- if (polarity == "pos") {
      "a_MTBLS10722_LC-MS_positive_reverse-phase_metabolite_profiling.txt"
    } else {
      "a_MTBLS10722_LC-MS_negative_reverse-phase_metabolite_profiling.txt"
    }
    mtbls_column_mapping <- list(
      sample_name = "Sample Name",
      file_name = "Raw Spectral Data File",
      sample_type = "Factor Value[Sample type]",
      injection_index = "Comment[injection_index]"
    )
    load_from_metaboLights(
      mtbls_id = MTBLS_ID,
      assay_name = assay_name,
      column_mapping = mtbls_column_mapping,
      filter_column = "Factor Value[Cohort]",
      filter_value = "Cohort_study",
      exclude_sample_types = "Blank",
      polarity = polarity
    )
  } else {
    stop("Invalid DATA_SOURCE: '", DATA_SOURCE,
         "'. Use 'local', 'sqlite', or 'metaboLights'.")
  }
}

message("Loading positive mode data...")
mse_pos <- .load_mode("pos", folders_pos, SEQ_FILE_POS)
message("Loading negative mode data...")
mse_neg <- .load_mode("neg", folders_neg, SEQ_FILE_NEG)
rm(.load_mode)

# Wrap both modes in a single dual-mode object (PosNegMsExp until findChromPeaks)
data <- PosNegMsExp(mse_pos, mse_neg)
message("✓ Loaded ", length(posExp(data)), " samples (pos) and ",
        length(negExp(data)), " samples (neg)")
sampleData(posExp(data))
```

```{r smoke-slice}
#| label: smoke-slice

# SMOKE mode: retain all QC samples (needed for RT alignment) and at most 2
# study samples per sample type, giving a fast but structurally complete run.
# Both modes are assumed to have the same sample ordering.
if (isTRUE(SMOKE)) {
    sd <- sampleData(posExp(data))
    qc_idx   <- which(sd$sample_type == "QC")
    other_types <- unique(sd$sample_type[sd$sample_type != "QC"])
    other_idx <- unlist(lapply(other_types, function(t) {
        head(which(sd$sample_type == t), 2)
    }))
    keep_idx <- sort(unique(c(qc_idx, other_idx)))
    data <- filterFile(data, file = keep_idx)
    message("SMOKE mode: subset to ", length(posExp(data)), " samples (",
            paste(sampleData(posExp(data))$sample_name, collapse = ", "), ")")
}
```

## Setup Color Palettes

Create color palettes for visualization. By default, colors are based on
`sample_type`. You can create additional palettes for other columns.

```{r setup-colors}
#| label: setup-colors

# Create palette for sample_type (same sample types in both modes)
palette_sample_type <- create_color_palette(
  posExp(data),
  column = "sample_type",
  custom_colors =
    c("QC" = "#999999",
      "M-VAMS" = "#377EB8",
      "W-DBS" = "#E41A1C",
      "C-qDBS" = "#4DAF4A",
      "Plasma" = "#984EA3"
    )
)

message("Color palette for sample_type:")
print(palette_sample_type)
```

# Dataset Investigation

## Chromatographic Data Visualization

```{r bpc-raw}
#| label: bpc-raw
#| fig-cap: "Base Peak Chromatogram (BPC) for positive (left) and negative (right) mode"

par(mfrow = c(1, 2))
plot_bpc(posExp(data), palette = palette_sample_type, main = "BPC - Positive Mode")
plot_bpc(negExp(data), palette = palette_sample_type, main = "BPC - Negative Mode")
```

::: {.callout-important}
## 📋 USER CHECKPOINT: RT Filter Range
Examine the BPC above and adjust `RT_FILTER_POS` / `RT_FILTER_NEG` in the
Configuration section if needed.
Current: POS **`r RT_FILTER_POS[1]`–`r RT_FILTER_POS[2]`** s,
NEG **`r RT_FILTER_NEG[1]`–`r RT_FILTER_NEG[2]`** s.
:::

```{r filter-rt}
#| label: filter-rt

# Filter both modes using perMode() to allow different ranges per polarity
data <- filterRt(data, rt = perMode(pos = RT_FILTER_POS, neg = RT_FILTER_NEG))
```

```{r bpc-filtered}
#| label: bpc-filtered
#| fig-cap: "BPC after retention time filtering (positive left, negative right)"

par(mfrow = c(1, 2))
plot_bpc(posExp(data), palette = palette_sample_type, main = "BPC - Positive (filtered)")
plot_bpc(negExp(data), palette = palette_sample_type, main = "BPC - Negative (filtered)")
```

## Reference Lipid EICs

Load and visualize the internal standards / reference lipids used for RT correction.

```{r load-is}
#| label: load-is
#| tbl-cap: "Reference lipids for RT correction"

# Load reference lipids for each mode
intern_standard_pos <- prepare_reference_lipids(
  IS_FILE_POS, rt_window_left = 30, rt_window_right = 15
)
intern_standard_neg <- prepare_reference_lipids(
  IS_FILE_NEG, rt_window_left = 30, rt_window_right = 20
)

# Extract EICs for both modes
eic_is_pos <- extract_is_eics(posExp(data), intern_standard_pos)
eic_is_neg <- extract_is_eics(negExp(data), intern_standard_neg)
fdata_pos  <- fData(eic_is_pos)
fdata_neg  <- fData(eic_is_neg)

message("Positive mode reference lipids:")
kable(intern_standard_pos[, c("short_name", "mz", "RT")], format = "pipe")
message("Negative mode reference lipids:")
kable(intern_standard_neg[, c("short_name", "mz", "RT")], format = "pipe")
```

```{r plot-is-eics}
#| label: plot-is-eics

# Save EIC plots for reference lipids (both modes)
plot_eic_batch(eic_is_pos, fdata_pos, palette = palette_sample_type,
               output_dir = folders_pos$eic_istd)
plot_eic_batch(eic_is_neg, fdata_neg, palette = palette_sample_type,
               output_dir = folders_neg$eic_istd)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Reference Lipid EICs
Check the EIC plots saved in **`r folders_pos$eic_istd`** (positive) and
**`r folders_neg$eic_istd`** (negative).

- The red dashed line indicates the expected RT
- Each peak should be detected in most/all samples
- Note any compounds that show poor signal or unexpected RT shifts
- Note Peak Width.
:::

---

Add ppm code?

# Data Preprocessing

## Chromatographic Peak Detection

```{r peak-detection-params}
#| label: peak-detection-params

# Define peak detection parameters
param_pd <- CentWaveParam(
  peakwidth = PEAK_WIDTH,
  ppm = PPM,
  integrate = 2,
  snthresh = SN_THRESHOLD,
  firstBaselineCheck = FALSE
)
param_pd
```

```{r run-peak-detection}
#| label: run-peak-detection

# Run peak detection on both modes simultaneously
data <- findChromPeaks(data, param = param_pd, chunkSize = CORES_NB)

# Update EICs with peak detection results for each mode
eic_is_pos <- extract_is_eics(posExp(data), intern_standard_pos)
fData(eic_is_pos) <- fdata_pos
eic_is_neg <- extract_is_eics(negExp(data), intern_standard_neg)
fData(eic_is_neg) <- fdata_neg

# Inline checks and snapshots for both modes
for (.pol in c("pos", "neg")) {
    .mse <- if (.pol == "pos") posExp(data) else negExp(data)
    n_peaks   <- nrow(chromPeaks(.mse))
    n_samples <- length(.mse)
    if (n_peaks == 0L)
        stop("No chromatographic peaks detected (", .pol, "). ",
             "Check CentWave parameters or data quality.")
    if (n_peaks / n_samples < 10)
        warning("Very few peaks per sample in ", .pol, " mode (",
                round(n_peaks / n_samples, 1), ").")
    save_snapshot(as.data.frame(chromPeaks(.mse)),
                  "step3_chrompeaks", .pol, SMOKE)
}
rm(.pol, .mse, n_peaks, n_samples)
```

```{r peaks-per-sample}
#| label: peaks-per-sample
#| tbl-cap: "Chromatographic peaks detected per sample (positive / negative)"

for (.pol in c("pos", "neg")) {
    .mse <- if (.pol == "pos") posExp(data) else negExp(data)
    message(.pol, " mode:")
    data.frame(
        sample_name = sampleData(.mse)$sample_name,
        peak_count  = as.integer(table(chromPeaks(.mse)[, "sample"]))
    ) |> t() |> kable(format = "pipe") |> print()
}
rm(.pol, .mse)
```

```{r save-peak-detection-plots}
#| label: save-peak-detection-plots

# Save peak detection result plots for both modes
plot_eic_batch(eic_is_pos, fdata_pos, palette = palette_sample_type,
               output_dir = folders_pos$peak_detection)
plot_eic_batch(eic_is_neg, fdata_neg, palette = palette_sample_type,
               output_dir = folders_neg$peak_detection)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Peak Detection Results
Review the plots in **`r folders_pos$peak_detection`** (positive) and
**`r folders_neg$peak_detection`** (negative).

- Each peak should be detected in most samples
- Check that peaks are not merged incorrectly
- Adjust `PEAK_WIDTH`, `PPM`, or `SN_THRESHOLD` if needed
:::

## Retention Time Alignment

```{r rt-alignment}
#| label: rt-alignment

# Both modes share the same sample structure — use pos mode to derive groups
QC_samples <- sampleData(posExp(data))$sample_type == "QC"

param_c <- PeakDensityParam(
  sampleGroups = QC_samples,
  minFraction = 0.9,
  binSize = 0.01, ppm = PPM,
  bw = 2
)
data <- groupChromPeaks(data, param = param_c)

# Perform alignment on both modes simultaneously
subset_idx <- which(QC_samples)
param_align <- PeakGroupsParam(
  minFraction = 0.9,
  extraPeaks = 50,
  span = 0.5,
  subsetAdjust = "average",
  subset = subset_idx
)

data <- adjustRtime(data, param = param_align)
```

```{r plot-alignment}
#| label: plot-alignment
#| fig-cap: "Retention time alignment (positive left, negative right)"

par(mfrow = c(1, 2))
for (.pol in c("pos", "neg")) {
    .mse <- if (.pol == "pos") posExp(data) else negExp(data)
    col_sample <- get_sample_colors(sampleData(.mse)$sample_type, palette_sample_type)
    plotAdjustedRtime(.mse, col = paste0(col_sample, "80"), peakGroupsPch = 1,
                      main = paste("RT alignment -", .pol))
    grid()
    legend("topright", col = palette_sample_type,
           legend = names(palette_sample_type), lty = 1, bty = "n")
}
rm(.pol, .mse, col_sample)
```

```{r apply-alignment}
#| label: apply-alignment

# Snapshot BEFORE applying for both modes
save_snapshot(adjustedRtime(posExp(data)), "step5_adjrt", "pos", SMOKE)
save_snapshot(adjustedRtime(negExp(data)), "step5_adjrt", "neg", SMOKE)

# Keep raw data for comparison, then apply alignment.
# Note: applyAdjustedRtime() is not an S4 generic in xcms so it cannot be
# dispatched on PosNegXcmsExp — call it explicitly on each mode slot.
data_raw <- dropAdjustedRtime(data)
data <- PosNegXcmsExp(
    applyAdjustedRtime(posExp(data)),
    applyAdjustedRtime(negExp(data))
)
```

## Correspondence Analysis

```{r correspondence}
#| label: correspondence

# Define sample groups (QCs as NA)
group_mse <- sampleData(posExp(data))$sample_type
group_mse[group_mse == "QC"] <- NA
group_mse <- as.factor(group_mse)

# Correspondence parameters
param_cor <- PeakDensityParam(
  sampleGroups = group_mse,
  minFraction = 0.5,
  binSize = 0.01,
  ppm = PPM,
  bw = 2
)

# Run correspondence on both modes simultaneously
data <- groupChromPeaks(data, param = param_cor)

for (.pol in c("pos", "neg")) {
    .mse <- if (.pol == "pos") posExp(data) else negExp(data)
    message(.pol, " correspondence: ", nrow(featureDefinitions(.mse)), " features")
    save_snapshot(as.data.frame(featureDefinitions(.mse)),
                  "step6_features", .pol, SMOKE)
}
rm(.pol, .mse)
```

## Gap Filling

```{r gap-filling}
#| label: gap-filling

# Capture pre-fill NA rates for both modes
na_before <- list(
    pos = sum(is.na(featureValues(posExp(data)))) / length(featureValues(posExp(data))) * 100,
    neg = sum(is.na(featureValues(negExp(data)))) / length(featureValues(negExp(data))) * 100
)

# Fill missing values in both modes simultaneously
data <- fillChromPeaks(data, param = ChromPeakAreaParam(), chunkSize = CORES_NB)

# Capture post-fill NA rates, report, and snapshot
for (.pol in c("pos", "neg")) {
    .mse     <- if (.pol == "pos") posExp(data) else negExp(data)
    na_after <- sum(is.na(featureValues(.mse))) / length(featureValues(.mse)) * 100
    message(.pol, " gap filling: ", round(na_before[[.pol]], 1),
            "% -> ", round(na_after, 1), "% NA")
    if (na_before[[.pol]] > 50)
        warning("High initial NA % in ", .pol, " mode (",
                round(na_before[[.pol]], 1), "%).")
    save_snapshot(list(na_before  = na_before[[.pol]],
                       na_after   = na_after,
                       n_features = nrow(featureDefinitions(.mse))),
                  "step7_gapfill", .pol, SMOKE)
}
rm(.pol, .mse, na_after)
```

::: {.callout-note}
## Gap Filling Guidance
- **< 30% NA before filling**: Good - typical for well-detected features
- **30-50% NA before filling**: Acceptable - may indicate lower abundance compounds
- **> 50% NA before filling**: Review peak detection parameters or sample quality
:::

## Extract Results and Save

```{r extract-results}
#| label: extract-results

# Extract per-mode SummarizedExperiment objects
mse_pos <- posExp(data)
mse_neg <- negExp(data)

res_pos <- quantify(mse_pos, method = "sum", filled = FALSE)
assays(res_pos)$raw_filled <- featureValues(mse_pos, method = "sum", filled = TRUE)

res_neg <- quantify(mse_neg, method = "sum", filled = FALSE)
assays(res_neg)$raw_filled <- featureValues(mse_neg, method = "sum", filled = TRUE)
```

```{r validate-preprocessing}
#| label: validate-preprocessing

validate_result(res_pos, assay_name = "raw", min_features = 50)
validate_result(res_neg, assay_name = "raw", min_features = 50)
```

```{r save-preprocessed}
#| label: save-preprocessed
#| eval: true

# Save preprocessed objects for both modes
for (.pol in c("pos", "neg")) {
    .mse     <- if (.pol == "pos") mse_pos else mse_neg
    .res     <- if (.pol == "pos") res_pos else res_neg
    .folders <- if (.pol == "pos") folders_pos else folders_neg

    mse_save_path <- file.path(.folders$objects, paste0("preprocessed_mse_", .pol))
    res_save_path <- file.path(.folders$objects, paste0("preprocessed_res_", .pol))

    if (DATA_SOURCE == "sqlite") {
        spectra(.mse) <- setBackend(spectra(.mse), MsBackendMemory())
        save(.mse, file = paste0(mse_save_path, ".RData"))
    } else {
        saveMsObject(.mse, AlabasterParam(path = mse_save_path))
    }

    saveRDS(.res, paste0(res_save_path, ".rds"))
}
rm(.pol, .mse, .res, .folders, mse_save_path, res_save_path)
```

---
# Skip Preprocessing (Optional)

Use this section to load previously saved objects if prepossessing was already
done.

```{r load-preprocessed}
#| label: load-preprocessed
#| eval: false
source("R/lipid_helpers.R")
source("R/pos_neg_exp.R")
load_lipid_packages(verbose = TRUE)
folders_pos <- setup_folders(polarity = "pos")
folders_neg <- setup_folders(polarity = "neg")

for (.pol in c("pos", "neg")) {
    .folders      <- if (.pol == "pos") folders_pos else folders_neg
    mse_save_path <- file.path(.folders$objects, paste0("preprocessed_mse_", .pol))
    res_save_path <- file.path(.folders$objects, paste0("preprocessed_res_", .pol))

    if (file.exists(paste0(mse_save_path, ".RData"))) {
        load(paste0(mse_save_path, ".RData"))
        assign(paste0("mse_", .pol), mse)
    } else {
        assign(paste0("mse_", .pol),
               readMsObject(XcmsExperiment(), AlabasterParam(path = mse_save_path)))
    }
    assign(paste0("res_", .pol), readRDS(paste0(res_save_path, ".rds")))
}
rm(.pol, .folders, mse_save_path, res_save_path)

intern_standard_pos <- prepare_reference_lipids(IS_FILE_POS)
intern_standard_neg <- prepare_reference_lipids(IS_FILE_NEG)
eic_is_pos <- extract_is_eics(mse_pos, intern_standard_pos)
eic_is_neg <- extract_is_eics(mse_neg, intern_standard_neg)
fdata_pos  <- fData(eic_is_pos)
fdata_neg  <- fData(eic_is_neg)

message("✓ Loaded preprocessed objects (pos + neg)")
```

---

# Annotation, Normalization, and QC (per mode)

Steps 8–10 run separately for each ionization mode. The loop below executes
the full annotation pipeline (RT correction → database matching → isotope
validation → ambiguity resolution → normalization → QC) for positive mode
followed by negative mode. Final objects are stored in `res_final_pos` and
`res_final_neg`.

::: {.callout-important}
## 📋 USER CHECKPOINT: Manual Ambiguity Resolution
After the loop completes, two Excel files per mode will have been created:

- `lipid_ambiguity_resolution_{pos|neg}.xlsx`
- `feature_ambiguity_resolution_{pos|neg}.xlsx`

Review and edit `keep_row` as needed, then re-run the **apply-curation** chunk.
:::

```{r annotation-loop}
#| label: annotation-loop

res_final_list <- list()

for (POLARITY in c("pos", "neg")) {

    mse     <- if (POLARITY == "pos") mse_pos else mse_neg
    res     <- if (POLARITY == "pos") res_pos else res_neg
    folders <- if (POLARITY == "pos") folders_pos else folders_neg
    IS_FILE          <- if (POLARITY == "pos") IS_FILE_POS  else IS_FILE_NEG
    DB_SHEET         <- if (POLARITY == "pos") DB_SHEET_POS else DB_SHEET_NEG
    RT_COL           <- if (POLARITY == "pos") RT_COL_POS   else RT_COL_NEG
    intern_standard  <- if (POLARITY == "pos") intern_standard_pos else intern_standard_neg
    eic_is           <- if (POLARITY == "pos") eic_is_pos else eic_is_neg

    message("\n========== Annotation: ", toupper(POLARITY), " mode ==========")

    # ------------------------------------------------------------------
    # Step 8 — Database RT Correction
    # ------------------------------------------------------------------
    group <- eic_is$sample_type
    group[group == "QC"] <- NA
    group <- as.factor(group)
    param_group <- PeakDensityParam(
        sampleGroups = group,
        minFraction = 0.5,
        binSize = 0.01, ppm = PPM, bw = 2
    )

    rt_fit <- fit_rt_correction(
        eic_is, intern_standard, param_group,
        poly_degree = 6,
        output_dir = folders$ref_lipid
    )
    summary(rt_fit$fit)
    validate_rt_correction(rt_fit, min_r_squared = 0.90, max_residual = 15)
    message(POLARITY, " RT correction R\u00b2 = ",
            round(summary(rt_fit$fit)$r.squared, 4))

    save_snapshot(list(coef      = coef(rt_fit$fit),
                       r_squared = summary(rt_fit$fit)$r.squared,
                       residuals = residuals(rt_fit$fit)),
                  "step8_rtcorrection", POLARITY, SMOKE)

    lipid_database <- prepare_lipid_database(
        db_path  = "LipidDatabase_R.xlsx",
        sheet    = DB_SHEET,
        polarity = POLARITY,
        rt_col   = RT_COL,
        rt_fit   = rt_fit
    )

    # ------------------------------------------------------------------
    # Step 9 — Annotation: rank 1 matching + isotope validation
    # ------------------------------------------------------------------
    match_result <- match_features_to_database(
        res, lipid_database,
        ppm    = MATCH_PPM,
        rt_tol = MATCH_RT_TOL
    )
    mtched_data <- match_result$mtched_data
    query       <- match_result$query
    lipid_db_r1 <- match_result$lipid_db_r1
    validate_matches(mtched_data, min_matches = 10, max_ppm_error = 15)

    iso_result  <- calculate_isotope_similarity(
        mse, mtched_data, polarity = POLARITY,
        isopeak_threshold    = 2,
        similarity_threshold = ISOPEAK_SIM_THRESHOLD
    )
    mtched_data <- iso_result$mtched_data
    message(POLARITY, " matches after isotope filter: ", nrow(mtched_data))

    if (nrow(mtched_data) == 0L)
        stop("No matches remain after isotope filter (", POLARITY, "). ",
             "Check ISOPEAK_SIM_THRESHOLD or data quality.")
    if (isTRUE(SMOKE) && nrow(mtched_data) < 5L)
        warning("Only ", nrow(mtched_data), " match(es) after isotope filter ",
                "in SMOKE/", POLARITY, " mode.")

    save_snapshot(mtched_data[, c("feature_id", "target_lipid_name_unique",
                                   "target_adduct", "target_rank",
                                   "isopeak_count", "isopeak_sim",
                                   "ppm_error", "score_rt")],
                  "step9_annotation", POLARITY, SMOKE)

    plot_isotope_patterns(
        mtched_data,
        iso_result$iso_spectra,
        iso_result$theoretical_spectra,
        folders$iso_pattern
    )

    # Adduct matching (rank > 1)
    mtched_data <- match_adducts(
        mtched_data, lipid_database, query,
        ppm = MATCH_PPM, rt_tol = 5
    )

    # ------------------------------------------------------------------
    # Step 10 — Ambiguity resolution
    # ------------------------------------------------------------------
    # Resolve lipid → multiple features (keep highest adduct count)
    key <- paste(mtched_data$target_lipid_name_unique,
                 mtched_data$target_adduct, sep = "_")
    rm_vector <- c()
    for (i in key[duplicated(key)]) {
        idxs  <- which(key == i)
        count <- mtched_data$adduct_ratio[idxs]
        if (which.max(count) == which.min(count)) next
        rm_vector <- c(rm_vector, setdiff(idxs, idxs[which.max(count)]))
    }
    if (length(rm_vector) > 0) {
        mtched_data <- mtched_data[-rm_vector, ]
        message("Removed ", length(rm_vector), " rows (lower adduct counts)")
    }

    # Resolve by RT score
    key       <- paste(mtched_data$target_lipid_name_unique,
                       mtched_data$target_adduct, sep = "_")
    dupe_keys    <- unique(key[duplicated(key)])
    rows_to_keep <- rep(TRUE, nrow(mtched_data))
    for (k in dupe_keys) {
        idxs       <- which(key == k)
        scores     <- abs(mtched_data$score_rt[idxs])
        best_score <- min(scores, na.rm = TRUE)
        to_remove  <- idxs[(scores - best_score) > 4]
        if (length(to_remove) > 0) rows_to_keep[to_remove] <- FALSE
    }
    mtched_data <- mtched_data[rows_to_keep, ]
    message("After RT score filter: ", nrow(mtched_data), " matches")

    mtched_data <- resolve_sm_isomers(mtched_data)

    ambiguities <- export_ambiguity_tables(
        mtched_data,
        lipid_file   = paste0("lipid_ambiguity_resolution_",   POLARITY, ".xlsx"),
        feature_file = paste0("feature_ambiguity_resolution_", POLARITY, ".xlsx")
    )

    # ------------------------------------------------------------------
    # Apply manual curation (reads per-mode Excel files)
    # ------------------------------------------------------------------
    lip_file <- paste0("lipid_ambiguity_resolution_",   POLARITY, ".xlsx")
    ft_file  <- paste0("feature_ambiguity_resolution_", POLARITY, ".xlsx")

    if (file.exists(lip_file)) {
        lipid_amb_res      <- read_excel(lip_file)
        lipid_amb_res$keep_row <- as.logical(lipid_amb_res$keep_row)
        mtch_to_be_rm      <- lipid_amb_res$ntch_idx[!lipid_amb_res$keep_row]
        mtched_data        <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm, ]
    }
    if (file.exists(ft_file)) {
        feature_amb_res      <- read_excel(ft_file)
        feature_amb_res$keep_row <- as.logical(feature_amb_res$keep_row)
        mtch_to_be_rm_ft     <- feature_amb_res$ntch_idx[!feature_amb_res$keep_row]
        mtched_data          <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm_ft, ]
    }

    # Merge remaining duplicated features
    mtched_data$nb_annotations <- 1
    fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
    for (i in unique(fids_ambiguous)) {
        idxs <- which(mtched_data$feature_id == i)
        if (length(idxs) > 1) {
            mtched_data$target_lipid_name_unique[idxs[1]] <-
                paste(unique(mtched_data$target_lipid_name_unique[idxs]), collapse = "; ")
            mtched_data$target_adduct[idxs[1]] <-
                paste(unique(mtched_data$target_adduct[idxs]), collapse = "; ")
            mtched_data$nb_annotations[idxs[1]] <- length(idxs)
            mtched_data <- mtched_data[-idxs[-1], ]
        }
    }
    message("Final matched lipids (", POLARITY, "): ", nrow(mtched_data))

    # Coverage summary
    n_unique_lipids <- length(unique(mtched_data$target_lipid_name_unique))
    n_db_lipids     <- length(unique(lipid_db_r1$lipid_name_unique))
    message(POLARITY, " database coverage: ", n_unique_lipids, " / ", n_db_lipids,
            " (", round(n_unique_lipids / n_db_lipids * 100, 1), "%)")

    counts <- table(mtched_data$`target_LIPID.CATEGORY..ABBREV.`)
    plot_lipid_donut(counts, title = paste("Lipid Distribution -", toupper(POLARITY)))

    # ------------------------------------------------------------------
    # Normalization
    # ------------------------------------------------------------------
    res <- apply_volume_correction(res, VOLUME_FACTORS,
                                   assay_name = "raw",
                                   new_assay_name = "raw_corr")
    res <- apply_volume_correction(res, VOLUME_FACTORS,
                                   assay_name = "raw_filled",
                                   new_assay_name = "corr_filled")

    annotated_res <- res[rownames(res) %in% mtched_data$feature_id, ]
    rowData(annotated_res) <- cbind(
        rowData(annotated_res),
        mtched_data[match(rownames(annotated_res), mtched_data$feature_id),
                    c("target_lipid.name", "target_lipid_name_unique",
                      "target_LIPID.CATEGORY..ABBREV.",
                      "target_LIPID.SUBCLASS..ABBREV.", "target_Adduct",
                      "target_IS_norm")]
    )
    message("Annotated result object (", POLARITY, "): ", nrow(annotated_res), " features")

    tmp <- apply(assay(annotated_res, "raw_filled"),  MARGIN = 1, na_unidis)
    assays(annotated_res)$raw_filled_imputed <- t(tmp)
    tmp <- apply(assay(annotated_res, "corr_filled"), MARGIN = 1, na_unidis)
    assays(annotated_res)$corr_filled_imputed <- t(tmp)

    annotated_res <- normalize_by_is(
        annotated_res,
        input_assay  = "corr_filled",
        output_assay = "ISnorm_filled",
        is_col       = "target_IS_norm",
        lipid_col    = "target_lipid.name"
    )
    tmp <- apply(assay(annotated_res, "ISnorm_filled"), MARGIN = 1, na_unidis)
    assays(annotated_res)$ISnorm_filled_imputed <- t(tmp)

    # ------------------------------------------------------------------
    # QC
    # ------------------------------------------------------------------
    res_unfilt    <- annotated_res
    annotated_res <- filter_by_qc_rsd(annotated_res, threshold = RSD_THRESHOLD)

    # PCA
    vals   <- assay(annotated_res, "ISnorm_filled_imputed") |>
        log2() |> t() |> scale(center = TRUE, scale = TRUE)
    pca_res <- prcomp(vals, scale = FALSE, center = FALSE)
    vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
    print(
        autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
            scale_color_manual(values = palette_sample_type) +
            theme_minimal() +
            ggtitle(paste("PCA - Normalized Data (", toupper(POLARITY), ")"))
    )

    # RLA
    col_sample_ann <- get_sample_colors(annotated_res$sample_type, palette_sample_type)
    par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))
    boxplot(rowRla(assay(annotated_res, "raw_filled"), group = annotated_res$sample_type),
            cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
            col = paste0(col_sample_ann, "80"), outline = FALSE,
            xaxt = "n", main = paste("Raw Data -", toupper(POLARITY)), boxwex = 1)
    grid(nx = NA, ny = NULL); abline(h = 0, lty = 3)
    boxplot(rowRla(assay(annotated_res, "ISnorm_filled"), group = annotated_res$sample_type),
            cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
            col = paste0(col_sample_ann, "80"), outline = FALSE,
            xaxt = "n", main = paste("Normalized -", toupper(POLARITY)), boxwex = 1)
    axis(side = 1, at = seq_len(ncol(annotated_res)),
         labels = annotated_res$sample_name, las = 2, cex.axis = 0.6)
    grid(nx = NA, ny = NULL); abline(h = 0, lty = 3)

    # CV summary
    index_study <- annotated_res$sample_type %in% c("M-VAMS", "W-DBS", "C-qDBS", "Plasma")
    index_QC    <- annotated_res$sample_type == "QC"
    sample_res  <- cbind(
        QC_raw    = rowRsd(assay(annotated_res, "raw_filled")[, index_QC],
                           na.rm = TRUE, mad = TRUE),
        QC_norm   = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_QC],
                           na.rm = TRUE, mad = TRUE),
        Study_raw  = rowRsd(assay(annotated_res, "raw_filled")[, index_study],
                            na.rm = TRUE, mad = TRUE),
        Study_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_study],
                            na.rm = TRUE, mad = TRUE)
    )
    res_df <- data.frame(
        QC_raw    = quantile(sample_res[, "QC_raw"],    na.rm = TRUE),
        QC_norm   = quantile(sample_res[, "QC_norm"],   na.rm = TRUE),
        Study_raw  = quantile(sample_res[, "Study_raw"],  na.rm = TRUE),
        Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
    )
    message(POLARITY, " CV summary:")
    print(kable(res_df, format = "pipe", digits = 3))

    # ------------------------------------------------------------------
    # Final results
    # ------------------------------------------------------------------
    res_qc    <- annotated_res[, annotated_res$sample_type == "QC"]
    res_final <- annotated_res[, annotated_res$sample_type != "QC"]
    rowData(res_final)$qc_cv <- assay(res_qc, "ISnorm_filled") |> rowRsd()

    saveRDS(res_final, file.path(folders$objects, paste0("final_res_", POLARITY, ".rds")))

    qc_cv_na_frac <- mean(is.na(rowData(res_final)$qc_cv))
    if (qc_cv_na_frac > 0.5)
        warning(round(qc_cv_na_frac * 100, 1), "% of features have NA qc_cv (",
                POLARITY, ") — likely caused by missing IS in SMOKE subset.")

    snap_cols <- intersect(
        c("mzmed", "rtmed", "target_lipid_name_unique", "target_Adduct", "qc_cv"),
        colnames(rowData(res_final))
    )
    save_snapshot(as.data.frame(rowData(res_final)[, snap_cols]),
                  "step10_final", POLARITY, SMOKE)

    res_final_list[[POLARITY]] <- res_final

} # end for POLARITY

res_final_pos <- res_final_list$pos
res_final_neg <- res_final_list$neg
message("\n✓ Annotation complete for both modes.")
```

## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```

---

# Reference

1. Database source: <https://doi.org/10.1016/j.jlr.2024.100671>
2. Original study: <https://doi.org/10.1016/j.microc.2025.113760>
