---
title: "R_lipidomicsOptimization"
format: html
---

# TODOS

So we organize a bit what is left to do. Also `!!!` can be found in the text
for more context or issues.

-   Improve annotation strategy.

-   Add figure and table captions (I have been adding them to the plots I think we are going to keep)

-   Clean up useless code.

-   output plots code

# Introduction

Goals:

-   Develop a vendor-independent, open-source and one-step targeted lipid
    processing and annotation workflow for high-confidence lipidomics.

-   Automated RT correction, improving accuracy and reproducibility.

-   Enhanced speed and efficiency in targeted lipid searches, including
    verification through adduct profile matching and isomeric overlaps.

-   A validated computational tool that supports multiple MS platforms, making
    lipidomic studies more accessible and standardized.

-   Apply the developed data-processing pipeline to two independant datasets.
    1)<https://doi.org/10.1016/j.microc.2025.113760>
    2)<https://doi.org/10.1021/acs.jproteome.5c00480>

1) The aim of this study was to explore the potential of three BμS devices for
reliable untargeted lipidomic profiling using reversed-phase ultra-high
performance liquid chromatography coupled with quadrupole-time-of-flight mass
spectrometry (RP-UHPLC-MS). Such devices –from the traditional non-volumetric
dried blood spot cards (DBS) to the recently developed quantitative DBS (qDBS),
and volumetric absorptive microsampling (VAMS)– were compared to conventional
plasma samples from ten healthy volunteers
(https://doi.org/10.1016/j.microc.2025.113760). In this novel study, a
comprehensive lipidomic workflow was applied, combining an all-in-one
extraction protocol and a single-step data processing and lipid annotation
using a targeted peak extraction strategy based on a human plasma lipid
database (https://doi.org/10.1016/j.jlr.2024.100671)

## Load needed libraries

```{r, eval=TRUE}
library(knitr)
library(readxl)
library(MsExperiment)
library(MsIO)
library(alabaster.se)
library(MsBackendMetaboLights)
library(SummarizedExperiment)
library(xcms)
library(Spectra)
library(MetaboCoreUtils)
library(limma)
library(matrixStats)
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggplot2)
library(ggfortify)
library(gridExtra)
library(BiocFileCache)
library(dbplyr)
library(AnnotationHub)
library(CompoundDb)
library(MetaboAnnotation)
library(ggVennDiagram)
library(UpSetR)
```

## Data Import

Load raw mzML data of the pilot study, this includes data of venous blood of 10
healthy individuals loaded in the following BuS devices: M-VAMS, C-qDBS, W-DBS,
also plasma samples.

OPTION 1: User should create the seq_pos.xlsx where injection order, sample name and 
sample_type have to be included. Also POS_ data need to have all the files 
in .mzml that where included in the excel file.

```{r, eval=FALSE}
# Load pos data
seq_pos <- read_xlsx(paste0("seq_pos.xlsx"),
                 col_names = TRUE) |> as.data.frame()
seq_pos$polarity <- "pos"    # create a column named polarity

spectra_files_pos <- file.path("POS_data", seq_pos$`file_name`)

mse_pos <- readMsExperiment(spectraFiles = spectra_files_pos,
                            sampleData = seq_pos)

sampleData(mse_pos)
```

OPTION 2: Below we show how to load from MetaboLights as an alternative to mzml files.

```{r, eval=TRUE}
param <- MetaboLightsParam(mtblsId = "MTBLS10722",
                           assayName = "a_MTBLS10722_LC-MS_positive_reverse-phase_metabolite_profiling.txt",
                           filePattern = ".mzML")

mse_pos <- readMsObject(MsExperiment(),
                        param,
                        keepOntology = FALSE,
                        keepProtocol = FALSE,
                        simplify = TRUE)
```

Some variable mapping needs to be done when importing from MetaboLights

```{r, eval=TRUE}
# Filter for the pilot study
mse_pos <- mse_pos[sampleData(mse_pos)$`Factor Value[Cohort]` == "Cohort_study"]

# Let's rename the column for easier access
sad <- sampleData(mse_pos)
sad <- sad[,c(1, 3, 8, 10)]
colnames(sad) <- c("sample_name", "file_name",  "sample_type", "injection_index")

#  Add injection index column
sad$polarity <- "pos"
sampleData(mse_pos) <- sad

## remove blank samples 
mse_pos <- mse_pos[sampleData(mse_pos)$sample_type != "Blank"]

## reorder the sample based on injection order
mse_pos <- mse_pos[order(sampleData(mse_pos)$injection_index), ]

##only phili - subset to make analysis faster
#mse_pos <- mse_pos[sampleData(mse_pos)$sample_type == "QC"]
```

\[Phili: Here i would explain maybe some column names so that other people
understand what is in them\]

Adjust the number of cores based on your machine. Suggested to use 1 or 2 below
your higher capacity.

```{r, eval=TRUE}
cores_nb <- 4  # Ajust here
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(cores_nb))
} else {
    register(SnowParam(cores_nb))
}
```

Define color per sample_type: M-VAMS, C-qDBS, W-DBS, QC, Plasma

```{r,eval=TRUE}
col_sample_type <- brewer.pal(9, name = "Set1")[c(9, 2, 1, 3, 4)]
names(col_sample_type) <- c("QC",   # grey
                          "M-VAMS", # blue
                          "W-DBS",  # red
                          "C-qDBS", # green
                          "Plasma") # purple
col_sample <- col_sample_type[sampleData(mse_pos)$sample_type]
```

# Dataset investigation

## Chromatographic data visualization

BPC plot of all samples (colored by device), including QCs.

```{r, eval=TRUE}
#| fig-cap: "BPC of all samples colored by phenotype"
#' Extract and plot BPC for full data
bpc <- chromatogram(mse_pos, aggregationFun = "max")

#png("BPC_pos.png", width = 2000, height = 1500, res = 300)

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6, 
       horiz = TRUE,
       bty = "n")

#dev.off()
```

Filter the data based on the retention time observed on the BPC. In this data set we filter
from 10 to 950 sec.

```{r, eval=TRUE}
#| fig-cap: "BPC after filtering retention time."
#' Filter the data based on retention time
mse_pos <- filterRt(mse_pos, c(10, 950))

# Plot BPC after filtering
bpc <- chromatogram(mse_pos, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6,
       horiz = TRUE, bty = "n")
```

```{r, eval=TRUE}
#| fig-cap: "Heatmap of TIC similarities. All plasma samples are group together."
#' Total ion chromatogram
tic <- chromatogram(mse_pos, aggregationFun = "sum") |>
  bin(binSize = 8)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

rownames(ticmap) <- colnames(ticmap) <- sampleData(mse_pos)$sample_name
ann <- data.frame(sample_type = sampleData(mse_pos)[, "sample_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
#png("heatmap_pos.png", width = 2000, height = 1500, res = 300)

pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = col_sample_type),
         show_colnames = FALSE,
         show_rownames = FALSE)

#dev.off()
```

## EICs

Upload the list of the reference lipids (internal standards form SPLAH LIPIDOMIX and also some endogenous compounds), that will be used to make retention time shift correction. Retention time must be in seconds.

```{r, eval=TRUE, warning = FALSE}
#| tbl.cap: "Set of reference lipids, including deuterated lipids from SPLASH LIPIDOMIX and some deuterated compounds that are in high abundance in plasma."
#' Load list of Internal standard SPLASH LIPIDOMIX
intern_standard <- read_xlsx("splashlipidomix_list.xlsx") |>
    as.data.frame()
intern_standard$rtmin <- intern_standard$RT - 30
intern_standard$rtmax <- intern_standard$RT + 15
intern_standard$mzmin <- intern_standard$mz - 0.008 
intern_standard$mzmax <- intern_standard$mz + 0.008
rownames(intern_standard) <- intern_standard$short_name

# Extract EICs for the list
eic_is <- chromatogram(
    mse_pos,
    rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
    mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

#' Add internal standard metadata
fData(eic_is)$mz <- intern_standard$mz
fData(eic_is)$rt <- intern_standard$RT
fData(eic_is)$rtmin <- intern_standard$rtmin
fData(eic_is)$rtmax <- intern_standard$rtmax
fData(eic_is)$short_name <- intern_standard$short_name
rownames(fData(eic_is)) <- intern_standard$short_name

fdata <- fData(eic_is)

fData(eic_is)[c("CAR", "LysoPC", "dLysoPC", "dLysoPE", "LPC", "SM",
                "dPG", "dSM", "PC", "dPC", "dPE", "SM1",
                "dDG", "HexCer", "TG", "dTG", "dCE"),
              c("short_name", "mz", "rt", "rtmin", "rtmax")] |>
    kable(format = "pipe")
```

We below plot the EICs for reference lipids (17 in total).The dotted red line represents the expected retention time of the peak, showing that the retention time shift varies depending on the isocratic range.

```{r, eval=TRUE}
#| fig.cap: "EIC of all reference lipids."

# Loop to go through all rows of the eic_is object and plot independent  
dr <- "./figures/EIC_internal_standards/"
if (!dir.exists(dr)) dir.create(dr, recursive = TRUE)
for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  png(paste0(dr, rownames(fData(eic_is))[i], ".png"), ## check if the path breaks
      width = 800, height = 600, res = 120)
  plot(eic,
       main = rownames(fData(eic_is))[i], 
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
  abline(v = fData(eic_is)$rt[i], col = "red", lty = 3)
  dev.off()
}
```


[Phili: do we need to keep this below ? now i have saved them locally]

```{r, eval=TRUE}
#|fig.cap: "Example: EIC of three selected deuterated lipids."
# Chromatogram example of three lipids from the list that was imported before
eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
#png("EIC_exampledeuteratedlipids.png", width = 2000, height = 1500, res = 300)

par(mfrow = c(1, 3), mar = c(4, 2, 2, 0.5))

plot(eic_deuteratedLysoPC, main = fData(eic_deuteratedLysoPC)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedLysoPC)$rt, col = "red", lty = 3)

plot(eic_deuteratedSM, main = fData(eic_deuteratedSM)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
     col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedSM)$rt, col = "red", lty = 3)

plot(eic_CAR, main = fData(eic_CAR)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_CAR)$rt, col = "red", lty = 3)

legend("topright", col = col_sample_type, legend = names(col_sample_type),
       lty = 1,
       bty = "n")

#dev.off()
```

# Data preprocessing

## Chromatographic peak detection

We use the deuterated LysoPC of the first sample to extract the *m/z*,
calculate the absolute difference and expressed in ppm.

```{r, eval=TRUE}
#' Restrict the data to signal from dLysoPC in the first sample
lpc <- mse_pos[1L] |>
  spectra() |>
  filterRt(rt = c(155, 170)) |>
  filterMzRange(mz = fData(eic_deuteratedLysoPC)["dLysoPC", 
                                                 c("mzmin", "mzmax")])

#' Show the number of peaks per m/z filtered spectra
lengths(lpc)
```

```{r, eval=TRUE}
#' Calculate the difference in m/z values between scans
mz_diff <- lpc |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(lpc))))
```

We therefore, choose a value close to the maximum within this range for
parameter `ppm`, i.e., 10 ppm, otherwise work with 2 `ppm` will be to low.

From previous observation of some of our ISTD we estimate the average peak
width to be around 4 to 8 seconds. We therefore set the `peakwidth` parameter
to `c(4, 8)`. \[Here link to figure above\]

We can now perform the chromatographic peak detection with the adapted settings
on our EICs, and test the parameters in some Lipids from the list. In this case
deuterated LysoPC, Sphingomelyn and Carnitine.

```{r, eval=TRUE}
#' Parameters adapted for chromatographic peak detection on EICs.
#param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
#                          snthresh = 2) # these were the initial parameters
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2, firstBaselineCheck = FALSE)

#' Evaluate on the dLysoPC ion
dLysoPC_test <- findChromPeaks(eic_deuteratedLysoPC, param = param_pd)
chromPeaks(dLysoPC_test)
```

```{r, eval=TRUE}
#' Evaluate on the Sphingomyelin ion
dSM_test <- findChromPeaks(eic_deuteratedSM, param = param_pd)
chromPeaks(dSM_test)
```

```{r}
#' Evaluate on the Sphingomyelin ion
CAR_test <- findChromPeaks(eic_CAR, param = param_pd)
chromPeaks(CAR_test)
```

BELOW IS A TEST to improve peak picking params - to be deleted.
I  tested it by subsetting for the 8 QC samples

```{r}
mse_pos[1] |>
  filterSpectra(filterRt, rt = c(190, 220)) |>
  filterSpectra(filterMzRange, mz = c(482.3, 482.4)) |>
  plot()

## Current param on the problematic compounds
mse_test <- mse_pos[1]
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2)
mse_test <- findChromPeaks(mse_test, param = param_pd, chunkSize = 2L)
fts <- chromatogram(mse_test,
                       rt = as.matrix(df_features[, c("rt_min", "rt_max")]),
                       mz = as.matrix(df_features[, c("mz_min", "mz_max")]))
chromPeaks(fts)
plot(fts)
## cp1 no 
## cp2 yes in 6/8 QCs ( i have some doubt this is not marged...)
## cp3 in 6/8 samples
## cp4 properly separated = refinement might merge. 


param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2, firstBaselineCheck = FALSE) #keep it
mse_test <- findChromPeaks(mse_test, param = param_pd, chunkSize = 2L)
fts <- chromatogram(mse_test,
                       rt = as.matrix(df_features[, c("rt_min", "rt_max")]),
                       mz = as.matrix(df_features[, c("mz_min", "mz_max")]))
chromPeaks(fts)
plot(fts)
## cp1 no 
## cp2 yes in 6/8 QCs ( i have some doubt this is not marged...)
## cp3 in 4/8 samples
## cp4 properly separated = refinement might merge. 

## sooo the original code works best actually

mse_test[1] |>
  filterSpectra(filterRt, rt = c(190, 220)) |>
  filterSpectra(filterMzRange, mz = c(482.3, 482.4)) |>
  plot()

## next i need ot make sure we don't merge cp2 and cp4.
## current code
param_msp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.0015,
                                        minProp = 0.75) ## reduce to 2 and 1s, and then test minprop
mse_test <- refineChromPeaks(mse_test, param = param_msp, chunkSize = 2L)
fts <- chromatogram(mse_test,
                       rt = as.matrix(df_features[c(2,4), c("rt_min", "rt_max")]),
                       mz = as.matrix(df_features[c(2,4), c("mz_min", "mz_max")]))
chromPeaks(fts[1]) ## still ok ? the rt boundary look ok, why is it wrong later on ?
chromPeaks(fts[2]) ## nope, so let's fix this. 
plot(fts)

## new
param_msp <- MergeNeighboringPeaksParam(expandRt = 2, expandMz = 0.0015,
                                        minProp = 0.95) ## reduce to 2 and 1s, and then test minprop
mse_test <- refineChromPeaks(mse_test, param = param_msp, chunkSize = 2L)
fts <- chromatogram(mse_test,
                       rt = as.matrix(df_features[c(2,4), c("rt_min", "rt_max")]),
                       mz = as.matrix(df_features[c(2,4), c("mz_min", "mz_max")]))
chromPeaks(fts[1]) 
chromPeaks(fts[2]) 
plot(fts)
```

My conclusion on the test is: 
- The current chosen param are the best in my opinion, i could not get better 
  results. 
- Sara needs to check that compound 1 was actually detected in her original 
  analysis because i could not see it in any of my tests.
- Regarding merging, I would advise to not run it actually, we merge a very low 
  amount of peaks in the end and the peaks that we don't want to merge seem to
  be merging.. (compound 4 for example). SO let's remove that step ? 
- I would advocate to re-increase the `sntresth` to 5 as well, to avoid picking
  noise. I think it was lowered to get the carnitine but it did not work in 
  the end no ? If it did then ignore my comment and keep it to 2


With the customized parameters, a chromatographic peak was detected in each
sample. Below, we use
the [`plot()`](https://rdrr.io/r/graphics/plot.default.html)  function on the
EICs to visualize these results.

```{r, eval=TRUE}
#| fig.cap: "EIC of three selected deuterated lipids after applying the optimized param selection."
#' Plot test chromatogram for deuterated LysoPC, Sphingomielyn and Carnitine.
par(mfrow = c(1, 3))
plot(dLysoPC_test, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
abline(v = fData(eic_deuteratedLysoPC)$rt, col = "red", lty = 3)

plot(dSM_test, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
abline(v = fData(eic_deuteratedSM)$rt[i], col = "red", lty = 3)

plot(CAR_test, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
abline(v = fData(eic_CAR)$rt, col = "red", lty = 3)

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

We can see a peak seems to be detected in each sample for both ions. This
indicates that our custom settings are suitable for our dataset. We now proceed
and apply them to the entire dataset, extracting EICs again for the same ions
to evaluate and confirm that chromatographic peak detection worked as expected.

```{r, eval=TRUE}
#' Using the same settings, but with default snthresh
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2, firstBaselineCheck = FALSE) # added november 4
## lowered the sntresth super low and still no detection of the carnitine peaks.
mse_pos <- findChromPeaks(mse_pos, param = param_pd, chunkSize = cores_nb)

eics_is_noprocess <- eic_is
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
```

```{r, eval=TRUE}
fData(eic_is) <- fData(eics_is_noprocess)
```

```{r}
# Plot to evaluate the chromatographic peak detection results.
rownames(fData(eic_is)) <- intern_standard$short_name

eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]

## Here add some other double peak compound that we failed to identify before.
```

```{r, eval=FALSE}
#| fig.cap: "EIC of three selected deuterated lipids after applying the chromatographic peak detection on the entire dataset."
par(mfrow = c(1, 3))
plot(eic_deuteratedLysoPC, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_deuteratedSM, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_CAR, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

After applying the cinematographic peak detection over the entire dataset, only
the first two peaks have been detected properly in all sample, but this does
not happen with the Carnitine (not even after changing the signal to noise
ratio, or opening the window for the m/z or rt)

## Refinment of chromatographic peaks

[Phili: I am testing for removal of refinment, this should help for the 
compounds annotation that elute very close to eachothers]

```{r, eval=FALSE}
#' set up the parameter
param_msp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.0015,
                                        minProp = 0.75)
```

```{r, eval=FALSE}
#' Apply on whole dataset
mse_pos <- refineChromPeaks(mse_pos, param = param_msp, chunkSize = cores_nb)
```

```{r, eval=FALSE}
chromPeakData(mse_pos)$merged |>
                         table()
```

Evaluate the results of the chromatographic peak detection on EICs of e.g.
internal standards or other compounds/ions known to be present in the samples.
Here we update our EICs objects for the internal standards with the results of
the chromatographic peak detection and refinement.

```{r, eval=FALSE}
eics_is_chrompeaks <- eic_is

eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fData(eics_is_chrompeaks)

eic_deuteratedLysoPC <- eic_is["dLysoPC", ]
eic_deuteratedSM <- eic_is["dSM", ]
eic_CAR <- eic_is["CAR"]
```

### Evaluate peak detection results

```{r, eval=FALSE}
# Define output directory (create if it doesn't exist)
dr <- "./figures/peak_detection_ref_lipid/"
if (!dir.exists(dr)) dir.create(dr)

# Loop through each row and save plots as PNG
for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i, ]
  filename <- file.path(dr, paste0(rownames(fData(eic_is))[i], ".png"))
  png(filename, width = 800, height = 600, res = 120)
  plot(
    eic,
    main = rownames(fData(eic_is))[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  abline(v = fData(eic_is)$rt[i], col = "red", lty = 3)
  dev.off()
}
```

All the compound plots in `eic_is` are now saved locally, these should be
checked before continuing. Each peak should be detected in most samples (if not
all). Also ensure that all the EICs that have 2 peak are properly separated.
Discuss with Johannes to see how to prevent merge. !!!

Number of chromatographic peaks per sample.

```{r, eval= TRUE}
#| tbl.cap: "Number of chromatographic peaks detected per sample."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(
  sample_name = sampleData(mse_pos)$sample_name,
  peak_count = as.integer(table(chromPeaks(mse_pos)[, "sample"]))) |>
  t() |>
  kable(format = "pipe")
```

A similar number of chromatographic peaks was identified within the various
samples of the data set. In all cases W-DBS has the lower number of features.
Which makes sense as less volume was injected therefore features of lower\
intensity would not be seen in W-DBS.

## Retention time alignment

To evaluate the presence of a shift we extract and plot below the BPC from the
QC samples.

```{r, eval=TRUE}
#| fig-cap: "BPC of QC samples."
#' Get QC samples
QC_samples <- sampleData(mse_pos)$sample_type == "QC"

#' extract BPC
mse_pos[QC_samples] |>
    chromatogram(aggregationFun = "max", chromPeaks = "none") |>
    plot(col = col_sample_type["QC"], main = "BPC of QC samples") |>
    grid()
```

Initial correspondence analysis. Parameters do not need to be fully optimized.
We base it on the QC samples.

```{r, eval=TRUE}
param_c <- PeakDensityParam(
  sampleGroups = sampleData(mse_pos)$sample_type == "QC",
  minFraction = 0.9,
  binSize = 0.01, ppm = 10,
  bw = 2)
mse_pos <- groupChromPeaks(mse_pos, param = param_c)
```

```{r, eval=TRUE}
#| fig-cap: "Initial correspondence analysis based on QC samples."
plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_c,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]], 20),
    peakPch = 16)
```

Below we perform the alignment of our data set based on retention times of
anchor peaks defined in the subset of QC samples.

```{r, eval=TRUE}
#' Define parameters of choice
subset <- which(sampleData(mse_pos)$sample_type == "QC")
param_c <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                           subsetAdjust = "average",
                           subset = subset)

#' Perform the alignment
mse_pos <- adjustRtime(mse_pos, param = param_c)
```

Alignment adjusted the retention times of all spectra in the data set, as well
as the retention times of all identified chromatographic peaks.

```{r, eval=TRUE}
#| fig-cap: "Retention time alignment results."
#| label: fig-alignment-results
#' Visualize alignment results
plotAdjustedRtime(mse_pos, col = paste0(col_sample, 80), peakGroupsPch = 1)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

Only little adjustments needed to be performed (shifts of at maximum 1.5 second
as can be seen in Figure @fig-alignment-results).

We can also compare the BPC before and after alignment.

```{r, eval = TRUE}
#' Get data before alignment
mse_pos_raw <- dropAdjustedRtime(mse_pos)

#' Apply the adjusted retention time to our dataset
mse_pos <- applyAdjustedRtime(mse_pos)
```

```{r, eval= TRUE}
#| fig-cap: "BPC before and after alignment."
#' Plot the BPC before and after alignment
par(mfrow = c(2, 1), mar = c(2, 1, 1, 0.5))
chromatogram(mse_pos_raw, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC before alignment", col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)

chromatogram(mse_pos, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC after alignment",
         col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)
```

There is no shift observed.

We next evaluate the impact of the alignment on the EICs of the selected
internal standards. We thus below first extract the ion chromatograms after
alignment, and subsequently plot them.

```{r, eval=TRUE}
#' Extract the EICs for the test ions before alignment
old_eic_dLysoPC <- eic_is["dLysoPC"]
old_eic_dSM <- eic_is["dSM"]
old_eic_CAR <- eic_is["CAR"]

#' Update the EICs
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fdata

#' Extract the EICs for the test ions
eic_dLysoPC <- eic_is["dLysoPC"]
eic_dSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
```

```{r, eval=TRUE}
#| fig-cap: "EIC of three selected deuterated lipids before and after alignment."
par(mfrow = c(2, 2), mar = c(4, 4.5, 2, 1))

plot(old_eic_dLysoPC, main = "dLysoPC before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(eic_dLysoPC, main = "dLysoPC after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(old_eic_dSM, main = "Sphingomyelin before alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")

plot(eic_dSM, main = "Sphingomyelin after alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

plot(old_eic_CAR, main = "CAR before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)

plot(eic_CAR, main = "CAR after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)
```

In addition to a visual inspection of the results, we also evaluate the impact
of the alignment by comparing the variance in retention times for internal
standards before and after alignment.

```{r, eval= TRUE}
#' Extract the matrix with all chromatographic peaks and add a column
#' with the ID of the chromatographic peak
chrom_peaks <- chromPeaks(mse_pos) |> as.data.frame()
chrom_peaks$peak_id <- rownames(chrom_peaks)

#' Define the parameters for the matching and filtering of the matches
p_1 <- MzRtParam(ppm = 50, toleranceRt = 10)
p_2 <- SingleMatchParam(duplicates = "top_ranked", column = "target_maxo",
                        decreasing = TRUE)

#' Iterate over samples and identify for each the chromatographic peaks
#' with similar m/z and retention time than the onse from the internal
#' standard, and extract among them the ID of the peaks with the
#' highest intensity.
intern_standard_peaks <- lapply(seq_along(mse_pos), function(i) {
    tmp <- chrom_peaks[chrom_peaks[, "sample"] == i, , drop = FALSE]

     # Convert to S4 DataFrames
    query  <- DataFrame(intern_standard)
    target <- DataFrame(tmp)


    mtch <- matchValues(query, target,
                        mzColname = c("mz", "mz"),
                        rtColname = c("RT", "rt"),
                        param = p_1)
    mtch <- filterMatches(mtch, p_2)
    mtch$target_peak_id
}) |>
    do.call(what = cbind)
```

We have now for each internal standard the ID of the chromatographic peak in
each sample that most likely represents signal from that ion. We can now
extract the retention times for these chromatographic peaks before and after
alignment.

```{r, eval= TRUE}
#' Define the index of the selected chromatographic peaks in the
#' full chromPeaks matrix
idx <- match(intern_standard_peaks, rownames(chromPeaks(mse_pos)))

#' Extract the raw retention times for these
rt_raw <- chromPeaks(mse_pos_raw)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))

#' Extract the adjusted retention times for these
rt_adj <- chromPeaks(mse_pos)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))
```

We can now evaluate the impact of the alignment on the retention times of
internal standards across the full data set:

```{r, eval=FALSE}
#| fig-cap: "Retetion time variation of internal standards before and after alignment."
list(all_raw = rowSds(rt_raw, na.rm = TRUE),
     all_adj = rowSds(rt_adj, na.rm = TRUE)
     ) |>
    vioplot()
grid()
```

## Correspondence

The goal of the correspondence analysis is to identify chromatographic peaks
that originate from the same types of ions in all samples of an experiment and
to group them into LC-MS *features*.

```{r, eval=TRUE}
#' Optimized parameter for the grouping and apply them to the test ions BP2
#' fix grouping, NA for QCs
group_mse <- sampleData(mse_pos)$sample_type
group_mse[group_mse =="QC"] <- NA
group_mse <- as.factor(group_mse)
param_cor <- PeakDensityParam(sampleGroups = group_mse, bw = 2)
param_cor
```

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
                    20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated LysoPC.

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedSM, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated SM.

```{r, eval=TRUE}
plotChromPeakDensity(
    eic_CAR, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_CAR)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_CAR)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters. Carnitine does not have any
chromatographic peak, check what is happening.

We can observe for the first two ions that the peaks are grouped accurately
into a single feature. So we proced to define the settings for the param\_ cor
and apply them to the whole data.

```{r, eval=TRUE}
#' Define the settings for the param_cor
param_cor <- PeakDensityParam(sampleGroups = group_mse,
                              minFraction = 0.5, binSize = 0.01, ppm = 10,
                              bw = 2)

## for phili tests, to be removed later: 
# param_cor <- PeakDensityParam(sampleGroups = rep(1, length(mse_pos)),
# minFraction = 0.5, binSize = 0.01, ppm = 10,
# bw = 2)

#' Apply to whole data
mse_pos <- groupChromPeaks(mse_pos, param = param_cor)
```

Evaluation of the correspondence results in other EIC in the IS list similar to
the deuterated LysoPC.

```{r, eval=TRUE}
#' Extract chromatogram for an m/z similar to the one of the labeled methionine
chr_test <- chromatogram(mse_pos,
                         mz = as.matrix(intern_standard["LysoPC",
                                                        c("mzmin", "mzmax")]),
                         rt = c(120, 184),
                         aggregationFun = "max")
plotChromPeakDensity(
    chr_test, simulate = FALSE,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(chr_test)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(chr_test)[, "sample"]], 20),
    peakPch = 16)
```

This corresponds to non-deuterated LysoPC and we can see that it can
distinguish between the two peaks.

```{r, eval=FALSE}
#| fig-cap: "Evaluation of the correspondance results in 17 EICs in the IS list."
# Loop to go through all rows of the eic_is object and plot independent after correspondance analysis

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 abline(v = fData(eic_is)$rt[i], col = "red", lty = 3)
}

# Assuming mse_pos is your raw data object (XCMSnExp) ERROR
mse_posp <- findChromPeaks(mse_pos, param = param)

# Extract chromatograms again with peaks included ERROR
eic_isp <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

for (i in seq_len(nrow(eic_isp))) {
  eicp <- eic_isp[i, ]
  plotChromPeakDensity(
    eicp, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(mse_pos)[, "sample"]], # peaks now exist
    peakBg = paste0(col_sample[chromPeaks(mse_pos)[, "sample"]], 20),
    peakPch = 16
  )
  grid()
  abline(v = fData(eic_icp)$rt[i], col = "red", lty = 3)
}
```

## Gap filling

We next examine how prevalent missing values are in our present dataset:

```{r}
#' Percentage of missing values
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100

idx_qc <- sampleData(mse_pos)$sample_type == "QC"
#' Percentage of missing values
sum(is.na(featureValues(mse_pos[idx_qc, keepFeatures = TRUE]))) /
    length(featureValues(mse_pos[idx_qc, keepFeatures = TRUE])) * 100
```

We can observe a substantial number of missing values values in our dataset.

Fill the missing values in the entire data set.

```{r}
#' Fill in the missing values in the whole dataset
mse_pos <- fillChromPeaks(mse_pos, param = ChromPeakAreaParam(),
                          chunkSize = cores_nb)

#' Percentage of missing values after gap-filling
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100
```

```{r}
#| fig-cap: "Detected vs. filled-in signal."
#' Get only detected signal in QC samples
vals_detect <- featureValues(mse_pos, filled = FALSE)[, QC_samples]

#' Get detected and filled-in signal
vals_filled <- featureValues(mse_pos)[, QC_samples]

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)

#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080")
grid()
abline(0, 1)
```

Figure X. Detected vs. filled-in signal.


## Save mse_pos and res object

```{r}
# save the mse_pos preprocessed object
saveMsObject(mse_pos, AlabasterParam(path = file.path("objects/preprocessed_mse_pos")))

# save the SummarizedExperiemnt
saveObject(res, file.path("objects/preprocessed_res")) 
```

# Database correction.

```{r, eval=TRUE}
#' Default parameter for the grouping and apply them to the test ions BPC
group <- eic_is$sample_type
group[group =="QC"] <- NA
param1 <- PeakDensityParam(sampleGroups = group , minFraction = 0.5,
                           binSize = 0.01, ppm = 10,
                           bw = 2)

## phili code 
# param1 <- PeakDensityParam(sampleGroups = rep(1, length(mse_pos)) , minFraction = 0.5,
#                            binSize = 0.01, ppm = 10,
#                            bw = 2)


eic_is_corr <- groupChromPeaks(eic_is, param = param1)
featureDefinitions(eic_is_corr[3, ]) ## Show johannes, why do we have a
## third peak, we can't even see on the plot. has no intensity.
## with xcms version >= 4.7.1 only 2 features are found.

exp_rt <- vector("numeric", length = nrow(eic_is_corr))
dr <- "figures/ref_lipid_image/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(eic_is_corr))) {
  table <- featureDefinitions(eic_is_corr[i,])
  if (!nrow(table)) {
    message("compound not detected: ", fData(eic_is_corr)$short_name[i])
    exp_rt[i] <- NA
    next
  }
  png(filename = paste0(dr, "feature_compound_", i, ".png"))
  plot(eic_is_corr[i,],
       main = rownames(fData(eic_is_corr))[i],
       cex.axis = 0.8,
       cex.main = 0.8)
  abline(v = fData(eic_is_corr)$rt[i], col = "red", lty = 3)
  dev.off()
  if (nrow(table) > 1) {
    idx_max <- which.max(rowSums(featureValues(eic_is_corr[i, ]),na.rm = TRUE))
    rt <- table[idx_max, ]$rtmed
  } else rt <- table$rtmed
  exp_rt[i] <- rt
}

## before going on, check the plots. the line should overlap what we expect.

# below is the reference lipid table
ref_rt <- intern_standard$RT # is the ref RT

# res is the exp RT
intern_standard$exp_rt <- exp_rt

diff_rt <- intern_standard$RT - intern_standard$exp_rt
quantile(diff_rt, na.rm = TRUE)

## Fit the linear model
fit <- lm(exp_rt ~ poly(ref_rt, 6, raw = TRUE) )
summary(fit)
plot(fit)
```

Sara will check if the shift see m to be similar to what she has or not.I did
not apply any RT shift correction, because my data had not to much shift on the
RT and we were able to do the match against the library. The average difference
is 9.6 seconds

Library fix

```{r}
lipid_database <- read_xlsx("LipidDatabase_R.xlsx", sheet = 4L)
## remove rows that have rank == 0, because they essentially do not exist
lipid_database <- lipid_database[lipid_database$rank != 0, ]
## convert rt
lipid_database$rt_sd <- as.numeric(lipid_database$`RT ESI(+) (min)`) * 60
lipid_database <- lipid_database[order(lipid_database$rt_sd), ]
## Correct naming so that all names are unique 
get_rt_rank <- function(x) {
  match(x, sort(unique(x)))
}

rt_ranks <- ave(
  lipid_database$`RT ESI(+) (min)`, 
  lipid_database$`lipid name`, 
  FUN = get_rt_rank
)
# Create the final unique name
lipid_database$lipid_name_unique <- paste(
  lipid_database$`lipid name`, 
  rt_ranks, 
  sep = "_"
)

is.unsorted(lipid_database$rt_sd) # check rtime are properly sorted. we want FALSE

rt_sd <- lipid_database$rt_sd

## finding values that need to be extrapolated.
in_range_idx <- which(rt_sd >= min(fit$fitted.values) &
                      rt_sd <= max(fit$fitted.values))

nrow(lipid_database) - length(in_range_idx)
## Adjust library retention times
corrected_rt <- rep(NA, length = nrow(lipid_database))
corrected_rt[in_range_idx] <- predict(
    fit, newdata = data.frame(ref_rt = rt_sd[in_range_idx]))
## Adjust values outside the fitted range
idx <- which(rt_sd < min(fit$fitted.values))
lidx <- length(idx)
if (lidx)
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[lidx + 1L] - corrected_rt[lidx + 1L])
idx <- which(rt_sd > max(fit$fitted.values))
if (length(idx))
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[idx[1L] - 1L] - corrected_rt[idx[1L] - 1L])

lipid_database$rt_adjusted <- corrected_rt

## difference between original and adjusted retention times
quantile(abs(rt_sd - corrected_rt))

# plot library 592 compounds rt  raw vs adj. then (on top) add the ref_rt vs exp_rt (20compounds)
```

```{r}
#| fig-cap: "lipid database rt vs. rt adjusted base on the model."
# scatterplot lipid_name and rt(seconds)
lipid_database$mz <- as.numeric(lipid_database$mz)

#lipid_database <- lipid_database %>% mutate(`LIPID CATEGORY [ABBREV]`= as.factor(`LIPID CATEGORY [ABBREV]`))

#ggplot(lipid_database, aes(x=rt_sec, y=mz, group=`LIPID CATEGORY [ABBREV]`)) +
#  geom_point(aes(color=`LIPID CATEGORY [ABBREV]`))+
#  ggtitle("rt_database")

plot(y = lipid_database$rt_sd, x =lipid_database$rt_adjusted, type = "l",
     xlab = "Adjusted rt",
     ylab = "Database rt")
points(intern_standard$RT ~ intern_standard$exp_rt, col = "red", pch = 16, cex = 1)
abline(v = min(fit$fitted.values), col = "blue", lwd = 1, lty = 2)
abline(v = max(fit$fitted.values), col = "blue", lwd = 1, lty =2)
grid()
abline(0, 1, lty = 2, col = "#00000040")
```
The Black line is build with the 592 compounds of the database and the red dots are the reference
lipids (17 between endogenous and deuterated)

# Annotate/ match
-   maybe separate the QCs into another object or do that later.

```{r}
#' extract results as a SummarizedExperiment
res <- quantify(mse_pos, method = "sum", filled = FALSE)
res

assays(res)$raw_filled <- featureValues(mse_pos, method = "sum",
                                        filled = TRUE)
```

```{r}
# Filter the res object setting a threshold of 1.5*10^3
res_filtered <- assay(res)[rowMeans(assay(res), na.rm = TRUE) > 1500, ]

sum(!rowMeans(assay(res), na.rm = TRUE) > 1500) #503 removed features
```


Match features (query) against (aligned) database (target).
DO nOT run for now
```{r, eval = FALSE}
# TGs annotation 
lipid_database_TG <- lipid_database[lipid_database$`LIPID SUBCLASS [ABBREV]` == "Triglycerides [TG]", ]

lipid_database_TG$rank[
  lipid_database_TG$Adduct == "Adduct[M+NH4]+"
] <- 1

lipid_database_TG$rank[
  lipid_database_TG$Adduct == "Adduct[M+C2H7N2]+"
] <- 2

lipid_database_TG2 <- lipid_database_TG[lipid_database_TG$rank == 1, ]
## Get the reference/databse data as a data.frame
target <- as.data.frame(lipid_database_TG2)
target$mz <- as.numeric(target$mz)

## Configure the matching between features (query) and database (target)
param_match <- MzRtParam(tolerance = 0.001,
                         ppm = 20,
                         toleranceRt = 5. #+/-5 sec 
                         ) ## maybe adapt to settings from the original approach
## Perform the matching; define also the column names in query and target that
## contain the values to be matched (i.e. the m/z and retention time values)
mtch <- matchValues(query, target,
                    param = param_match,
                    mzColname = c("mzmed", "mz"),
                    rtColname = c("rtmed", "rt_adjusted"))
mtch

mtch <- mtch[whichQuery(mtch)]
mtched_data <- matchedData(mtch) |>
    as.data.frame()  ## can do some stuff with that table

## Get the features that match multiple targets
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]


# table_amb
table_amb <- mtched_data[mtched_data$feature_id %in% fids_ambiguous,
                         c("feature_id", "mzmed", "rtmed", "target_lipid.name",
                           "target_Adduct", "ppm_error", "score_rt","target_rank")]
table_amb <- table_amb[order(table_amb$feature_id, table_amb$rtmed), ]
rownames(table_amb) <- NULL
library(pander)
pandoc.table(table_amb, split.table = Inf, style = "rmarkdown")

#12 amb and 56 obs whe using amonium as predominant adduct 
#12 amb more ppm error from 3 to 18ppm and 72 obs when acetamidine is predominant adduct
#we have 136 lipids in the databse
write_xlsx(mtched_data, path = "TGs-Amonium.xlsx")
```

CAN BE DELETED
```{r}
#examples amb TGs
d <- featureChromatograms(mse_pos, features = "FT17527") #TG(53:4)acetamidina no isotopic patter in qualitative 
plot(d)
d1 <- chromatogram(mse_pos, mz = c(927.804, 927.822), rt = c(830, 845)) 
plot(d1)

d_nh4 <- featureChromatograms(mse_pos, features = "FT16255") #TG(53:4)amonium isotopic pattern in qualitative
plot(d_nh4)
d_nh4_1 <- chromatogram(mse_pos, mz = c(886.785, 886.792), rt = c(810,840)) 
plot(d_nh4_1)


par(mfrow = c(1, 2))
plot(d1)
plot(d_nh4_1)

par(mfrow = c(1, 2))
plot(d)
plot(d_nh4)


t <- featureChromatograms(mse_pos, features = "FT16322") #TG(53:3)amonium 
plot(t)
t1 <- chromatogram(mse_pos, mz = c(888.798, 888.804), rt = c(835,850)) 
plot(t1)

```
Match features (query) against (aligned) database (target).

```{r}
## Extract the feature definitions as a data.frame
query <- as.data.frame(rowData(res))
query$feature_id <- rownames(rowData(res))
## Subset the lipid library for rank 1 adducts only 
lipid_database_rank1 <- lipid_database[lipid_database$rank == 1, ] #242 annotations with both param_pd

## Get the reference/databse data as a data.frame
target <- as.data.frame(lipid_database_rank1)
target$mz <- as.numeric(target$mz)

## Configure the matching between features (query) and database (target)
param_match <- MzRtParam(tolerance = 0.001,
                         ppm = 20,
                         toleranceRt = 5. #+/-5 sec 
                         ) ## maybe adapt to settings from the original approach
## Perform the matching; define also the column names in query and target that
## contain the values to be matched (i.e. the m/z and retention time values)
mtch <- matchValues(query, target,
                    param = param_match,
                    mzColname = c("mzmed", "mz"),
                    rtColname = c("rtmed", "rt_adjusted"))
mtch
```

From the in total `r nrow(query)` LC-MS features `r length(whichQuery(mtch))`
were matched to at least one lipid in the database. In total
`r length(whichTarget(mtch))` lipids from the database were matched (from in
total `r nrow(target)` compounds/lipids in the database).

We next evaluate the matching results. We first identify features that match
more than one database entry (i.e., the same feature matching to two or more
different lipids/compounds).

```{r}
## only keep query that has matched
mtch <- mtch[whichQuery(mtch)]
mtched_data <- matchedData(mtch) |>
    as.data.frame()  ## can do some stuff with that table

mtched_data$feature_id <- sub("\\..*$", "", rownames(mtched_data))
sum(duplicated(mtched_data$feature_id))
```

Before checking for feature duplication (whih is mostly an issue of the database or can easily be resolved based on score)
I also want to check if one lipid matches multiple feautres as it could
indicate an issue with the processing above. bigger problem

```{r}
sum(duplicated(paste0(mtched_data$target_lipid_name_unique, mtched_data$target_Adduct, sep = "_"))) 
## this above is necessary as some adduct are both rank 1 for the same lipid.
## I have 12, that's a problem. 
amblip <- mtched_data[paste0(mtched_data$target_lipid_name_unique, mtched_data$target_Adduct, sep = "_") %in%
                        paste0(mtched_data$target_lipid_name_unique[duplicated(paste0(mtched_data$target_lipid_name_unique, mtched_data$target_Adduct, sep = "_"))],
                               mtched_data$target_Adduct[duplicated(paste0(mtched_data$target_lipid_name_unique, mtched_data$target_Adduct, sep = "_"))],
                               sep = "_"), ]

amblip

plot(featureChromatograms(mse_pos[1:2, keepFeatures = TRUE], features = c("FT02887", "FT02888"))) ## well it's actually very different fts 
## ok so after checking a lot of them are just proper ambiguity. reducing the ppm threshold would help to be honest. but otherwise it's either score based deicsion or manually (but even manually sometimes we just can tell when it's really betweeen 2 features.)

## for now i will keep the one with the highest score 

mtched_data <- mtched_data[order(mtched_data$score), ]
mtched_data <- mtched_data[!duplicated(paste0(mtched_data$target_lipid_name_unique, mtched_data$target_Adduct, sep = "_")), ]
```

The table below shows the ambiguous (multi-matching) features.

```{r, results = "asis"}
## Get the features that match multiple targets
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
sum(duplicated(mtched_data$feature_id))
table_amb <- mtched_data[mtched_data$feature_id %in% fids_ambiguous,
                         c("feature_id", "mzmed", "rtmed", "target_lipid_name_unique",
                           "target_Adduct", "ppm_error", "score_rt","target_rank", "target_mz", "target_rt_adjusted")]
table_amb <- table_amb[order(table_amb$feature_id), ]
##export
rownames(table_amb) <- NULL
library(writexl)
write_xlsx(table_amb, path = "features-matching-multiple-lipids.xlsx")

## let's check it
library(pander)
pandoc.table(table_amb, split.table = Inf, style = "rmarkdown")

## when checking honestly, i do not have a great way to deal with this. 
## it's either hte rt between the 2 lipids are sooo close < 4 seconds, which is less than the peak width so there's just now way ## and in this case i would group the annotation 
## in another case we have a bit larger difference in rt (still very small though), and in that case i would keep the one that matches the best in rt. 
## we could increase the rt threshold and make some peak match to a further one (as we saw in some case we have a large shift), and then again filter the ambiguity based on rt score.
## but hoenstly if the RT is SO different HOW do we know if this is rteally the right compound.., without MS2 confirmation this is a reach and i would not trust it. 
## one confirmation  could be, the adduct we are checking below. we could check this, afterward on in final only keep the ambiguity where we can fiund more adduct. this would confirm the RT. Sara let's discuss this. i kinda like that idea. 
## 
## ## for now below i just keep the one with the best rt score. will change that based on our discussion
sorted_data <- mtched_data[order(mtched_data$feature_id, mtched_data$score_rt), ]
mtched_data_unique <- sorted_data[!duplicated(sorted_data$feature_id), ]
```




## Matching adducts 
testing another matching
- matching rtmed of the feature definiton, we expect adducts to have the same rt.
- i also tried using rtmin and rtmax but this gave me better results both in term of number AND ambiguity.

I feel like the code can be simplified very much 

```{r}
## i'm "correcting" the duplication by taking the ones that have the lowest
##  RT error, this need to be changed later.
query <- featureDefinitions(mse_pos)
## now i'm taking the RT ranges of the features 
target_round2 <- featureDefinitions(mse_pos)[mtched_data_unique$feature_id,"rtmed", drop = FALSE]
target_round2$lipid_name <- mtched_data_unique$target_lipid_name_unique[
    match(rownames(target_round2), mtched_data_unique$feature_id)]

## subset the data for some lipids of interest
target_round2_subset <- lipid_database[lipid_database$lipid_name_unique %in% target_round2$lipid_name, ]

## remove rank 1 
target_round2_subset <- target_round2_subset[target_round2_subset$rank != 1, ]
target_round2_subset$rtmed <- target_round2$rtmed[
    match(target_round2_subset$lipid_name_unique, target_round2$lipid_name)]

target_round2_subset <- as.data.frame(target_round2_subset)
## Configure the matching between features (query) and database (target)
param_match <- MzRtParam(tolerance = 0.001,
                         ppm = 20,
                         toleranceRt = 1)
mtch <- matchValues(query, target_round2_subset,
                    param = param_match,
                    mzColname = c("mzmed", "mz"),
                    rtColname = c("rtmed", "rtmed"))
mtch

mtch <- mtch[whichQuery(mtch)]
mtched_data_round2 <- matchedData(mtch) |>
    as.data.frame()  

mtched_data_round2$feature_id <- sub("\\..*$", "", rownames(mtched_data_round2))

##  % match 
length(unique(mtched_data_round2$feature_id)) / nrow(target_round2_subset) * 100

## table of duplicated features 
ambfts <- mtched_data_round2[mtched_data_round2$feature_id %in% 
                                 mtched_data_round2$feature_id[duplicated(mtched_data_round2$feature_id)], ]

#plot(featureChromatograms(mse_pos[1:2, keepFeatures = TRUE], features = "FT08860"))
## ye i mean the rt (wether raw or from the table is so close, there is no way we can know)
## some are VERY close isomers, no idea how to deal with that. 

## table of duplicated lipids - i am not 100% sure this is right...
sum(duplicated(paste0(mtched_data_round2$target_lipid_name_unique, mtched_data_round2$target_Adduct, sep = "_"))) 

## honestly this is not bad. 

## let's merge with the first matching results to have a full table
colnames(mtched_data_round2)[!colnames(mtched_data_round2) %in% colnames(mtched_data)]
mtched_data_full <- rbind(mtched_data, mtched_data_round2[, colnames(mtched_data)])
```


```{r}
## example of peak non detected before changing the centwave parameter _pd: 
## it has low intensity, we need to refine preprocessing. 
eg1 <- chromatogram(mse_pos, mz = c(482.315, 482.333), rt = c(193, 210)) #LPE(0:0/18:0)
plot(eg1)

eg2 <- chromatogram(mse_pos, mz = c(510.390, 510.440), rt = c(200, 230)) #LPC(O-0:0/18:0)
plot(eg2)

eg3 <- chromatogram(mse_pos, mz = c(510.380, 510.388), rt = c(175, 190)) # LPC(O-0:0/18:1) // LPC(P-0:0/18:0) only 1 peak?
plot(eg3)
featureDefinitions(eg3)

eg4 <- chromatogram(mse_pos, mz = c(929.833, 929.836), rt = c(855, 875)) #TG(53:3)
plot(eg4)
featureDefinitions(eg4) # wrong rtmin rtmax in the definition ???????

eg5 <- chromatogram(mse_pos, mz = c(787.549, 787.673), rt = c(705, 720)) #SM(d40:1) # set a ppm error on amb table is 4.1 and 787.7 m/z
plot(eg5)
featureDefinitions(eg5)

```

```{r}
#plot some features
a <- featureChromatograms(mse_pos, features = "FT17595") #TG(53:3)
plot(a)
a1 <- chromatogram(mse_pos, mz = c(929.826, 929.847), rt = c(855, 875)) 
plot(a1)

b <- featureChromatograms(mse_pos, features = "FT13173") #SM(d18:2/23:0)
c <- featureChromatograms(mse_pos, features = "FT13172") #SM(d18:2/23:0) some peaks are 10 sec wide
par(mfrow = c(1, 2))
plot(b)
plot(c)
b1 <- chromatogram(mse_pos, mz = c(799.536, 799.671), rt = c(690, 710)) 
plot(b1)


d <- featureChromatograms(mse_pos, features = "FT17527") #TG(53:4)acetamidina 
plot(d)
d1 <- chromatogram(mse_pos, mz = c(927.804, 927.822), rt = c(830, 850)) 
plot(d1)

e <- featureChromatograms(mse_pos, features = "FT08851")
f <- featureChromatograms(mse_pos, features = "FT08853")
g <- featureChromatograms(mse_pos, features = "FT08344")
par(mfrow = c(2, 2))
plot(d,main = "Gráfico D")
plot(e, main = "Gráfico E")
plot(f, main = "Gráfico F")
plot(g, main = "Gráfico G")


h <- featureChromatograms(mse_pos, features = "FT02422") # intensity MG 
j <- featureChromatograms(mse_pos, features = "FT02607")
par(mfrow = c(1, 2))
plot(h,main = "Gráfico H")
plot(j, main = "Gráfico J")

k <- featureChromatograms(mse_pos, features = "FT03637") # intensity PG 1400
plot(k, main = "Gráfico k")
```

```{r}
df_subset <- subset(mtched_data, grepl("ISTD", target_ORIGIN, ignore.case = TRUE))

# Loop to plot all the compounds in the df_subset these are the internal standards
dr <- "./figures/ISTD_mtched_data"
if (!dir.exists(dr)) dir.create(dr, recursive = TRUE)

# Loop through each row and save plots as PNG
for (i in rownames(df_subset)) {
    #Check for the presence of "."
  if (grepl("\\.", i)) {
    original_i <- i
    i <- sub("\\..*$", "", i)  # remove everything after the first dot
    warning(sprintf(
      "Feature '%s' was also matched to something else. Using '%s' instead.",
      original_i, i
    ), call. = FALSE)
  } 
  
  ISTD <- featureChromatograms(mse_pos, features = i)

  # Define filename using the row name
  filename <- file.path(dr, paste0(i, ".png"))

  # Open PNG device
  png(filename, width = 800, height = 600, res = 120)

  # Plot chromatogram
  plot(
    ISTD,
    main = rownames(df_subset)[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  abline(v = df_subset[i, "rtmed"], col = "red", lty = 3)

  # Close device
  dev.off()
}

cat("All chromatograms saved in:", dr, "\n")

```



Can we delete what's below ? sara i will let you deal with that. 

```{r}
## find features IDs of ambiguous matches
tmp <- table(mtch$feature_id)
tmp <- tmp[tmp > 1]
fids <- names(tmp)

## Check the first one
tmp1 <- mtched_data[mtched_data$feature_id == fids[1],
                   c("feature_id", "rtmed", "mzmed", "target_lipid name",
                     "ppm_error", "score_rt", "target_rt_adjusted",
                     "target_rt_sd")] |>
  as.data.frame()
tmp_eic1 <- featureChromatograms(mse_pos, features = fids[1L])
plot(tmp_eic1)

tmp2 <- mtched_data[mtched_data$feature_id == fids[2],
                     c("feature_id", "rtmed", "mzmed", "target_lipid name",
                       "ppm_error", "score_rt", "target_rt_adjusted",
                       "target_rt_sd")] |>
  as.data.frame()
tmp_eic2 <- featureChromatograms(mse_pos, features = fids[2L])
plot(tmp_eic2)

par(mfrow = c(2, 1))
plot(tmp_eic1)
plot(tmp_eic2)

tmp <- mtched_data[mtched_data$feature_id %in% fids,
                     c("feature_id", "rtmed", "mzmed", "target_lipid name",
                       "ppm_error", "score_rt", "target_rt_adjusted",
                       "target_rt_sd")] |>
  as.data.frame()
tmp_eic <- featureChromatograms(mse_pos, features = fids[5])
plot(tmp_eic)


#subset the ISTD

sara <- subset(lipid_database_rank1, grepl("ISTD", ORIGIN, ignore.case = TRUE))
# we don have a match with Tg istd and we should have it 

chromatogram(mse_pos, mz = c(482.29, 482.31)) -> test
plot(test)
# check the feautrea that match 2 compounds. check score, plotSpectraMirror.
# plot using the mtch object.
# scoring and othe rinfo check the mtch_data

## need tofilter them out somehow.


## import information from mtch_data dataframe to res object using the feature ID present in both table.
mtched_data$`target_lipid name`
rownames(mtched_data)

```

## Plots with the annotated compounds (mtched_data), the feature abundance of each sample, can be found in assay(res)

```{r}
## remove duplicated form the
length(unique(mtched_data$target_lipid.name)) ## number of lipid matched, ONLY 206

length(unique(mtched_data$target_LIPID.CATEGORY..ABBREV.))

# Lipid Distribution barplot
# Count data
counts <- table(mtched_data$target_LIPID.CATEGORY..ABBREV.)
bar_colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854")

# Rename categories for readability
names(counts) <- c("Fatty Acyls [FA]",
                   "Glycerolipids [GL]",
                   "Glycerophospholipids [GP]",
                   "Sphingolipids [SP]",
                   "Sterol Lipids [ST]")

# Adjust plot margins and font size
par(mar = c(5, 7, 4, 2))  # increase left margin for labels

bp <- barplot(counts,
              horiz = TRUE,
              col = bar_colors,
              main = "Lipid Distribution",
              xlab = "Number of Lipids",
              las = 1,
              cex.names = 0.9,
              border = NA)

text(counts, bp, labels = counts, pos = 4, cex = 0.8, col = "black")
legend("topright",
       legend = names(counts),
       fill = bar_colors,
       title = "Lipid Category",
       cex = 0.6,
       border = NA,
       bty = "n")
```

-   Donut chart showing the lipid class with the total quantity of lipids
    detected

```{r}
# Donnut chart of lipid distribution
counts <- table(mtched_data$target_LIPID.CATEGORY..ABBREV.)

names(counts) <- c("Fatty Acyls [FA]",
                   "Glycerolipids [GL]",
                   "Glycerophospholipids [GP]",
                   "Sphingolipids [SP]",
                   "Sterol Lipids [ST]")

bar_colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854")
labels <- counts
par(mar = c(1, 1, 2, 1))

# Plot
pie(counts,
    col = bar_colors,
    labels = labels,     # solo los números
    border = "white",
    main = "Lipid Distribution",
    cex = 0.9)

symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

# Leyend
legend("topright",
       legend = names(counts),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Lipid Category",
       cex = 0.8)
```

```{r}
# pie chart for lipid subclass. HERE ADD A NEW COLUMN WITH THE SHORT NAME
counts1 <- table(mtched_data$`target_LIPID SUBCLASS [ABBREV]`)

n_cat <- length(counts1)
base_colors <- brewer.pal(9, "Set1")
bar_colors <- rep(base_colors, length.out = n_cat)

# Labels: only counts
labels <- counts1

# Adjust margins
par(mar = c(1, 1, 2, 1))

# Create pie (donut) chart
pie(counts1,
    col = bar_colors,
    labels = labels,      # show only counts
    border = "white",
    main = paste("Category Distribution (n =", sum(counts), ")"),
    cex = 0.9)

# Add white circle in center for donut effect
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

legend("topright",
       legend = names(counts1),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Categories",
       cex = 0.4)

```

-   heatmap with the significant metabolites found

-   

later on discuss how to deal with features that matches multiple compound.

we could concatenate the annotation to the summarized experiment object. Do the
normalization, quantitation, plotting after,..

# Data normalization 

```{r}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "raw_filled"), MARGIN = 1, na_unidis)
assays(annotated_res)$raw_filled_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

## Cembio workflow

-   normalization followed at CEMBIO for this study:
  - missing values imputation (KNN algorithm)
  - Normalization per volume (multiply whatman samples- factor 2.5 and plama samples - factor 1.08) - DONE
  - Normalization per internal standard (for each lipid subclass dividing per each IS) - DONE

     CAR with C17 sphinganine
     CE with Chol Ester (18:1(d7))
     Cer, Hexcer and SM with SM(18:1/18:1(d9))
     DG with DAG(15:0/18:1(d7))
     FA with FA(16:0(d31))
     LPC with LysoPC(18:1(d7))  
     LPE with LysoPE(18:1(d7))
     PC with PC(15:0/18:1(d7))
     PE with PE(15:0/18:1(d7))
     TG with TAG(15:0/18:1(d7/15:0))
     All this is now on the column IS_ norm, there I indicate the lipid name of the IS that we should use to make the division 
     
  - Remove lipids with a cv in QCs >30% - done

Here before we imput the missing values we have to normalize per volume as
follows: The abundance of all W-DBS samples should be multiply per 2.5 and
Plasma samples per 1.08

```{r}
factors <- ifelse(grepl("W", colnames(res)), 2.5,
             ifelse(grepl("P", colnames(res)), 1.08, 1))
assay(res, "raw_corr") <- sweep(assay(res), MARGIN = 2, 
                                STATS = factors, FUN = "*")
assay(res, "raw_filled_corr") <- sweep(assay(res, "raw_filled"),
                                       MARGIN = 2, STATS = factors, FUN = "*")

print("Assays now in res object:")
print(names(assays(res)))

## Check the changes in one sample (CAN BE DELETED)
cbind(
  original = assay(res, "raw")["FT00001", "D10W_pos.mzML"],
  scaled   = assay(res, "raw_corr")["FT00001", "D10W_pos.mzML"]
)


# Below i create a subset of the res object with only annotated features
mtched_features_list <- mtched_data_unique$feature_id
annotated_res <- res[rownames(res) %in% mtched_data_unique$feature_id, ] #only 279 id out of 299, aprox 20 duplicates
rowData(annotated_res) <- cbind(rowData(annotated_res),
                                 mtched_data_unique[match(rownames(annotated_res),
                                                   mtched_data_unique$feature_id),
                                             c("target_lipid.name",
                                               "target_LIPID.CATEGORY..ABBREV.",
                                               "target_LIPID.SUBCLASS..ABBREV.",
                                               "target_Adduct", "target_IS_norm"
                                             )])
## above if we have different lengths, it's because of duplicated features in the annotation table.

# Below we keep the res object but append annotation info for the annotated features
rowData(res) <- cbind(rowData(res),
                         mtched_data_unique[match(rownames(res),
                                           mtched_data_unique$feature_id),
                                     c("target_lipid.name",
                                       "target_LIPID.CATEGORY..ABBREV.",
                                       "target_LIPID.SUBCLASS..ABBREV.",
                                       "target_Adduct", "target_IS_norm"
                                       )])
```


```{r}
#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "raw_filled_corr"), MARGIN = 1, na_unidis)
assays(annotated_res)$raw_filled_corr_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "raw_filled_corr_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

- Normalization per internal standard (for each lipid subclass dividing per each IS)

     CAR with C17 sphinganine
     CE with Chol Ester (18:1(d7))
     Cer, Hexcer and SM with SM(18:1/18:1(d9))
     DG with DAG(15:0/18:1(d7))
     FA with FA(16:0(d31))
     LPC with LysoPC(18:1(d7))  
     LPE with LysoPE(18:1(d7))
     PC with PC(15:0/18:1(d7))
     PE with PE(15:0/18:1(d7))
     TG with TAG(15:0/18:1(d7/15:0))
     All this is now on the column IS_ norm, there I indicate the lipid name of the IS that we should use to make the division

Again need to simplify the code ( phili will do it here)
```{r}
## this mapping is there: 
rowData(annotated_res)$target_IS_norm

## the internal standanrd nae is here: 
rowData(annotated_res)$target_lipid.name

## now we can make a loop to divide each feature by the IS appropiate
assay(annotated_res, "raw_filled_ISnorm") <- assay(annotated_res, "raw_filled_corr") # create new assay to store the IS normalized data

#check that the is have been detected before anything. 
rowData(annotated_res)$target_IS_norm %in% rowData(annotated_res)$target_lipid.name
## C17 Sphinganine missing


for (i in seq_len(nrow(annotated_res))) {
    is_name <- rowData(annotated_res)$target_IS_norm[i]
    if (!is_name %in% rowData(annotated_res)$target_lipid.name) {
        warning(sprintf("IS '%s' for feature '%s' not found in the dataset. Skipping IS normalization for this feature.",
                        is_name, rownames(annotated_res)[i]))
        next
    }
    if (!is.na(is_name)) {
        is_idx <- which(rowData(annotated_res)$target_lipid.name == is_name)
        if (length(is_idx) == 1) {
            assay(annotated_res, "raw_filled_ISnorm")[i, ] <-
                assay(annotated_res, "raw_filled_corr")[i, ] /
                assay(annotated_res, "raw_filled_corr")[is_idx, ]
        } else {
            warning(sprintf("IS '%s' not found or duplicated for feature '%s'. Skipping IS normalization for this feature.",
                            is_name, rownames(annotated_res)[i]))
        }
    } else {
        warning(sprintf("No IS specified for feature '%s'. Skipping IS normalization for this feature.",
                        rownames(annotated_res)[i]))
    }
}

## quickly check res ## fix evaluation
mat_corr <- assay(annotated_res, "raw_filled_corr")
mat_norm <- assay(annotated_res, "raw_filled_ISnorm")
vals_before <- as.numeric(mat_corr)
vals_after  <- as.numeric(mat_norm)

# Filter out zeros/NAs for logging to avoid errors
vals_before_log <- log2(vals_before[vals_before > 0 & !is.na(vals_before)])
vals_after_log  <- log2(vals_after[vals_after > 0 & !is.na(vals_after)])

vioplot(list(Before = vals_before_log, After = vals_after_log),
        main = "Global Intensity Dist. (Log10)",
        ylab = "Log10 Intensity",
        col = c("mistyrose", "lightcyan"))
```

```{r}
#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "raw_filled_ISnorm"), MARGIN = 1, na_unidis)
assays(annotated_res)$raw_filled_ISnorm_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "raw_filled_ISnorm_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

- Remove lipids with a cv in QCs >30% - quality control 


```{r}
#' Number of features before filtering
nrow(annotated_res)
```

```{r}
 #' keep unfiltered object
res_unfilt <- annotated_res
```

```{r}
# RSD filter 
rsd_filter <- RsdFilter(threshold = 0.3,
                        qcIndex = annotated_res$sample_type == "QC")

annotated_res <- filterFeatures(annotated_res, filter = rsd_filter)
```


```{r}
#' Number of features after analysis
nrow(annotated_res)
```

```{r}
#' Percentage left: end/beginning
nrow(annotated_res)/nrow(res_unfilt) * 100
```

From there on i guess we don't need QCs anymore

```{r}
res_qc <- annotated_res[, annotated_res$sample_type == "QC"]
res <- annotated_res[, annotated_res$sample_type != "QC"]

## redefine colours for the new data set without QCs
col_sample <- col_sample[names(col_sample) != "QC"]
col_sample_type <- col_sample_type[names(col_sample_type) != "QC"]
```

```{r}
#' Calculate the QC's CV and add as feature variable to the data set
rowData(res)$qc_cv <- assay(res_qc, "norm") |>
               rowRsd()
```

```{r}
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_ISnorm_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

FROM HERE statistics metabonaut
we need to discuss if more is necessary and how much they want to deviate from their workflow

```{r}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(res, "raw_filled"), MARGIN = 1, na_unidis)
assays(res)$raw_filled_imputed <- t(tmp)
```

## Initial quality assessment

-   plot also just the annotated data
-   calcualte the missing values Principal Component Analysis

```{r}
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

Figure X. PCA of the data colored by sample_type.Volumetric devices profiles
are clearly different from W-DBS and plasma.

Intensity evaluation

```{r}
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))

par(mar = c(0.2, 4.5, 0.2, 3))
barplot(apply(assay(res, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected peaks", xaxt = "n", space = 0.012)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sample, 80), border = col_sample)
points(colMedians(log2(assay(res, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

replot this after normalisation too Number of detected peaks and feature
abundances.

```{r}
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res, "raw_filled"),
                            f = res$sample_type, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_sample, 80), ylab = "RLA",
        border = col_sample, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

RLA plot for the raw data and filled data.

## Between sample normalisation

Media scaling, Here should I apply the volume and IS normalization???

```{r}
#' Compute median and generate normalization factor
mdns <- apply(assay(res, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(res)$norm <- sweep(assay(res, "raw_filled"), MARGIN = 2, nf_mdn, '/')
assays(res)$norm_imputed <- sweep(assay(res, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn, '/')
```

## Assessing overall effectiveness of the normalization approach

Principal component analysis

```{r}
#' Data before normalization
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_raw <- autoplot(pca_res, data = vals_st,
                    colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)

#' Data after normalization
vals_norm <- apply(assay(res, "norm"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)

pca_res_norm <- prcomp(vals_norm, scale = FALSE, center = FALSE)
vals_st_norm <- cbind(vals_norm, sample_type = res$sample_type)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)

grid.arrange(pca_raw, pca_adj, ncol = 1)
```

write PC1 PC2 interpretation

PC1 and PC2 of the data before and after normalization

```{r}
pca_raw <- autoplot(pca_res, data = vals_st ,
                    colour = 'sample_type', x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    colour = 'sample_type', x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)

grid.arrange(pca_raw, pca_adj, ncol = 1)
```

PC3 and PC4 of the data before and after normalization

Intensity evaluation

```{r}
par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))

boxplot(rowRla(assay(res, "raw_filled"), group = res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Raw data", boxwex = 1)
grid(nx = NA, ny = NULL)
legend("topright", inset = c(0, -0.2), col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE,
       ncol = 3, cex = 0.7, bty = "n")
abline(h = 0, lty=3, lwd = 1, col = "black")

boxplot(rowRla(assay(res, "norm"), group = res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normallized data", cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = res$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3, lwd = 1, col = "black")
```

fix RLA plot before and after normalization - plot also intensity of IS before
and after normalization (log 2)

Coefficient of variation

```{r}
#' Calculate the CV values
index_study <- res$sample_type %in% c("M-VAMS", "W-DBS", "C-qDBS", "Plasma")
index_QC <- res$sample_type == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(res, "norm")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(res, "raw_filled")[, index_study],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(res, "norm")[, index_study],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe")
```

Distribution of cv values across samples for the raw and normalized data

the Cv in study sample decrease, usually that's not ideal but here might be
driven by the whatman intensity correction (check this)



# Overlaping features

```{r}
#create a logical matrix to be able to make the upsetplot
df_logical <- !is.na(assay(res, "raw_filled"))

# Extract group code (letter before "_pos.mzML")
code <- sub(".*([MWCP])_pos\\.mzML$", "\\1", colnames(df_logical))

# Map codes to descriptive names
code_map <- c(
  M = "M-VAMS",
  C = "C-qDBS",
  W = "W-DBS",
  P = "Plasma"
)
groups <- code_map[code]

# Collapse by group using 70% threshold
result <- sapply(unique(groups), function(g) {
  # number of samples in this group
  n <- sum(groups == g)
  # proportion of TRUEs across samples in this group
  rowMeans(df_logical[, groups == g, drop = FALSE]) >= 0.7
})

# Result: logical matrix compounds x groups
result[1:5, ]

#make the upsetplot
binary_data <-apply(result, MARGIN = 2, as.integer)
upset(as.data.frame(binary_data), nset = 6, sets = colnames(result), keep.order = TRUE)
```

Number of Features in common between the devices.

# Differential abundance analysis

```{r}
#' Define the colors for the plot
col_sample <- col_sample_type[res$sample_type]

#' Log transform and scale the data for PCA analysis
vals <- assay(res, "norm_imputed") |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- cbind(vals, sample_type = res$sample_type)
autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)


```

PCA of the data after normalization and quality control

The samples clearly separate by study group on PCA indicating that there are
differences in the metabolite profiles between the groups, speacially Plasma
and de BuS devices are separated by the first component.
