---
title: "R_lipidomicsOptimization"
format: html

---

# TODOS

So we organize a bit what is left to do. Also `!!!` can be found in the text
for more context or issues.


-   Library fix chunk have some errors
-   ERROR in check formula
-   chem_checked ERROR
-   Matching loop (rank 2, 3 etc.)
-   Clean up useless code.

-   output plots code



# Introduction

Goals:

-   Develop a vendor-independent, open-source and one-step targeted lipid
    processing and annotation workflow for high-confidence lipidomics.

-   Automated RT correction, improving accuracy and reproducibility.

-   Enhanced speed and efficiency in targeted lipid searches, including
    verification through adduct profile matching and isomeric overlaps.

-   A validated computational tool that supports multiple MS platforms, making
    lipidomic studies more accessible and standardized.

-   Apply the developed data-processing pipeline to two independant datasets.
    1)<https://doi.org/10.1016/j.microc.2025.113760>
    2)<https://doi.org/10.1021/acs.jproteome.5c00480>

1) The aim of this study was to explore the potential of three BμS devices for
reliable untargeted lipidomic profiling using reversed-phase ultra-high
performance liquid chromatography coupled with quadrupole-time-of-flight mass
spectrometry (RP-UHPLC-MS). Such devices –from the traditional non-volumetric
dried blood spot cards (DBS) to the recently developed quantitative DBS (qDBS),
and volumetric absorptive microsampling (VAMS)– were compared to conventional
plasma samples from ten healthy volunteers
(https://doi.org/10.1016/j.microc.2025.113760). In this novel study, a
comprehensive lipidomic workflow was applied, combining an all-in-one
extraction protocol and a single-step data processing and lipid annotation
using a targeted peak extraction strategy based on a human plasma lipid
database (https://doi.org/10.1016/j.jlr.2024.100671)

## Load needed libraries

```{r}
#| eval: true
library(knitr)
library(readxl)
library(MsExperiment)
library(MsIO)
library(alabaster.se)
library(MsBackendMetaboLights)
library(SummarizedExperiment)
library(xcms)
library(Spectra)
#BiocManager::install("RforMassSpectrometry/MetaboCoreUtils")
library(MetaboCoreUtils)
library(limma)
library(matrixStats)
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggplot2)
library(ggfortify)
library(gridExtra)
library(BiocFileCache)
library(dbplyr)
library(AnnotationHub)
library(CompoundDb)
library(MetaboAnnotation)
library(ggVennDiagram)
library(UpSetR)

run_plot <- FALSE
```

## Data Import

Load raw mzML data of the pilot study, this includes data of venous blood of 10
healthy individuals loaded in the following BuS devices: M-VAMS, C-qDBS, W-DBS,
also plasma samples.

User should create the seq_neg.xlsx where injection order, sample name and
sample_type have to be included. Also neg_ data need to have all the files
in .mzml that where included in the excel file.

```{r}
#| eval: true
# Load neg data
seq_neg <- read_xlsx(paste0("seq_neg.xlsx"),
                 col_names = TRUE) |> as.data.frame()

spectra_files_neg <- file.path("NEG_data", seq_neg$`file_name`)
```

Store the MS data into a SQLite database.

```{r}
#| eval: false
## Only run once
library(MsBackendSql)
library(RSQLite)
dbf <- normalizePath(file.path("NEG_data", "pilot_neg.sqlite"))
if (file.exists(dbf)) unlink(dbf)
con <- dbConnect(SQLite(), dbf)
createMsBackendSqlDatabase(dbcon = con, spectra_files_neg)
dbDisconnect(con)
```


```{r}
#| eval: true
library(MsBackendSql)
library(RSQLite)
dbf <- normalizePath(file.path("NEG_data", "pilot_neg.sqlite"))

s <- Spectra(dbname = dbf, source = MsBackendOfflineSql(), drv = SQLite())
s$base_file <- basename(s$dataOrigin)
mse_neg <- MsExperiment(spectra = s, sampleData = seq_neg)
mse_neg <- linkSampleData(
  mse_neg, with = "sampleData.file_name = spectra.base_file")
```

Below we show how to load from MetaboLights as an alternative to mzml files.

```{r}
#| eval: false
param <- MetaboLightsParam(mtblsId = "MTBLS10722",
                           assayName = "a_MTBLS10722_LC-MS_negative_reverse-phase_metabolite_profiling.txt",
                           filePattern = ".mzML")

mse_neg <- readMsObject(MsExperiment(),
                        param,
                        keepOntology = FALSE,
                        keepProtocol = FALSE,
                        simplify = TRUE)
```

Some variable mapping needs to be done when importing from MetaboLights

```{r}
#| eval: false
# Filter for the pilot study
mse_neg <- mse_neg[sampleData(mse_neg)$`Factor Value[Cohort]` == "Cohort_study"]

# Let's rename the column for easier access
sad <- sampleData(mse_neg)
sad <- sad[,c(1, 3, 8, 10)]
colnames(sad) <- c("sample_name", "file_name",  "sample_type", "injection_index")

#  Add injection index column
sad$polarity <- "neg"
sampleData(mse_neg) <- sad

## remove blank samples
mse_neg <- mse_neg[sampleData(mse_neg)$sample_type != "Blank"]

## reorder the sample based on injection order
mse_neg <- mse_neg[order(sampleData(mse_neg)$injection_index), ]

##only phili - subset to make analysis faster
# mse_neg <- mse_neg[sampleData(mse_neg)$sample_type == "QC"]
```

\[Phili: Here i would explain maybe some column names so that other people
understand what is in them\]

Adjust the number of cores based on your machine. Suggested to use 1 or 2 below
your higher capacity.

```{r}
#| eval: true
cores_nb <- 4  # Ajust here
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(cores_nb))
} else {
    register(SnowParam(cores_nb))
}
```

Define color per sample_type: M-VAMS, C-qDBS, W-DBS, QC, Plasma

```{r}
#| eval: true
col_sample_type <- brewer.pal(9, name = "Set1")[c(9, 2, 1, 3, 4)]
names(col_sample_type) <- c("QC",   # grey
                          "M-VAMS", # blue
                          "W-DBS",  # red
                          "C-qDBS", # green
                          "Plasma") # purple
col_sample <- col_sample_type[sampleData(mse_neg)$sample_type]
```

# Dataset investigation

## Chromatographic data visualization

BPC plot of all samples (colored by device), including QCs.

```{r}
#| eval: true
#| fig-cap: "BPC of all samples colored by phenotype"
#' Extract and plot BPC for full data
bpc <- chromatogram(mse_neg, aggregationFun = "max")

#png("BPC_neg.png", width = 2000, height = 1500, res = 300)

plot(bpc, col = paste0(col_sample, 80), main = "BPC ", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6,
       horiz = TRUE,
       bty = "n")

#dev.off()
```

Filter the data based on the retention time observed on the BPC. In this data set we filter
from 10 to 800 sec.

```{r}
#| eval: true
#' Filter the data based on retention time
mse_neg <- filterRt(mse_neg, c(10, 800))
```

```{r}
#| eval: true
#| fig-cap: "BPC after filtering retention time."
# Plot BPC after filtering
bpc <- chromatogram(mse_neg, aggregationFun = "max") # ERROR when , aggregationFun = "max" ## it works for me, maybe it's your mac screwing up like last time ?

plot(bpc, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6,
       horiz = TRUE, bty = "n")
```


```{r}
#| eval: false
#| fig-cap: "Heatmap of TIC similarities. All plasma samples are group together."
#' Total ion chromatogram
tic <- chromatogram(mse_neg, aggregationFun = "sum") |>
  bin(binSize = 8)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

rownames(ticmap) <- colnames(ticmap) <- sampleData(mse_neg)$sample_name
ann <- data.frame(sample_type = sampleData(mse_neg)[, "sample_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
#png("heatmap_neg.png", width = 2000, height = 1500, res = 300)

pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = col_sample_type),
         show_colnames = FALSE,
         show_rownames = FALSE)

#dev.off()
```

## EICs

Upload the list of the reference lipids (internal standards form SPLAH LIPIDOMIX and also some endogenous compounds), that will be used to make retention time shift correction. Retention time must be in seconds.

```{r, r load-is}
#| eval: true
#| warning: false
#| tbl.cap: "Set of reference lipids, including deuterated lipids from SPLASH LIPIDOMIX and some deuterated compounds that are in high abundance in plasma."
#' Load list of Internal standard SPLASH LIPIDOMIX
intern_standard <- read_xlsx("neg_splashlipidomix_list.xlsx") |>
    as.data.frame()
intern_standard$rtmin <- intern_standard$RT - 30
intern_standard$rtmax <- intern_standard$RT + 20
intern_standard$mzmin <- intern_standard$mz - 0.008
intern_standard$mzmax <- intern_standard$mz + 0.008
rownames(intern_standard) <- intern_standard$short_name

# Extract EICs for the list
eic_is <- chromatogram(
    mse_neg,
    rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
    mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

#' Add internal standard metadata
fData(eic_is)$mz <- intern_standard$mz
fData(eic_is)$rt <- intern_standard$RT
fData(eic_is)$rtmin <- intern_standard$rtmin
fData(eic_is)$rtmax <- intern_standard$rtmax
fData(eic_is)$short_name <- intern_standard$short_name
rownames(fData(eic_is)) <- intern_standard$short_name

fdata <- fData(eic_is)

fData(eic_is)[c("dFA", "dPI"),
              c("short_name", "mz", "rt", "rtmin", "rtmax")] |>
    kable(format = "pipe")
```

We below plot the EICs for reference lipids (17 in total).The dotted red line represents the expected retention time of the peak, showing that the retention time shift varies depending on the isocratic range.

```{r}
#| eval: true
#| fig.cap: "EIC of all reference lipids."
# Loop to go through all rows of the eic_is object and plot independent

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 abline(v = fData(eic_is)$rt[i], col = "red", lty = 3)
}
```

```{r}
#| eval: true
#| fig.cap: "Example: EIC of three selected deuterated lipids."
# Chromatogram example of three lipids from the list that was imported before
eic_FA <- eic_is["dFA"]
eic_PI <- eic_is["dPI"]

#png("EIC_exampledeuteratedlipids.png", width = 2000, height = 1500, res = 300)

par(mfrow = c(1, 2), mar = c(4, 2, 2, 0.5))

plot(eic_FA, main = fData(eic_FA)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_FA)$rt, col = "red", lty = 3)

plot(eic_PI, main = fData(eic_PI)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
     col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_PI)$rt, col = "red", lty = 3)


legend("topright", col = col_sample_type, legend = names(col_sample_type),
       lty = 1,
       bty = "n")

#dev.off()
```

# Data preprocessing

## Chromatographic peak detection

We use the deuterated LysoPC of the first sample to extract the *m/z*,
calculate the absolute difference and expressed in ppm.

```{r}
#| eval: true
#' Restrict the data to signal from dLysoPC in the first sample
lpc <- mse_neg[1L] |>
  spectra() |>
  filterRt(rt = c(209, 230)) |>
  filterMzRange(mz = fData(eic_FA)["dFA", c("mzmin", "mzmax")])

#' Show the number of peaks per m/z filtered spectra
lengths(lpc)
```

```{r}
#| eval: true
#' Calculate the difference in m/z values between scans
mz_diff <- lpc |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(lpc))))
```

We therefore, choose a value close to the maximum within this range for
parameter `ppm`, i.e., 10 ppm, otherwise work with 2 `ppm` will be to low.

From previous observation of some of our ISTD we estimate the average peak
width to be around 4 to 8 seconds. We therefore set the `peakwidth` parameter
to `c(4, 8)`. \[Here link to figure above\]

We can now perform the chromatographic peak detection with the adapted settings
on our EICs, and test the parameters in some Lipids from the list. In this case
deuterated LysoPC, Sphingomelyn and Carnitine.

```{r}
#| eval: true
#' Parameters adapted for chromatographic peak detection on EICs.
#param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
#                          snthresh = 2) # these were the initial parameters
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2, firstBaselineCheck = FALSE)
```

```{r}
#| eval: `run_plot`

#' Evaluate on the dLysoPC ion
FA_test <- findChromPeaks(eic_FA, param = param_pd)
chromPeaks(FA_test)

#' Evaluate on the PI ion
PI_test <- findChromPeaks(eic_PI, param = param_pd)
chromPeaks(PI_test)
```

With the customized parameters, a chromatographic peak was detected in each
sample. Below, we use
the [`plot()`](https://rdrr.io/r/graphics/plot.default.html)  function on the
EICs to visualize these results.

```{r}
#| eval: `run_plot`
#| fig.cap: "EIC of two selected deuterated lipids after applying the optimized param selection."
#' Plot test chromatogram for deuterated LysoPC, Sphingomielyn and Carnitine.
par(mfrow = c(1, 2))
plot(FA_test, main = fData(FA_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(FA_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
abline(v = fData(eic_FA)$rt, col = "red", lty = 3)

plot(PI_test, main = fData(PI_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(PI_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
abline(v = fData(eic_PI)$rt[i], col = "red", lty = 3)


legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

We can see a peak seems to be detected in each sample for both ions. This
indicates that our custom settings are suitable for our dataset. We now proceed
and apply them to the entire dataset, extracting EICs again for the same ions
to evaluate and confirm that chromatographic peak detection worked as expected.

```{r}
#| eval: true
#' Using the same settings, but with default snthresh
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2, firstBaselineCheck = FALSE)

mse_neg <- findChromPeaks(mse_neg, param = param_pd, chunkSize = cores_nb)

eics_is_noprocess <- eic_is
eic_is <- chromatogram(mse_neg,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
```

```{r}
#| eval: true
fData(eic_is) <- fData(eics_is_noprocess)
```

```{r}
# Plot to evaluate the chromatographic peak detection results.
rownames(fData(eic_is)) <- intern_standard$short_name

eic_FA <- eic_is["dFA"]
eic_PI <- eic_is["dPI"]

## Here add some other double peak compound that we failed to identify before.
```

```{r}
#| eval: true
#| fig.cap: "EIC of three selected deuterated lipids after applying the chromatographic peak detection on the entire dataset."
par(mfrow = c(1, 2))
plot(eic_FA, main = fData(FA_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(FA_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_PI, main = fData(PI_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(PI_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

After applying the cinematographic peak detection over the entire dataset, only
the first two peaks have been detected properly in all sample, but this does
not happen with the Carnitine (not even after changing the signal to noise
ratio, or opening the window for the m/z or rt)

## Refinment of chromatographic peaks

```{r}
#| eval: false
#' set up the parameter
param_msp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.0015,
                                        minProp = 0.75)
```

```{r}
#| eval: false
#' Apply on whole dataset
mse_neg <- refineChromPeaks(mse_neg, param = param_msp, chunkSize = cores_nb)
```

```{r}
#| eval: false
chromPeakData(mse_neg)$merged |>
                         table()
```

Evaluate the results of the chromatographic peak detection on EICs of e.g.
internal standards or other compounds/ions known to be present in the samples.
Here we update our EICs objects for the internal standards with the results of
the chromatographic peak detection and refinement.

```{r}
#| eval: false
eics_is_chrompeaks <- eic_is

eic_is <- chromatogram(mse_neg,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fData(eics_is_chrompeaks)

eic_deuteratedLysoPC <- eic_is["dFA", ]
eic_deuteratedSM <- eic_is["dPI", ]
```

### Evaluate peak detection results

```{r}
#| eval: false
# Define output directory (create if it doesn't exist)
outdir <- "neg_peak_detection_ref_lipid"
if (!dir.exists(outdir)) dir.create(outdir)

# Loop through each row and save plots as PNG
for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i, ]
  filename <- file.path(outdir, paste0(rownames(fData(eic_is))[i], ".png"))
  png(filename, width = 800, height = 600, res = 120)
  plot(
    eic,
    main = rownames(fData(eic_is))[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  # abline(v = median(rtime(eic)), col = "red", lty = 3) ## i have fixed that 3 times now please check the pos code.
  dev.off()
}
```

All the compound plots in `eic_is` are now saved locally, these should be
checked before continuing. Each peak should be detected in most samples (if not
all). Also ensure that all the EICs that have 2 peak are properly separated.
Discuss with Johannes to see how to prevent merge. !!!

Number of chromatographic peaks per sample.

```{r, eval= TRUE}
#| tbl.cap: "Number of chromatographic peaks detected per sample."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(
  sample_name = sampleData(mse_neg)$sample_name,
  peak_count = as.integer(table(chromPeaks(mse_neg)[, "sample"]))) |>
  t() |>
  kable(format = "pipe")
```

A similar number of chromatographic peaks was identified within the various
samples of the data set. In all cases W-DBS has the lower number of features.
Which makes sense as less volume was injected therefore features of lower\
intensity would not be seen in W-DBS.

## Retention time alignment

To evaluate the presence of a shift we extract and plot below the BPC from the
QC samples.

```{r}
#| eval: false
#| fig-cap: "BPC of QC samples."
#' Get QC samples
QC_samples <- sampleData(mse_neg)$sample_type == "QC"

#' extract BPC
mse_neg[QC_samples] |>
    chromatogram(aggregationFun = "max", chromPeaks = "none") |>
    plot(col = col_sample_type["QC"], main = "BPC of QC samples") |>
    grid()
```

Initial correspondence analysis. Parameters do not need to be fully optimized.
We base it on the QC samples.

```{r, eval=TRUE}
param_c <- PeakDensityParam(
  sampleGroups = sampleData(mse_neg)$sample_type == "QC",
  minFraction = 0.9,
  binSize = 0.01, ppm = 10,
  bw = 2)
mse_neg <- groupChromPeaks(mse_neg, param = param_c)
```

```{r}
#| eval: false
#| fig-cap: "Initial correspondence analysis based on QC samples."
plotChromPeakDensity(
    eic_FA, param = param_c,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_FA)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_FA)[, "sample"]], 20),
    peakPch = 16)
```

Below we perform the alignment of our data set based on retention times of
anchor peaks defined in the subset of QC samples.

```{r, eval=TRUE}
#' Define parameters of choice
subset <- which(sampleData(mse_neg)$sample_type == "QC")
param_c <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                           subsetAdjust = "average",
                           subset = subset)

#' Perform the alignment
mse_neg <- adjustRtime(mse_neg, param = param_c)
```

Alignment adjusted the retention times of all spectra in the data set, as well
as the retention times of all identified chromatographic peaks.

```{r, eval=run_plot}
#| fig-cap: "Retention time alignment results."
#| label: fig-alignment-results
#' Visualize alignment results
plotAdjustedRtime(mse_neg, col = paste0(col_sample, 80), peakGroupsPch = 1)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

Only little adjustments needed to be performed (shifts of at maximum 1.5 second
as can be seen in Figure @fig-alignment-results).

We can also compare the BPC before and after alignment.

```{r, eval = TRUE}
#' Get data before alignment
mse_neg_raw <- dropAdjustedRtime(mse_neg)

#' Apply the adjusted retention time to our dataset
mse_neg <- applyAdjustedRtime(mse_neg)
```

```{r}
#| eval: false
#| fig-cap: "BPC before and after alignment."
#' Plot the BPC before and after alignment
par(mfrow = c(2, 1), mar = c(2, 1, 1, 0.5))
chromatogram(mse_neg_raw, chromPeaks = "none") |>
    plot(main = "BPC before alignment", col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)

chromatogram(mse_neg, chromPeaks = "none") |>
    plot(main = "BPC after alignment",
         col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)
```

There is no shift observed.

We next evaluate the impact of the alignment on the EICs of the selected
internal standards. We thus below first extract the ion chromatograms after
alignment, and subsequently plot them.

```{r}
#| eval= false
#' Extract the EICs for the test ions before alignment
old_eic_FA <- eic_is["dFA"]
old_eic_PI<- eic_is["dPI"]

#' Update the EICs
eic_is <- chromatogram(mse_neg,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fdata

#' Extract the EICs for the test ions
eic_FA <- eic_is["dFA"]
eic_PI <- eic_is["dPI"]
```

```{r}
#| fig-cap: "EIC of three selected deuterated lipids before and after alignment."
par(mfrow = c(2, 2), mar = c(4, 4.5, 2, 1))

plot(old_eic_FA, main = "FA before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dFA", "RT"], col = "red", lty = 3)

plot(eic_FA, main = "FA after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dFA", "RT"], col = "red", lty = 3)

plot(old_eic_PI, main = "PI before alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dPI", "RT"], col = "red", lty = 3)

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")

plot(eic_PI, main = "PI after alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dPI", "RT"], col = "red", lty = 3)
```

In addition to a visual inspection of the results, we also evaluate the impact
of the alignment by comparing the variance in retention times for internal
standards before and after alignment.

```{r}
#' Extract the matrix with all chromatographic peaks and add a column
#' with the ID of the chromatographic peak
chrom_peaks <- chromPeaks(mse_neg) |> as.data.frame()
chrom_peaks$peak_id <- rownames(chrom_peaks)

#' Define the parameters for the matching and filtering of the matches
p_1 <- MzRtParam(ppm = 50, toleranceRt = 10)
p_2 <- SingleMatchParam(duplicates = "top_ranked", column = "target_maxo",
                        decreasing = TRUE)

#' Iterate over samples and identify for each the chromatographic peaks
#' with similar m/z and retention time than the onse from the internal
#' standard, and extract among them the ID of the peaks with the
#' highest intensity.
intern_standard_peaks <- lapply(seq_along(mse_neg), function(i) {
    tmp <- chrom_peaks[chrom_peaks[, "sample"] == i, , drop = FALSE]

     # Convert to S4 DataFrames
    query  <- DataFrame(intern_standard)
    target <- DataFrame(tmp)


    mtch <- matchValues(query, target,
                        mzColname = c("mz", "mz"),
                        rtColname = c("RT", "rt"),
                        param = p_1)
    mtch <- filterMatches(mtch, p_2)
    mtch$target_peak_id
}) |>
    do.call(what = cbind)
```

We have now for each internal standard the ID of the chromatographic peak in
each sample that most likely represents signal from that ion. We can now
extract the retention times for these chromatographic peaks before and after
alignment.

```{r}
#' Define the index of the selected chromatographic peaks in the
#' full chromPeaks matrix
idx <- match(intern_standard_peaks, rownames(chromPeaks(mse_neg)))

#' Extract the raw retention times for these
rt_raw <- chromPeaks(mse_neg_raw)[idx, "rt"] |>
    matrix(ncol = length(mse_neg_raw))

#' Extract the adjusted retention times for these
rt_adj <- chromPeaks(mse_neg)[idx, "rt"] |>
    matrix(ncol = length(mse_neg_raw))
```

We can now evaluate the impact of the alignment on the retention times of
internal standards across the full data set:

```{r}
#| fig-cap: "Retention time variation of internal standards before and after alignment."
list(all_raw = rowSds(rt_raw, na.rm = TRUE),
     all_adj = rowSds(rt_adj, na.rm = TRUE)
     ) |>
    vioplot()
grid()
```

## Correspondence

The goal of the correspondence analysis is to identify chromatographic peaks
that originate from the same types of ions in all samples of an experiment and
to group them into LC-MS *features*.

```{r}
#' Optimized parameter for the grouping and apply them to the test ions BP2
#' fix grouping, NA for QCs
group_mse <- sampleData(mse_neg)$sample_type
group_mse[group_mse =="QC"] <- NA
group_mse <- as.factor(group_mse)
#' group for phili
# group_mse <- rep(1, length(mse_neg))
param_cor <- PeakDensityParam(sampleGroups = group_mse, bw = 2)
param_cor
```

```{r}
plotChromPeakDensity(
    eic_FA, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_FA)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_FA)[, "sample"]],
                    20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated LysoPC.

```{r}
plotChromPeakDensity(
    eic_PI, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_PI)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_PI)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated PI.


We can observe for the first two ions that the peaks are grouped accurately
into a single feature. So we proced to define the settings for the param\_ cor
and apply them to the whole data.

```{r, eval=TRUE}
#' Define the settings for the param_cor
param_cor <- PeakDensityParam(sampleGroups = group_mse,
                              minFraction = 0.5, binSize = 0.01, ppm = 10,
                              bw = 2)

#' Apply to whole data
mse_neg <- groupChromPeaks(mse_neg, param = param_cor)
```

Evaluation of the correspondence results in other EIC in the IS list similar to
the deuterated FA.

```{r}
#| eval: false
#' Extract chromatogram for an m/z similar to the one of the labeled
chr_test <- chromatogram(mse_neg,
                         mz = as.matrix(intern_standard["FA",
                                                        c("mzmin", "mzmax")]),
                         rt = c(370, 390),
                         aggregationFun = "max")
plotChromPeakDensity(
    chr_test,param = param_cor, simulate = FALSE,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(chr_test)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(chr_test)[, "sample"]], 20),
    peakPch = 16)
```

This corresponds to non-deuterated LysoPC and we can see that it can
distinguish between the two peaks.

```{r}
#| eval: false
#| fig-cap: "Evaluation of the correspondance results in  EICs in the IS list."
# Loop to go through all rows of the eic_is object and plot independent after correspondance analysis

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 abline(v = fData(eic_is)$rt[i], col = "red", lty = 3)
}
```

## Gap filling

We next examine how prevalent missing values are in our present dataset:

```{r}
#' Percentage of missing values.
sum(is.na(featureValues(mse_neg))) /
    length(featureValues(mse_neg)) * 100

idx_qc <- sampleData(mse_neg)$sample_type == "QC"
#' Percentage of missing values
sum(is.na(featureValues(mse_neg[idx_qc, keepFeatures = TRUE]))) /
    length(featureValues(mse_neg[idx_qc, keepFeatures = TRUE])) * 100
```

We can observe a substantial number of missing values values in our dataset.

Fill the missing values in the entire data set.

```{r}
#' Fill in the missing values in the whole dataset
mse_neg <- fillChromPeaks(mse_neg, param = ChromPeakAreaParam(),
                          chunkSize = cores_nb)

#' Percentage of missing values after gap-filling
sum(is.na(featureValues(mse_neg))) /
    length(featureValues(mse_neg)) * 100
```

```{r}
#| fig-cap: "Detected vs. filled-in signal."

QC_samples <- which(sampleData(mse_neg)$sample_type == "QC")
#' Get only detected signal in QC samples
vals_detect <- featureValues(mse_neg, filled = FALSE)[, QC_samples]

#' Get detected and filled-in signal
vals_filled <- featureValues(mse_neg)[, QC_samples]

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)

#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080")
grid()
abline(0, 1)
```

```{r}
#' extract results as a SummarizedExperiment
res <- quantify(mse_neg, method = "sum", filled = FALSE)
res

assays(res)$raw_filled <- featureValues(mse_neg, method = "sum",
                                        filled = TRUE)
```

```{r}
# Filter the res object setting a threshold of 1.5*10^3
res_filtered <- assay(res)[rowMeans(assay(res), na.rm = TRUE) > 1500, ]

sum(!rowMeans(assay(res), na.rm = TRUE) > 1500)
```

## Save mse_neg and res object

```{r}
#| eval: true
# save the mse_neg preprocessed object
spectra(mse_neg) <- setBackend(spectra(mse_neg), MsBackendMemory())
spectra(mse_neg) <- applyProcessing(spectra(mse_neg))
saveMsObject(mse_neg, AlabasterParam(path = file.path("objects/preprocessed_mse_neg")))
save(mse_neg, file = file.path("objects/preprocessed_mse_neg.RData")) ## just in case 
# save the SummarizedExperiemnt
res@metadata <- list()
saveObject(res, file.path("objects/preprocessed_res_neg"))
```

The chunk before is if preprocessing was already run and we just want to load
the objects again and continue the rest of the analysis from here.
```{r}
#| eval: false
#mse_neg <- readMsObject(XcmsExperiment(), AlabasterParam(path = file.path("objects/preprocessed_mse_neg")))
# the saving did not work for some reason. fro now just saved as Rdata object
load("objects/preprocessed_mse_neg.RData")
res <- readObject(file.path("objects/preprocessed_res_neg"))

intern_standard <- read_xlsx("neg_splashlipidomix_list.xlsx") |>
    as.data.frame()
intern_standard$rtmin <- intern_standard$RT - 30
intern_standard$rtmax <- intern_standard$RT + 20
intern_standard$mzmin <- intern_standard$mz - 0.008
intern_standard$mzmax <- intern_standard$mz + 0.008
rownames(intern_standard) <- intern_standard$short_name

# Extract EICs for the list
eic_is <- chromatogram(
    mse_neg,
    rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
    mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

#' Add internal standard metadata
fData(eic_is)$mz <- intern_standard$mz
fData(eic_is)$rt <- intern_standard$RT
fData(eic_is)$rtmin <- intern_standard$rtmin
fData(eic_is)$rtmax <- intern_standard$rtmax
fData(eic_is)$short_name <- intern_standard$short_name
rownames(fData(eic_is)) <- intern_standard$short_name

fdata <- fData(eic_is)
```

# Database correction

```{r}
#' Default parameter for the grouping and apply them to the test ions BPC
group <- eic_is$sample_type
group[group =="QC"] <- NA

# for phili
# group <- rep(1, ncol(eic_is))
param1 <- PeakDensityParam(sampleGroups = group , minFraction = 0.5,
                           binSize = 0.01, ppm = 10,
                           bw = 2)

eic_is_corr <- groupChromPeaks(eic_is, param = param1)

exp_rt <- vector("numeric", length = nrow(eic_is_corr))
dr <- "./ref_lipid_image_neg/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(eic_is_corr))) {
  table <- featureDefinitions(eic_is_corr[i,])
  if (!nrow(table)) {
    message("compound not detected: ", fData(eic_is_corr)$short_name[i])
    exp_rt[i] <- NA
    next
  }
  png(filename = paste0(dr, "feature_compound_", i, ".png"))
  plot(eic_is_corr[i,],
       main = rownames(fData(eic_is_corr))[i],
       cex.axis = 0.8,
       cex.main = 0.8)
  abline(v = fData(eic_is_corr)$rt[i], col = "red", lty = 3)
  dev.off()
  if (nrow(table) > 1) {
    idx_max <- which.max(rowSums(featureValues(eic_is_corr[i, ])))
    rt <- table[idx_max, ]$rtmed
  } else rt <- table$rtmed
  exp_rt[i] <- rt
}


## before going on, check the plots. the line should overlap what we expect.

# below is the reference lipid table
ref_rt <- intern_standard$RT # is the ref RT

# res is the exp RT
intern_standard$exp_rt <- exp_rt
diff_rt <- intern_standard$RT - intern_standard$exp_rt
quantile(diff_rt)

## Fit the linear model
fit <- lm(exp_rt ~ poly(ref_rt, 6, raw = TRUE) )
summary(fit)
plot(fit)
```

Sara will check if the shift see m to be similar to what she has or not.I did
not apply any RT shift correction, because my data had not to much shift on the
RT and we were able to do the match against the library. The average difference
is 9.6 seconds

Library fix

```{r}
lipid_database <- read_xlsx("LipidDatabase_R.xlsx", sheet = 5L)

lipid_database <- lipid_database[lipid_database$rank != 0, ]
lipid_database$rt_sd <- as.numeric(lipid_database$`RT ESI(–) (min)`) * 60
lipid_database <- lipid_database[order(lipid_database$rt_sd), ]

get_rt_rank <- function(x) {
  match(x, sort(unique(x)))
}

rt_ranks <- ave(
  lipid_database$`RT ESI(–) (min)`,
  lipid_database$`lipid name`,
  FUN = get_rt_rank
)

lipid_database$lipid_name_unique <- paste(
  lipid_database$`lipid name`,
  rt_ranks,
  sep = "_"
)

lipid_database$adduct <- gsub("Adduct", "", lipid_database$Adduct)

lookup_ads <- data.frame(
  name = c(
    "[M-H]-",
    "[M+CH3COO]-",
    "[M+HCOO]-",
    "[M+Cl]-",
    "[M-H+(CH3COONa)]-",
    "[M+HCOO+(CH3COONa)]-",
    "[M+CH3COO+(CH3COONa)]-",
    "[M+CH3COO+(CH3COONa)2]-",
    "[M+CH3COO+(CH3COONa)3]-"
  ),
  mass_multi = 1.0000000,
  mass_add = c(
    -1.007276,
    59.01385,
    44.99820,
    34.96940,
    81.00551,
    127.0110,
    141.0267,
    223.0395,
    305.0524
  ),
  formula_add = c(
    "C0",
    "CH3COO",
    "HCOO",
    "Cl",
    "C2H3O2Na",
    "C3H4O4Na",
    "C4H6O4Na",
    "C6H9O6Na2",
    "C8H12O8Na3"
  ),
  formula_sub = c(
    "H", "C0", "C0", "C0", "H", "C0", "C0", "C0", "C0"
  ),
  charge = -1,
  positive = FALSE,
  stringsAsFactors = FALSE
)

lipid_database$fix_form <- gsub("D([0-9]+)", "[2H\\1]", lipid_database$`MOLECULAR FORMULA`)
lipid_database <- head(lipid_database, -11)

custom_ads <- lookup_ads[match(lipid_database$adduct, lookup_ads$name), ]
row.names(custom_ads) <- NULL
sladduct <- split(custom_ads, seq(nrow(custom_ads)))

form <- mapply(
  FUN = adductFormula,
  lipid_database$fix_form,
  adduct = sladduct,
  standardize = TRUE
)

form <- gsub("\\[2H([0-9]+)\\]", "D\\1", form)

lipid_database$adduct_formula <- unname(gsub("^\\[|\\].*$", "", form))

is.unsorted(lipid_database$rt_sd) # check rtime are properly sorted. We want FALSE

rt_sd <- lipid_database$rt_sd

## finding values that need to be extrapolated.
in_range_idx <- which(rt_sd >= min(fit$fitted.values) &
                      rt_sd <= max(fit$fitted.values))

nrow(lipid_database) - length(in_range_idx)
## Adjust library retention times
corrected_rt <- rep(NA, length = nrow(lipid_database))
corrected_rt[in_range_idx] <- predict(
    fit, newdata = data.frame(ref_rt = rt_sd[in_range_idx]))
## Adjust values outside the fitted range
idx <- which(rt_sd < min(fit$fitted.values))
lidx <- length(idx)
if (lidx)
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[lidx + 1L] - corrected_rt[lidx + 1L])
idx <- which(rt_sd > max(fit$fitted.values))
if (length(idx))
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[idx[1L] - 1L] - corrected_rt[idx[1L] - 1L])

lipid_database$rt_adjusted <- corrected_rt

## difference between original and adjusted retention times
quantile(abs(rt_sd - corrected_rt))
```


```{r}
#| fig-cap: "lipid database rt vs. rt adjusted base on the model."
# scatterplot lipid_name and rt(seconds)
lipid_database$mz <- as.numeric(lipid_database$mz)

plot(y = lipid_database$rt_sd, x =lipid_database$rt_adjusted, type = "l",
     xlab = "Adjusted rt",
     ylab = "Database rt")
points(intern_standard$RT ~ intern_standard$exp_rt, col = "red", pch = 16, cex = 1)
abline(v = min(fit$fitted.values), col = "blue", lwd = 1, lty = 2)
abline(v = max(fit$fitted.values), col = "blue", lwd = 1, lty =2)
grid()
abline(0, 1, lty = 2, col = "#00000040")
```
The Black line is build with the compounds of the database and the red dots are the reference
lipids (7 deuterated)


# Annotate/ match

## rank1 matching

Match features (query) against (aligned) database (target).

```{r}
## Prepare Query (Experimental Features)
query <- as.data.frame(rowData(res))
query$feature_id <- rownames(query)

## Prepare Target (Lipid Database - Rank 1 only)
lipid_db_r1 <- lipid_database[lipid_database$rank == 1, ]
target <- as.data.frame(lipid_db_r1)
target$mz <- as.numeric(target$mz)

## Configure parameters
## Note on PPM: 20 ppm is standard for Orbitrap/Q-TOF. Only lower to 5-10 if
## you have internal calibration or very high-res data.
param_match <- MzRtParam(
    tolerance = 0.001,      # Absolute m/z tolerance
    ppm = 20,               # Relative m/z tolerance
    toleranceRt = 20        # +/- 20 seconds (Total window size)
)

mtch <- matchValues(query, target,
                    param = param_match,
                    mzColname = c("mzmed", "mz"),
                    rtColname = c("rtmed", "rt_adjusted"))

## Filter object to keep only matches
mtch <- mtch[whichQuery(mtch)]

## Extract data and clean IDs
mtched_data <- as.data.frame(matchedData(mtch))
mtched_data$feature_id <- sub("\\..*$", "", rownames(mtched_data))

## Quick stats
cat("Unique Features Matched:", length(unique(mtched_data$feature_id)), "\n")
cat("Database Coverage (%):",
    length(unique(mtched_data$target_lipid_name_unique)) /
    length(unique(lipid_db_r1$lipid_name_unique)) * 100, "\n")

nrow(mtched_data)
tail(mtched_data)
```

From the in total `r nrow(query)` LC-MS features `r length(whichQuery(mtch))`
were matched to at least one lipid in the database. In total
`r length(whichTarget(mtch))` lipids from the database were matched (from in
total `r nrow(target)` compounds/lipids in the database).


## Calculate isotope pattern similarity

```{r}
library(enviPat)
data(isotopes)

cat("Matches before filter:", nrow(mtched_data), "\n")

##  EXTRACT EXPERIMENTAL MS1 PATTERNS
spectra(mse_neg) <- setBackend(spectra(mse_neg), MsBackendMemory())

## Extract spectra for matched features only
sp <- featureSpectra(mse_neg, msLevel = 1L, skipFilled = TRUE,
                     features = unique(mtched_data$feature_id),
                     method = "closest_rt")

## Combine spectra per feature to get a clean representative spectrum
csp <- combineSpectra(sp, f = sp$feature_id, p = sp$feature_id,
                      peaks = "intersect", ppm = 2)

## Filter for peaks belonging to the isotope pattern
iso_spectra <- spectrapply(csp, function(z) {
    idx <- isotopologues(peaksData(z)[[1L]], ppm = 10, seedMz = z$feature_mzmed)

    # If standard isotopologues fails, try simple filtering around the seed
    if (length(idx) == 1L) {
        addProcessing(z, function(x, i = idx[[1L]], ...) x[i, , drop = FALSE]) |>
            applyProcessing()
    } else {
        filterMzValues(z, mz = z$feature_mzmed, ppm = 10, tolerance = 0) |>
            applyProcessing()
    }
}) |>
    concatenateSpectra() |>
    scalePeaks(by = max)

## CALCULATE THEORETICAL PATTERNS
## Check formulas
chem_checked <- check_chemform(isotopes, mtched_data$target_adduct_formula)

## Generate patterns (Vectorized by charge)
ip <- isopattern(
    isotopes,
    chem_checked$new_formula,
    threshold = 0.001,
    charge = -1,  ##  here hardcoed -1 for negative mode
    rel_to = 0
)

## Convert enviPat list to Spectra object for comparison
isopattern_to_spectra <- function(x) {
    df <- data.frame(msLevel = 1L, formula = names(x))
    df$mz <- lapply(x, function(z) z[, 1L])
    df$intensity <- lapply(x, function(z) z[, 2L])
    Spectra(df)

}
theoretical_spectra <- isopattern_to_spectra(ip)

## Align indices between Experimental (iso_spectra) and Theoretical (mtched_data)
## We map experimental spectra to the match table based on feature_id
match_indices <- match(mtched_data$feature_id, iso_spectra$feature_id)
mtched_data$isopeak_count <- lengths(iso_spectra)[match_indices]

## Calculate similarity
## subset theoretical to match the order/length of experimental
mtched_data$isopeak_sim <- diag(
    compareSpectra(iso_spectra[match_indices],
                   theoretical_spectra,
                   ppm = 20)
)

## Filter: Keep if >1 peak OR similarity >= 0.78
mtched_data <- subset(mtched_data, isopeak_count > 1 & isopeak_sim >= 0.78)

cat("Matches after filter:", nrow(mtched_data), "\n")
tail(mtched_data)
```

The plotting can take a bit of time (and space), so do it only if needed.

```{r}
# plot
dr <- file.path("figures/iso_pattern_check/")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

for (i in seq_len(nrow(mtched_data))) {

  cmp <- paste0(mtched_data$target_lipid_name_unique[i],
                "_", mtched_data$target_adduct[i])

  cmp <- gsub("[():/ +]", "", cmp)

  out_file <- file.path(dr, paste0(cmp, "_", iso_spectra$feature_id[i], ".png"))
  png(filename = out_file, width = 8, height = 8, units = "cm", res = 600, pointsize = 4)
  plotSpectraMirror(iso_spectra[i],
                    theoretical_spectra[i], ppm = 20,
                    main = paste(cmp, iso_spectra$feature_id[i], sep = " - "))

  dev.off()
}

## if you want to check for a specific feature/lipid, i would advise to ctrl+F
## for the feature in the folder and then visually check which lipid (in the
## file name)
```

inital filtering please explain
We decided to stay with the compounds that have at least 2 isotopic peak and also more than 0.78 similarity

```{r}
## number of lipids matching > 1 feature
sum(duplicated(paste0(mtched_data$target_lipid_name_unique,
                      mtched_data$target_Adduct, sep = "_")))

## number of features matching > 1 lipid.
sum(duplicated(mtched_data$feature_id))
```


## Matching Adducts (Rank > 1)

- Returning the max abundance for each adduct
  can be found in the featureValues(mse_neg) and the rownames are the feature ids
- a ratio of adduct detected/ total number of adducts possible for that lipid.

```{r}
mtched_data$ntch_idx <- seq_len(nrow(mtched_data))
# We select the Rank 1 matches we want to find adducts for.
link_table <- mtched_data[, c("feature_id", "target_lipid_name_unique",
                              "rtmed", "ntch_idx")]
colnames(link_table) <- c("ref_feature_id", "lipid_name_unique",
                          "rt_experimental", "ntch_idx")

# Get all potential adducts (Rank > 1) from the database
db_adducts <- lipid_database[lipid_database$rank != 1, ]

# We are creating a custom target list specific to your experimental features.
target_r2 <- merge(as.data.frame(db_adducts), link_table,
                   by = "lipid_name_unique")

# Ensure numeric type
target_r2$mz <- as.numeric(target_r2$mz)

param_adduct <- MzRtParam(
    tolerance = 0.001,
    ppm = 20,
    toleranceRt = 5
)

# Match Query (features) against our Custom Target (target_r2)
mtch_adducts <- matchValues(query, target_r2,
                            param = param_adduct,
                            mzColname = c("mzmed", "mz"),
                            rtColname = c("rtmed", "rt_experimental"))

mtch_adducts <- mtch_adducts[whichQuery(mtch_adducts)]

## Extract results
adduct_hits <- as.data.frame(matchedData(mtch_adducts))
adduct_hits$feature_id <- sub("\\..*$", "", rownames(adduct_hits))


## below also compute isotope pattern similarity for adducts found
sp <- featureSpectra(mse_neg, msLevel = 1L, skipFilled = TRUE,
                     features = unique(adduct_hits$feature_id),
                     method = "closest_rt")
csp <- combineSpectra(sp, f = sp$feature_id, p = sp$feature_id,
                      peaks = "intersect", ppm = 2)
iso_spectra <- spectrapply(csp, function(z) {
    idx <- isotopologues(peaksData(z)[[1L]], ppm = 10, seedMz = z$feature_mzmed)

    # If standard isotopologues fails, try simple filtering around the seed
    if (length(idx) == 1L) {
        addProcessing(z, function(x, i = idx[[1L]], ...) x[i, , drop = FALSE]) |>
            applyProcessing()
    } else {
        filterMzValues(z, mz = z$feature_mzmed, ppm = 10, tolerance = 0) |>
            applyProcessing()
    }
}) |>
    concatenateSpectra() |>
    scalePeaks(by = max)
chem_checked <- check_chemform(isotopes, adduct_hits$target_adduct_formula)
ip <- isopattern(
    isotopes,
    chem_checked$new_formula,
    threshold = 0.001,
    charge = - 1,  ##  hardcoded 1 could be a problem
    rel_to = 0
)
theoretical_spectra <- isopattern_to_spectra(ip)
match_indices <- match(adduct_hits$feature_id, iso_spectra$feature_id)
adduct_hits$isopeak_count <- lengths(iso_spectra)[match_indices]
adduct_hits$isopeak_sim <- diag(
    compareSpectra(iso_spectra[match_indices],
                   theoretical_spectra,
                   ppm = 20)
)

## filter
adduct_hits <- subset(adduct_hits, isopeak_count > 1 & isopeak_sim >= 0.78)
counts_per_mtch <- table(adduct_hits$target_ntch_idx)
total_count <- table(target_r2$ntch_idx)
names(total_count) <- as.character(names(total_count))

## make a ratio
ratios_per_math <- table(adduct_hits$target_ntch_idx) /
  total_count[names(table(adduct_hits$target_ntch_idx))]

# Map these counts back to the main dataframe
mtched_data$adduct_ratio <- 0
mtched_data$adduct_count <- 0
mtched_data$adduct_count[as.numeric(names(counts_per_mtch))] <- as.numeric(counts_per_mtch)
mtched_data$adduct_ratio[as.numeric(names(ratios_per_math))] <- as.numeric(ratios_per_math)

message("Total Rank 1 Matches: ", nrow(mtched_data))
message("Matches with at least one extra adduct: ", sum(mtched_data$adduct_ratio > 0))
```

I  want to check if one lipid matches multiple feautres as it could
indicate an issue with the processing above. bigger problem

```{r}
cols_amb <- c("feature_id", "target_mz", "mzmed", "target_rt_adjusted",
              "rtmed", "target_lipid_name_unique", "target_adduct",
              "ppm_error", "score_rt", "target_rank",  "score",
              "isopeak_count", "isopeak_sim", "adduct_ratio", "ntch_idx")

# Create a key for Lipid + Adduct
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")

rm_vector <- c()
for (i in key[duplicated(key)]) {
  idxs <- which(key == i)
  count <- mtched_data$adduct_ratio[idxs]
  if (which.max(count) == which.min(count)) next
  best_idx <- idxs[which.max(count)]

  # Mark all other rows for removal
  idxs_to_remove <- setdiff(idxs, best_idx)
  rm_vector <- c(rm_vector, idxs_to_remove)
}
cat("Removing ", length(rm_vector), " rows with lower adduct counts.\n")
mtched_data <- mtched_data[-rm_vector, ]

sum(duplicated(paste0(mtched_data$target_lipid_name_unique,
                      mtched_data$target_Adduct, sep = "_"))) #DELETE

key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
dupe_keys <- unique(key[duplicated(key)])
rows_to_keep <- rep(TRUE, nrow(mtched_data))

# 3. Iterate and Flag Rows to Remove
for (k in dupe_keys) {
  idxs <- which(key == k)
  scores <- abs(mtched_data$score_rt[idxs])
  best_score <- min(scores, na.rm = TRUE)
  # Logic:
  # If (Score - BestScore) > 4, it is significantly worse -> Remove it.
  # If (Score - BestScore) <= 4, it is "ambiguously close" -> Keep it.
  to_remove <- idxs[ (scores - best_score) > 4 ]

  if (length(to_remove) > 0) {
    rows_to_keep[to_remove] <- FALSE
  }
}

# 4. Apply Filter
mtched_data <- mtched_data[rows_to_keep, ]
message("Removed ", sum(!rows_to_keep), " rows with significantly worse RT scores.")
nrow(mtched_data)
```


```{r}
## check amblip again
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
is_duplicated <- key %in% key[duplicated(key)]
# Subset the dataframe
amblip <- mtched_data[is_duplicated, cols_amb]
cols_amb[!cols_amb %in% colnames(mtched_data)]
amblip
```

The table below shows the ambiguous (multi-matching) features.

```{r, results = "asis"}
## Get the features that match multiple targets
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
sum(duplicated(mtched_data$feature_id))
table_amb <- mtched_data[mtched_data$feature_id %in% fids_ambiguous, cols_amb]
table_amb <- table_amb[order(table_amb$feature_id), ]

## let's check it
table_amb

## same as before
## first based on adduct
rm_vector <- c()
for (i in unique(fids_ambiguous)) {
  idxs <- which(mtched_data$feature_id == i)
  # Get the index of the row with the maximum adduct_count
  count <- mtched_data$adduct_ratio[idxs]
  if (which.max(count) == which.min(count)) next
  best_idx <- idxs[which.max(count)]

  # Mark all other rows for removal
  idxs_to_remove <- setdiff(idxs, best_idx)
  rm_vector <- c(rm_vector, idxs_to_remove)
}
cat("Removing ", length(rm_vector), " rows with lower adduct counts.\n")
mtched_data <- mtched_data[-rm_vector, ]
nrow(mtched_data)

## then based on score_rt
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
rm_vector <- c()
for (i in unique(fids_ambiguous)) {
  idxs <- which(mtched_data$feature_id == i)
  score <- abs(mtched_data$score_rt[idxs])
  best_score <- min(score, na.rm = TRUE)

  # Logic:
  # If (Score - BestScore) > 4, it is significantly worse -> Remove it.
  # If (Score - BestScore) <= 4, it is "ambiguously close" -> Keep it.
  to_remove <- idxs[ (score - best_score) > 4 ]
  if (length(to_remove) > 0) {
    rm_vector <- c(rm_vector, to_remove)
  }
}
cat("Removing ", length(rm_vector), " rows with significantly worse RT scores.\n")
mtched_data <- mtched_data[-rm_vector, ]

## duplciate lipid:
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
is_duplicated <- key %in% key[duplicated(key)]
length(sum(is_duplicated))

## duplciated feature:
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
sum(duplicated(mtched_data$feature_id))
```

## Dealing with SM1 SM2

```{r}
# A helper to extract the "Base Name" of a lipid (ignoring the 0:0 position)
# e.g., "LPC(16:0/0:0)_1" -> "LPC(16:0)_1"
# e.g., "LPC(0:0/16:0)_1" -> "LPC(16:0)_1"
get_lipid_base <- function(names) {
  # Remove "0:0/" or "/0:0" from the string
  # This normalizes both isomers to the same string
  clean <- gsub("0:0/|/0:0", "", names)
  return(clean)
}

# A helper to check if a name is SM2 (starts with 0:0/ or contains (0:0/)
is_sm2 <- function(name) {
  grepl("0:0/", name, fixed = TRUE)
}

# A helper to check if a name is SM1 (ends with /0:0 or contains /0:0))
is_sm1 <- function(name) {
  grepl("/0:0", name, fixed = TRUE)
}

df_clean <- mtched_data
df_clean$lipid_base_group <- get_lipid_base(df_clean$target_lipid_name_unique)
df_clean$keep_row <- TRUE

unique_species <- unique(df_clean$lipid_base_group)

message("Processing ", length(unique_species), " lipid species...")

for (species in unique_species) {
  idx <- which(df_clean$lipid_base_group == species)
  sub_df <- df_clean[idx, ]
  has_sm1 <- any(is_sm1(sub_df$target_lipid_name_unique))
  has_sm2 <- any(is_sm2(sub_df$target_lipid_name_unique))

  if (has_sm1 && has_sm2) {

    # 3. Identify unique Retention Times (Features) found for this species
    unique_rts <- sort(unique(round(sub_df$rtmed, 2)))

    if (length(unique_rts) >= 2) {

      rt_early <- unique_rts[1]                  # First peak
      rt_late  <- unique_rts[length(unique_rts)] # Last peak

      for (i in idx) {
        row_rt <- round(df_clean$rtmed[i], 2)
        row_name <- df_clean$target_lipid_name_unique[i]
        if (row_rt == rt_early && is_sm1(row_name)) {
          df_clean$keep_row[i] <- FALSE
        }
        if (row_rt == rt_late && is_sm2(row_name)) {
          df_clean$keep_row[i] <- FALSE
        }
      }

    } else {
      # No Separation (Only 1 peak found)
      # We have one feature matching BOTH SM1 and SM2.
      # Rule: Keep SM1, Remove SM2.

      for (i in idx) {
        if (is_sm2(df_clean$target_lipid_name_unique[i])) {
          df_clean$keep_row[i] <- FALSE
        }
      }
    }
  }
}

final_resolved_data <- df_clean[df_clean$keep_row, ]
final_resolved_data$lipid_base_group <- NULL
final_resolved_data$keep_row <- NULL

# Check stats
message("Original rows: ", nrow(mtched_data))
message("Cleaned rows:  ", nrow(final_resolved_data))

# Update main matched data
mtched_data <- final_resolved_data
mtched_data$ntch_idx <- seq_len(nrow(mtched_data))
```


## final amb check

```{r}
## lipid amb
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
is_duplicated <- key %in% key[duplicated(key)]
amblip <- mtched_data[is_duplicated, cols_amb]
amblip
amblip$keep_row <- TRUE

## fts amb
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
table_amb <- mtched_data[mtched_data$feature_id %in% fids_ambiguous,
                         cols_amb]
table_amb <- table_amb[order(table_amb$feature_id), ]
table_amb
table_amb$keep_row <- TRUE

library(writexl)
write_xlsx(table_amb, path = "feature_ambiguity_resolution.xlsx")
write_xlsx(amblip, path = "lipid_ambiguity_resolution.xlsx")
```

please fix the ambiguity manually in the excel files above by changing to
FALSE the rows you want to remove.
The lipid ambiguity must be a 1:1 match.
the feature can be a 1:many match but we will then just merge the annotation into one.

```{r}
# Remove rows based on manual curation
# Lipid Ambiguity
lipid_amb_res <- read_excel("lipid_ambiguity_resolution.xlsx")
mtch_to_be_rm <- lipid_amb_res$ntch_idx[!lipid_amb_res$keep_row]
mtched_data <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm, ]
# Feature Ambiguity
feature_amb_res <- read_excel("feature_ambiguity_resolution.xlsx")
mtch_to_be_rm_ft <- feature_amb_res$ntch_idx[!feature_amb_res$keep_row]
mtched_data <- mtched_data[!mtched_data$ntch_idx %in% mtch_to_be_rm_ft, ]

## check for duplicates again
mtched_data$nb_annotations <- 1
key <- paste(mtched_data$target_lipid_name_unique,
             mtched_data$target_adduct, sep = "_")
is_duplicated <- key %in% key[duplicated(key)]
if (any(is_duplicated)) {
  warning("There are still duplicated Lipid + Adduct entries after manual curation. Please fix this")
} else {
  message("No duplicated Lipid + Adduct entries remain.")
}
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
if (length(fids_ambiguous) > 0) {
  warning("There are still duplicated Feature entries after manual curation. They will be merged into one annotation.")
  # Merge annotations for duplicated features
  for (i in unique(fids_ambiguous)) {
    idxs <- which(mtched_data$feature_id == i)
    if (length(idxs) > 1) {
      # Combine lipid names and adducts
      combined_lipid_names <- paste(unique(mtched_data$target_lipid_name_unique[idxs]), collapse = "; ")
      combined_adducts <- paste(unique(mtched_data$target_adduct[idxs]), collapse = "; ")

      # Update the first occurrence
      mtched_data$target_lipid_name_unique[idxs[1]] <- combined_lipid_names
      mtched_data$target_adduct[idxs[1]] <- combined_adducts

      # Remove other occurrences
      mtched_data <- mtched_data[-idxs[-1], ]
      # Update the number of annotations
      mtched_data$nb_annotations[idxs[1]] <- length(idxs)
    }
  }
} else {
  message("No duplicated Feature entries remain.")
}

## final table
mtched_data
```



```{r}
## summary of coverage (accurate now)
length(unique(mtched_data$target_lipid_name_unique)) / 592 * 100
## i think this number might be a bit biased negatively because we merge annotations.
sum(mtched_data$nb_annotations)/592 * 100 # is that right ? 
```

```{r}
df_subset <- subset(mtched_data, grepl("ISTD", target_ORIGIN, ignore.case = TRUE))

# Loop to plot all the compounds in the df_subset these are the internal standards
outdir <- "ISTD_mtched_data_neg"
if (!dir.exists(outdir)) dir.create(outdir)

# Loop through each row and save plots as PNG
for (i in rownames(df_subset)) {
    #Check for the presence of "."
  if (grepl("\\.", i)) {
    original_i <- i
    i <- sub("\\..*$", "", i)  # remove everything after the first dot
    warning(sprintf(
      "Feature '%s' was also matched to something else. Using '%s' instead.",
      original_i, i
    ), call. = FALSE)
  }

  ISTD <- featureChromatograms(mse_neg, features = i)

  # Define filename using the row name
  filename <- file.path(outdir, paste0(i, ".png"))

  # Open PNG device
  png(filename, width = 800, height = 600, res = 120)

  # Plot chromatogram
  plot(
    ISTD,
    main = rownames(df_subset)[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  # abline(v = median(rtime(eic)), col = "red", lty = 3)

  # Close device
  dev.off()
}

cat("All chromatograms saved in:", outdir, "\n")
```

## Plots with the annotated compounds (mtched_data), the feature abundance of each sample, can be found in assay(res)

```{r}
## remove duplicated form the
length(unique(mtched_data$target_lipid_name_unique)) ## number of lipid matched, 283
```


TO BE DELETED

```{r}
#| eval: false
## example of peak non detected:
## it has low intensity, we need to refine preprocessing.
eg1 <- chromatogram(mse_neg, mz = c(499.035, 499.044)) #LPE(0:0/18:0)
plot(eg1)

eg2 <- chromatogram(mse_neg, mz = c(828.562, 828.569), rt = c(360, 380)) #dPI
plot(eg2)

eg3 <- chromatogram(mse_neg, mz = c(510.380, 510.388), rt = c(175, 190)) # LPC(O-0:0/18:1) // LPC(P-0:0/18:0) only 1 peak?
plot(eg3)
featureDefinitions(eg3)

eg4 <- chromatogram(mse_neg, mz = c(929.833, 929.836), rt = c(855, 875)) #TG(53:3)
plot(eg4)
featureDefinitions(eg4) # wrong rtmin rtmax in the definition ???????

eg5 <- chromatogram(mse_neg, mz = c(787.549, 787.673), rt = c(705, 720)) #SM(d40:1) # set a ppm error on amb table is 4.1 and 787.7 m/z
plot(eg5)
featureDefinitions(eg5)

```

```{r}
# eval: false
#plot some features
a <- featureChromatograms(mse_neg, features = "FT13941") 
plot(a)
featureDefinitions(a)
a1 <- chromatogram(mse_neg, mz = c(929.833, 929.836))
plot(a1)
r <- featureData(a1)
b <- featureChromatograms(mse_neg, features = "FT03107")
par(mfrow = c(1, 2))
plot(a)
plot(b)

d <- featureChromatograms(mse_neg, features = "FT08345")
e <- featureChromatograms(mse_neg, features = "FT08851")
f <- featureChromatograms(mse_neg, features = "FT08853")
g <- featureChromatograms(mse_neg, features = "FT08344")
par(mfrow = c(2, 2))
plot(d,main = "Gráfico D")
plot(e, main = "Gráfico E")
plot(f, main = "Gráfico F")
plot(g, main = "Gráfico G")


h <- featureChromatograms(mse_neg, features = "FT02422") # intensity MG
j <- featureChromatograms(mse_neg, features = "FT02607")
par(mfrow = c(1, 2))
plot(h,main = "Gráfico H")
plot(j, main = "Gráfico J")

k <- featureChromatograms(mse_neg, features = "FT03637") # intensity PG 1400
plot(k, main = "Gráfico k")
```


## Plots with the annotated compounds (mtched_data), the feature abundance of each sample, can be found in assay(res)

-   Donut chart showing the lipid class with the total quantity of lipids
    detected

```{r}
# Donnut chart of lipid distribution
counts <- table(mtched_data$target_LIPID.CATEGORY..ABBREV.)

names(counts) <- c("Glycerolipids [GL]",
                   "Glycerophospholipids [GP]",
                   "Sphingolipids [SP]")

bar_colors <- c("#fc8d62", "#8da0cb", "#e78ac3")
labels <- counts
par(mar = c(1, 1, 2, 1))

# Plot
pie(counts,
    col = bar_colors,
    labels = labels,     # solo los números
    border = "white",
    main = "Lipid Distribution",
    cex = 0.9)

symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

# Leyend
legend("topright",
       legend = names(counts),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Lipid Category",
       cex = 0.8)
```

```{r}
# pie chart for lipid subclass. HERE ADD A NEW COLUMN WITH THE SHORT NAME
counts1 <- table(mtched_data$target_LIPID.SUBCLASS..ABBREV.)

n_cat <- length(counts1)
base_colors <- brewer.pal(9, "Set1")
bar_colors <- rep(base_colors, length.out = n_cat)

# Labels: only counts
labels <- counts1

# Adjust margins
par(mar = c(1, 1, 2, 1))

# Create pie (donut) chart
pie(counts1,
    col = bar_colors,
    labels = labels,      # show only counts
    border = "white",
    main = paste("Category Distribution (n =", sum(counts), ")"),
    cex = 0.9)

# Add white circle in center for donut effect
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

legend("topright",
       legend = names(counts1),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Categories",
       cex = 0.4)

```



-   heatmap with the significant metabolites found

-

# Data normalization
move the chunck that describes the annotated_res object
```{r}
factors <- ifelse(grepl("W", colnames(res)), 2.5,
             ifelse(grepl("P", colnames(res)), 1.08, 1))
assay(res, "raw_corr") <- sweep(assay(res), MARGIN = 2,
                                STATS = factors, FUN = "*")
assay(res, "corr_filled") <- sweep(assay(res, "raw_filled"),
                                       MARGIN = 2, STATS = factors, FUN = "*")

print("Assays now in res object:")
print(names(assays(res)))

## Check the changes in one sample (CAN BE DELETED)
# cbind(
#   original = assay(res, "raw")["FT00001", "D10W_pos.mzML"],
#   scaled   = assay(res, "raw_corr")["FT00001", "D10W_pos.mzML"]
# )


# Below i create a subset of the res object with only annotated features
mtched_features_list <- mtched_data$feature_id
annotated_res <- res[rownames(res) %in% mtched_data$feature_id, ] #only 279 id out of 299, aprox 20 duplicates
rowData(annotated_res) <- cbind(rowData(annotated_res),
                                 mtched_data[match(rownames(annotated_res),
                                                   mtched_data$feature_id),
                                             c("target_lipid.name",
                                               "target_LIPID.CATEGORY..ABBREV.",
                                               "target_LIPID.SUBCLASS..ABBREV.",
                                               "target_Adduct", "target_IS_norm"
                                             )])
## above if we have different lengths, it's because of duplicated features in the annotation table.

# Below we keep the res object but append annotation info for the annotated features
rowData(res) <- cbind(rowData(res),
                         mtched_data[match(rownames(res),
                                           mtched_data$feature_id),
                                     c("target_lipid.name",
                                       "target_LIPID.CATEGORY..ABBREV.",
                                       "target_LIPID.SUBCLASS..ABBREV.",
                                       "target_Adduct", "target_IS_norm"
                                       )])
```

## Initial quality assessment

```{r}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "raw_filled"), MARGIN = 1, na_unidis)
assays(annotated_res)$raw_filled_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

Figure X. PCA of the data colored by sample_type.Volumetric devices profiles
are clearly different from W-DBS and plasma.


Intensity evaluation

```{r}
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))

par(mar = c(0.2, 4.5, 0.2, 3))
barplot(apply(assay(res, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected peaks", xaxt = "n", space = 0.012)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sample, 80), border = col_sample)
points(colMedians(log2(assay(res, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

Less features in plasma..


```{r}
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res, "raw_filled"),
                            f = res$sample_type, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_sample, 80), ylab = "RLA",
        border = col_sample, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

RLA plot for the raw data and filled data.


## Cembio workflow

The only thing missing is this:
  - missing values imputation (KNN algorithm)
    - We usually do it using uniform distribution, but honeslty i'm not sure
     what is best.

Here before we impute the missing values we have to normalize per volume as
follows: The abundance of all W-DBS samples should be multiply per 2.5 and
Plasma samples per 1.08






HERE WAS BEFORE


```{r}
#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "corr_filled"), MARGIN = 1, na_unidis)
assays(annotated_res)$corr_filled_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "corr_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

- Normalization per internal standard (for each lipid subclass dividing per each IS)

     CAR with C17 sphinganine
     CE with Chol Ester (18:1(d7))
     Cer, Hexcer and SM with SM(18:1/18:1(d9))
     DG with DAG(15:0/18:1(d7))
     FA with FA(16:0(d31))
     LPC with LysoPC(18:1(d7))
     LPE with LysoPE(18:1(d7))
     PC with PC(15:0/18:1(d7))
     PE with PE(15:0/18:1(d7))
     TG with TAG(15:0/18:1(d7/15:0))
     All this is now on the column IS_ norm, there I indicate the lipid name of the IS that we should use to make the division

Again need to simplify the code ( phili will do it here)
```{r}
## this mapping is there:
rowData(annotated_res)$target_IS_norm

## the internal standanrd nae is here:
rowData(annotated_res)$target_lipid.name

## now we can make a loop to divide each feature by the IS appropiate
assay(annotated_res, "ISnorm_filled") <- assay(annotated_res, "corr_filled")

#check that the is have been detected before anything.
rowData(annotated_res)$target_IS_norm %in% rowData(annotated_res)$target_lipid.name
## C17 Sphinganine missing


for (i in seq_len(nrow(annotated_res))) {
    is_name <- rowData(annotated_res)$target_IS_norm[i]
    if (!is_name %in% rowData(annotated_res)$target_lipid.name) {
        warning(sprintf("IS '%s' for feature '%s' not found in the dataset. Skipping IS normalization for this feature.",
                        is_name, rownames(annotated_res)[i]))
        next
    }
    if (rowData(annotated_res)$target_lipid.name[i] == is_name)
      next
    if (!is.na(is_name)) {
        is_idx <- which(rowData(annotated_res)$target_lipid.name == is_name)
        if (length(is_idx) == 1) {
            assay(annotated_res, "ISnorm_filled")[i, ] <-
                assay(annotated_res, "corr_filled")[i, ] /
                assay(annotated_res, "corr_filled")[is_idx, ]
        } else {
            warning(sprintf("IS '%s' not found or duplicated for feature '%s'. Skipping IS normalization for this feature.",
                            is_name, rownames(annotated_res)[i]))
        }
    } else {
        warning(sprintf("No IS specified for feature '%s'. Skipping IS normalization for this feature.",
                        rownames(annotated_res)[i]))
    }
}

## quickly check res ## fix evaluation
mat_corr <- assay(annotated_res, "corr_filled")
mat_norm <- assay(annotated_res, "ISnorm_filled")
vals_before <- as.numeric(mat_corr)
vals_after  <- as.numeric(mat_norm)

# Filter out zeros/NAs for logging to avoid errors
vals_before_log <- log2(vals_before[vals_before > 0 & !is.na(vals_before)])
vals_after_log  <- log2(vals_after[vals_after > 0 & !is.na(vals_after)])

vioplot(list(Before = vals_before_log, After = vals_after_log),
        main = "Global Intensity Dist. (Log10)",
        ylab = "Log10 Intensity",
        col = c("mistyrose", "lightcyan"))
```

```{r}
#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(annotated_res, "ISnorm_filled"), MARGIN = 1, na_unidis)
assays(annotated_res)$ISnorm_filled_imputed <- t(tmp)
```

```{r}
#' Log2 transform and scale data
vals <- assay(annotated_res, "ISnorm_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = annotated_res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```


## Quality control
- Remove lipids with a cv in QCs >30% - quality control


```{r}
#' Number of features before filtering
nrow(annotated_res)
```

```{r}
 #' keep unfiltered object
res_unfilt <- annotated_res
```

```{r}
# RSD filter
rsd_filter <- RsdFilter(threshold = 0.3,
                        qcIndex = annotated_res$sample_type == "QC")

annotated_res <- filterFeatures(annotated_res, filter = rsd_filter)
```


```{r}
#' Number of features after analysis
nrow(annotated_res)
```

```{r}
#' Percentage left: end/beginning
nrow(annotated_res)/nrow(res_unfilt) * 100
```

## Assessing overall effectiveness of the normalization approach

Intensity evaluation

```{r}
par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))

boxplot(rowRla(assay(annotated_res, "raw_filled"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Raw data", boxwex = 1)
grid(nx = NA, ny = NULL)
legend("topright", inset = c(0, -0.2), col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE,
       ncol = 3, cex = 0.7, bty = "n")
abline(h = 0, lty=3, lwd = 1, col = "black")

boxplot(rowRla(assay(annotated_res, "ISnorm_filled"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normallized data", cex.main = 1)
axis(side = 1, at = seq_len(ncol(annotated_res)), labels = annotated_res$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3, lwd = 1, col = "black")
```

fix RLA plot before and after normalization - plot also intensity of IS before
and after normalization (log 2)

Coefficient of variation

```{r}
#' Calculate the CV values
index_study <- annotated_res$sample_type %in% c("M-VAMS", "W-DBS", "C-qDBS", "Plasma")
index_QC <- annotated_res$sample_type == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_study],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_study],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe")
```
## test median scaling

```{r}
#' Compute median and generate normalization factor
mdns <- apply(assay(annotated_res, "ISnorm_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(annotated_res)$median_scaling <- sweep(assay(annotated_res,
                                                    "ISnorm_filled"),
                                              MARGIN = 2, nf_mdn, '/')
assays(annotated_res)$median_scaling_imputed <- sweep(assay(annotated_res,
                                                            "ISnorm_filled_imputed"),
                                                      MARGIN = 2,
                                  nf_mdn, '/')
```

Intensity evaluation

```{r}
par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))

boxplot(rowRla(assay(annotated_res, "raw_filled"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Raw data", boxwex = 1)
grid(nx = NA, ny = NULL)
legend("topright", inset = c(0, -0.2), col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE,
       ncol = 3, cex = 0.7, bty = "n")
abline(h = 0, lty=3, lwd = 1, col = "black")

boxplot(rowRla(assay(annotated_res, "median_scaling"), group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normallized data", cex.main = 1)
axis(side = 1, at = seq_len(ncol(annotated_res)), labels = annotated_res$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3, lwd = 1, col = "black")
```

fix RLA plot before and after normalization - plot also intensity of IS before
and after normalization (log 2)

Coefficient of variation

```{r}
#' Calculate the CV values
index_study <- annotated_res$sample_type %in% c("M-VAMS", "W-DBS", "C-qDBS", "Plasma")
index_QC <- annotated_res$sample_type == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(annotated_res, "median_scaling")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(annotated_res, "raw_filled")[, index_study],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(annotated_res, "median_scaling")[, index_study],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe")
```
mm the median scaling seem to actually flatten the study sample variation...
but it is quiote high so not sure hwat to think. I'll continue without fro now.

From there on i guess we don't need QCs anymore

```{r}
res_qc <- annotated_res[, annotated_res$sample_type == "QC"]
res <- annotated_res[, annotated_res$sample_type != "QC"]

## redefine colours for the new data set without QCs
col_sample <- col_sample[names(col_sample) != "QC"]
col_sample_type <- col_sample_type[names(col_sample_type) != "QC"]
```

```{r}
#' Calculate the QC's CV and add as feature variable to the data set
rowData(res)$qc_cv <- assay(res_qc, "ISnorm_filled") |>
               rowRsd()
```


# Overlaping features

```{r}
#create a logical matrix to be able to make the upsetplot
df_logical <- !is.na(assay(res, "raw_filled"))

# Extract group code (letter before "_pos.mzML")
code <- sub(".*([MWCP])_pos\\.mzML$", "\\1", colnames(df_logical))

# Map codes to descriptive names
code_map <- c(
  M = "M-VAMS",
  C = "C-qDBS",
  W = "W-DBS",
  P = "Plasma"
)
groups <- code_map[code]

# Collapse by group using 70% threshold
result <- sapply(unique(groups), function(g) {
  # number of samples in this group
  n <- sum(groups == g)
  # proportion of TRUEs across samples in this group
  rowMeans(df_logical[, groups == g, drop = FALSE]) >= 0.7
})

# Result: logical matrix compounds x groups
result[1:5, ]

#make the upsetplot
binary_data <-apply(result, MARGIN = 2, as.integer)
upset(as.data.frame(binary_data), nset = 6, sets = colnames(result), keep.order = TRUE)
```



bELOW IS THE CODE THAT WILL BE DONE after we merge pos and neg together
before we need to group the adducts for the same compound.
Number of Features in common between the devices.

# Differential abundance analysis

```{r}
#' Define the colors for the plot
col_sample <- col_sample_type[res$sample_type]

#' Log transform and scale the data for PCA analysis
vals <- assay(res, "ISnorm_filled_imputed") |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- cbind(vals, sample_type = res$sample_type)
autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
```

PCA of the data after normalization and quality control, without QCs
- Plotting with highlighting different colors would be good.

The samples clearly separate by study group on PCA indicating that there are
differences in the metabolite profiles between the groups, specially Plasma
and de BuS devices are separated by the first component.

```{r}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

age <- res$age
sample_type <- factor(res$sample_type)
design <- model.matrix(~ sample_type) ## here need to add  + sex, but that is not in the current metadata..

#' Fit the linear model to the data, explaining metabolite
#' concentrations by phenotype and age.
fit <- lmFit(log2(assay(res, "ISnorm_filled_imputed")), design = design)
fit <- eBayes(fit)
```

Below plot stuff for each, this is random code, it runs but i don't know if
this is interesting or not.

```{r}
tmp <- data.frame(
    coef.vams = fit$coefficients[, "sample_typeM-VAMS"],
    pvalue.vams = fit$p.value[, "sample_typeM-VAMS"],
    adjp.vams = p.adjust(fit$p.value[, "sample_typeM-VAMS"], method = "BH"),
    avg.vams = rowMeans(
        log2(assay(res, "ISnorm_filled_imputed")[, res$sample_type == "M-VAMS"])),
    avg.CTR = rowMeans(
        log2(assay(res, "ISnorm_filled_imputed")[, res$sample_type == "C-qDBS"]))
)
tmp$significant.vams <- tmp$adjp.vams < 0.05
#' Add the results to the object's rowData
rowData(res) <- cbind(rowData(res), tmp)
```

the thing above can be build for each device. and a heatmap can be build on it.

```{r}
#' Plot volcano plot of the statistical results
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.vams, -log10(rowData(res)$adjp.vams),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.vams)) {
    points(rowData(res)$coef.vams[rowData(res)$significant.vams],
           -log10(rowData(res)$adjp.vams[rowData(res)$significant.vams]),
           col = "#0000ffcc")
}
```

Cant hen have a table of significant features, build a heatmap,....
