---
title: "R_lipidomicsOptimization"
format: html
---

# TODOS

So we organize a bit what is left to do. Also `!!!` can be found in the text for more context or issues.

-   Write intro
-   Test peak picking and refinement parameters to improve peak detection
    -   find problem areas \[Sara\], add them to peak picking code chunks
    -   test param
    -   Rerun preprocessing
-   Implement MetaboLights use properly.
-   Fix abline in all the plotting code \[Phili\]
-   Matching loop (rank 2, 3 etc.)
-   Add figure and table captions
-   Clean up useless code.
- Implement the filterFeatures RSd code

- output plots code 
- step by step explanation of normalisation
- 

# Introduction

Goals:

-   Develop a vendor-independent, open-source and one-step targeted lipid processing and annotation workflow for high-confidence lipidomics.

-   Automated RT correction, improving accuracy and reproducibility.

-   Enhanced speed and efficiency in targeted lipid searches, including verification through adduct profile matching and isomeric overlaps.

-   A validated computational tool that supports multiple MS platforms, making lipidomic studies more accessible and standardized.

-   Apply the developed data-processing pipeline to two independant datasets. 1)<https://doi.org/10.1016/j.microc.2025.113760> 2)<https://doi.org/10.1021/acs.jproteome.5c00480>

1\) The aim of this study was to explore the potential of three BμS devices for reliable untargeted lipidomic profiling using reversed-phase ultra-high performance liquid chromatography coupled with quadrupole-time-of-flight mass spectrometry (RP-UHPLC-MS). Such devices –from the traditional non-volumetric dried blood spot cards (DBS) to the recently developed quantitative DBS (qDBS), and volumetric absorptive microsampling (VAMS)– were compared to conventional plasma samples from ten healthy volunteers (https://doi.org/10.1016/j.microc.2025.113760). In this novel study, a comprehensive lipidomic workflow was applied, combining an all-in-one extraction protocol and a single-step data processing and lipid annotation using a targeted peak extraction strategy based on a human plasma lipid database (https://doi.org/10.1016/j.jlr.2024.100671)

## Load needed libraries

```{r, eval=TRUE}
library(knitr)
library(readxl)
library(MsExperiment)
library(MsIO)
library(alabaster.se)
library(MsBackendMetaboLights)
library(SummarizedExperiment)
library(xcms)
library(Spectra)
library(MetaboCoreUtils)
library(limma)
library(matrixStats)
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggplot2)
library(ggfortify)
library(gridExtra)
library(BiocFileCache)
library(dbplyr)
library(AnnotationHub)
library(CompoundDb)
library(MetaboAnnotation)
library(ggVennDiagram)
library(UpSetR)
```

## Data Import

Load raw mzML data of the pilot study, this includes data of venous blood of 10 healthy individuals loaded in the following BuS devices: M-VAMS, C-qDBS, W-DBS, also plasma samples.

```{r, eval=TRUE}
# Load pos data
seq_pos <- read_xlsx(paste0("seq_pos.xlsx"),
                 col_names = TRUE) |> as.data.frame()
seq_pos$polarity <- "pos"    # create a column named polarity

spectra_files_pos <- file.path("POS_data", seq_pos$`file_name`)

mse_pos <- readMsExperiment(spectraFiles = spectra_files_pos,
                            sampleData = seq_pos)

sampleData(mse_pos)
```

Below we show how to load from MetaboLights as an alternative to mzml files.

```{r, eval=FALSE}
param <- MetaboLightsParam(mtblsId = "MTBLS10722",
                           assayName = "a_MTBLS10722_LC-MS_positive_reverse-phase_metabolite_profiling.txt",
                           filePattern = ".mzML")

mse_pos <- readMsObject(MsExperiment(),
                        param,
                        keepOntology = FALSE,
                        keepProtocol = FALSE,
                        simplify = TRUE)
```

Some variable mapping needs to be done when importing from MetaboLights

TODO !!!

```{r, eval=FALSE}
# Let's rename the column for easier access
colnames(sampleData(mse_pos)) <- c("sample_name", "derived_spectra_data_file",
                                "metabolite_asssignment_file",
                                "source_name",
                                "organism",
                                "blood_sample_type",
                                "sample_type", "age", "unit", "phenotype")

# Add "QC" to the phenotype of the QC samples
sampleData(mse_pos)$phenotype[sampleData(mse_pos)$sample_name == "POOL"] <- "QC"
sampleData(mse_pos)$sample_name[sampleData(mse_pos)$sample_name == "POOL" ] <- c("POOL1", "POOL2", "POOL3", "POOL4")

#  Add injection index column
sampleData(mse_pos)$injection_index <- seq_len(nrow(sampleData(mse_pos)))

#let's look at the updated sample data
sampleData(mse_pos)[, c("derived_spectra_data_file",
                     "phenotype", "sample_name", "age",
                     "injection_index")] |>
    kable(format = "pipe")
```

\[Phili: Here i would explain maybe some column names so that other people understand what is in them\]

Adjust the number of cores based on your machine. Suggested to use 1 or 2 below your higher capacity.

```{r, eval=TRUE}
cores_nb <- 4  # Ajust here
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(cores_nb))
} else {
    register(SnowParam(cores_nb))
}
```

Define color per sample_type: M-VAMS, C-qDBS, W-DBS, QC, Plasma

```{r,eval=TRUE}
col_sample_type <- brewer.pal(9, name = "Set1")[c(9, 2, 1, 3, 4)]
names(col_sample_type) <- c("QC",   # grey
                          "M-VAMS", # blue
                          "W-DBS",  # red
                          "C-qDBS", # green
                          "Plasma") # purple
col_sample <- col_sample_type[sampleData(mse_pos)$sample_type]
```

# Dataset investigation

## Chromatographic data visualization

BPC plot of all samples (colored by device), including QCs.

```{r, eval=TRUE}
#| fig-cap: "BPC of all samples colored by phenotype"
#' Extract and plot BPC for full data
bpc <- chromatogram(mse_pos, aggregationFun = "max")

#png("BPC_pos.png", width = 2000, height = 1500, res = 300)

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6, 
       horiz = TRUE,
       bty = "n")

#dev.off()
```

Filter the data based on the RT observed on the BPC. In this data set we filter from 10 to 950 sec.

```{r, eval=TRUE}
#| fig-cap: "BPC after filtering retention time."
#' Filter the data based on retention time
mse_pos <- filterRt(mse_pos, c(10, 950))

# Plot BPC after filtering
bpc <- chromatogram(mse_pos, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6,
       horiz = TRUE, bty = "n")
```

```{r, eval=TRUE}
#| fig-cap: "Heatmap of TIC similarities. All plasma samples are group together."
#' Total ion chromatogram
tic <- chromatogram(mse_pos, aggregationFun = "sum") |>
  bin(binSize = 8)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

rownames(ticmap) <- colnames(ticmap) <- sampleData(mse_pos)$sample_name
ann <- data.frame(sample_type = sampleData(mse_pos)[, "sample_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
#png("heatmap_pos.png", width = 2000, height = 1500, res = 300)

pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = col_sample_type),
         show_colnames = FALSE,
         show_rownames = FALSE)

#dev.off()
```

## EICs

Upload the list of the internal standards and also some endogenous compounds that will be used to make RT shift correction. RT must be in seconds.

```{r, eval=TRUE, warning = FALSE}
#| tbl.cap: "Set of reference lipids, including deuterated lipids from SPLASH LIPIDOMIX and some deuterated compounds that are in high abundance in plasma."
#' Load list of Internal standard SPLASH LIPIDOMIX
intern_standard <- read_xlsx("splashlipidomix_list.xlsx") |>
    as.data.frame()
intern_standard$rtmin <- intern_standard$RT - 30
intern_standard$rtmax <- intern_standard$RT + 20
intern_standard$mzmin <- intern_standard$mz - 0.008 
intern_standard$mzmax <- intern_standard$mz + 0.008
rownames(intern_standard) <- intern_standard$short_name

# Extract EICs for the list
eic_is <- chromatogram(
    mse_pos,
    rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
    mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

#' Add internal standard metadata
fData(eic_is)$mz <- intern_standard$mz
fData(eic_is)$rt <- intern_standard$RT
fData(eic_is)$rtmin <- intern_standard$rtmin
fData(eic_is)$rtmax <- intern_standard$rtmax
fData(eic_is)$short_name <- intern_standard$short_name
rownames(fData(eic_is)) <- intern_standard$short_name

fdata <- fData(eic_is)

#' Summary of IS information
summary(fdata)
head(fdata)

fData(eic_is)[c("CAR", "LysoPC", "dLysoPC", "dLysoPE", "LPC", "SM",
                "dPG", "dSM", "PC", "dPC", "dPE", "SM1",
                "dDG", "HexCer", "TG", "dTG", "dCE"),
              c("short_name", "mz", "rt", "rtmin", "rtmax")] |>
    kable(format = "pipe")
```

We below plot the EICs for reference lipids (17 in total): isotope labeled lipids from splash lipidomix and some endogenous lipids.

```{r, eval=TRUE}
#| fig.cap: "EIC of all reference lipids."
# Loop to go through all rows of the eic_is object and plot independent  

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 #abline(v = median(rtime(eic)), col = "red", lty = 3)
}
```

```{r, eval=TRUE}
#|fig.cap: "EIC of three selected deuterated lipids."
# Chromatogram example of three lipids from the list that was imported before
eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
#png("EIC_exampledeuteratedlipids.png", width = 2000, height = 1500, res = 300)

par(mfrow = c(1, 3), mar = c(4, 2, 2, 0.5))

plot(eic_deuteratedLysoPC, main = fData(eic_deuteratedLysoPC)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedLysoPC)$rt, col = "red", lty = 3)

plot(eic_deuteratedSM, main = fData(eic_deuteratedSM)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
     col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedSM)$rt, col = "red", lty = 3)

plot(eic_CAR, main = fData(eic_CAR)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()

legend("topright", col = col_sample_type, legend = names(col_sample_type),
       lty = 1,
       bty = "n")

#dev.off()
```

```{r}
# Extract some feature examples where we are having issues
# features "FT03637", "FT03866", "FT03865", "FT12793"
feature_name <- c("cp1", "cp2", "cp3", "cp4")
mz_min <- c(482.315, 510.390, 510.380, 929.833)
mz_max <- c(482.333,510.440, 510.388, 929.836 )
rt_min <- c(193, 200, 175, 855)
rt_max <- c(210, 230, 190, 875 )

# Combine into a data frame
df_features <- data.frame(
  feature_name = feature_name,
  mz_min = mz_min,
  mz_max = mz_max,
  rt_min = rt_min,
  rt_max = rt_max
)

# Extract EIC from the problematic features
eic_features <- chromatogram(
    mse_pos,
    rt = as.matrix(df_features[, c("rt_min", "rt_max")]),
    mz = as.matrix(df_features[, c("mz_min", "mz_max")]))

# Add the cp name into the EIC_ features object 
fData(eic_features)$cp_name <- df_features$feature_name
rownames(fData(eic_features)) <- df_features$feature_name

eic_cp1 <- eic_features["cp1"]
eic_cp2 <- eic_features["cp2"]
eic_cp3 <- eic_features["cp3"]
eic_cp4 <- eic_features["cp4"]

par(mfrow = c(2, 2), mar = c(4, 2, 2, 0.5))

plot(eic_cp1, main = fData(eic_cp1)$cp_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_cp1)$rt, col = "red", lty = 3)

plot(eic_cp2, main = fData(eic_cp2)$cp_name,
     cex.axis = 0.8,
     cex.main = 0.8,
     col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_cp2)$rt, col = "red", lty = 3)

plot(eic_cp3, main = fData(eic_cp3)$cp_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()

plot(eic_cp4, main = fData(eic_cp4)$cp_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()

legend("topright", col = col_sample_type, legend = names(col_sample_type),
       lty = 1,
       bty = "n")
```

# Data preprocessing

## Chromatographic peak detection

We use the deuterated LysoPC of the first sample to extract the *m/z*, calculate the absolute difference and expressed in ppm.

```{r, eval=TRUE}
#' Restrict the data to signal from dLysoPC in the first sample
lpc <- mse_pos[1L] |>
  spectra() |>
  filterRt(rt = c(155, 170)) |>
  filterMzRange(mz = fData(eic_deuteratedLysoPC)["dLysoPC", 
                                                 c("mzmin", "mzmax")])

#' Show the number of peaks per m/z filtered spectra
lengths(lpc)
```

```{r, eval=TRUE}
#' Calculate the difference in m/z values between scans
mz_diff <- lpc |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(lpc))))
```

We therefore, choose a value close to the maximum within this range for
parameter `ppm`, i.e., 10 ppm, otherwise work with 2 `ppm` will be to low.

From previous observation of some of our ISTD we estimate the average peak 
width to be around 4 to 8 seconds. We therefore set the `peakwidth` parameter 
to `c(4, 8)`. \[Here link to figure above\]

We can now perform the chromatographic peak detection with the adapted settings on our EICs, and test the parameters in some Lipids from the list. In this case deuterated LysoPC, Sphingomelyn and Carnitine.

```{r, eval=TRUE}
#' Parameters adapted for chromatographic peak detection on EICs.
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2)

#' Evaluate on the dLysoPC ion
dLysoPC_test <- findChromPeaks(eic_deuteratedLysoPC, param = param_pd)
chromPeaks(dLysoPC_test)
```

```{r, eval=TRUE}
#' Evaluate on the Sphingomyelin ion
dSM_test <- findChromPeaks(eic_deuteratedSM, param = param_pd)
chromPeaks(dSM_test)
```

```{r}
#' Evaluate on the Sphingomyelin ion
CAR_test <- findChromPeaks(eic_CAR, param = param_pd)
chromPeaks(CAR_test)
```

It now seems that we have a peak detected in each sample for the first two ions, !!!

With the customized parameters, a chromatographic peak was detected in each sample. Below, we use the [`plot()`](https://rdrr.io/r/graphics/plot.default.html)  function on the EICs to visualize these results.

```{r, eval=TRUE}
#| fig.cap: "EIC of three selected deuterated lipids after applying the optimized param selection."
#' Plot test chromatogram for deuterated LysoPC, Sphingomielyn and Carnitine.
par(mfrow = c(1, 3))
plot(dLysoPC_test, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(dSM_test, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(CAR_test, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

We can see a peak seems to be detected in each sample for both ions. This indicates that our custom settings are suitable for our dataset. We now proceed and apply them to the entire dataset, extracting EICs again for the same ions to evaluate and confirm that chromatographic peak detection worked as expected.

```{r, eval=TRUE}
#' Using the same settings, but with default snthresh
param_pd <- CentWaveParam(peakwidth = c(4, 8), ppm = 10, integrate = 2,
                          snthresh = 2)
## lowered the sntresth super low and still no detection of the carnitine peaks.
mse_pos <- findChromPeaks(mse_pos, param = param_pd, chunkSize = cores_nb)

#' Update EIC internal standard object
#' Below tested a larger mz window to see if the carnitine gets detected but it just highlights all the noise. 
# new_mz <- intern_standard[, c("mzmin", "mzmax")]
# new_mz$mzmin <- new_mz$mzmin - 0.01
# new_mz$mzmax <- new_mz$mzmax + 0.01

eics_is_noprocess <- eic_is
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
```

```{r, eval=FALSE}
fData(eic_is) <- fData(eics_is_noprocess)
```

```{r}
# Plot to evaluate the chromatographic peak detection results.
rownames(fData(eic_is)) <- intern_standard$short_name

eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]

## Here add some other double peak compound that we failed to identify before.
```

```{r, eval=FALSE}
#| fig.cap: "EIC of three selected deuterated lipids after applying the chromatographic peak detection on the entire dataset."
par(mfrow = c(1, 3))
plot(eic_deuteratedLysoPC, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_deuteratedSM, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_CAR, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

After applying the cinematographic peak detection over the entire dataset, only the first two peaks have been detected properly in all sample, but this does not happen with the Carnitine (not even after changing the signal to noise ratio, or opening the window for the m/z or rt)

## Refinment of chromatographic peaks

```{r, eval=TRUE}
#' set up the parameter
param_msp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.0015,
                                        minProp = 0.75)
```

```{r, eval=TRUE}
#' Apply on whole dataset
mse_pos <- refineChromPeaks(mse_pos, param = param_msp, chunkSize = cores_nb)
```

```{r}
chromPeakData(mse_pos)$merged |>
                         table()
```

Evaluate the results of the chromatographic peak detection on EICs of e.g. internal standards or other compounds/ions known to be present in the samples. Here we update our EICs objects for the internal standards with the results of the chromatographic peak detection and refinement.

```{r, eval=FALSE}
eics_is_chrompeaks <- eic_is

eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fData(eics_is_chrompeaks)

eic_deuteratedLysoPC <- eic_is["dLysoPC", ]
eic_deuteratedSM <- eic_is["dSM", ]
eic_CAR <- eic_is["CAR"]
```

### Evaluate peak detection results

```{r, eval=FALSE}
# Define output directory (create if it doesn't exist)
outdir <- "peak_detection_ref_lipid"
if (!dir.exists(outdir)) dir.create(outdir)

# Loop through each row and save plots as PNG
for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i, ]
  filename <- file.path(outdir, paste0(rownames(fData(eic_is))[i], ".png"))
  png(filename, width = 800, height = 600, res = 120)
  plot(
    eic,
    main = rownames(fData(eic_is))[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  # abline(v = median(rtime(eic)), col = "red", lty = 3)
  dev.off()
}
```

All the compound plots in `eic_is` are now saved locally, these should be 
checked before continuing. Each peak should be detected in most samples 
(if not all). Also ensure that all the EICs that have 2 peak are properly
separated. Discuss with Johannes to see how to prevent merge. !!!

Number of chromatographic peaks per sample.

```{r, eval= TRUE}
#| tbl.cap: "Number of chromatographic peaks detected per sample."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(
  sample_name = sampleData(mse_pos)$sample_name,
  peak_count = as.integer(table(chromPeaks(mse_pos)[, "sample"]))) |>
  t() |>
  kable(format = "pipe")
```

A similar number of chromatographic peaks was identified within the various samples of the data set. In all cases W-DBS has the lower number of features. Which makes sense as less volume was injected therefore features of lower\
intensity would not be seen im W-DBS.

## Retention time alignment

To evaluate the presence of a shift we extract and plot below the BPC from the QC samples.

```{r, eval=TRUE}
#| fig-cap: "BPC of QC samples."
#' Get QC samples
QC_samples <- sampleData(mse_pos)$sample_type == "QC"

#' extract BPC
mse_pos[QC_samples] |>
    chromatogram(aggregationFun = "max", chromPeaks = "none") |>
    plot(col = col_sample_type["QC"], main = "BPC of QC samples") |>
    grid()
```

Initial correspondence analysis. Parameters do not need to be fully optimized. We base it on the QC samples.

```{r, eval=TRUE}
param_c <- PeakDensityParam(
  sampleGroups = sampleData(mse_pos)$sample_type == "QC",
  minFraction = 0.9,
  binSize = 0.01, ppm = 10,
  bw = 2)
mse_pos <- groupChromPeaks(mse_pos, param = param_c)
```

```{r, eval=TRUE}
#| fig-cap: "Initial correspondence analysis based on QC samples."
plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_c,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]], 20),
    peakPch = 16)
```

Below we perform the alignment of our data set based on retention times of anchor peaks defined in the subset of QC samples.

```{r, eval=TRUE}
#' Define parameters of choice
subset <- which(sampleData(mse_pos)$sample_type == "QC")
param_c <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                           subsetAdjust = "average",
                           subset = subset)

#' Perform the alignment
mse_pos <- adjustRtime(mse_pos, param = param_c)
```

Alignment adjusted the retention times of all spectra in the data set, as well as the retention times of all identified chromatographic peaks.

```{r, eval=TRUE}
#| fig-cap: "Retention time alignment results."
#| label: fig-alignment-results
#' Visualize alignment results
plotAdjustedRtime(mse_pos, col = paste0(col_sample, 80), peakGroupsPch = 1)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

Only little adjustments needed to be performed (shifts of at maximum 1.5 second as can be seen in Figure @fig-alignment-results).

We can also compare the BPC before and after alignment.

```{r, eval = TRUE}
#' Get data before alignment
mse_pos_raw <- dropAdjustedRtime(mse_pos)

#' Apply the adjusted retention time to our dataset
mse_pos <- applyAdjustedRtime(mse_pos)
```

```{r, eval= TRUE}
#| fig-cap: "BPC before and after alignment."
#' Plot the BPC before and after alignment
par(mfrow = c(2, 1), mar = c(2, 1, 1, 0.5))
chromatogram(mse_pos_raw, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC before alignment", col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)

chromatogram(mse_pos, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC after alignment",
         col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)
```

There is no shift observed.

We next evaluate the impact of the alignment on the EICs of the selected internal standards. We thus below first extract the ion chromatograms after alignment, and subsequently plot them.

```{r, eval=TRUE}
#' Extract the EICs for the test ions before alignment
old_eic_dLysoPC <- eic_is["dLysoPC"]
old_eic_dSM <- eic_is["dSM"]
old_eic_CAR <- eic_is["CAR"]

#' Update the EICs
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fdata

#' Extract the EICs for the test ions
eic_dLysoPC <- eic_is["dLysoPC"]
eic_dSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
```

```{r, eval=TRUE}
#| fig-cap: "EIC of three selected deuterated lipids before and after alignment."
par(mfrow = c(2, 2), mar = c(4, 4.5, 2, 1))

plot(old_eic_dLysoPC, main = "dLysoPC before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(eic_dLysoPC, main = "dLysoPC after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(old_eic_dSM, main = "Sphingomyelin before alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")

plot(eic_dSM, main = "Sphingomyelin after alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

plot(old_eic_CAR, main = "CAR before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)

plot(eic_CAR, main = "CAR after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)
```

In addition to a visual inspection of the results, we also evaluate the impact of the alignment by comparing the variance in retention times for internal standards before and after alignment.

```{r, eval= TRUE}
#' Extract the matrix with all chromatographic peaks and add a column
#' with the ID of the chromatographic peak
chrom_peaks <- chromPeaks(mse_pos) |> as.data.frame()
chrom_peaks$peak_id <- rownames(chrom_peaks)

#' Define the parameters for the matching and filtering of the matches
p_1 <- MzRtParam(ppm = 50, toleranceRt = 10)
p_2 <- SingleMatchParam(duplicates = "top_ranked", column = "target_maxo",
                        decreasing = TRUE)

#' Iterate over samples and identify for each the chromatographic peaks
#' with similar m/z and retention time than the onse from the internal
#' standard, and extract among them the ID of the peaks with the
#' highest intensity.
intern_standard_peaks <- lapply(seq_along(mse_pos), function(i) {
    tmp <- chrom_peaks[chrom_peaks[, "sample"] == i, , drop = FALSE]

     # Convert to S4 DataFrames
    query  <- DataFrame(intern_standard)
    target <- DataFrame(tmp)


    mtch <- matchValues(query, target,
                        mzColname = c("mz", "mz"),
                        rtColname = c("RT", "rt"),
                        param = p_1)
    mtch <- filterMatches(mtch, p_2)
    mtch$target_peak_id
}) |>
    do.call(what = cbind)
```

We have now for each internal standard the ID of the chromatographic peak in each sample that most likely represents signal from that ion. We can now extract the retention times for these chromatographic peaks before and after alignment.

```{r, eval= TRUE}
#' Define the index of the selected chromatographic peaks in the
#' full chromPeaks matrix
idx <- match(intern_standard_peaks, rownames(chromPeaks(mse_pos)))

#' Extract the raw retention times for these
rt_raw <- chromPeaks(mse_pos_raw)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))

#' Extract the adjusted retention times for these
rt_adj <- chromPeaks(mse_pos)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))
```

We can now evaluate the impact of the alignment on the retention times of internal standards across the full data set:

```{r, eval=FALSE}
list(all_raw = rowSds(rt_raw, na.rm = TRUE),
     all_adj = rowSds(rt_adj, na.rm = TRUE)
     ) |>
    vioplot()
grid()
```

Figure X. Retetion time variation of internal standards before and after alignment.

## Correspondence

The goal of the correspondence analysis is to identify chromatographic peaks that originate from the same types of ions in all samples of an experiment and to group them into LC-MS *features*.

```{r, eval=TRUE}
#' Optimized parameter for the grouping and apply them to the test ions BP2
#' fix grouping, NA for QCs
group_mse <- sampleData(mse_pos)$sample_type
group_mse[group_mse =="QC"] <- NA
group_mse <- as.factor(group_mse)
param_cor <- PeakDensityParam(sampleGroups = group_mse, bw = 2)
param_cor
```

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
                    20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated LysoPC.

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedSM, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated SM.

```{r, eval=TRUE}
plotChromPeakDensity(
    eic_CAR, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_CAR)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_CAR)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters. Carnitine does not have any chromatographic peak, check what is happening.

We can observe for the first two ions that the peaks are grouped accurately into a single feature. So we proced to define the settings for the param\_ cor and apply them to the whole data.

```{r, eval=TRUE}
#' Define the settings for the param_cor
param_cor <- PeakDensityParam(sampleGroups = group_mse,
                              minFraction = 0.5, binSize = 0.01, ppm = 10,
                              bw = 2)

#' Apply to whole data
mse_pos <- groupChromPeaks(mse_pos, param = param_cor)
```

Evaluation of the correspondence results in other EIC in the IS list similar to the deuterated LysoPC.

```{r, eval=TRUE}
#' Extract chromatogram for an m/z similar to the one of the labeled methionine
chr_test <- chromatogram(mse_pos,
                         mz = as.matrix(intern_standard["LysoPC",
                                                        c("mzmin", "mzmax")]),
                         rt = c(120, 184),
                         aggregationFun = "max")
plotChromPeakDensity(
    chr_test, simulate = FALSE,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(chr_test)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(chr_test)[, "sample"]], 20),
    peakPch = 16)
```

This corresponds to non-deuterated LysoPC and we can see that it can distinguish between the two peaks.

```{r, eval=FALSE}
#### TEST for review Evaluation of the correspondance results in 17 EICs in the IS list.



# Loop to go through all rows of the eic_is object and plot independent after correspondance analysis

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 #abline(v = median(rtime(eic)), col = "red", lty = 3)
}





param_peak <-


# Assuming lcms1 is your raw data object (XCMSnExp)
mse_posp <- findChromPeaks(mse_pos, param = param)

# Extract chromatograms again with peaks included
eic_isp <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

for (i in seq_len(nrow(eic_isp))) {
  eicp <- eic_isp[i, ]
  plotChromPeakDensity(
    eicp, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(mse_pos)[, "sample"]], # peaks now exist
    peakBg = paste0(col_sample[chromPeaks(mse_pos)[, "sample"]], 20),
    peakPch = 16
  )
  grid()
}
```

## Gap filling

We next examine how prevalent missing values are in our present dataset:

```{r}
#' Percentage of missing values
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100

idx_qc <- sampleData(mse_pos)$sample_type == "QC"
#' Percentage of missing values
sum(is.na(featureValues(mse_pos[idx_qc, keepFeatures = TRUE]))) /
    length(featureValues(mse_pos[idx_qc, keepFeatures = TRUE])) * 100
```

We can observe a substantial number of missing values values in our dataset.

Fill the missing values in the entire data set.

```{r}
#' Fill in the missing values in the whole dataset
mse_pos <- fillChromPeaks(mse_pos, param = ChromPeakAreaParam(),
                          chunkSize = cores_nb)

#' Percentage of missing values after gap-filling
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100
```

```{r}
#' Get only detected signal in QC samples
vals_detect <- featureValues(mse_pos, filled = FALSE)[, QC_samples]

#' Get detected and filled-in signal
vals_filled <- featureValues(mse_pos)[, QC_samples]

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)

#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080")
grid()
abline(0, 1)
```

Figure X. Detected vs. filled-in signal.

# Library correction.

```{r, eval=FALSE}
#' Default parameter for the grouping and apply them to the test ions BPC
group <- eic_is$sample_type
group[group =="QC"] <- NA
param1 <- PeakDensityParam(sampleGroups = group , minFraction = 0.5,
                           binSize = 0.01, ppm = 10,
                           bw = 2)

eic_is_corr <- groupChromPeaks(eic_is, param = param1)
featureDefinitions(eic_is_corr[3, ]) ## Show johannes, why do we have a
## third peak, we can't even see on the plot. has no intensity.
## with xcms version >= 4.7.1 only 2 features are found.

exp_rt <- vector("numeric", length = nrow(eic_is_corr))
dr <- "./ref_lipid_image/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(eic_is_corr))) {
  table <- featureDefinitions(eic_is_corr[i,])
  if (!nrow(table)) {
    message("compound not detected: ", fData(eic_is_corr)$short_name[i])
    exp_rt[i] <- NA
    next
  }
  png(filename = paste0(dr, "feature_compound_", i, ".png"))
  plot(eic_is_corr[i,],
       main = rownames(fData(eic_is_corr))[i],
       cex.axis = 0.8,
       cex.main = 0.8)
  #abline(v = median(rtime(eic_is_corr)), col = "red", lty = 3) #ADD abline
  dev.off()
  if (nrow(table) > 1) {
    idx_max <- which.max(rowSums(featureValues(eic_is_corr[i, ])))
    rt <- table[idx_max, ]$rtmed
  } else rt <- table$rtmed
  exp_rt[i] <- rt
}


## before going on, check the plots. the line should overlap what we expect.

# below is the reference lipid table
ref_rt <- intern_standard$RT # is the ref RT

# res is the exp RT
intern_standard$exp_rt <- exp_rt

diff_rt <- intern_standard$RT - intern_standard$exp_rt
quantile(diff_rt)

## Fit the linear model
fit <- lm(exp_rt ~ poly(ref_rt, 6, raw = TRUE) )
summary(fit)
plot(fit)
```

Sara will check if the shift see m to be similar to what she has or not.I did not apply any RT shift correction, because my data had not to much shift on the RT and we were able to do the match against the library. The average difference is 9.6 seconds

Library fix

```{r}
lipid_database <- read_xlsx("LipidDatabase_R.xlsx", sheet = 4L)
lipid_database$rt_sd <- as.numeric(lipid_database$`RT ESI(+) (min)`) * 60
lipid_database <- lipid_database[order(lipid_database$rt_sd), ]

is.unsorted(lipid_database$rt_sd) # check rtime are properly sorted. we want FALSE

rt_sd <- lipid_database$rt_sd

## finding values that need to be extrapolated.
in_range_idx <- which(rt_sd >= min(fit$fitted.values) &
                      rt_sd <= max(fit$fitted.values))

nrow(lipid_database) - length(in_range_idx)
## Adjust library retention times
corrected_rt <- rep(NA, length = nrow(lipid_database))
corrected_rt[in_range_idx] <- predict(
    fit, newdata = data.frame(ref_rt = rt_sd[in_range_idx]))
## Adjust values outside the fitted range
idx <- which(rt_sd < min(fit$fitted.values))
lidx <- length(idx)
if (lidx)
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[lidx + 1L] - corrected_rt[lidx + 1L])
idx <- which(rt_sd > max(fit$fitted.values))
if (length(idx))
    corrected_rt[idx] <- rt_sd[idx] - (rt_sd[idx[1L] - 1L] - corrected_rt[idx[1L] - 1L])

lipid_database$rt_adjusted <- corrected_rt

## difference between original and adjusted retention times
quantile(abs(rt_sd - corrected_rt))

# plot library 500 compounds rt  raw vs adj. then (on top) add the ref_rt vs exp_rt (20compounds)
```

```{r}
# scatterplot lipid_name and rt(seconds)
lipid_database$mz <- as.numeric(lipid_database$mz)

#lipid_database <- lipid_database %>% mutate(`LIPID CATEGORY [ABBREV]`= as.factor(`LIPID CATEGORY [ABBREV]`))

#ggplot(lipid_database, aes(x=rt_sec, y=mz, group=`LIPID CATEGORY [ABBREV]`)) +
#  geom_point(aes(color=`LIPID CATEGORY [ABBREV]`))+
#  ggtitle("rt_database")

plot(y = lipid_database$rt_sd, x =lipid_database$rt_adjusted, type = "l")
points(intern_standard$RT ~ intern_standard$exp_rt, col = "red", pch = 16, cex = 1)
abline(v = min(fit$fitted.values), col = "blue", lwd = 1, lty = 2)
abline(v = max(fit$fitted.values), col = "blue", lwd = 1, lty =2)
grid()
abline(0, 1, lty = 2, col = "#00000040")
```

Figure X. lipid database rt vs. rt adjusted base on the model. The Blak line is build with the 592 compounds of the database and the red dots are the reference lipids (17 between endognous and deuterated)

## missing steps ??

These steps could be done after annotation, as we are in a semi-targeted workflow

-   plot some pCA

-   normalization

    -   per volume
    -   per internal standard

-   filtering (not sure we need that)

# Annotate/ match

-   maybe separate the QCs into another object or do that later.

```{r}
#' extract results as a SummarizedExperiment
res <- quantify(mse_pos, method = "sum", filled = FALSE)
res

assays(res)$raw_filled <- featureValues(mse_pos, method = "sum",
                                        filled = TRUE)
```

```{r}
# Filter the res object setting a threshold of 1.5*10^3
res_filtered <- assay(res)[rowMeans(assay(res), na.rm = TRUE) > 1500, ]

sum(!rowMeans(assay(res), na.rm = TRUE) > 1500) #503 removed features

```

## Save mse_pos and res object

```{r}
# save the mse_pos preprocessed object
saveMsObject(mse_pos, AlabasterParam(path = file.path("objects/preprocessed_mse_pos")))

# save the SummarizedExperiemnt
saveObject(res, file.path("objects/preprocessed_res")) 
```

Match features (query) against (aligned) library (target).

```{r}
## Extract the feature definitions as a data.frame
query <- as.data.frame(rowData(res))
query$feature_id <- rownames(rowData(res))
#subset the lipid library for rank 1 adducts only 
lipid_database_rank1 <- lipid_database[lipid_database$rank == 1, ]


## Get the reference/databse data as a data.frame
target <- as.data.frame(lipid_database_rank1)
target$mz <- as.numeric(target$mz)
## Configure the matching between features (query) and database (target)
param_match <- MzRtParam(tolerance = 0.001,
                         ppm = 20,
                         toleranceRt = 5. #+/-5 sec 
                         ) ## maybe adapt to settings from the original approach
## Perform the matching; define also the column names in query and target that
## contain the values to be matched (i.e. the m/z and retention time values)
mtch <- matchValues(query, target,
                    param = param_match,
                    mzColname = c("mzmed", "mz"),
                    rtColname = c("rtmed", "rt_adjusted"))
mtch
```

From the in total `r nrow(query)` LC-MS features `r length(whichQuery(mtch))` were matched to at least one lipid in the database. In total `r length(whichTarget(mtch))` lipids from the database were matched (from in total `r nrow(target)` compounds/lipids in the database).

We next evaluate the matching results. We first identify features that match more than one database entry (i.e., the same feature matching to two or more different lipids/compounds).

```{r}
## only keep query that has matched
mtch <- mtch[whichQuery(mtch)]
mtched_data <- matchedData(mtch) |>
    as.data.frame()  ## can do some stuff with that table

## Get the features that match multiple targets
fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
```

The table below shows the ambiguous (multi-matching) features.

```{r, results = "asis"}
table_amb <- mtched_data[mtched_data$feature_id %in% fids_ambiguous,
                         c("feature_id", "mzmed", "rtmed", "target_lipid.name",
                           "target_Adduct", "ppm_error", "score_rt")]
table_amb <- table_amb[order(table_amb$feature_id, table_amb$rtmed), ]
rownames(table_amb) <- NULL
library(pander)
pandoc.table(table_amb, split.table = Inf, style = "rmarkdown")
library(writexl)
write_xlsx(table_amb, path = "features-matching-multiple-lipids.xlsx")
```

```{r}
## example of peak non detected: 
## it has low intensity, we need to refine preprocessing. 
eg1 <- chromatogram(mse_pos, mz = c(482.315, 482.333), rt = c(193, 210)) #LPE(0:0/18:0)
plot(eg1)

eg2 <- chromatogram(mse_pos, mz = c(510.390, 510.440), rt = c(200, 230)) #LPC(O-0:0/18:0)
plot(eg2)

eg3 <- chromatogram(mse_pos, mz = c(510.380, 510.388), rt = c(175, 190)) # LPC(O-0:0/18:1) // LPC(P-0:0/18:0) only 1 peak?
plot(eg3)
featureDefinitions(eg3)

eg4 <- chromatogram(mse_pos, mz = c(929.833, 929.836), rt = c(855, 875)) #TG(53:3)
plot(eg4)
featureDefinitions(eg4) # wrong rtmin rtmax in the definition ???????

eg5 <- chromatogram(mse_pos, mz = c(787.549, 787.673), rt = c(705, 720)) #SM(d40:1) # set a ppm error on amb table is 4.1 and 787.7 m/z
plot(eg5)
featureDefinitions(eg5)

```

```{r}
#plot some features
a <- featureChromatograms(mse_pos, features = "FT06046")
b <- featureChromatograms(mse_pos, features = "FT06045")
par(mfrow = c(1, 2))
plot(a)
plot(b)

d <- featureChromatograms(mse_pos, features = "FT08345")
e <- featureChromatograms(mse_pos, features = "FT08851")
f <- featureChromatograms(mse_pos, features = "FT08853")
g <- featureChromatograms(mse_pos, features = "FT08344")
par(mfrow = c(2, 2))
plot(d,main = "Gráfico D")
plot(e, main = "Gráfico E")
plot(f, main = "Gráfico F")
plot(g, main = "Gráfico G")


h <- featureChromatograms(mse_pos, features = "FT02422") # intensity MG 
j <- featureChromatograms(mse_pos, features = "FT02607")
par(mfrow = c(1, 2))
plot(h,main = "Gráfico H")
plot(j, main = "Gráfico J")

k <- featureChromatograms(mse_pos, features = "FT03637") # intensity PG 1400
plot(k, main = "Gráfico k")
```

```{r}
df_subset <- subset(mtched_data, grepl("ISTD", target_ORIGIN, ignore.case = TRUE))

# Loop to plot all the compounds in the df_subset these are the internal standards
outdir <- "ISTD_mtched_data"
if (!dir.exists(outdir)) dir.create(outdir)

# Loop through each row and save plots as PNG
for (i in rownames(df_subset)) {
    #Check for the presence of "."
  if (grepl("\\.", i)) {
    original_i <- i
    i <- sub("\\..*$", "", i)  # remove everything after the first dot
    warning(sprintf(
      "Feature '%s' was also matched to something else. Using '%s' instead.",
      original_i, i
    ), call. = FALSE)
  } 
  
  ISTD <- featureChromatograms(mse_pos, features = i)

  # Define filename using the row name
  filename <- file.path(outdir, paste0(i, ".png"))

  # Open PNG device
  png(filename, width = 800, height = 600, res = 120)

  # Plot chromatogram
  plot(
    ISTD,
    main = rownames(df_subset)[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  # abline(v = median(rtime(eic)), col = "red", lty = 3)

  # Close device
  dev.off()
}

cat("All chromatograms saved in:", outdir, "\n")

```

Next we get the database entries that match multiple features (i.e., two different features matching the same lipid/compound).

```{r}
lipid_adducts <- paste0(mtched_data$target_lipid.name, "_",
                        mtched_data$target_Adduct)
lipid_ambiguous <- lipid_adducts[duplicated(lipid_adducts)]
table_lips <- mtched_data[lipid_adducts %in% lipid_ambiguous,
                         c("feature_id", "mzmed", "rtmed", "target_lipid.name",
                           "target_Adduct", "ppm_error", "score_rt")]
table_lips <- table_lips[order(table_lips$target_lipid.name, table_lips$rtmed), ]
rownames(table_lips) <- NULL
pandoc.table(table_lips, split.table = Inf, style = "rmarkdown")
write_xlsx(table_amb, path = "lipids-matching-multiple-features.xlsx")
```

??? why is there DG(34:2) twice in the table, with different retentions times? same retention time and differen adduct

```{r}
## find features IDs of ambiguous matches
tmp <- table(mtch$feature_id)
tmp <- tmp[tmp > 1]
fids <- names(tmp)

## Check the first one
tmp1 <- mtched_data[mtched_data$feature_id == fids[1],
                   c("feature_id", "rtmed", "mzmed", "target_lipid name",
                     "ppm_error", "score_rt", "target_rt_adjusted",
                     "target_rt_sd")] |>
  as.data.frame()
tmp_eic1 <- featureChromatograms(mse_pos, features = fids[1L])
plot(tmp_eic1)

tmp2 <- mtched_data[mtched_data$feature_id == fids[2],
                     c("feature_id", "rtmed", "mzmed", "target_lipid name",
                       "ppm_error", "score_rt", "target_rt_adjusted",
                       "target_rt_sd")] |>
  as.data.frame()
tmp_eic2 <- featureChromatograms(mse_pos, features = fids[2L])
plot(tmp_eic2)

par(mfrow = c(2, 1))
plot(tmp_eic1)
plot(tmp_eic2)

tmp <- mtched_data[mtched_data$feature_id %in% fids,
                     c("feature_id", "rtmed", "mzmed", "target_lipid name",
                       "ppm_error", "score_rt", "target_rt_adjusted",
                       "target_rt_sd")] |>
  as.data.frame()
tmp_eic <- featureChromatograms(mse_pos, features = fids[5])
plot(tmp_eic)


#subset the ISTD

sara <- subset(lipid_database_rank1, grepl("ISTD", ORIGIN, ignore.case = TRUE))
# we don have a match with Tg istd and we should have it 

chromatogram(mse_pos, mz = c(482.29, 482.31)) -> test
plot(test)
# check the feautrea that match 2 compounds. check score, plotSpectraMirror.
# plot using the mtch object.
# scoring and othe rinfo check the mtch_data

## need tofilter them out somehow.







## import information from mtch_data dataframe to res object using the feature ID present in both table.
mtched_data$`target_lipid name`
rownames(mtched_data)

```

## Plots with the annotated compounds (mtched_data), the feature abundance of each sample, can be found in assay(res)

```{r}
## remove duplicated form the
length(unique(mtched_data$target_lipid.name)) ## number of lipid matched, ONLY 206

length(unique(mtched_data$target_LIPID.CATEGORY..ABBREV.))

# Lipid Distribution barplot
# Count data
counts <- table(mtched_data$target_LIPID.CATEGORY..ABBREV.)
bar_colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854")

# Rename categories for readability
names(counts) <- c("Fatty Acyls [FA]",
                   "Glycerolipids [GL]",
                   "Glycerophospholipids [GP]",
                   "Sphingolipids [SP]",
                   "Sterol Lipids [ST]")

# Adjust plot margins and font size
par(mar = c(5, 7, 4, 2))  # increase left margin for labels

bp <- barplot(counts,
              horiz = TRUE,
              col = bar_colors,
              main = "Lipid Distribution",
              xlab = "Number of Lipids",
              las = 1,
              cex.names = 0.9,
              border = NA)

text(counts, bp, labels = counts, pos = 4, cex = 0.8, col = "black")
legend("topright",
       legend = names(counts),
       fill = bar_colors,
       title = "Lipid Category",
       cex = 0.6,
       border = NA,
       bty = "n")
```

-   Donut chart showing the lipid class with the total quantity of lipids detected

```{r}
# Donnut chart of lipid distribution
counts <- table(mtched_data$target_LIPID.CATEGORY..ABBREV.)

names(counts) <- c("Fatty Acyls [FA]",
                   "Glycerolipids [GL]",
                   "Glycerophospholipids [GP]",
                   "Sphingolipids [SP]",
                   "Sterol Lipids [ST]")

bar_colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854")
labels <- counts
par(mar = c(1, 1, 2, 1))

# Plot
pie(counts,
    col = bar_colors,
    labels = labels,     # solo los números
    border = "white",
    main = "Lipid Distribution",
    cex = 0.9)

symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

# Leyend
legend("topright",
       legend = names(counts),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Lipid Category",
       cex = 0.8)
```

```{r}
# pie chart for lipid subclass. HERE ADD A NEW COLUMN WITH THE SHORT NAME
counts1 <- table(mtched_data$`target_LIPID SUBCLASS [ABBREV]`)

n_cat <- length(counts1)
base_colors <- brewer.pal(9, "Set1")
bar_colors <- rep(base_colors, length.out = n_cat)

# Labels: only counts
labels <- counts1

# Adjust margins
par(mar = c(1, 1, 2, 1))

# Create pie (donut) chart
pie(counts1,
    col = bar_colors,
    labels = labels,      # show only counts
    border = "white",
    main = paste("Category Distribution (n =", sum(counts), ")"),
    cex = 0.9)

# Add white circle in center for donut effect
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")

legend("topright",
       legend = names(counts1),
       fill = bar_colors,
       border = NA,
       bty = "n",
       title = "Categories",
       cex = 0.4)

```

-   heatmap with the significant metabolites found

-   

later on discuss how to deal with features that matches multiple compound.

we could concatenate the annotation to the summarized experiment object. Do the normalization, quantitation, plotting after,..

# Data normalization

Here before we imput the missing values we have to normalize per volume as follows: The abundance of all W-DBS samples should be multiply per 2.5 and Plasma samples per 1.08


```{r}
assay_data <- assay(res)  # table with the features and their abundances 16800 rows
assay_data <- as.data.frame(assay_data)
mtched_data # table with the annotations and also the name of the features

# create a new column in each dataset containing the rownames 
assay_data$feature_id <- rownames(assay_data)
mtched_data$feature_id <- rownames(mtched_data)

# Create a join_df that contains the annotated lipids and the areas
library(dplyr)
join_df <- mtched_data %>%
  left_join(assay_data, by = "feature_id")

# if the sample name includes W, multiply per 2.5, si el nombre de la muestra contiene P mayuscula multiply per 1.08
# volume normalization 
# 
factors <- ifelse(grepl("\\.mzML", colnames(join_df)) & grepl("W", colnames(join_df)), 2.5,
           ifelse(grepl("\\.mzML", colnames(join_df)) & grepl("P", colnames(join_df)), 1.08, 1))

normalized_df_vol <- join_df
mzml_cols <- grep("\\.mzML", colnames(normalized_df_vol), value = TRUE)
# Convert only those to numeric
normalized_df_vol[mzml_cols] <- lapply(normalized_df_vol[mzml_cols], function(x) as.numeric(as.character(x)))

normalized_df_vol[mzml_cols] <- Map(function(col, f) col * f,
                     join_df[mzml_cols],
                     factors[match(mzml_cols, colnames(normalized_df_vol))])

#check the changes in one sample (CAN BE DELETED)
cbind(
  original = join_df$D10W_pos.mzML,
  scaled   = normalized_df_vol$D10W_pos.mzML
)








# create a new vector containing the features that have a match
mtched_features_list <- mtched_data$feature_id
# subset the res object only with the annotated compounds
#subset_res <- subset(res,subset = mtched_features_list)
subset_res <- res[rownames(res2) %in% mtched_features_list, ] #only 174 id out of 269

##code above need to be fixed so mtchd_features does not have duplciated featureID. 






# Crear un nuevo objeto SummarizedExperiment con los datos modificados
#se_scaled <- SummarizedExperiment(
 # assays = list(counts = assay_data_scaled),
  #rowData = rowData(se),
  #colData = colData(se)
#)

```
```{r}
## match group to 
colData(res)$lipid_group 
```



```{r}
# subset_res 
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(subset_res, "raw_filled"), MARGIN = 1, na_unidis)
assays(subset_res)$raw_filled_imputed <- t(tmp)
```

```{r}
# subset_res
#' Log2 transform and scale data
vals <- assay(subset_res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```


DESDE AQUI estadistica metabonout
```{r}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(res, "raw_filled"), MARGIN = 1, na_unidis)
assays(res)$raw_filled_imputed <- t(tmp)
```
## Initial quality assessment

-   plot also just the annotated data
-   calcualte the missing values Principal Component Analysis

```{r}
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'sample_type',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
grid.arrange(pca_12, pca_34, ncol = 1)
```

Figure X. PCA of the data colored by sample_type.Volumetric devices profiles are clearly different from W-DBS and plasma.

Intensity evaluation

```{r}
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))

par(mar = c(0.2, 4.5, 0.2, 3))
barplot(apply(assay(res, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected peaks", xaxt = "n", space = 0.012)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sample, 80), border = col_sample)
points(colMedians(log2(assay(res, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

replot this after normalisation too Number of detected peaks and feature abundances.

```{r}
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res, "raw_filled"),
                            f = res$sample_type, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_sample, 80), ylab = "RLA",
        border = col_sample, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

RLA plot for the raw data and filled data.

## Between sample normalisation

Media scaling, Here should I apply the volume and IS normalization???

```{r}
#' Compute median and generate normalization factor
mdns <- apply(assay(res, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(res)$norm <- sweep(assay(res, "raw_filled"), MARGIN = 2, nf_mdn, '/')
assays(res)$norm_imputed <- sweep(assay(res, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn, '/')
```

## Assessing overall effectiveness of the normalization approach

Principal component analysis

```{r}
#' Data before normalization
vals_st <- cbind(vals, sample_type = res$sample_type)
pca_raw <- autoplot(pca_res, data = vals_st,
                    colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)

#' Data after normalization
vals_norm <- apply(assay(res, "norm"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)

pca_res_norm <- prcomp(vals_norm, scale = FALSE, center = FALSE)
vals_st_norm <- cbind(vals_norm, sample_type = res$sample_type)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)

grid.arrange(pca_raw, pca_adj, ncol = 1)
```

write PC1 PC2 interpretation

PC1 and PC2 of the data before and after normalization

```{r}
pca_raw <- autoplot(pca_res, data = vals_st ,
                    colour = 'sample_type', x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    colour = 'sample_type', x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_sample_type)

grid.arrange(pca_raw, pca_adj, ncol = 1)
```

PC3 and PC4 of the data before and after normalization

Intensity evaluation

```{r}
par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))

boxplot(rowRla(assay(res, "raw_filled"), group = res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Raw data", boxwex = 1)
grid(nx = NA, ny = NULL)
legend("topright", inset = c(0, -0.2), col = col_sample_type,
       legend = names(col_sample_type), lty=1, lwd = 2, xpd = TRUE,
       ncol = 3, cex = 0.7, bty = "n")
abline(h = 0, lty=3, lwd = 1, col = "black")

boxplot(rowRla(assay(res, "norm"), group = res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample,
        col = paste0(col_sample, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normallized data", cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = res$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3, lwd = 1, col = "black")
```

fix RLA plot before and after normalization - plot also intensity of IS before and after normalization (log 2)

Coefficient of variation

```{r}
#' Calculate the CV values
index_study <- res$sample_type %in% c("M-VAMS", "W-DBS", "C-qDBS", "Plasma")
index_QC <- res$sample_type == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(res, "norm")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(res, "raw_filled")[, index_study],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(res, "norm")[, index_study],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe")
```

Distribution of cv values across samples for the raw and normalized data

the Cv in study sample decrease, usually that's not ideal but here might be driven by the whatman intensity correction (check this)

# Quality control: Feature prefiltering

\[Move this section above\]

```{r}
#' Number of features before filtering
nrow(res)
```

```{r}
 #' keep unfiltered object
res_unfilt <- res
```

```{r}
#' Compute and filter based on the Dratio
filter_dratio <- DratioFilter(threshold = 0.3,
                              qcIndex = res$sample_type == "QC",
                              studyIndex = res$sample_type != "QC",
                              mad = TRUE)
res <- filterFeatures(res, filter = filter_dratio, assay = "norm_imputed")
```

no Dratio: because Qc is baased on different matrices. instead rsd, check ?RsdFilter

```{r}
#' Number of features after analysis
nrow(res)
```

```{r}
#' Percentage left: end/beginning
nrow(res)/nrow(res_unfilt) * 100
```

```{r}
res_qc <- res[, res$sample_type == "QC"]
res <- res[, res$sample_type != "QC"]
```

```{r}
#' Calculate the QC's CV and add as feature variable to the data set
rowData(res)$qc_cv <- assay(res_qc, "norm") |>
               rowRsd()
```

# Overlaping features

```{r}
#create a logical matrix to be able to make the upsetplot
df_logical <- !is.na(assay(res, "raw_filled"))

# Extract group code (letter before "_pos.mzML")
code <- sub(".*([MWCP])_pos\\.mzML$", "\\1", colnames(df_logical))

# Map codes to descriptive names
code_map <- c(
  M = "M-VAMS",
  C = "C-qDBS",
  W = "W-DBS",
  P = "Plasma"
)
groups <- code_map[code]

# Collapse by group using 70% threshold
result <- sapply(unique(groups), function(g) {
  # number of samples in this group
  n <- sum(groups == g)
  # proportion of TRUEs across samples in this group
  rowMeans(df_logical[, groups == g, drop = FALSE]) >= 0.7
})

# Result: logical matrix compounds x groups
result[1:5, ]

#make the upsetplot
binary_data <-apply(result, MARGIN = 2, as.integer)
upset(as.data.frame(binary_data), nset = 6, sets = colnames(result), keep.order = TRUE)
```

Number of Features in common between the devices.

# Differential abundance analysis

```{r}
#' Define the colors for the plot
col_sample <- col_sample_type[res$sample_type]

#' Log transform and scale the data for PCA analysis
vals <- assay(res, "norm_imputed") |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- cbind(vals, sample_type = res$sample_type)
autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)


```

PCA of the data after normalization and quality control

The samples clearly separate by study group on PCA indicating that there are differences in the metabolite profiles between the groups, speacially Plasma and de BuS devices are separated by the first component.

