---
title: "R_lipidomicsOptimization"
format: html
---

## Load needed libraries

```{r, eval=TRUE}
library(knitr)
library(readxl)
library(MsExperiment)
library(MsIO)
library(alabaster.se)
library(MsBackendMetaboLights)
library(SummarizedExperiment)
library(xcms)
library(Spectra)
library(MetaboCoreUtils)
library(limma) 
library(matrixStats)
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggplot2)
library(ggfortify)      
library(gridExtra)
library(BiocFileCache)
library(dbplyr)
library(AnnotationHub) 
library(CompoundDb)    
library(MetaboAnnotation)
```

## Import the sample data for the analysis

Load raw mzML data of the pilot study, this includes data of venous blood of 10 healthy individuals loaded in the following BuS devices: M-VAMS, C-qDBS, W-DBS, also plasma samples.

```{r, eval=TRUE}
# Load pos data
seq_pos <- read_xlsx(paste0("seq_pos.xlsx"), 
                 col_names = TRUE) |> as.data.frame()
seq_pos$polarity <- "pos"    # create a column named polarity
# load neg data
#seq_neg <- read_xlsx(paste0("/seq_neg.xlsx"), 
#                 col_names = TRUE) |> as.data.frame()
#seq_neg$polarity <- "neg"

spectra_files_pos <- file.path("POS_data", seq_pos$`file_name`)

mse_pos <- readMsExperiment(spectraFiles = spectra_files_pos,
                        sampleData = seq_pos)

sampleData(mse_pos)
```

Define the number of cores you want to work with

```{r, eval=TRUE}
# Requires library(BiocParallel)
cores_nb <- 3           # Adjust the number of cores based on your machine
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(cores_nb))
} else {
    register(SnowParam(cores_nb))
}
```

Define color per sample_type: M-VAMS, C-qDBS, W-DBS, QC, Plasma

```{r,eval=TRUE}
#' Define colors for the different phenotypes
col_sample_type <- brewer.pal(9, name = "Set1")[c(9, 2, 1, 3, 4)]
names(col_sample_type) <- c("QC",   # grey
                          "M-VAMS", # blue
                          "W-DBS",  # red
                          "C-qDBS", # green
                          "Plasma") # purple
col_sample <- col_sample_type[sampleData(mse_pos)$sample_type]

#' Access Spectra Object
spectra(mse_pos)
```

## Chromatographic data visualization

BPC plot of all samples (colored by device), including QCs.

```{r, eval=TRUE}
#' Extract and plot BPC for full data
bpc <- chromatogram(mse_pos, aggregationFun = "max")

#png("BPC_pos.png", width = 2000, height = 1500, res = 300)

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6, horiz = TRUE,
       bty = "n")

#dev.off()
```

Filter the data based on the RT observed on the BPC.

```{r, eval=TRUE}
#' Filter the data based on retention time
mse_pos <- filterRt(mse_pos, c(10, 950))

# Plot BPC after filtering
bpc <- chromatogram(mse_pos, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, lwd = 2, cex=0.6,horiz = TRUE, bty = "n")

```

```{r, eval=TRUE}
#' Total ion chromatogram
tic <- chromatogram(mse_pos, aggregationFun = "sum") |>
  bin(binSize = 2) #increase to 8

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

rownames(ticmap) <- colnames(ticmap) <- sampleData(mse_pos)$sample_name
ann <- data.frame(sample_type = sampleData(mse_pos)[, "sample_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
#png("heatmap_pos.png", width = 2000, height = 1500, res = 300)

pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = col_sample_type),
         show_colnames = FALSE,
         show_rownames = FALSE)

#dev.off()
```

Heatmap of TIC similarities. All plasma samples are group together.

## EICs

Upload the list of the internal standards and also some endogenous compounds that will be used to make RT shift correction.

```{r, eval=TRUE}
#' Load list of Internal standard SPLASH LIPIDOMIX
intern_standard <- read_xlsx("splashlipidomix_list.xlsx")
intern_standard$rtmin <- intern_standard$RT - 30 
intern_standard$rtmax <- intern_standard$RT + 20
intern_standard$mzmin <- intern_standard$mz - 0.008 #0.01 was suggested
intern_standard$mzmax <- intern_standard$mz + 0.008
rownames(intern_standard) <- intern_standard$short_name

# Extract EICs for the list
eic_is <- chromatogram(
    mse_pos,
    rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
    mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

#' Add internal standard metadata
fData(eic_is)$mz <- intern_standard$mz
fData(eic_is)$rt <- intern_standard$RT
fData(eic_is)$rtmin <- intern_standard$rtmin
fData(eic_is)$rtmax <- intern_standard$rtmax
fData(eic_is)$short_name <- intern_standard$short_name
rownames(fData(eic_is)) <- intern_standard$short_name

fdata <- fData(eic_is)

#' Summary of IS information
summary(fdata)
head(fdata)

fData(eic_is)[c("CAR", "LysoPC", "dLysoPC", "dLysoPE", "LPC", "SM", 
  "dPG", "dSM", "PC", "dPC", "dPE", "SM1", 
  "dDG", "HexCer", "TG", "dTG", "dCE"), 
              c("short_name", "mz", "rt", "rtmin", "rtmax")] |>
    kable(format = "pipe")
```

We loop below plot the EICs for reference lipids (17 in total): isotope labeled lipids from splash lipidomix and some endogenous lipids.

```{r, eval=TRUE}
# Loop to go through all rows of the eic_is object and plot independent  

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add 
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 #abline(v = median(rtime(eic)), col = "red", lty = 3)
}
```

```{r, eval=TRUE}
# Chromatogram example of three lipids from the list that was imported before
eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
#png("EIC_exampledeuteratedlipids.png", width = 2000, height = 1500, res = 300)
 
par(mfrow = c(1, 3), mar = c(4, 2, 2, 0.5))

plot(eic_deuteratedLysoPC, main = fData(eic_deuteratedLysoPC)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedLysoPC)$rt, col = "red", lty = 3)

plot(eic_deuteratedSM, main = fData(eic_deuteratedSM)$short_name, 
     cex.axis = 0.8, 
     cex.main = 0.8,
     col = paste0(col_sample, 80))
grid()
abline(v = fData(eic_deuteratedSM)$rt, col = "red", lty = 3)

plot(eic_CAR, main = fData(eic_CAR)$short_name,
     cex.axis = 0.8,
     cex.main = 0.8,
      col = paste0(col_sample, 80))
grid()

legend("topright", col = col_sample_type, legend = names(col_sample_type), 
       lty = 1,
       bty = "n")
 
#dev.off()
```

## **Data preprocessing**

### **Chromatographic peak detection**

Below we apply the CentWave algorithm with its default settings on the EICs of deuterated LysoPC, Sphingomyelin and Carnitine ions and evaluate the results.

```{r, eval=TRUE}
param_pd <- CentWaveParam()  #' Use default Centwave parameter for peak detection
param_pd                     #' Look at the default parameters
```

```{r, eval=FALSE}
#' Evaluate for Deuterated LysoPC
dLysoPC_test <- findChromPeaks(eic_deuteratedLysoPC, param = param_pd)
chromPeaks(dLysoPC_test)
```

```{r, eval=FALSE}
#' Evaluate for Deuterated SM
dSM_test <- findChromPeaks(eic_deuteratedSM, param = param_pd)
chromPeaks(dSM_test)
```

```{r}
#' Evaluate for CAR
CAR_test <- findChromPeaks(eic_CAR, param = param_pd)
chromPeaks(CAR_test)
```

We use the deuterated LysoPC of the first sample to extract the *m/z*, calculate the absolute difference and expressed in ppm.

```{r, eval=TRUE}
#' Restrict the data to signal from dLysoPC in the first sample
lpc <- mse_pos[1L] |>
  spectra() |>
  filterRt(rt = c(155, 170)) |>
  filterMzRange(mz = fData(eic_deuteratedLysoPC)["dLysoPC", c("mzmin", "mzmax")])

#' Show the number of peaks per m/z filtered spectra
lengths(lpc)
```

```{r, eval=TRUE}
#' Calculate the difference in m/z values between scans
mz_diff <- lpc |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(lpc))))
```

We therefore, choose a value close to the maximum within this range for parameter `ppm`, i.e., 10 ppm, otherwise work with 2 `ppm` will be to low.

We can now perform the chromatographic peak detection with the adapted settings on our EICs, and test the parameters in some Lipids from the list. In this case deuterated LysoPC, Sphingomelyn and Carnitine.

```{r, eval=TRUE}
#' Parameters adapted for chromatographic peak detection on EICs.
param_pd <- CentWaveParam(peakwidth = c(1, 8), ppm = 10, integrate = 2,
                       snthresh = 2)

#' Evaluate on the dLysoPC ion
dLysoPC_test <- findChromPeaks(eic_deuteratedLysoPC, param = param_pd)
chromPeaks(dLysoPC_test)
```

```{r, eval=TRUE}
#' Evaluate on the Sphingomyelin ion
dSM_test <- findChromPeaks(eic_deuteratedSM, param = param_pd)
chromPeaks(dSM_test)
```

```{r}
#' Evaluate on the Sphingomyelin ion
CAR_test <- findChromPeaks(eic_CAR, param = param_pd)
chromPeaks(CAR_test)
```

With the customized parameters, a chromatographic peak was detected in each sample. Below, we use the [`plot()`](https://rdrr.io/r/graphics/plot.default.html) function on the EICs to visualize these results.

```{r, eval=TRUE}
#' Plot test chromatogram for deuterated LysoPC, Sphingomielyn and Carnitine.
par(mfrow = c(1, 3))
plot(dLysoPC_test, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(dSM_test, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(CAR_test, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

Chromatographic peak detection on EICs.

We can see a peak seems to be detected in each sample for both ions. This indicates that our custom settings are suitable for our dataset. We now proceed and apply them to the entire dataset, extracting EICs again for the same ions to evaluate and confirm that chromatographic peak detection worked as expected.

```{r, eval=FALSE}
#' Using the same settings, but with default snthresh
param_pd <- CentWaveParam(peakwidth = c(1, 8), ppm = 10, integrate = 2, snthresh = 2) ## lowered the sntresth super low and still no detection of the carnitine peaks. 
mse_pos <- findChromPeaks(mse_pos, param = param_pd, chunkSize = 2)

#' Update EIC internal standard object
#' 
#' Below tested a larger mz window to see if the carnitine gets detected but it jsut highlights all the noise. 
# new_mz <- intern_standard[, c("mzmin", "mzmax")]
# new_mz$mzmin <- new_mz$mzmin - 0.01
# new_mz$mzmax <- new_mz$mzmax + 0.01

eics_is_noprocess <- eic_is
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
```

```{r, eval=FALSE}
fData(eic_is) <- fData(eics_is_noprocess)
```

```{r, eval=FALSE}
# Plot to evaluate the chromatographic peak detection results.
rownames(fData(eic_is)) <- intern_standard$short_name

eic_deuteratedLysoPC <- eic_is["dLysoPC"]
eic_deuteratedSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]

par(mfrow = c(1, 3))
plot(eic_deuteratedLysoPC, main = fData(dLysoPC_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dLysoPC_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_deuteratedSM, main = fData(dSM_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(dSM_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()

plot(eic_CAR, main = fData(CAR_test)$short_name,
     col = paste0(col_sample, 80),
     peakBg = paste0(col_sample, 40)[chromPeaks(CAR_test)[, "column"]],
     cex.main = 0.8, cex.axis = 0.8)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

After applying the cinematographic peak detection over the entire dataset, only the first two peaks have been detected properly in all sample, but this does not happen with the Carnitine (not even after changing the signal to noise ratio, or opening the window for the m/z or rt)

**Merge split peaks**

```{r, eval=TRUE}
#' set up the parameter
param_msp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.0015,
                                    minProp = 0.75)
```

```{r, eval=TRUE}
#' Apply on whole dataset
mse_pos <- refineChromPeaks(mse_pos, param = param_msp, chunkSize = 5)
```

```{r}
chromPeakData(mse_pos)$merged |>
                      table()
```

Evaluate the results of the chromatographic peak detection on EICs of e.g. internal standards or other compounds/ions known to be present in the samples. Here we update our EICs objects for the internal standards with the results of the chromatographic peak detection and refinement.

```{r, eval=FALSE}
eics_is_chrompeaks <- eic_is

eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fData(eics_is_chrompeaks)

eic_deuteratedLysoPC <- eic_is["dLysoPC", ]
eic_deuteratedSM <- eic_is["dSM", ]
eic_CAR <- eic_is["CAR"]
```
```{r, eval=FALSE}

# Define output directory (create if it doesn't exist)
outdir <- "peak_detection_ref_lipid"
if (!dir.exists(outdir)) dir.create(outdir)

# Loop through each row and save plots as PNG
for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i, ]

  # Define filename using the row name
  filename <- file.path(outdir, paste0(rownames(fData(eic_is))[i], ".png"))

  # Open PNG device
  png(filename, width = 800, height = 600, res = 120)

  # Plot chromatogram
  plot(
    eic,
    main = rownames(fData(eic_is))[i],  # use row names as titles
    cex.axis = 0.8,
    cex.main = 0.8,
    col = paste0(col_sample, 80)
  )
  grid()
  # abline(v = median(rtime(eic)), col = "red", lty = 3)

  # Close device
  dev.off()
}

cat("All chromatograms saved in:", outdir, "\n")
```

All the compound plots in eic_is are saved locally 

and also run `eic_is` in the console again to ensure that all the eic that have 2 peak are properly separated. Discuss with Johannes to see how to prevent merge.

Number of chromatographic peaks per sample.

```{r, eval= TRUE}
#' Count the number of peaks per sample and summarize them in a table.
data.frame(sample_name = sampleData(mse_pos)$sample_name,
           peak_count = as.integer(table(chromPeaks(mse_pos)[, "sample"]))) |>
  t() |>
  kable(format = "pipe")
```

A similar number of chromatographic peaks was identified within the various samples of the data set. In all cases W-DBS has the lower number of features.

### **Retention time alignment**

To evaluate the presence of a shift we extract and plot below the BPC from the QC samples.

```{r, eval=TRUE}
#' Get QC samples
QC_samples <- sampleData(mse_pos)$sample_type == "QC"

#' extract BPC
mse_pos[QC_samples] |>
    chromatogram(aggregationFun = "max", chromPeaks = "none") |>
    plot(col = col_sample_type["QC"], main = "BPC of QC samples") |>
    grid()
```

Initial correspondence analysis. Parameters do not need to be fully optimized.

```{r, eval=TRUE}
# Initial correspondence analysis. # run for every device
param_c <- PeakDensityParam(sampleGroups = sampleData(mse_pos)$sample_type == "QC",
                          minFraction = 0.9,
                          binSize = 0.01, ppm = 10,
                          bw = 2)
mse_pos <- groupChromPeaks(mse_pos, param = param_c)

plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_c,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]], 20),
    peakPch = 16)
```

Below we perform the alignment of our data set based on retention times of anchor peaks defined in the subset of QC samples.

```{r, eval=TRUE}
#' Define parameters of choice
subset <- which(sampleData(mse_pos)$sample_type == "QC")
param_c <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                         subsetAdjust = "average",
                         subset = subset)

#' Perform the alignment
mse_pos <- adjustRtime(mse_pos, param = param_c)
```

Alignment adjusted the retention times of all spectra in the data set, as well as the retention times of all identified chromatographic peaks.

```{r, eval=TRUE}
#' Visualize alignment results
plotAdjustedRtime(mse_pos, col = paste0(col_sample, 80), peakGroupsPch = 1)
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")
```

Only little adjustments needed to be performed (shifts of at maximum 1.5 second as can be seen in the plot above).

We can also compare the BPC before and after alignment.

```{r, eval = TRUE}
#' Get data before alignment
mse_pos_raw <- dropAdjustedRtime(mse_pos)

#' Apply the adjusted retention time to our dataset
mse_pos_raw <- applyAdjustedRtime(mse_pos)
```

```{r, eval= TRUE}
#' Plot the BPC before and after alignment
par(mfrow = c(2, 1), mar = c(2, 1, 1, 0.5))
chromatogram(mse_pos_raw, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC before alignment", col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)

chromatogram(mse_pos, aggregationFun = "max", chromPeaks = "none") |>
    plot(main = "BPC after alignment",
         col = paste0(col_sample, 80))
grid()
legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n", horiz = TRUE)
```

There is no shift observed.

We next evaluate the impact of the alignment on the EICs of the selected internal standards. We thus below first extract the ion chromatograms after alignment, and subsequently plot them.

```{r, eval=TRUE}
#' Extract the EICs for the test ions before alignment
old_eic_dLysoPC <- eic_is["dLysoPC"]
old_eic_dSM <- eic_is["dSM"]
old_eic_CAR <- eic_is["CAR"]

#' Update the EICs
eic_is <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eic_is) <- fdata

#' Extract the EICs for the test ions
eic_dLysoPC <- eic_is["dLysoPC"]
eic_dSM <- eic_is["dSM"]
eic_CAR <- eic_is["CAR"]
```

```{r, eval=TRUE}
par(mfrow = c(2, 2), mar = c(4, 4.5, 2, 1))

plot(old_eic_dLysoPC, main = "dLysoPC before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(eic_dLysoPC, main = "dLysoPC after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dLysoPC", "RT"], col = "red", lty = 3)

plot(old_eic_dSM, main = "Sphingomyelin before alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

legend("topright", col = col_sample_type,
       legend = names(col_sample_type), lty = 1, bty = "n")

plot(eic_dSM, main = "Sphingomyelin after alignment",
     peakType = "none", col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["dSM", "RT"], col = "red", lty = 3)

plot(old_eic_CAR, main = "CAR before alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)

plot(eic_CAR, main = "CAR after alignment", peakType = "none",
     col = paste0(col_sample, 80))
grid()
abline(v = intern_standard["CAR", "RT"], col = "red", lty = 3)
```

In addition to a visual inspection of the results, we also evaluate the impact of the alignment by comparing the variance in retention times for internal standards before and after alignment.

```{r, eval= TRUE}
#' Extract the matrix with all chromatographic peaks and add a column
#' with the ID of the chromatographic peak
chrom_peaks <- chromPeaks(mse_pos) |> as.data.frame()
chrom_peaks$peak_id <- rownames(chrom_peaks)

#' Define the parameters for the matching and filtering of the matches
p_1 <- MzRtParam(ppm = 50, toleranceRt = 10)
p_2 <- SingleMatchParam(duplicates = "top_ranked", column = "target_maxo",
                        decreasing = TRUE)

#' Iterate over samples and identify for each the chromatographic peaks
#' with similar m/z and retention time than the onse from the internal
#' standard, and extract among them the ID of the peaks with the
#' highest intensity.
intern_standard_peaks <- lapply(seq_along(mse_pos), function(i) {
    tmp <- chrom_peaks[chrom_peaks[, "sample"] == i, , drop = FALSE]
    
     # Convert to S4 DataFrames
    query  <- DataFrame(intern_standard)
    target <- DataFrame(tmp)
    
    
    mtch <- matchValues(query, target,
                        mzColname = c("mz", "mz"),
                        rtColname = c("RT", "rt"),
                        param = p_1)
    mtch <- filterMatches(mtch, p_2)
    mtch$target_peak_id
}) |>
    do.call(what = cbind)
```

We have now for each internal standard the ID of the chromatographic peak in each sample that most likely represents signal from that ion. We can now extract the retention times for these chromatographic peaks before and after alignment.

```{r, eval= TRUE}
#' Define the index of the selected chromatographic peaks in the
#' full chromPeaks matrix
idx <- match(intern_standard_peaks, rownames(chromPeaks(mse_pos)))

#' Extract the raw retention times for these
rt_raw <- chromPeaks(mse_pos_raw)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))

#' Extract the adjusted retention times for these
rt_adj <- chromPeaks(mse_pos)[idx, "rt"] |>
    matrix(ncol = length(mse_pos_raw))
```

We can now evaluate the impact of the alignment on the retention times of internal standards across the full data set:

```{r, eval=FALSE}
list(all_raw = rowSds(rt_raw, na.rm = TRUE),
     all_adj = rowSds(rt_adj, na.rm = TRUE)
     ) |>
    vioplot()
grid()
```

### **Correspondence**

The goal of the correspondence analysis is to identify chromatographic peaks that originate from the same types of ions in all samples of an experiment and to group them into LC-MS *features*.

```{r, eval=TRUE}
#' Optimized parameter for the grouping and apply them to the test ions BPC, note that bw=2
#' fix grouping, NA for QCs
group_mse <- sampleData(mse_pos)$sample_type
group_mse[group_mse =="QC"] <- NA 
group_mse <- as.factor(group_mse)
param_cor <- PeakDensityParam(sampleGroups = group_mse, bw = 2)
param_cor
```

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedLysoPC, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedLysoPC)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated LysoPC.

```{r,eval=TRUE}
plotChromPeakDensity(
    eic_deuteratedSM, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_deuteratedSM)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters, deuterated SM.

```{r, eval=TRUE}
plotChromPeakDensity(
    eic_CAR, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eic_CAR)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eic_CAR)[, "sample"]], 20),
    peakPch = 16)
```

Correspondence analysis with optimized parameters. Carnitine does not have any chromatographic peak, check what is happening.

We can observe for the first two ions that the peaks are grouped accurately into a single feature. So we proced to define the settings for the param\_ cor and apply them to the whole data.

```{r, eval=TRUE}
#' Define the settings for the param_cor
param_cor <- PeakDensityParam(sampleGroups = group_mse,
                          minFraction = 0.5, binSize = 0.01, ppm = 10,
                          bw = 2)

#' Apply to whole data
mse_pos <- groupChromPeaks(mse_pos, param = param_cor)
```

Evaluation of the correspondence results in other EIC in the IS list similar to the deuterated LysoPC.

```{r, eval=TRUE}
#' Extract chromatogram for an m/z similar to the one of the labeled methionine
chr_test <- chromatogram(mse_pos,
                         mz = as.matrix(intern_standard["LysoPC",
                                                        c("mzmin", "mzmax")]),
                         rt = c(120, 184),
                         aggregationFun = "max")
plotChromPeakDensity(
    chr_test, simulate = FALSE,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(chr_test)[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(chr_test)[, "sample"]], 20),
    peakPch = 16)
```

This corresponds to non-deuterated LysoPC and we can see that it can distinguish between the two peaks.

```{r, eval=FALSE}
#### TEST for review Evaluation of the correspondance results in 17 EICs in the IS list.



# Loop to go through all rows of the eic_is object and plot independent after correspondance analysis 

for (i in seq_len(nrow(eic_is))) {
  eic <- eic_is[i,]
  #png add 
  plot(eic,
       main = rownames(fData(eic_is))[i], #use the rawnames as tittles
       cex.axis = 0.8,
       cex.main = 0.8,
       col = paste0(col_sample, 80)
  )
  grid()
 #abline(v = median(rtime(eic)), col = "red", lty = 3)
}





param_peak <- 


# Assuming lcms1 is your raw data object (XCMSnExp)
mse_posp <- findChromPeaks(mse_pos, param = param)

# Extract chromatograms again with peaks included
eic_isp <- chromatogram(mse_pos,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))

for (i in seq_len(nrow(eic_isp))) {
  eicp <- eic_isp[i, ]
  plotChromPeakDensity(
    eicp, param = param_cor,
    col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(mse_pos)[, "sample"]], # peaks now exist
    peakBg = paste0(col_sample[chromPeaks(mse_pos)[, "sample"]], 20),
    peakPch = 16
  )
  grid()
}
```

### Gap filling

We next examine how prevalent missing values are in our present dataset:

```{r}
#' Percentage of missing values
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100

idx_qc <- sampleData(mse_pos)$sample_type == "QC"
#' Percentage of missing values
sum(is.na(featureValues(mse_pos[idx_qc, keepFeatures = TRUE]))) /
    length(featureValues(mse_pos[idx_qc, keepFeatures = TRUE])) * 100
```

We can observe a substantial number of missing values values in our dataset. 

Fill the missing values in the entire data set.

```{r}
#' Fill in the missing values in the whole dataset
mse_pos <- fillChromPeaks(mse_pos, param = ChromPeakAreaParam(), chunkSize = 5)

#' Percentage of missing values after gap-filling
sum(is.na(featureValues(mse_pos))) /
    length(featureValues(mse_pos)) * 100
```

```{r}
#' Get only detected signal in QC samples
vals_detect <- featureValues(mse_pos, filled = FALSE)[, QC_samples]

#' Get detected and filled-in signal
vals_filled <- featureValues(mse_pos)[, QC_samples]

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)

#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080")
grid()
abline(0, 1)
```

## Library correction

```{r, eval=FALSE}
#' Default parameter for the grouping and apply them to the test ions BPC
group <- eic_is$sample_type
group[group =="QC"] <- NA 
param1 <- PeakDensityParam(sampleGroups = group , bw = 2)
eic_is_corr <- groupChromPeaks(eic_is, param = param1)
featureDefinitions(eic_is_corr[6,]) ## to be removed
fData(eic_is_corr) <- fData(eic_is)

exp_rt <- vector("numeric", length = nrow(eic_is_corr))
dr <- "./ref_lipid_image/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(eic_is_corr))) {
  table <- featureDefinitions(eic_is_corr[i,])
  if (!nrow(table)) {
    message("compound not detected: ", fData(eic_is_corr)$short_name[i])
    exp_rt[i] <- NA
    next
  }
  png(filename = paste0(dr, "feature_compound_", i, ".png"))
  plot(eic_is_corr[i,]) # improve code here
  dev.off()
  if (nrow(table) > 1)
    rt <- table[which.max(table$rtmed), "rtmed"]
  else rt <- table$rtmed
  exp_rt[i] <- rt
}

# below is the reference lipid table 
ref_rt <- intern_standard$RT # is the ref RT 

# res is the exp RT 
intern_standard$exp_rt <- exp_rt

diff_rt <- intern_standard$RT - intern_standard$exp_rt
mean(diff_rt, na.rm = TRUE)

## Fit the model 
## Not doing loess because we need to do extrapolation later on.
# fit <- loess(exp_rt ~ ref_rt, span = 0.5)

fit <- lm(exp_rt ~ poly(ref_rt, 2, raw = TRUE) ) ## limma ?, number of degrees ? 
summary(fit)
plot(fit)
```

Sara will check if the shift see m to be similar to what she has or not.

Library fix

```{r}
lipid_database <- read_xlsx("LipidDatabase_R.xlsx", sheet = 4L)
rt_sd <- as.numeric(lipid_database$`RT ESI(+) (min)`) * 60
corrected_rt <- predict(fit, newdata = data.frame(ref_rt = rt_sd))
## is the warning a problem ? maybe i'm missing something.
# why is corrected the length of the fit points ? 

lipid_database$rt_adjusted <- corrected_rt 

rt_sd - corrected_rt

mean(rt_sd - corrected_rt)
```

## missing steps ?? 
These steps could be done after annotation, as we are in a semi-targeted workflow

- plot some PCA
- normalization
  - per volume 
  - per internal standard
- filtering (not sure we need that) 

## Annotate/ match 

First we need to create a summarizedExperiement()

- maybe separate the QCs into another object or do that later. 

```{r}
res <- quantify(mse_pos, method = "sum", filled = FALSE)
res
save(res, "")

assays(res)$raw_filled <- featureValues(mse_pos, method = "sum",
                                        filled = TRUE)
```


```{r}
query <- as.data.frame(rowData(res))
target <- as.data.frame(lipid_database)
target$mz <- as.numeric(target$mz)
param_match <- MzRtParam(tolerance = 0.001, ppm = 10, toleranceRt = 4) ## to be refined
mtch <- matchValues(query, target, param = param_match, mzColname = c("mzmed", "mz"), rtColname = c("rtmed", "rt_adjusted"))

mtch

## only keep query that has matched 
mtch <- mtch[whichQuery(mtch)]
mtched_data <- matchedData(mtch) ## can do some stuff with that table 

length(unique(mtched_data$`target_lipid name`)) ## number of lipid matched

length(unique(mtched_data$`target_LIPID SUBCLASS [ABBREV]`)) 
```

List of plots that would be nice
- 
- 


later on discuss how to deal with features tha tmatches multiple compound.

we could concatenate the annotation to the summarized experiment object. 
Do the normalization, quantitation, plotting after,.. 


## Statistical analysis 
