---
title: "POS&NEG"
format: html
editor: source
---

## What to do after we have a res object for positive and negative ionization mode?

Put both objects together and add an extra column that specify the ionization mode of detection 

```{r}

```

  Rule to eliminate duplicates: look in the data the duplicated entries for lipid name and pick up the entry that has the higher abundance, make sure that the retention times are similar 

  Then perform univariate and multivariate analysis? not sure if is this what we want 

# Downstream analysis:

# Overlaping features

```{r}
#create a logical matrix to be able to make the upsetplot
df_logical <- !is.na(assay(res, "raw_filled"))

# Extract group code (letter before "_pos.mzML")
code <- sub(".*([MWCP])_pos\\.mzML$", "\\1", colnames(df_logical))

# Map codes to descriptive names
code_map <- c(
  M = "M-VAMS",
  C = "C-qDBS",
  W = "W-DBS",
  P = "Plasma"
)
groups <- code_map[code]

# Collapse by group using 70% threshold
result <- sapply(unique(groups), function(g) {
  # number of samples in this group
  n <- sum(groups == g)
  # proportion of TRUEs across samples in this group
  rowMeans(df_logical[, groups == g, drop = FALSE]) >= 0.7
})

# Result: logical matrix compounds x groups
result[1:5, ]

#make the upsetplot
binary_data <-apply(result, MARGIN = 2, as.integer)
upset(as.data.frame(binary_data), nset = 6, sets = colnames(result), keep.order = TRUE)
```

## Data distribution 

```{r}
#' Define the colors for the plot
col_sample <- col_sample_type[res$sample_type]

#' Log transform and scale the data for PCA analysis
vals <- assay(res, "ISnorm_filled_imputed") |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- cbind(vals, sample_type = res$sample_type)
autoplot(pca_res, data = vals_st , colour = 'sample_type', scale = 0) +
    scale_color_manual(values = col_sample_type)
```

PCA of the data after normalization and quality control, without QCs
- Plotting with highlighting different colors would be good.

  
  
