---
title: "POS&NEG"
format: html
editor: source
---

## What to do after we have a res object for positive and negative ionization mode?

Put both objects together and add an extra column that specify the ionization mode of detection 

```{r}
#| label: load-packages
library(alabaster.se)
library(SummarizedExperiment)
library(UpSetR)
library(ggfortify)
```

### Load positive and negative mode results

```{r}
#| label: load-data
# Read the positive and negative results
res_pos <- readObject("objects/final_res_pos")
res_neg <- readObject("objects/final_res_neg")

cat("Positive mode:", nrow(res_pos), "features x", ncol(res_pos), "samples\n")
cat("Negative mode:", nrow(res_neg), "features x", ncol(res_neg), "samples\n")
```

### Add ionization mode annotation

```{r}
#| label: add-mode
# Add ionization mode column to rowData
rowData(res_pos)$ionization_mode <- "positive"
rowData(res_neg)$ionization_mode <- "negative"
```

### Align samples between modes

```{r}
#| label: align-samples
# Harmonize sample names (remove _pos/_neg suffix for matching)
pos_order <- sub("_pos\\.mzML$", "", colnames(res_pos))
neg_order <- sub("_neg\\.mzML$", "", colnames(res_neg))

# Reorder neg to match pos sample order
neg_reorder_idx <- match(pos_order, neg_order)
res_neg <- res_neg[, neg_reorder_idx]

# Verify order matches
stopifnot(all(pos_order == sub("_neg\\.mzML$", "", colnames(res_neg))))
```

### Identify and handle duplicate lipids

Rule to eliminate duplicates: for duplicated lipid names between modes, pick the entry with higher abundance, ensuring retention times are similar (<30 sec difference). We use `target_lipid_name_unique` to properly handle isomers.

```{r}
#| label: find-duplicates
# Find common lipids between pos and neg using unique names (handles isomers)
common_lipids <- intersect(
    rowData(res_pos)$target_lipid_name_unique,
    rowData(res_neg)$target_lipid_name_unique
)

cat("Number of common lipids:", length(common_lipids), "\n")

# Create comparison data frame
comparison_df <- data.frame(
    lipid_name_unique = common_lipids,
    pos_idx = match(common_lipids, rowData(res_pos)$target_lipid_name_unique),
    neg_idx = match(common_lipids, rowData(res_neg)$target_lipid_name_unique)
)

# Calculate mean abundances and retention times
comparison_df$pos_mean_abundance <- sapply(comparison_df$pos_idx, function(i) {
    mean(assay(res_pos, "ISnorm_filled_imputed")[i, ], na.rm = TRUE)
})
comparison_df$neg_mean_abundance <- sapply(comparison_df$neg_idx, function(i) {
    mean(assay(res_neg, "ISnorm_filled_imputed")[i, ], na.rm = TRUE)
})

comparison_df$pos_rt <- rowData(res_pos)$rtmed[comparison_df$pos_idx]
comparison_df$neg_rt <- rowData(res_neg)$rtmed[comparison_df$neg_idx]
comparison_df$rt_diff <- abs(comparison_df$pos_rt - comparison_df$neg_rt)

# Flag similar RT (within 30 seconds)
comparison_df$rt_similar <- comparison_df$rt_diff < 30

# Decide which mode to keep (higher abundance wins)
comparison_df$keep_mode <- ifelse(
    comparison_df$pos_mean_abundance >= comparison_df$neg_mean_abundance,
    "positive", "negative"
)

cat("\nRetention time differences summary:\n")
summary(comparison_df$rt_diff)

cat("\nLipids with dissimilar RT (>30s, kept in both modes):\n")
comparison_df[!comparison_df$rt_similar, c("lipid_name_unique", "pos_rt", "neg_rt", "rt_diff")]
```

### Remove duplicates

```{r}
#| label: remove-duplicates
# Positive: remove common lipids where neg has higher abundance AND similar RT
pos_remove_idx <- comparison_df$pos_idx[
    comparison_df$keep_mode == "negative" & comparison_df$rt_similar
]

# Negative: remove common lipids where pos has higher abundance AND similar RT
neg_remove_idx <- comparison_df$neg_idx[
    comparison_df$keep_mode == "positive" & comparison_df$rt_similar
]

cat("Removing", length(pos_remove_idx), "duplicates from positive mode\n")
cat("Removing", length(neg_remove_idx), "duplicates from negative mode\n")

# Create filtered datasets
res_pos_filtered <- res_pos[-pos_remove_idx, ]
res_neg_filtered <- res_neg[-neg_remove_idx, ]

cat("\nPositive after filtering:", nrow(res_pos_filtered), "features\n")
cat("Negative after filtering:", nrow(res_neg_filtered), "features\n")
```

### Combine into single SummarizedExperiment

```{r}
#| label: combine-results
# Standardize sample names (remove polarity suffix)
colnames(res_pos_filtered) <- sub("_pos\\.mzML$", "", colnames(res_pos_filtered))
colnames(res_neg_filtered) <- sub("_neg\\.mzML$", "", colnames(res_neg_filtered))

# Standardize colData columns that differ between modes
colData(res_pos_filtered)$file_name <- sub("_pos\\.mzML$", "", colData(res_pos_filtered)$file_name)
colData(res_neg_filtered)$file_name <- sub("_neg\\.mzML$", "", colData(res_neg_filtered)$file_name)

# Remove mode-specific columns from colData
colData(res_pos_filtered)$spectraOrigin <- NULL
colData(res_neg_filtered)$spectraOrigin <- NULL
colData(res_pos_filtered)$polarity <- NULL
colData(res_neg_filtered)$polarity <- NULL

# Ensure ionization_mode is in rowData (should have been added earlier)
if (!"ionization_mode" %in% colnames(rowData(res_pos_filtered))) {
    rowData(res_pos_filtered)$ionization_mode <- "positive"
}
if (!"ionization_mode" %in% colnames(rowData(res_neg_filtered))) {
    rowData(res_neg_filtered)$ionization_mode <- "negative"
}

# Combine row-wise (features are rows)
res_combined <- rbind(res_pos_filtered, res_neg_filtered)

cat("Combined dataset:", nrow(res_combined), "features x", ncol(res_combined), "samples\n")
cat("Features from positive mode:", sum(rowData(res_combined)$ionization_mode == "positive"), "\n")
cat("Features from negative mode:", sum(rowData(res_combined)$ionization_mode == "negative"), "\n")
```

### Save combined results

```{r}
#| label: save-combined
saveObject(res_combined, "objects/res_combined")
```

### Summary table of duplicate handling

```{r}
#| label: duplicate-summary
# Create summary of what happened to duplicates
dup_summary <- comparison_df |>
    dplyr::mutate(
        decision = dplyr::case_when(
            !rt_similar ~ "Kept both (different RT)",
            keep_mode == "positive" ~ "Kept positive (higher abundance)",
            keep_mode == "negative" ~ "Kept negative (higher abundance)"
        )
    )

cat("Duplicate resolution summary:\n")
table(dup_summary$decision)

# Show which lipids were kept in which mode
cat("\nDetailed duplicate resolution:\n")
dup_summary[, c("lipid_name_unique", "decision",
 "pos_mean_abundance", "neg_mean_abundance")]
``` 

# Downstream analysis:

## Overlapping features across sample types

```{r}
#| label: upset-plot
# Create a logical matrix to be able to make the upsetplot
df_logical <- !is.na(assay(res_combined, "raw_filled"))

# Extract group code from sample names
code <- sub(".*([MWCP])$", "\\1", colnames(df_logical))

# Map codes to descriptive names
code_map <- c(
    M = "M-VAMS",
    C = "C-qDBS",
    W = "W-DBS",
    P = "Plasma"
)
groups <- code_map[code]

# Collapse by group using 70% threshold
result <- sapply(unique(groups), function(g) {
    n <- sum(groups == g)
    rowMeans(df_logical[, groups == g, drop = FALSE]) >= 0.7
})

# Make the upsetplot
binary_data <- apply(result, MARGIN = 2, as.integer)
upset(as.data.frame(binary_data), nsets = 6, sets = colnames(result), keep.order = TRUE)
```

## Data distribution 

```{r}
#| label: pca-setup
# Define colors for sample types
col_sample_type <- c(
    "QC" = "#1b9e77",
    "sample" = "#d95f02"
)

# Get sample type from column data
res_combined$sample_type <- ifelse(
    grepl("^QC", colnames(res_combined)), "QC", "sample"
)
col_sample <- col_sample_type[res_combined$sample_type]
```

```{r}
#| label: pca-all
# Log transform and scale the data for PCA analysis
# Convert to matrix first to avoid DelayedArray issues
vals <- assay(res_combined, "ISnorm_filled_imputed") |>
    as.matrix() |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- as.data.frame(vals)
vals_st$sample_type <- res_combined$sample_type

autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
    scale_color_manual(values = col_sample_type) +
    labs(title = "PCA - All samples (Positive + Negative combined)")
```

### PCA without QCs

```{r}
#| label: pca-no-qc
# Filter out QC samples
is_sample <- res_combined$sample_type == "sample"
res_samples <- res_combined[, is_sample]

# Extract sample type information (M-VAMS, C-qDBS, W-DBS, Plasma)
sample_code <- sub("^D\\d{2}([MWCP])$", "\\1", colnames(res_samples))
sample_type_full <- c(
    M = "M-VAMS",
    C = "C-qDBS",
    W = "W-DBS",
    P = "Plasma"
)[sample_code]

# Define colors for sample types
col_sample_full <- c(
    "M-VAMS" = "#e41a1c",
    "C-qDBS" = "#377eb8",
    "W-DBS" = "#4daf4a",
    "Plasma" = "#984ea3"
)

# Log transform and scale
vals_samples <- assay(res_samples, "ISnorm_filled_imputed") |>
    as.matrix() |>
    t() |>
    log2() |>
    scale(center = TRUE, scale = TRUE)

pca_samples <- prcomp(vals_samples, scale = FALSE, center = FALSE)

vals_df <- as.data.frame(vals_samples)
vals_df$sample_type <- sample_type_full

autoplot(pca_samples, data = vals_df, colour = "sample_type", scale = 0) +
    scale_color_manual(values = col_sample_full) +
    labs(title = "PCA - Samples only (without QCs)")
```

### PCA colored by ionization mode

```{r}
#| label: pca-by-mode
# Transpose to get features as observations
vals_features <- assay(res_samples, "ISnorm_filled_imputed") |>
    as.matrix() |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE) |>
    t()

pca_features <- prcomp(vals_features, scale = FALSE, center = FALSE)

feature_df <- data.frame(
    PC1 = pca_features$x[, 1],
    PC2 = pca_features$x[, 2],
    ionization_mode = rowData(res_samples)$ionization_mode
)

ggplot(feature_df, aes(x = PC1, y = PC2, color = ionization_mode)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("positive" = "#e41a1c", "negative" = "#377eb8")) +
    labs(title = "PCA of features colored by ionization mode") +
    theme_bw()
```

  
  
