---
title: "METFORMIN-HIIE Study — Stage 1: LC-MS Data Preprocessing"
subtitle: "Data import and validation, Peak detection, RT alignment, correspondence and gap filling"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "METFORMIN-HIIE"
format:
  html:
    output-file: "METFORMIN-HIIE_Preprocessing_pos.html"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

```{r}
library("tictoc")
tic("Quarto compilation")
```

# Overview

::: {.callout-tip}
## Study-specific configuration
This document performs **positive ionization mode** preprocessing for the
high intensity interval exercise **METFORMIN-HIIE study** 
(exercise vs resting comparison). The corresponding negative ionization mode 
workflow is available in the `negative/` folder.
:::

This document handles the preprocessing of Untargeted LC-MS lipidomics data:

1. **Data Import** — Imports raw `.mzML` data from (local files, SQLite, or MetaboLights).
2. **Dataset Investigation** — Visualizes base peak chromatograms (BPC) and extracted ion chromatograms (EIC) for reference lipids.
3. **Chromatographic Peak Detection** — Detects chromatographic features using the CentWave algorithm.
4. **Retention Time Alignment** — Aligns retention time across samples using QC-based peak groups.
5. **Correspondence** — Groups corresponding feature across samples.
6. **Gap Filling** — Fill missing peak values.

The output is a preprocessed `MsExperiment` and `SummarizedExperiment` saved to
the `objects/` folder, and used as input for lipid annotation in `Annotation_pos.qmd`.

## Folder Hierarchy

```
METFORMIN-HIIE_study/
├── LipidDatabase_R.xlsx                 # SRM 1950-derived Lipid database (shared by both polarities)
├── positive/                            # Positive ionization mode (this folder)
│   ├── Preprocessing_pos.qmd            # Stage 1 (this file)
│   ├── Annotation_pos.qmd               # Stage 2
│   ├── seq_pos_METFORMIN-HIIE.xlsx      # Sample sequence
│   ├── pos_lipid_reference_set.xlsx     # Lipid reference set
│   ├── data/                            # .mzML files
│   ├── objects/                         # Saved R objects (auto-generated)
│   └── figures/                         # Output figures (auto-generated)
├── negative/                            # Negative ionization mode
│   ├── Preprocessing_neg.qmd
│   ├── Annotation_neg.qmd
│   └── ...
└── POS_NEG_merge.qmd                   # Step 3: Cross-Mode Data Integration
```

## Required Input Files

Before running this workflow, ensure you have:

1. **LC-MS raw data** (`.mzML` files) in the `data/` folder
2. **Sample sequence file** (`seq_pos_METFORMIN-HIIE.xlsx`) in this folder
3. **SRM 1950-derived lipid database** (`LipidDatabase_R.xlsx`) in the parent study folder
4. **Lipid Reference Set** (`pos_lipid_reference_set.xlsx`) in this folder

---

# Configuration

```{r config}
#| label: configuration

# ============================================================================
# METFORMIN-HIIE STUDY CONFIGURATION
# ============================================================================

# Project root: relative path from this file to the main project directory
PROJECT_ROOT <- "../../.."

# Study identifier
STUDY_ID <- "METFORMIN-HIIE"

# Ionization mode: "pos" or "neg"
POLARITY <- "pos"

# Input data source: "local", "sqlite", or "metaboLights"
DATA_SOURCE <- "sqlite"

# MetaboLights study ID (only used if DATA_SOURCE = "metaboLights")
MTBLS_ID <- NULL

# SQLite database path (only used if DATA_SOURCE = "sqlite")
SQLITE_DB <- NULL

# Number of CPU cores for parallel processing (use 1-2 below your max)
CORES_NB <- 4

# Expected sample types in the sequence file
EXPECTED_SAMPLE_TYPES <- c("QC", "Exercise", "Resting")

# Color palette for sample types
SAMPLE_TYPE_COLORS <- c(
  "QC" = "#999999",
  "Exercise" = "#377EB8",
  "Resting" = "#E41A1C"
)

# --- MetaboLights options (only used if DATA_SOURCE = "metaboLights") ---
MTBLS_ASSAY_POS <- NULL
MTBLS_ASSAY_NEG <- NULL

MTBLS_COLUMN_MAPPING <- list(
  sample_name     = "Sample Name",
  file_name       = "Raw Spectral Data File",
  sample_type     = "Factor Value[Sample type]",
  injection_index = "Comment[injection_index]"
)

MTBLS_FILTER_COLUMN <- NULL
MTBLS_FILTER_VALUE  <- NULL
MTBLS_EXCLUDE_TYPES <- c("Blank")

# Retention time filter range (seconds) - adjust based on your BPC
RT_FILTER_MIN <- 10
RT_FILTER_MAX <- 950

# Chromatographic peak detection parameters
PEAK_WIDTH <- c(4, 8)
PPM <- 10
SN_THRESHOLD <- 2

# EIC window setting
eic_rt_window_left <- 40
eic_rt_window_right <- 20
```

# Setup

## Load Packages and Helper Functions

```{r load-helpers}
#| label: load-helpers

source(file.path(PROJECT_ROOT, "R/lipid_helpers.R"))
source(file.path(PROJECT_ROOT, "R/create_sqlite_database.R"))
load_lipid_packages(verbose = TRUE)
```

## Setup Folder Structure

```{r setup-folders}
#| label: setup-folders

folders <- setup_folders(polarity = POLARITY)

SEQ_FILE <- paste0("seq_", POLARITY, "_", STUDY_ID, ".xlsx")
IS_FILE <- if (POLARITY == "pos") "pos_lipid_reference_set.xlsx" else "neg_lipid_reference_set.xlsx"
DATA_SUFFIX <- paste0("_", POLARITY, ".mzML")

# Setup parallel processing
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(CORES_NB))
} else {
  register(SnowParam(CORES_NB))
}
```

---

# Data Import

Load LC-MS data based on the selected `DATA_SOURCE` option.

::: {.panel-tabset}

## Local mzML Files

Load data from local `.mzML` files in your `data/` folder.

## SQLite Database

Load from a SQLite database. If the database does not exist yet, it will be
automatically created from the local mzML files.

## MetaboLights

Download data directly from MetaboLights (requires internet connection).

:::

## Validate Input Files

```{r validate-inputs}
#| label: validate-inputs

validate_metadata(
  SEQ_FILE,
  expected_sample_types = EXPECTED_SAMPLE_TYPES,
  data_dir = if (DATA_SOURCE == "local") folders$data else NULL
)

validate_reference_lipids(IS_FILE)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Input Validation
Review the output above for any **errors** or **warnings**.
Fix all reported issues in the input files **before continuing**.
:::

```{r load-data}
#| label: load-data

seq_data <- read_xlsx(SEQ_FILE, col_names = TRUE) |> as.data.frame()
seq_data$polarity <- POLARITY

if (DATA_SOURCE == "local") {
  message("Loading data from local mzML files...")
  spectra_files <- file.path(folders$data, seq_data$file_name)

  missing_files <- spectra_files[!file.exists(spectra_files)]
  if (length(missing_files) > 0) {
    stop("Missing files:\n", paste(missing_files, collapse = "\n"))
  }

  mse <- readMsExperiment(spectraFiles = spectra_files, sampleData = seq_data)

} else if (DATA_SOURCE == "sqlite") {
  if (is.null(SQLITE_DB)) {
    SQLITE_DB <- file.path(folders$data, paste0(STUDY_ID, "_", POLARITY, ".sqlite"))
  }
  if (!file.exists(SQLITE_DB)) {
    message("SQLite database not found at '", SQLITE_DB, "'. Creating from mzML files...")
    create_sqlite_database(
      study_id  = STUDY_ID,
      polarity  = POLARITY,
      seq_file  = SEQ_FILE,
      data_dir  = folders$data,
      output_db = SQLITE_DB
    )
  }
  mse <- load_from_sqlite(SQLITE_DB, seq_data)

} else if (DATA_SOURCE == "metaboLights") {
  assay_name <- if (POLARITY == "pos") MTBLS_ASSAY_POS else MTBLS_ASSAY_NEG
  if (is.null(assay_name)) stop("Set MTBLS_ASSAY_POS / MTBLS_ASSAY_NEG in the Configuration section.")

  mse <- load_from_metaboLights(
    mtbls_id = MTBLS_ID,
    assay_name = assay_name,
    column_mapping = MTBLS_COLUMN_MAPPING,
    filter_column = MTBLS_FILTER_COLUMN,
    filter_value = MTBLS_FILTER_VALUE,
    exclude_sample_types = MTBLS_EXCLUDE_TYPES,
    polarity = POLARITY
  )

} else {
  stop("Invalid DATA_SOURCE: '", DATA_SOURCE, "'. Use 'local', 'sqlite', or 'metaboLights'.")
}

message("✓ Loaded ", length(mse), " samples")
sampleData(mse)
```

## Setup Color Palettes

```{r setup-colors}
#| label: setup-colors

palette_sample_type <- create_color_palette(
  mse,
  column = "sample_type",
  custom_colors = SAMPLE_TYPE_COLORS
)

message("Color palette for sample_type:")
print(palette_sample_type)
```

# Dataset Investigation

## Chromatographic Data Visualization

```{r bpc-raw}
#| label: bpc-raw
#| fig-cap: "Base Peak Chromatogram (BPC) of all samples colored by sample type"

plot_bpc(mse, palette = palette_sample_type, main = "BPC - All Samples")
```

::: {.callout-important}
## 📋 USER CHECKPOINT: RT Filter Range
Review the BPC above and adjust `RT_FILTER_MIN` and `RT_FILTER_MAX` in the
**Configuration** section if needed. The current RT filter range is
**`r RT_FILTER_MIN`** to **`r RT_FILTER_MAX`** seconds.
:::

```{r filter-rt}
#| label: filter-rt

mse <- filterRt(mse, c(RT_FILTER_MIN, RT_FILTER_MAX))
```

```{r bpc-filtered}
#| label: bpc-filtered
#| fig-cap: "BPC after retention time filtering"

plot_bpc(mse, palette = palette_sample_type,
         main = "BPC - After RT Filtering")
```

## Reference Lipid EIC

```{r load-is}
#| label: load-is
#| tbl-cap: "Reference lipids for RT adjustment"

intern_standard <- prepare_reference_lipids(
  IS_FILE,
  rt_window_left = eic_rt_window_left,
  rt_window_right = eic_rt_window_right
)

eic_sample_subset <- sampleData(mse)$sample_type == "QC"

eic_is <- extract_is_eics(mse, intern_standard,
                          sample_subset = eic_sample_subset)
fdata <- fData(eic_is)

kable(intern_standard[, c("short_name", "mz", "RT")], format = "pipe")
```

::: {.callout-note}
## Subsetting EIC to speed up RT-window tuning

`extract_is_eics()` accepts an optional `sample_subset` argument to restrict
the chromatogram extraction to a subset of samples. This is useful when
iterating on `rt_window_left` / `rt_window_right` on large datasets, as
extracting EIC for all samples can be slow.

Here, for the METFORMIN-HIIE dataset we demonstrate the feature by plotting EIC **only for QC samples**
(`eic_sample_subset <- sampleData(mse)$sample_type == "QC"`). Any vector
accepted by `mse[, i]` subsetting works — logical, integer index, or names.
Set `sample_subset = NULL` (or omit) to use all samples.
:::

```{r plot-is-eics}
#| label: plot-is-eics

plot_eic_batch(eic_is, fdata, palette = palette_sample_type,
               output_dir = folders$eic_is)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: EIC of Lipid reference set 
Review the EIC plots saved in **`r folders$eic_is`** and adjust `eic_rt_window_left` and `eic_rt_window_right` in the
**Configuration** section if needed.

- The red dashed line indicates the expected RT.
- Each peak should be detected in most or all samples.
- Note compounds showing poor signal or unexpected RT shifts.
- Check the Peak Width and confirm that they are consistent with the expected range.
:::

---

# Data Preprocessing

## Chromatographic Peak Detection

```{r peak-detection-params}
#| label: peak-detection-params

param_pd <- CentWaveParam(
  peakwidth = PEAK_WIDTH,
  ppm = PPM,
  integrate = 2,
  snthresh = SN_THRESHOLD,
  firstBaselineCheck = FALSE
)
param_pd
```

```{r run-peak-detection}
#| label: run-peak-detection

mse <- findChromPeaks(mse, param = param_pd, chunkSize = CORES_NB)

eic_is <- extract_is_eics(mse, intern_standard,
                          sample_subset = eic_sample_subset)
fData(eic_is) <- fdata
```

```{r peaks-per-sample}
#| label: peaks-per-sample
#| tbl-cap: "Number of chromatographic peaks detected per sample"

data.frame(
  sample_name = sampleData(mse)$sample_name,
  peak_count = as.integer(table(chromPeaks(mse)[, "sample"]))
) |> t() |> kable(format = "pipe")
```

::: {.callout-note}
This is the second EIC plot in the file (after peak detection). If you
adjust `rt_window_left` / `rt_window_right` when tuning the first plot
above, make sure the same values are reflected here — both extractions
must use matching windows for the before/after comparison to be valid.
:::

```{r save-peak-detection-plots}
#| label: save-peak-detection-plots

plot_eic_batch(eic_is, fdata, palette = palette_sample_type,
               output_dir = folders$peak_detection)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Peak Detection Results
Review the plots in **`r folders$peak_detection`**.

- Confirm that expected peaks are detected consistently across most samples.
- Check that peaks are not merged incorrectly.
- If the peak detection is not satisfactory, adjust `PEAK_WIDTH`, `PPM`, or `SN_THRESHOLD` in the **configuration** section and rerun the peak-detection step.
:::

## Retention Time Alignment

```{r rt-alignment}
#| label: rt-alignment

QC_samples <- sampleData(mse)$sample_type == "QC"

param_c <- PeakDensityParam(
  sampleGroups = QC_samples,
  minFraction = 0.9,
  binSize = 0.01, ppm = PPM,
  bw = 2
)
mse <- groupChromPeaks(mse, param = param_c)

subset_idx <- which(QC_samples)
param_align <- PeakGroupsParam(
  minFraction = 0.9,
  extraPeaks = 50,
  span = 0.5,
  subsetAdjust = "average",
  subset = subset_idx
)

mse <- adjustRtime(mse, param = param_align)
```

```{r plot-alignment}
#| label: plot-alignment
#| fig-cap: "Retention time alignment results"

col_sample <- get_sample_colors(sampleData(mse)$sample_type, palette_sample_type)
plotAdjustedRtime(mse, col = paste0(col_sample, "80"), peakGroupsPch = 1)
grid()
legend("topright", col = palette_sample_type,
       legend = names(palette_sample_type), lty = 1, bty = "n")
```

```{r apply-alignment}
#| label: apply-alignment

mse_raw <- dropAdjustedRtime(mse)
mse <- applyAdjustedRtime(mse)
```

## Correspondence Analysis

```{r correspondence}
#| label: correspondence

group_mse <- sampleData(mse)$sample_type
group_mse[group_mse == "QC"] <- NA
group_mse <- as.factor(group_mse)

param_cor <- PeakDensityParam(
  sampleGroups = group_mse,
  minFraction = 0.5,
  binSize = 0.01,
  ppm = PPM,
  bw = 2
)

mse <- groupChromPeaks(mse, param = param_cor)

message("Correspondence complete: ", nrow(featureDefinitions(mse)), " features")
```

## Gap Filling

```{r gap-filling}
#| label: gap-filling

na_before <- sum(is.na(featureValues(mse))) / length(featureValues(mse)) * 100

mse <- fillChromPeaks(mse, param = ChromPeakAreaParam(), chunkSize = CORES_NB)

na_after <- sum(is.na(featureValues(mse))) / length(featureValues(mse)) * 100

message("Gap filling: ", round(na_before, 1), "% -> ", round(na_after, 1), "% NA")

if (na_before > 50) {
  warning("High initial missing value percentage (", round(na_before, 1),
          "%). Consider reviewing peak detection parameters or sample quality.")
}
```

::: {.callout-note}
## Gap Filling Guidance
- **< 30% NA before filling**: Good — typical for well-detected features
- **30-50% NA before filling**: Acceptable — may indicate lower abundance compounds
- **> 50% NA before filling**: Review peak detection parameters or sample quality
:::

## Extract Results and Save

```{r extract-results}
#| label: extract-results

res <- quantify(mse, method = "sum", filled = FALSE)
assays(res)$raw_filled <- featureValues(mse, method = "sum", filled = TRUE)
```

```{r validate-preprocessing}
#| label: validate-preprocessing

validate_result(res, assay_name = "raw", min_features = 50)
```

```{r save-preprocessed}
#| label: save-preprocessed

mse_save_path <- file.path(folders$objects,
                           paste0(STUDY_ID, "_preprocessed_mse_", POLARITY))
res_save_path <- file.path(folders$objects,
                           paste0(STUDY_ID, "_preprocessed_res_", POLARITY))

replace_saved_directory(
  mse_save_path,
  function(path) saveMsObject(mse, AlabasterParam(path = path))
)

res@metadata <- list()
replace_saved_directory(res_save_path, function(path) saveObject(res, path))

message("✓ Preprocessed objects saved to: ", folders$objects)
message("  - MsExperiment: ", basename(mse_save_path))
message("  - SummarizedExperiment: ", basename(res_save_path))
message("\nNext step: Run Annotation.qmd for RT adjustment and annotation.")
```

---

## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```

```{r}
toc()
```

---

# Reference

1. Metabonaut tutorial: <https://rformassspectrometry.github.io/Metabonaut/articles/a-end-to-end-untargeted-metabolomics.html#retention-time-alignment>
