---
title: "Pilot Study — Targeted Lipid Annotation"
subtitle: "RT correction, annotation, normalization and quality control (POS + NEG)"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "pilot"
format:
  html:
    output-file: "pilot_Annotation.html"
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

# Overview

::: {.callout-tip}
## Study-specific instance
This is the **combined** annotation for the **pilot study** (BμS device
comparison), running positive and negative ionization modes in parallel
loops. Expects preprocessed objects from `Preprocessing.qmd`.
:::

This is the second step of the targeted lipidomics workflow. It handles
**RT correction, annotation, normalization and QC** for targeted lipidomic
profiling using LC-MS data. It expects preprocessed objects from
`Preprocessing.qmd`.


## Required Input Files

1. **Preprocessed objects** from `Preprocessing.qmd` in
   `positive/objects/` and `negative/objects/`
2. **Lipid database** (`LipidDatabase_R.xlsx`) in this folder
3. **Reference lipid lists** (`positive/pos_lipid_reference_set.xlsx`,
   `negative/neg_lipid_reference_set.xlsx`)

---

# Configuration

::: {.callout-important}
## @SARA:
Maybe we can create a shared configuration document that is loaded from both `Preprocessing.qmd` and `Annotation.qmd` to avoid duplicating configuration chunks
:::


Set the analysis parameters here. `STUDY_ID` and `SAVE_PREFIX` must match
what was used in `Preprocessing.qmd`.

```{r config}
#| label: configuration

# ============================================================================
# PILOT STUDY CONFIGURATION
# ============================================================================

# Project root: relative path to the repository root (R/ helpers live there)
PROJECT_ROOT <- "../.."

# Study identifier: must match the value used in Preprocessing.qmd
STUDY_ID <- "pilot"

# Prefix for saved objects — must match the value used in Preprocessing.qmd
SAVE_PREFIX <- "posNegExp"

# Number of CPU cores for parallel processing
CORES_NB <- 4

# m/z tolerance used during preprocessing (needed for grouping parameters)
PPM <- 10

# Matching parameters
MATCH_PPM             <- 20
MATCH_RT_TOL          <- 20
ISOPEAK_SIM_THRESHOLD <- 0.78

# QC filtering
RSD_THRESHOLD <- 0.3

# Volume correction factors (specific to pilot study)
VOLUME_FACTORS <- list(
    "W" = 2.5,   # W-DBS: smaller volume injected
    "P" = 1.08   # Plasma: slight adjustment
)

# Expected sample types (for validation and CV summary)
EXPECTED_SAMPLE_TYPES <- c("QC", "M-VAMS", "W-DBS", "C-qDBS", "Plasma")

# Color palette for sample types
SAMPLE_TYPE_COLORS <- c(
    "QC"     = "#999999",
    "M-VAMS" = "#377EB8",
    "W-DBS"  = "#E41A1C",
    "C-qDBS" = "#4DAF4A",
    "Plasma" = "#984EA3"
)

# Regression snapshot directory (relative to project root).
# "tests/snapshots_annotation"     → golden baselines (committed, never overwrite)
# "tests/snapshots_annotation_dev" → written during active development / refactor
SNAPSHOT_DIR <- file.path(PROJECT_ROOT, "tests/snapshots_annotation_dev")

# --- Per-mode configuration ---
IS_FILE_POS  <- "positive/pos_lipid_reference_set.xlsx"
IS_FILE_NEG  <- "negative/neg_lipid_reference_set.xlsx"
DB_SHEET_POS <- 4L
DB_SHEET_NEG <- 5L
RT_COL_POS   <- "RT ESI(+) (min)"
RT_COL_NEG   <- "RT ESI(\u2013) (min)"
```

---

# Setup

## Load Packages and Helper Functions

```{r load-helpers}
#| label: load-helpers

source(file.path(PROJECT_ROOT, "R/pos_neg_exp.R"))
source(file.path(PROJECT_ROOT, "R/lipid_helpers.R"))
source(file.path(PROJECT_ROOT, "R/snapshot_helpers.R"))
load_lipid_packages(verbose = TRUE)
```

## Setup Paths and Parallel Processing

```{r setup-paths}
#| label: setup-paths

OBJECTS_DIR_POS     <- "positive/objects"
OBJECTS_DIR_NEG     <- "negative/objects"
REF_LIPID_DIR_POS   <- "positive/figures/ref_lipid_image"
REF_LIPID_DIR_NEG   <- "negative/figures/ref_lipid_image"
ISO_PATTERN_DIR_POS <- "positive/figures/iso_pattern_check"
ISO_PATTERN_DIR_NEG <- "negative/figures/iso_pattern_check"

for (.d in c(
    OBJECTS_DIR_POS, OBJECTS_DIR_NEG,
    REF_LIPID_DIR_POS, REF_LIPID_DIR_NEG,
    ISO_PATTERN_DIR_POS, ISO_PATTERN_DIR_NEG
)) {
    dir.create(.d, recursive = TRUE, showWarnings = FALSE)
}
rm(.d)

if (.Platform$OS.type == "unix") {
    register(MulticoreParam(CORES_NB))
} else {
    register(SnowParam(CORES_NB))
}
```

## Validate Lipid Database

```{r validate-database}
#| label: validate-database

for (.pol in c("pos", "neg")) {
    validate_lipid_database(
        db_path  = "LipidDatabase_R.xlsx",
        sheet    = if (.pol == "pos") DB_SHEET_POS else DB_SHEET_NEG,
        polarity = .pol,
        rt_col   = if (.pol == "pos") RT_COL_POS else RT_COL_NEG
    )
}
rm(.pol)
```

---

# Load Preprocessed Data

Load the preprocessed `XcmsExperiment` and `SummarizedExperiment` objects
generated by `Preprocessing.qmd`.

```{r load-preprocessed}
#| label: load-preprocessed

for (POLARITY in c("pos", "neg")) {
    .obj_dir      <- if (POLARITY == "pos") OBJECTS_DIR_POS else OBJECTS_DIR_NEG
    mse_save_path <- file.path(.obj_dir,
                               paste0(SAVE_PREFIX, "_preprocessed_mse_", POLARITY))
    res_save_path <- file.path(.obj_dir,
                               paste0(SAVE_PREFIX, "_preprocessed_res_", POLARITY))

    if (file.exists(paste0(mse_save_path, ".RData"))) {
        load(paste0(mse_save_path, ".RData"))
        assign(paste0("mse_", POLARITY), mse)
    } else {
        assign(paste0("mse_", POLARITY),
               readMsObject(XcmsExperiment(), AlabasterParam(path = mse_save_path)))
    }
    assign(paste0("res_", POLARITY), readObject(res_save_path))

    message("\u2713 Loaded ", POLARITY, " \u2014 ",
            length(get(paste0("mse_", POLARITY))), " samples, ",
            nrow(get(paste0("res_", POLARITY))), " features")
}
rm(POLARITY, .obj_dir, mse_save_path, res_save_path)

# Load reference lipids and extract EICs for both modes
intern_standard_pos <- prepare_reference_lipids(
    IS_FILE_POS, rt_window_left = 30, rt_window_right = 15
)
intern_standard_neg <- prepare_reference_lipids(
    IS_FILE_NEG, rt_window_left = 30, rt_window_right = 20
)
eic_is_pos <- extract_is_eics(mse_pos, intern_standard_pos)
eic_is_neg <- extract_is_eics(mse_neg, intern_standard_neg)
fdata_pos  <- fData(eic_is_pos)
fdata_neg  <- fData(eic_is_neg)

# Color palette (same sample structure for both modes)
palette_sample_type <- create_color_palette(
    mse_pos,
    column        = "sample_type",
    custom_colors = SAMPLE_TYPE_COLORS
)
```

```{r init-state, include=FALSE}
#| label: init-state

rt_fit_list              <- list()
lipid_db_list            <- list()
mtched_data_list             <- list()
mtched_data_precuration_list <- list()
iso_result_list              <- list()
res_corr_list            <- list()
res_unfilt_list          <- list()
annotated_res_list       <- list()
res_final_list           <- list()
```

---

# Database RT Correction

Fit a polynomial model to correct library retention times based on observed
shifts in reference lipids.

```{r rt-correction-fit}
#| label: rt-correction-fit
#| fig-cap: "RT correction model diagnostic plots"

for (POLARITY in c("pos", "neg")) {
    intern_standard <- if (POLARITY == "pos") intern_standard_pos else intern_standard_neg
    eic_is          <- if (POLARITY == "pos") eic_is_pos else eic_is_neg
    .ref_lipid_dir  <- if (POLARITY == "pos") REF_LIPID_DIR_POS else REF_LIPID_DIR_NEG

    message("\n=== RT Correction: ", toupper(POLARITY), " ===")

    group <- eic_is$sample_type
    group[group == "QC"] <- NA
    group <- as.factor(group)
    param_group <- PeakDensityParam(
        sampleGroups = group,
        minFraction  = 0.5,
        binSize      = 0.01,
        ppm          = PPM,
        bw           = 2
    )

    rt_fit <- fit_rt_correction(
        eic_is, intern_standard, param_group,
        poly_degree = 6,
        output_dir  = .ref_lipid_dir
    )
    print(summary(rt_fit$fit))
    validate_rt_correction(rt_fit, min_r_squared = 0.90, max_residual = 15)
    message(POLARITY, " RT correction R\u00b2 = ",
            round(summary(rt_fit$fit)$r.squared, 4))

    rt_fit_list[[POLARITY]] <- rt_fit
}
rm(POLARITY, intern_standard, eic_is, .ref_lipid_dir,
   group, param_group, rt_fit)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: RT Correction Model

Review the diagnostic plots in **`r REF_LIPID_DIR_POS`** (positive) and
**`r REF_LIPID_DIR_NEG`** (negative).

- Red points (reference lipids) should align closely with the fitted curve
- **R² > 0.95** is expected for a good fit; values below 0.90 indicate
  poor calibration
- **Residuals** should be randomly distributed around zero with no
  systematic patterns
- Typical residual range: ±10 seconds; larger deviations suggest outlier
  reference compounds
:::

## Load and Correct Lipid Database

The `prepare_lipid_database()` function handles all database preprocessing:

1. Filter by rank - Removes rank=0 (unconfirmed) entries
2. RT conversion - Minutes → seconds
3. Unique naming - Appends RT rank for lipids with multiple isomers
4. Adduct processing - Calculates theoretical m/z using `adductFormula()`
5. Deuterium handling - Converts D↔[2H] notation for formula calculation
6. NA removal - Cleans empty rows from Excel import
7. RT correction - Applies polynomial model from reference lipids

```{r load-database}
#| label: load-database

for (POLARITY in c("pos", "neg")) {
    DB_SHEET <- if (POLARITY == "pos") DB_SHEET_POS else DB_SHEET_NEG
    RT_COL   <- if (POLARITY == "pos") RT_COL_POS   else RT_COL_NEG
    rt_fit   <- rt_fit_list[[POLARITY]]

    lipid_database <- prepare_lipid_database(
        db_path  = "LipidDatabase_R.xlsx",
        sheet    = DB_SHEET,
        polarity = POLARITY,
        rt_col   = RT_COL,
        rt_fit   = rt_fit
    )

    lipid_db_list[[POLARITY]] <- lipid_database
}
rm(POLARITY, DB_SHEET, RT_COL, rt_fit, lipid_database)
```

```{r plot-rt-correction}
#| label: plot-rt-correction
#| fig-cap: "Database RT vs. adjusted RT with reference lipids overlay"

for (POLARITY in c("pos", "neg")) {
    rt_fit         <- rt_fit_list[[POLARITY]]
    lipid_database <- lipid_db_list[[POLARITY]]
    intern_standard <- if (POLARITY == "pos") intern_standard_pos
                       else intern_standard_neg

    plot(y = lipid_database$rt_sd, x = lipid_database$rt_adjusted, type = "l",
         xlab = "Adjusted RT", ylab = "Database RT",
         main = paste("RT Correction -", toupper(POLARITY)))
    points(intern_standard$RT ~ rt_fit$exp_rt, col = "red", pch = 16, cex = 1)
    abline(v = min(rt_fit$fit$fitted.values), col = "blue", lwd = 1, lty = 2)
    abline(v = max(rt_fit$fit$fitted.values), col = "blue", lwd = 1, lty = 2)
    grid()
    abline(0, 1, lty = 2, col = "#00000040")
    legend("topleft", legend = c("Database compounds", "Reference lipids"),
           col = c("black", "red"), lty = c(1, NA), pch = c(NA, 16), bty = "n")
}
rm(POLARITY, rt_fit, lipid_database, intern_standard)
```

---

# Annotation / Matching

## Rank 1 Matching

Match features against the lipid database using primary adducts (rank 1).
Matching uses both m/z (ppm tolerance) and RT (time tolerance) constraints.

```{r rank1-matching}
#| label: rank1-matching

annotation <- PosNegAnnotation(
    pos = match_features_to_database(res_pos, lipid_db_list[["pos"]],
                                     ppm = MATCH_PPM, rt_tol = MATCH_RT_TOL),
    neg = match_features_to_database(res_neg, lipid_db_list[["neg"]],
                                     ppm = MATCH_PPM, rt_tol = MATCH_RT_TOL)
)

snap_cols <- c("feature_id", "target_lipid_name_unique", "target_adduct",
               "ppm_error", "score", "score_rt")
for (.pol in c("pos", "neg")) {
    .a <- if (.pol == "pos") posMode(annotation) else negMode(annotation)
    validate_matches(annotMatches(.a), min_matches = 10, max_ppm_error = 15)
    save_snapshot(annotMatches(.a)[, snap_cols], "annot_rank1", .pol)
}
rm(.pol, .a, snap_cols)
```

## Isotope Pattern Validation

```{r isotope-validation}
#| label: isotope-validation

iso_combined    <- calculate_isotope_similarity(
    PosNegXcmsExp(mse_pos, mse_neg),
    annotation,
    isopeak_threshold    = 2,
    similarity_threshold = ISOPEAK_SIM_THRESHOLD
)
annotation      <- iso_combined$annotation
iso_result_list <- iso_combined$iso_results

snap_cols <- c("feature_id", "target_lipid_name_unique",
               "isopeak_count", "isopeak_sim")
for (POLARITY in c("pos", "neg")) {
    .annot <- if (POLARITY == "pos") posMode(annotation) else negMode(annotation)
    message(POLARITY, " matches after isotope filter: ",
            nrow(annotMatches(.annot)))
    save_snapshot(annotMatches(.annot)[, snap_cols], "annot_isotope", POLARITY)
}
rm(iso_combined, POLARITY, .annot, snap_cols)
```

```{r plot-isotope-patterns}
#| label: plot-isotope-patterns

for (POLARITY in c("pos", "neg")) {
    .iso_dir   <- if (POLARITY == "pos") ISO_PATTERN_DIR_POS else ISO_PATTERN_DIR_NEG
    .annot     <- if (POLARITY == "pos") posMode(annotation) else negMode(annotation)
    iso_result <- iso_result_list[[POLARITY]]

    plot_isotope_patterns(
        annotMatches(.annot),
        iso_result$iso_spectra,
        iso_result$theoretical_spectra,
        .iso_dir
    )
}
rm(POLARITY, .iso_dir, .annot, iso_result)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Isotope Pattern Validation

Review the mirror plots in **`r ISO_PATTERN_DIR_POS`** (positive) and
**`r ISO_PATTERN_DIR_NEG`** (negative).

- Experimental (top) should match theoretical (bottom) patterns
- Check compounds with similarity scores near the threshold; adjust
  `ISOPEAK_SIM_THRESHOLD` if needed
- Note any false positive matches where patterns clearly don't align
:::

## Adduct Matching (Rank > 1)

After rank 1 matching, search for secondary adducts that support the
identification. Finding multiple adducts for the same lipid increases
annotation confidence.

```{r adduct-matching}
#| label: adduct-matching

annotation <- match_adducts(annotation, ppm = MATCH_PPM, rt_tol = 5)

snap_cols <- c("feature_id", "target_lipid_name_unique", "target_adduct",
               "adduct_ratio")
save_snapshot(annotMatches(posMode(annotation))[, snap_cols], "annot_adducts", "pos")
save_snapshot(annotMatches(negMode(annotation))[, snap_cols], "annot_adducts", "neg")
rm(snap_cols)
```

## Resolve Ambiguities

```{r resolve-ambiguities}
#| label: resolve-ambiguities

annotation <- resolve_annotation_ambiguities(annotation)
```

## SM1/SM2 Isomer Resolution

```{r sm-resolution}
#| label: sm-resolution

annotation <- resolve_sm_isomers(annotation)

mtched_data_precuration_list <- list(
    pos = annotMatches(posMode(annotation)),
    neg = annotMatches(negMode(annotation))
)

snap_cols <- c("feature_id", "target_lipid_name_unique", "target_adduct",
               "score", "ntch_idx")
save_snapshot(annotMatches(posMode(annotation))[, snap_cols], "annot_resolved", "pos")
save_snapshot(annotMatches(negMode(annotation))[, snap_cols], "annot_resolved", "neg")
rm(snap_cols)
```

## Export Ambiguity Tables for Manual Curation

Export tables for manual review of ambiguous matches:

- Lipid ambiguities: one lipid matches multiple features
- Feature ambiguities: one feature matches multiple lipids (will be merged
  if not resolved)

```{r export-ambiguity}
#| label: export-ambiguity

for (POLARITY in c("pos", "neg")) {
    ambiguities <- export_ambiguity_tables(
        mtched_data_precuration_list[[POLARITY]],
        lipid_file   = paste0("lipid_ambiguity_resolution_",   POLARITY, ".xlsx"),
        feature_file = paste0("feature_ambiguity_resolution_", POLARITY, ".xlsx")
    )
}
rm(POLARITY, ambiguities)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Manual Ambiguity Resolution

Two Excel files per mode have been created in this folder:

- `lipid_ambiguity_resolution_{pos|neg}.xlsx` — one lipid matches multiple
  features (DEFAULT: all `keep_row = TRUE`)
- `feature_ambiguity_resolution_{pos|neg}.xlsx` — one feature matches
  multiple lipids (unresolved entries will be merged)

**Action Required:**

- Open each file
- Set `keep_row = FALSE` for entries you want to remove
- Save the files
- Re-run the **apply-curation** chunk below to apply your decisions
:::

## Apply Manual Curation

```{r apply-curation}
#| label: apply-curation

for (POLARITY in c("pos", "neg")) {
    # Always start from the pre-curation snapshot for re-runnability
    mtched_data <- mtched_data_precuration_list[[POLARITY]]

    lip_file <- paste0("lipid_ambiguity_resolution_",   POLARITY, ".xlsx")
    ft_file  <- paste0("feature_ambiguity_resolution_", POLARITY, ".xlsx")

    if (file.exists(lip_file)) {
        lipid_amb_res          <- read_excel(lip_file)
        lipid_amb_res$keep_row <- as.logical(lipid_amb_res$keep_row)
        mtch_to_be_rm          <- lipid_amb_res$ntch_idx[!lipid_amb_res$keep_row]
        mtched_data            <- mtched_data[
            !mtched_data$ntch_idx %in% mtch_to_be_rm, ]
    }
    if (file.exists(ft_file)) {
        feature_amb_res          <- read_excel(ft_file)
        feature_amb_res$keep_row <- as.logical(feature_amb_res$keep_row)
        mtch_to_be_rm_ft         <- feature_amb_res$ntch_idx[
            !feature_amb_res$keep_row]
        mtched_data              <- mtched_data[
            !mtched_data$ntch_idx %in% mtch_to_be_rm_ft, ]
    }

    # Merge remaining duplicated features
    mtched_data$nb_annotations <- 1
    fids_ambiguous <- mtched_data$feature_id[duplicated(mtched_data$feature_id)]
    for (i in unique(fids_ambiguous)) {
        idxs <- which(mtched_data$feature_id == i)
        if (length(idxs) > 1) {
            mtched_data$target_lipid_name_unique[idxs[1]] <-
                paste(unique(mtched_data$target_lipid_name_unique[idxs]),
                      collapse = "; ")
            mtched_data$target_adduct[idxs[1]] <-
                paste(unique(mtched_data$target_adduct[idxs]),
                      collapse = "; ")
            mtched_data$nb_annotations[idxs[1]] <- length(idxs)
            mtched_data <- mtched_data[-idxs[-1], ]
        }
    }

    message("Final matched lipids (", POLARITY, "): ", nrow(mtched_data))
    mtched_data_list[[POLARITY]] <- mtched_data
}
rm(POLARITY, mtched_data, lip_file, ft_file, fids_ambiguous)
```

## Annotation Summary

```{r annotation-summary}
#| label: annotation-summary

for (POLARITY in c("pos", "neg")) {
    mtched_data  <- mtched_data_list[[POLARITY]]
    lipid_db_r1  <- lipid_db_list[[POLARITY]][lipid_db_list[[POLARITY]]$rank == 1, ]

    n_unique_lipids <- length(unique(mtched_data$target_lipid_name_unique))
    n_db_lipids     <- length(unique(lipid_db_r1$lipid_name_unique))
    coverage        <- n_unique_lipids / n_db_lipids * 100

    message(POLARITY, " final annotation: ", n_unique_lipids, " unique lipids (",
            round(coverage, 1), "% database coverage)")

    counts <- table(mtched_data$`target_LIPID.CATEGORY..ABBREV.`)
    print(plot_lipid_donut(
        counts,
        title = paste("Annotated Lipid Distribution -", toupper(POLARITY))
    ))
}
rm(POLARITY, mtched_data, lipid_db_r1, n_unique_lipids, n_db_lipids, coverage,
   counts)
```

---

# Data Normalization

## Volume Correction

```{r volume-correction}
#| label: volume-correction

res_both <- PosNegSumExp(res_pos, res_neg)
res_both <- apply_volume_correction(res_both, VOLUME_FACTORS,
                                    assay_name     = "raw",
                                    new_assay_name = "raw_corr")
res_both <- apply_volume_correction(res_both, VOLUME_FACTORS,
                                    assay_name     = "raw_filled",
                                    new_assay_name = "corr_filled")
res_corr_list <- list(pos = posMode(res_both), neg = negMode(res_both))
rm(res_both)
```

## Create Annotated Result Object

```{r annotated-res}
#| label: annotated-res

annotated_res_both <- annotate_features(
    PosNegSumExp(res_corr_list[["pos"]], res_corr_list[["neg"]]),
    mtched_data_list
)
annotated_res_list <- list(pos = posMode(annotated_res_both),
                           neg = negMode(annotated_res_both))
rm(annotated_res_both)
```

## Missing Value Imputation

We impute here only to perform PCA plotting.

```{r imputation}
#| label: imputation

for (POLARITY in c("pos", "neg")) {
    annotated_res <- annotated_res_list[[POLARITY]]

    tmp <- apply(assay(annotated_res, "raw_filled"),  MARGIN = 1, na_unidis)
    assays(annotated_res)$raw_filled_imputed <- t(tmp)

    tmp <- apply(assay(annotated_res, "corr_filled"), MARGIN = 1, na_unidis)
    assays(annotated_res)$corr_filled_imputed <- t(tmp)

    annotated_res_list[[POLARITY]] <- annotated_res
}
rm(POLARITY, annotated_res, tmp)
```

## Internal Standard Normalization

```{r is-normalization}
#| label: is-normalization

annotated_res_both <- PosNegSumExp(
    annotated_res_list[["pos"]],
    annotated_res_list[["neg"]]
)
annotated_res_both <- normalize_by_is(
    annotated_res_both,
    input_assay  = "corr_filled",
    output_assay = "ISnorm_filled",
    is_col       = "target_IS_norm",
    lipid_col    = "target_lipid.name"
)

for (POLARITY in c("pos", "neg")) {
    annotated_res <- if (POLARITY == "pos") posMode(annotated_res_both) else
        negMode(annotated_res_both)

    tmp <- apply(assay(annotated_res, "ISnorm_filled"), MARGIN = 1, na_unidis)
    assays(annotated_res)$ISnorm_filled_imputed <- t(tmp)

    save_snapshot(
        list(
            n_features     = nrow(annotated_res),
            feature_ids    = rownames(annotated_res),
            is_assignments = as.character(rowData(annotated_res)$target_IS_norm)
        ),
        "annot_isnorm", POLARITY
    )
    annotated_res_list[[POLARITY]] <- annotated_res
}
rm(POLARITY, annotated_res, tmp, annotated_res_both)
```

---

# Quality Control

## RSD Filtering

```{r rsd-filter}
#| label: rsd-filter

res_unfilt_list <- annotated_res_list

annotated_res_both <- PosNegSumExp(
    annotated_res_list[["pos"]],
    annotated_res_list[["neg"]]
)
annotated_res_both <- filter_by_qc_rsd(annotated_res_both,
                                        threshold = RSD_THRESHOLD)
annotated_res_list[["pos"]] <- posMode(annotated_res_both)
annotated_res_list[["neg"]] <- negMode(annotated_res_both)
rm(annotated_res_both)
```

## Quality Assessment Plots

```{r pca-plot}
#| label: pca-plot
#| fig-cap: "PCA of normalized data colored by sample type"

for (POLARITY in c("pos", "neg")) {
    annotated_res <- annotated_res_list[[POLARITY]]

    vals    <- assay(annotated_res, "ISnorm_filled_imputed") |>
        log2() |> t() |> scale(center = TRUE, scale = TRUE)
    pca_res <- prcomp(vals, scale = FALSE, center = FALSE)
    vals_st <- cbind(vals, sample_type = annotated_res$sample_type)

    print(
        autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
            scale_color_manual(values = palette_sample_type) +
            theme_minimal() +
            ggtitle(paste("PCA - Normalized Data (", toupper(POLARITY), ")"))
    )
}
rm(POLARITY, annotated_res, vals, pca_res, vals_st)
```

```{r rla-plot}
#| label: rla-plot
#| fig-cap: "Relative Log Abundance before and after normalization"

for (POLARITY in c("pos", "neg")) {
    annotated_res  <- annotated_res_list[[POLARITY]]
    col_sample_ann <- get_sample_colors(annotated_res$sample_type,
                                        palette_sample_type)

    par(mfrow = c(2, 1), mar = c(3.5, 4.5, 2.5, 1))
    boxplot(
        rowRla(assay(annotated_res, "raw_filled"),
               group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
        col = paste0(col_sample_ann, "80"), outline = FALSE,
        xaxt = "n", main = paste("Raw Data -", toupper(POLARITY)), boxwex = 1
    )
    grid(nx = NA, ny = NULL)
    abline(h = 0, lty = 3)

    boxplot(
        rowRla(assay(annotated_res, "ISnorm_filled"),
               group = annotated_res$sample_type),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_sample_ann,
        col = paste0(col_sample_ann, "80"), outline = FALSE,
        xaxt = "n",
        main = paste("Normalized Data -", toupper(POLARITY)), boxwex = 1
    )
    axis(side = 1, at = seq_len(ncol(annotated_res)),
         labels = annotated_res$sample_name, las = 2, cex.axis = 0.6)
    grid(nx = NA, ny = NULL)
    abline(h = 0, lty = 3)
}
rm(POLARITY, annotated_res, col_sample_ann)
```

## CV Summary

```{r cv-summary}
#| label: cv-summary
#| tbl-cap: "Coefficient of variation summary"

for (POLARITY in c("pos", "neg")) {
    annotated_res <- annotated_res_list[[POLARITY]]
    index_study   <- annotated_res$sample_type != "QC"
    index_QC      <- annotated_res$sample_type == "QC"

    sample_res <- cbind(
        QC_raw    = rowRsd(assay(annotated_res, "raw_filled")[, index_QC],
                           na.rm = TRUE, mad = TRUE),
        QC_norm   = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_QC],
                           na.rm = TRUE, mad = TRUE),
        Study_raw  = rowRsd(assay(annotated_res, "raw_filled")[, index_study],
                            na.rm = TRUE, mad = TRUE),
        Study_norm = rowRsd(assay(annotated_res, "ISnorm_filled")[, index_study],
                            na.rm = TRUE, mad = TRUE)
    )

    res_df <- data.frame(
        QC_raw    = quantile(sample_res[, "QC_raw"],    na.rm = TRUE),
        QC_norm   = quantile(sample_res[, "QC_norm"],   na.rm = TRUE),
        Study_raw  = quantile(sample_res[, "Study_raw"],  na.rm = TRUE),
        Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
    )

    message(POLARITY, " CV summary:")
    print(kable(res_df, format = "pipe", digits = 3))
}
rm(POLARITY, annotated_res, index_study, index_QC, sample_res, res_df)
```

---

# Final Results

```{r remove-qc}
#| label: remove-qc

for (POLARITY in c("pos", "neg")) {
    annotated_res <- annotated_res_list[[POLARITY]]
    .obj_dir      <- if (POLARITY == "pos") OBJECTS_DIR_POS else OBJECTS_DIR_NEG

    res_qc    <- annotated_res[, annotated_res$sample_type == "QC"]
    res_final <- annotated_res[, annotated_res$sample_type != "QC"]
    rowData(res_final)$qc_cv <- assay(res_qc, "ISnorm_filled") |> rowRsd()

    final_save_path <- file.path(.obj_dir,
                                 paste0(SAVE_PREFIX, "_final_res_", POLARITY))
    if (dir.exists(final_save_path)) unlink(final_save_path, recursive = TRUE)
    saveObject(res_final, final_save_path)
    message("\u2713 Saved final results (", POLARITY, ") to: ", .obj_dir)

    snap_cols <- intersect(
        c("target_lipid_name_unique", "target_LIPID.CATEGORY..ABBREV.",
          "target_IS_norm", "qc_cv"),
        colnames(rowData(res_final))
    )
    save_snapshot(as.data.frame(rowData(res_final)[, snap_cols]),
                  "annot_final", POLARITY)

    res_final_list[[POLARITY]] <- res_final
}
rm(POLARITY, annotated_res, .obj_dir, res_qc, res_final, final_save_path,
   snap_cols)

res_final_pos <- res_final_list$pos
res_final_neg <- res_final_list$neg
message("\n\u2713 Annotation complete for both modes.")
```

## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```

---

# Reference

1. Database source: <https://doi.org/10.1016/j.jlr.2024.100671>
2. Original study (dataset 1): <https://doi.org/10.1016/j.microc.2025.113760>
3. Metabonaut tutorial: <https://rformassspectrometry.github.io/Metabonaut/articles/a-end-to-end-untargeted-metabolomics.html#retention-time-alignment>
4. Validation study (dataset 2): <https://pubs.acs.org/doi/10.1021/acs.jproteome.5c00480>
