---
title: "Pilot Study — POS & NEG Mode Merge"
subtitle: "Combine positive and negative ionization mode results"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "pilot"
format:
  html:
    output-file: "pilot_POS_NEG_merge.html"
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

# Overview

::: {.callout-tip}
## Study-specific instance
This is a configured copy of `generic_workflow/POS_NEG_merge.qmd` for the
**pilot study**. Run after completing `Annotation.qmd`.
:::

This document merges positive and negative mode annotation results into a
single `SummarizedExperiment`. It handles:

1. Adding ionization mode metadata
2. Aligning samples between modes
3. Resolving duplicate lipids (keeps higher abundance when RT is similar)
4. Lipid coverage visualization
5. PCA analysis of the combined dataset

# Configuration

```{r config}

# ============================================================================
# PILOT STUDY CONFIGURATION
# ============================================================================

# Study identifier: must match what was used in the annotation workflow
STUDY_ID <- "pilot"

# Prefix used by the unified Preprocessing.qmd / Annotation.qmd workflows.
SAVE_PREFIX <- "posNegExp"

# Paths to the final result objects (from Annotation.qmd)
RES_POS_PATH <- file.path("positive", "objects", paste0(SAVE_PREFIX, "_final_res_pos"))
RES_NEG_PATH <- file.path("negative", "objects", paste0(SAVE_PREFIX, "_final_res_neg"))

# Output path for combined result
RES_COMBINED_PATH <- file.path("objects", paste0(SAVE_PREFIX, "_res_combined"))

# RT difference threshold for duplicate resolution (seconds)
RT_DIFF_THRESHOLD <- 30
```

# Setup

```{r}
#| label: load-packages
library(alabaster.se)
library(SummarizedExperiment)
library(UpSetR)
library(ggfortify)
library(ggplot2)
library(RColorBrewer)

source("../../R/pos_neg_merge.R")
```

# Load Results

```{r}
#| label: load-data
if (!dir.exists(RES_POS_PATH)) {
    stop("Positive final result not found: ", RES_POS_PATH,
         "\nRun Annotation.qmd first, or update SAVE_PREFIX / RES_POS_PATH.")
}
if (!dir.exists(RES_NEG_PATH)) {
    stop("Negative final result not found: ", RES_NEG_PATH,
         "\nRun Annotation.qmd first, or update SAVE_PREFIX / RES_NEG_PATH.")
}

res_pos <- readObject(RES_POS_PATH)
res_neg <- readObject(RES_NEG_PATH)

cat("Positive mode:", nrow(res_pos), "features x", ncol(res_pos), "samples\n")
cat("Negative mode:", nrow(res_neg), "features x", ncol(res_neg), "samples\n")
```

# Add Ionization Mode

```{r}
#| label: add-mode
rowData(res_pos)$ionization_mode <- "positive"
rowData(res_neg)$ionization_mode <- "negative"
```

# Align Samples

```{r}
#| label: align-samples
aligned <- align_neg_to_pos_samples(res_pos, res_neg)
res_neg <- aligned$res_neg
pos_order <- aligned$pos_order
```

# Resolve Duplicate Lipids

```{r}
#| label: find-duplicates
comparison_df <- find_pos_neg_duplicates(
    res_pos = res_pos,
    res_neg = res_neg,
    rt_diff_threshold = RT_DIFF_THRESHOLD
)

cat("Number of common lipids:", nrow(comparison_df), "\n")

cat("\nRetention time differences summary:\n")
summary(comparison_df$rt_diff)

cat("\nLipids with dissimilar RT (kept in both modes):\n")
comparison_df[!comparison_df$rt_similar,
              c("lipid_name_unique", "pos_rt", "neg_rt", "rt_diff")]
```

```{r}
#| label: remove-duplicates
pos_remove_idx <- comparison_df$pos_idx[
    comparison_df$keep_mode == "negative" & comparison_df$rt_similar
]
neg_remove_idx <- comparison_df$neg_idx[
    comparison_df$keep_mode == "positive" & comparison_df$rt_similar
]

cat("Removing", length(pos_remove_idx), "duplicates from positive mode\n")
cat("Removing", length(neg_remove_idx), "duplicates from negative mode\n")

res_pos_filtered <- if (length(pos_remove_idx)) res_pos[-pos_remove_idx, ] else res_pos
res_neg_filtered <- if (length(neg_remove_idx)) res_neg[-neg_remove_idx, ] else res_neg
```

# Combine

```{r}
#| label: combine-results
colnames(res_pos_filtered) <- strip_mode_suffix(colnames(res_pos_filtered))
colnames(res_neg_filtered) <- strip_mode_suffix(colnames(res_neg_filtered))

if ("file_name" %in% colnames(colData(res_pos_filtered))) {
    colData(res_pos_filtered)$file_name <- strip_mode_suffix(colData(res_pos_filtered)$file_name)
}
if ("file_name" %in% colnames(colData(res_neg_filtered))) {
    colData(res_neg_filtered)$file_name <- strip_mode_suffix(colData(res_neg_filtered)$file_name)
}

res_pos_filtered <- drop_col_data_columns(res_pos_filtered,
                                          c("spectraOrigin", "polarity"))
res_neg_filtered <- drop_col_data_columns(res_neg_filtered,
                                          c("spectraOrigin", "polarity"))

if (!"ionization_mode" %in% colnames(rowData(res_pos_filtered)))
    rowData(res_pos_filtered)$ionization_mode <- "positive"
if (!"ionization_mode" %in% colnames(rowData(res_neg_filtered)))
    rowData(res_neg_filtered)$ionization_mode <- "negative"

res_combined <- rbind(res_pos_filtered, res_neg_filtered)

cat("Combined dataset:", nrow(res_combined), "features x", ncol(res_combined), "samples\n")
cat("Features from positive mode:", sum(rowData(res_combined)$ionization_mode == "positive"), "\n")
cat("Features from negative mode:", sum(rowData(res_combined)$ionization_mode == "negative"), "\n")
```

# Save

```{r}
#| label: save-combined
dir.create(dirname(RES_COMBINED_PATH), recursive = TRUE, showWarnings = FALSE)
if (dir.exists(RES_COMBINED_PATH)) unlink(RES_COMBINED_PATH, recursive = TRUE)
saveObject(res_combined, RES_COMBINED_PATH)
```

# Lipid Coverage

```{r}
#| label: lipid-coverage
#| fig-cap: "Lipid class distribution in the combined POS+NEG dataset"

counts_cat <- table(rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`)
bar_colors_cat <- rep(c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854"),
                      length.out = length(counts_cat))

par(mar = c(1, 1, 2, 1))
pie(counts_cat, col = bar_colors_cat, labels = counts_cat,
    border = "white",
    main = paste0("Combined Lipid Distribution (n = ", sum(counts_cat), ")"),
    cex = 0.9)
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")
legend("topright", legend = names(counts_cat), fill = bar_colors_cat,
       border = NA, bty = "n", title = "Lipid Category", cex = 0.8)
```

```{r}
#| label: lipid-coverage-subclass
#| fig-cap: "Lipid subclass distribution in the combined dataset"

counts_sub <- table(rowData(res_combined)$`target_LIPID.SUBCLASS..ABBREV.`)
sub_colors <- rep(brewer.pal(9, "Set1"), length.out = length(counts_sub))

par(mar = c(1, 1, 2, 1))
pie(counts_sub, col = sub_colors, labels = counts_sub,
    border = "white",
    main = paste0("Subclass Distribution (n = ", sum(counts_sub), ")"),
    cex = 0.9)
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")
legend("topright", legend = names(counts_sub), fill = sub_colors,
       border = NA, bty = "n", title = "Lipid Subclass", cex = 0.5)
```

```{r}
#| label: lipid-coverage-by-mode
#| fig-cap: "Lipid categories by ionization mode"

coverage_df <- as.data.frame(
  table(
    category = rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`,
    mode = rowData(res_combined)$ionization_mode
  )
)

ggplot(coverage_df, aes(x = category, y = Freq, fill = mode)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("positive" = "#e41a1c", "negative" = "#377eb8")) +
  labs(title = "Lipid Categories by Ionization Mode",
       x = "Lipid Category", y = "Count", fill = "Mode") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Duplicate Resolution Summary

```{r}
#| label: duplicate-summary
dup_summary <- comparison_df |>
    dplyr::mutate(
        decision = dplyr::case_when(
            !rt_similar ~ "Kept both (different RT)",
            keep_mode == "positive" ~ "Kept positive (higher abundance)",
            keep_mode == "negative" ~ "Kept negative (higher abundance)"
        )
    )

cat("Duplicate resolution summary:\n")
table(dup_summary$decision)

dup_summary[, c("lipid_name_unique", "decision",
                "pos_mean_abundance", "neg_mean_abundance")]
```

# PCA Analysis

```{r}
#| label: pca-setup
res_combined$sample_type <- ifelse(
    grepl("^QC", colnames(res_combined)), "QC", "sample"
)
```

```{r}
#| label: pca-all
#| fig-cap: "PCA of all samples (POS + NEG combined)"

col_sample_type <- c("QC" = "#1b9e77", "sample" = "#d95f02")

vals <- assay(res_combined, "ISnorm_filled_imputed") |>
    as.matrix() |> t() |> log2() |> scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- as.data.frame(vals)
vals_st$sample_type <- res_combined$sample_type

autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
    scale_color_manual(values = col_sample_type) +
    labs(title = "PCA - All samples (Positive + Negative combined)")
```

```{r}
#| label: pca-by-mode
#| fig-cap: "PCA of features colored by ionization mode"

is_sample <- res_combined$sample_type == "sample"
res_samples <- res_combined[, is_sample]

vals_features <- assay(res_samples, "ISnorm_filled_imputed") |>
    as.matrix() |> log2() |> t() |> scale(center = TRUE, scale = TRUE) |> t()
pca_features <- prcomp(vals_features, scale = FALSE, center = FALSE)

feature_df <- data.frame(
    PC1 = pca_features$x[, 1],
    PC2 = pca_features$x[, 2],
    ionization_mode = rowData(res_samples)$ionization_mode
)

ggplot(feature_df, aes(x = PC1, y = PC2, color = ionization_mode)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("positive" = "#e41a1c", "negative" = "#377eb8")) +
    labs(title = "PCA of features colored by ionization mode") +
    theme_bw()
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
