---
title: "Pilot Study — LC-MS Data Preprocessing"
subtitle: "Peak detection, RT alignment, correspondence and gap filling (POS + NEG)"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "pilot"
format:
  html:
    output-file: "pilot_Preprocessing.html"
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

# Overview

::: {.callout-tip}
## Study-specific instance
This is the **combined** preprocessing for the **pilot study** (BμS device
comparison), running positive and negative ionization modes simultaneously via
`PosNegExp`. See `positive/` and `negative/` for the single-mode counterparts.
:::

This document handles the preprocessing of LC-MS lipidomics data:

1. **Data Import** — Load raw `.mzML` data (local files, SQLite, or MetaboLights)
2. **Dataset Investigation** — BPC visualization and reference lipid EICs
3. **Chromatographic Peak Detection** — CentWave algorithm
4. **Retention Time Alignment** — QC-based peak group alignment
5. **Correspondence** — Feature grouping across samples
6. **Gap Filling** — Fill missing peak values

Output is saved to `positive/objects/` and `negative/objects/`, ready for
`Annotation.qmd`.

## The `PosNegExp` Framework

This workflow uses the `PosNegExp` S4 class system (defined in `R/pos_neg_exp.R`)
to run both ionization modes in a single pipeline with minimal code duplication.

**Classes**

| Class | Wraps | Lifecycle |
|---|---|---|
| `PosNegMsExp` | Two `MsExperiment` | From data load until `findChromPeaks()` |
| `PosNegXcmsExp` | Two `XcmsExperiment` | After `findChromPeaks()` onwards |

**Accessors** — extract individual modes at any point:

```r
posMode(mse)   # returns the positive-mode MsExperiment / XcmsExperiment
negMode(mse)   # returns the negative-mode counterpart
```

**Dispatched xcms functions** — called once, applied to both modes automatically:

```r
findChromPeaks(), filterRt(), groupChromPeaks(),
adjustRtime(), fillChromPeaks()
```

**`perMode()`** — pass mode-specific arguments to a dispatched function:

```r
filterRt(mse, rt = perMode(pos = c(10, 950), neg = c(10, 800)))
```

**Limitation** — `applyAdjustedRtime()` is not yet dispatched; it must be called
explicitly per mode and the result re-wrapped:

```r
mse <- PosNegXcmsExp(applyAdjustedRtime(posMode(mse)),
                     applyAdjustedRtime(negMode(mse)))
```

## Folder Hierarchy

```
pilot_study/
├── LipidDatabase_R.xlsx
├── Preprocessing.qmd           # Step 1: This file (combined POS + NEG)
├── Annotation.qmd              # Step 2: Combined annotation
├── POS_NEG_merge.qmd           # Step 3: Combine POS + NEG results
├── positive/
│   ├── seq_pos_pilot.xlsx              # Sample sequence
│   ├── pos_lipid_reference_set.xlsx    # Reference lipids
│   ├── data/                           # .mzML files (symlink → POS_data/)
│   ├── objects/                        # Saved R objects (auto-generated)
│   └── figures/                        # Output figures (auto-generated)
└── negative/
    ├── seq_neg_pilot.xlsx              # Sample sequence
    ├── neg_lipid_reference_set.xlsx    # Reference lipids
    ├── data/                           # .mzML files (symlink → NEG_data/)
    ├── objects/                        # Saved R objects (auto-generated)
    └── figures/                        # Output figures (auto-generated)
```

---

::: {.callout-important}
## @SARA:
maybe the folder structure can be simplified by having xlsx at the top level and
folders `data`, `objects` and `figures` with subfolders `positive` and `negative`
inside them, e.g. `data/positive` and `data/negative` (instead of `negative/data`,
etc.)
:::

## Required Input Files

Before running this workflow, ensure you have:

1. **Sequence files** `positive/seq_pos_pilot.xlsx` and `negative/seq_neg_pilot.xlsx`
2. **Reference lipid lists** `positive/pos_lipid_reference_set.xlsx` and `negative/neg_lipid_reference_set.xlsx`
3. **Raw data** (`.mzML` files) in `positive/data/` and `negative/data/`
4. **Lipid database** (`LipidDatabase_R.xlsx`) in this folder

---

# Configuration

```{r config}
#| label: configuration

# ============================================================================
# PILOT STUDY CONFIGURATION
# ============================================================================

# Project root: relative path to the repository root (R/ helpers live there)
PROJECT_ROOT <- "../.."

# Study identifier
STUDY_ID <- "pilot"

# Prefix for saved objects — distinguishes PosNegExp PoC outputs from
# per-mode outputs (which use STUDY_ID alone as prefix)
SAVE_PREFIX <- "posNegExp"

# Data source: "local", "sqlite", or "metaboLights"
DATA_SOURCE <- "local"

# MetaboLights study ID (only used if DATA_SOURCE = "metaboLights")
MTBLS_ID <- "MTBLS10722"

# SQLite database paths (only used if DATA_SOURCE = "sqlite")
SQLITE_DB_POS <- NULL
SQLITE_DB_NEG <- NULL

# Number of CPU cores for parallel processing
CORES_NB <- 4

# Expected sample types (for validation)
EXPECTED_SAMPLE_TYPES <- c("QC", "M-VAMS", "W-DBS", "C-qDBS", "Plasma")

# Color palette for sample types
SAMPLE_TYPE_COLORS <- c(
  "QC"     = "#999999",
  "M-VAMS" = "#377EB8",
  "W-DBS"  = "#E41A1C",
  "C-qDBS" = "#4DAF4A",
  "Plasma" = "#984EA3"
)

# --- MetaboLights options (only used if DATA_SOURCE = "metaboLights") ---
MTBLS_ASSAY_POS <- "a_MTBLS10722_LC-MS_positive_reverse-phase_metabolite_profiling.txt"
MTBLS_ASSAY_NEG <- "a_MTBLS10722_LC-MS_negative_reverse-phase_metabolite_profiling.txt"

MTBLS_COLUMN_MAPPING <- list(
  sample_name     = "Sample Name",
  file_name       = "Raw Spectral Data File",
  sample_type     = "Factor Value[Sample type]",
  injection_index = "Comment[injection_index]"
)

MTBLS_FILTER_COLUMN <- "Factor Value[Cohort]"
MTBLS_FILTER_VALUE  <- "Cohort_study"
MTBLS_EXCLUDE_TYPES <- c("Blank")

# Retention time filter range (seconds) — may differ by polarity
RT_FILTER_POS <- c(10, 950)
RT_FILTER_NEG <- c(10, 800)

# Chromatographic peak detection parameters
PEAK_WIDTH   <- c(4, 8)  # min and max peak width in seconds
PPM          <- 10       # m/z tolerance in ppm
SN_THRESHOLD <- 2        # signal-to-noise threshold

# --- Per-mode input files ---
SEQ_FILE_POS <- "positive/seq_pos_pilot.xlsx"
SEQ_FILE_NEG <- "negative/seq_neg_pilot.xlsx"
IS_FILE_POS  <- "positive/pos_lipid_reference_set.xlsx"
IS_FILE_NEG  <- "negative/neg_lipid_reference_set.xlsx"
```

---

# Setup

## Load Packages and Helper Functions

```{r load-helpers}
#| label: load-helpers

source(file.path(PROJECT_ROOT, "R/lipid_helpers.R"))
source(file.path(PROJECT_ROOT, "R/pos_neg_exp.R"))
source(file.path(PROJECT_ROOT, "R/create_sqlite_database.R"))
load_lipid_packages(verbose = TRUE)
```

## Setup Paths and Parallel Processing

```{r setup-paths}
#| label: setup-paths

# Data directories (symlinked to POS_data/ and NEG_data/ at repo root)
DATA_DIR_POS <- "positive/data"
DATA_DIR_NEG <- "negative/data"

# Output directories
OBJECTS_DIR_POS  <- "positive/objects"
OBJECTS_DIR_NEG  <- "negative/objects"
EIC_ISTD_DIR_POS <- "positive/figures/EIC_internal_standards"
EIC_ISTD_DIR_NEG <- "negative/figures/EIC_internal_standards"
PEAK_DET_DIR_POS <- "positive/figures/peak_detection_ref_lipid"
PEAK_DET_DIR_NEG <- "negative/figures/peak_detection_ref_lipid"

# Create output directories
for (.d in c(OBJECTS_DIR_POS, OBJECTS_DIR_NEG,
             EIC_ISTD_DIR_POS, EIC_ISTD_DIR_NEG,
             PEAK_DET_DIR_POS, PEAK_DET_DIR_NEG)) {
  dir.create(.d, recursive = TRUE, showWarnings = FALSE)
}
rm(.d)

# Setup parallel processing
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(CORES_NB))
} else {
  register(SnowParam(CORES_NB))
}
```

---

# Data Import

Load MS data based on the selected `DATA_SOURCE` option.

::: {.panel-tabset}

## Local mzML Files

Load data from local `.mzML` files in `positive/data/` and `negative/data/`.

## SQLite Database

Load from a pre-built SQLite database. Set `SQLITE_DB_POS` / `SQLITE_DB_NEG`
in the Configuration section, or leave `NULL` to auto-generate from local files.

## MetaboLights

Download data directly from MetaboLights (requires internet connection).

:::

## Validate Input Files

```{r validate-inputs}
#| label: validate-inputs

for (.pol in c("pos", "neg")) {
  .seq_file <- if (.pol == "pos") SEQ_FILE_POS else SEQ_FILE_NEG
  .is_file  <- if (.pol == "pos") IS_FILE_POS  else IS_FILE_NEG
  .data_dir <- if (.pol == "pos") DATA_DIR_POS else DATA_DIR_NEG

  validate_metadata(
    .seq_file,
    expected_sample_types = EXPECTED_SAMPLE_TYPES,
    data_dir = if (DATA_SOURCE == "local") .data_dir else NULL
  )
  validate_reference_lipids(.is_file)
}
rm(.pol, .seq_file, .is_file, .data_dir)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Input Validation
Check the output above for any **errors** or **warnings**.
Fix any reported problems in your input files **before** continuing.
:::

```{r load-data}
#| label: load-data

.load_mode <- function(polarity, data_dir, seq_file) {
  seq_data <- read_xlsx(seq_file, col_names = TRUE) |> as.data.frame()
  seq_data$polarity <- polarity

  if (DATA_SOURCE == "local") {
    spectra_files <- file.path(data_dir, seq_data$file_name)
    missing_files <- spectra_files[!file.exists(spectra_files)]
    if (length(missing_files) > 0)
      stop("Missing files (", polarity, "):\n", paste(missing_files, collapse = "\n"))
    readMsExperiment(spectraFiles = spectra_files, sampleData = seq_data)

  } else if (DATA_SOURCE == "sqlite") {
    sqlite_db <- if (polarity == "pos") SQLITE_DB_POS else SQLITE_DB_NEG
    if (is.null(sqlite_db))
      sqlite_db <- file.path(data_dir, paste0(STUDY_ID, "_", polarity, ".sqlite"))
    if (!file.exists(sqlite_db)) {
      message("SQLite database not found. Creating from mzML files...")
      create_sqlite_database(
        study_id  = STUDY_ID,
        polarity  = polarity,
        seq_file  = seq_file,
        data_dir  = data_dir,
        output_db = sqlite_db
      )
    }
    load_from_sqlite(sqlite_db, seq_data)

  } else if (DATA_SOURCE == "metaboLights") {
    assay_name <- if (polarity == "pos") MTBLS_ASSAY_POS else MTBLS_ASSAY_NEG
    load_from_metaboLights(
      mtbls_id             = MTBLS_ID,
      assay_name           = assay_name,
      column_mapping       = MTBLS_COLUMN_MAPPING,
      filter_column        = MTBLS_FILTER_COLUMN,
      filter_value         = MTBLS_FILTER_VALUE,
      exclude_sample_types = MTBLS_EXCLUDE_TYPES,
      polarity             = polarity
    )
  } else {
    stop("Invalid DATA_SOURCE: '", DATA_SOURCE,
         "'. Use 'local', 'sqlite', or 'metaboLights'.")
  }
}

message("Loading positive mode data...")
mse_pos <- .load_mode("pos", DATA_DIR_POS, SEQ_FILE_POS)
message("Loading negative mode data...")
mse_neg <- .load_mode("neg", DATA_DIR_NEG, SEQ_FILE_NEG)
rm(.load_mode)
```

::: {.callout-caution}
## `PosNegMsExp`: wrapping both modes
`PosNegMsExp` holds both `MsExperiment` objects in a single container so that
subsequent xcms calls automatically process positive and negative mode together.
`posMode()` / `negMode()` extract each mode at any time.
:::

```{r wrap-pos-neg}
#| label: wrap-pos-neg

mse <- PosNegMsExp(mse_pos, mse_neg)
message("✓ Loaded ", length(posMode(mse)), " samples (pos) and ",
        length(negMode(mse)), " samples (neg)")
sampleData(posMode(mse))
```

## Setup Color Palettes

```{r setup-colors}
#| label: setup-colors

# Both modes share the same sample structure — derive palette from pos mode
palette_sample_type <- create_color_palette(
  posMode(mse),
  column        = "sample_type",
  custom_colors = SAMPLE_TYPE_COLORS
)

message("Color palette for sample_type:")
print(palette_sample_type)
```

---

# Dataset Investigation

## Chromatographic Data Visualization

```{r bpc-raw}
#| label: bpc-raw
#| fig-cap: "Base Peak Chromatogram — positive (left) and negative (right)"

par(mfrow = c(1, 2))
plot_bpc(posMode(mse), palette = palette_sample_type, main = "BPC - Positive Mode")
plot_bpc(negMode(mse), palette = palette_sample_type, main = "BPC - Negative Mode")
```

::: {.callout-important}
## 📋 USER CHECKPOINT: RT Filter Range
Examine the BPCs above and adjust `RT_FILTER_POS` / `RT_FILTER_NEG` in the
Configuration section if needed.
Current: POS **`r RT_FILTER_POS[1]`–`r RT_FILTER_POS[2]`** s,
NEG **`r RT_FILTER_NEG[1]`–`r RT_FILTER_NEG[2]`** s.
:::

::: {.callout-caution}
## `perMode()`: mode-specific arguments in a single call
`perMode(pos = ..., neg = ...)` is an S3 container that tells the dispatched
method to apply different values to each mode. Here it passes distinct RT ranges
to a single `filterRt()` call that operates on both modes simultaneously.
:::

```{r filter-rt}
#| label: filter-rt

mse <- filterRt(mse, rt = perMode(pos = RT_FILTER_POS, neg = RT_FILTER_NEG))
```

```{r bpc-filtered}
#| label: bpc-filtered
#| fig-cap: "BPC after retention time filtering"

par(mfrow = c(1, 2))
plot_bpc(posMode(mse), palette = palette_sample_type, main = "BPC - Positive (filtered)")
plot_bpc(negMode(mse), palette = palette_sample_type, main = "BPC - Negative (filtered)")
```

## Reference Lipid EICs

```{r load-is}
#| label: load-is
#| tbl-cap: "Reference lipids for RT correction"

intern_standard_pos <- prepare_reference_lipids(
  IS_FILE_POS, rt_window_left = 30, rt_window_right = 15
)
intern_standard_neg <- prepare_reference_lipids(
  IS_FILE_NEG, rt_window_left = 30, rt_window_right = 20
)

eic_is_pos <- extract_is_eics(posMode(mse), intern_standard_pos)
eic_is_neg <- extract_is_eics(negMode(mse), intern_standard_neg)
fdata_pos  <- fData(eic_is_pos)
fdata_neg  <- fData(eic_is_neg)

message("Positive mode reference lipids:")
kable(intern_standard_pos[, c("short_name", "mz", "RT")], format = "pipe")
message("Negative mode reference lipids:")
kable(intern_standard_neg[, c("short_name", "mz", "RT")], format = "pipe")
```

```{r plot-is-eics}
#| label: plot-is-eics

plot_eic_batch(eic_is_pos, fdata_pos, palette = palette_sample_type,
               output_dir = EIC_ISTD_DIR_POS)
plot_eic_batch(eic_is_neg, fdata_neg, palette = palette_sample_type,
               output_dir = EIC_ISTD_DIR_NEG)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Reference Lipid EICs
Check the EIC plots saved in **`r EIC_ISTD_DIR_POS`** (positive) and
**`r EIC_ISTD_DIR_NEG`** (negative).

- The red dashed line indicates the expected RT
- Each peak should be detected in most/all samples
- Note any compounds with poor signal or unexpected RT shifts
- Note peak width to inform `PEAK_WIDTH`
:::

---

# Data Preprocessing

## Chromatographic Peak Detection

```{r peak-detection-params}
#| label: peak-detection-params

param_pd <- CentWaveParam(
  peakwidth          = PEAK_WIDTH,
  ppm                = PPM,
  integrate          = 2,
  snthresh           = SN_THRESHOLD,
  firstBaselineCheck = FALSE
)
param_pd
```

::: {.callout-caution}
## Dispatched xcms functions
`findChromPeaks()`, `groupChromPeaks()`, `adjustRtime()`, and `fillChromPeaks()`
are all dispatched on `PosNegXcmsExp`. A single call processes both modes in
sequence; the result is automatically re-wrapped in a `PosNegXcmsExp`.
:::

```{r run-peak-detection}
#| label: run-peak-detection

mse <- findChromPeaks(mse, param = param_pd, chunkSize = CORES_NB)

# Update EICs with detected peaks for each mode
eic_is_pos <- extract_is_eics(posMode(mse), intern_standard_pos)
fData(eic_is_pos) <- fdata_pos
eic_is_neg <- extract_is_eics(negMode(mse), intern_standard_neg)
fData(eic_is_neg) <- fdata_neg
```

```{r peaks-per-sample}
#| label: peaks-per-sample
#| results: asis

for (.pol in c("pos", "neg")) {
  .mse <- if (.pol == "pos") posMode(mse) else negMode(mse)
  df <- data.frame(
    sample_name = sampleData(.mse)$sample_name,
    peak_count  = as.integer(table(chromPeaks(.mse)[, "sample"]))
  )
  cat("\n\n**", toupper(.pol), "mode — peaks per sample**\n\n", sep = "")
  cat(knitr::kable(df, format = "pipe"), sep = "\n")
  cat("\n")
}
rm(.pol, .mse, df)
```

```{r save-peak-detection-plots}
#| label: save-peak-detection-plots

plot_eic_batch(eic_is_pos, fdata_pos, palette = palette_sample_type,
               output_dir = PEAK_DET_DIR_POS)
plot_eic_batch(eic_is_neg, fdata_neg, palette = palette_sample_type,
               output_dir = PEAK_DET_DIR_NEG)
```

::: {.callout-important}
## 📋 USER CHECKPOINT: Peak Detection Results
Review the plots in **`r PEAK_DET_DIR_POS`** (positive) and
**`r PEAK_DET_DIR_NEG`** (negative).

- Each peak should be detected in most samples
- Check that peaks are not merged incorrectly
- Adjust `PEAK_WIDTH`, `PPM`, or `SN_THRESHOLD` if needed
:::

## Retention Time Alignment

```{r rt-alignment}
#| label: rt-alignment

# Both modes share the same sample structure — use pos to derive groups
QC_samples <- sampleData(posMode(mse))$sample_type == "QC"

param_c <- PeakDensityParam(
  sampleGroups = QC_samples,
  minFraction  = 0.9,
  binSize      = 0.01,
  ppm          = PPM,
  bw           = 2
)
mse <- groupChromPeaks(mse, param = param_c)

subset_idx  <- which(QC_samples)
param_align <- PeakGroupsParam(
  minFraction  = 0.9,
  extraPeaks   = 50,
  span         = 0.5,
  subsetAdjust = "average",
  subset       = subset_idx
)

mse <- adjustRtime(mse, param = param_align)
```

```{r plot-alignment}
#| label: plot-alignment
#| fig-cap: "Retention time alignment (positive left, negative right)"

par(mfrow = c(1, 2))
for (.pol in c("pos", "neg")) {
  .mse       <- if (.pol == "pos") posMode(mse) else negMode(mse)
  col_sample <- get_sample_colors(sampleData(.mse)$sample_type, palette_sample_type)
  plotAdjustedRtime(.mse, col = paste0(col_sample, "80"), peakGroupsPch = 1,
                    main = paste("RT alignment -", .pol))
  grid()
  legend("topright", col = palette_sample_type,
         legend = names(palette_sample_type), lty = 1, bty = "n")
}
rm(.pol, .mse, col_sample)
```

::: {.callout-caution}
## `applyAdjustedRtime()`: explicit per-mode call
`applyAdjustedRtime()` is not yet an S4 generic dispatched on `PosNegXcmsExp`.
It must be called on each mode separately and the results re-wrapped manually.
:::

```{r apply-alignment}
#| label: apply-alignment

mse_raw <- dropAdjustedRtime(mse)
mse <- PosNegXcmsExp(
  applyAdjustedRtime(posMode(mse)),
  applyAdjustedRtime(negMode(mse))
)
```

## Correspondence Analysis

```{r correspondence}
#| label: correspondence

group_mse <- sampleData(posMode(mse))$sample_type
group_mse[group_mse == "QC"] <- NA
group_mse <- as.factor(group_mse)

param_cor <- PeakDensityParam(
  sampleGroups = group_mse,
  minFraction  = 0.5,
  binSize      = 0.01,
  ppm          = PPM,
  bw           = 2
)

mse <- groupChromPeaks(mse, param = param_cor)

for (.pol in c("pos", "neg")) {
  .mse <- if (.pol == "pos") posMode(mse) else negMode(mse)
  message(.pol, " correspondence: ", nrow(featureDefinitions(.mse)), " features")
}
rm(.pol, .mse)
```

## Gap Filling

```{r gap-filling}
#| label: gap-filling

na_before <- list(
  pos = sum(is.na(featureValues(posMode(mse)))) /
          length(featureValues(posMode(mse))) * 100,
  neg = sum(is.na(featureValues(negMode(mse)))) /
          length(featureValues(negMode(mse))) * 100
)

mse <- fillChromPeaks(mse, param = ChromPeakAreaParam(), chunkSize = CORES_NB)

for (.pol in c("pos", "neg")) {
  .mse     <- if (.pol == "pos") posMode(mse) else negMode(mse)
  na_after <- sum(is.na(featureValues(.mse))) / length(featureValues(.mse)) * 100
  message(.pol, " gap filling: ", round(na_before[[.pol]], 1),
          "% -> ", round(na_after, 1), "% NA")
  if (na_before[[.pol]] > 50)
    warning("High initial NA % in ", .pol, " mode (",
            round(na_before[[.pol]], 1), "%).")
}
rm(.pol, .mse, na_after)
```

::: {.callout-note}
## Gap Filling Guidance
- **< 30% NA before filling**: Good — typical for well-detected features
- **30-50% NA before filling**: Acceptable — may indicate lower abundance compounds
- **> 50% NA before filling**: Review peak detection parameters or sample quality
:::

## Extract Results and Save

```{r extract-results}
#| label: extract-results

mse_pos <- posMode(mse)
mse_neg <- negMode(mse)

res_pos <- quantify(mse_pos, method = "sum", filled = FALSE)
assays(res_pos)$raw_filled <- featureValues(mse_pos, method = "sum", filled = TRUE)

res_neg <- quantify(mse_neg, method = "sum", filled = FALSE)
assays(res_neg)$raw_filled <- featureValues(mse_neg, method = "sum", filled = TRUE)
```

```{r validate-preprocessing}
#| label: validate-preprocessing

validate_result(res_pos, assay_name = "raw", min_features = 50)
validate_result(res_neg, assay_name = "raw", min_features = 50)
```

```{r save-preprocessed}
#| label: save-preprocessed

for (.pol in c("pos", "neg")) {
  .mse      <- if (.pol == "pos") mse_pos else mse_neg
  .res      <- if (.pol == "pos") res_pos else res_neg
  .obj_dir  <- if (.pol == "pos") OBJECTS_DIR_POS else OBJECTS_DIR_NEG

  mse_save_path <- file.path(.obj_dir,
                             paste0(SAVE_PREFIX, "_preprocessed_mse_", .pol))
  res_save_path <- file.path(.obj_dir,
                             paste0(SAVE_PREFIX, "_preprocessed_res_", .pol))

  if (DATA_SOURCE == "sqlite") {
    spectra(.mse) <- setBackend(spectra(.mse), MsBackendMemory())
    save(.mse, file = paste0(mse_save_path, ".RData"))
  } else {
    if (dir.exists(mse_save_path)) unlink(mse_save_path, recursive = TRUE)
    saveMsObject(.mse, AlabasterParam(path = mse_save_path))
  }

  .res@metadata <- list()
  if (dir.exists(res_save_path)) unlink(res_save_path, recursive = TRUE)
  saveObject(.res, res_save_path)

  message("✓ Saved ", .pol, " objects to: ", .obj_dir)
}
rm(.pol, .mse, .res, .obj_dir, mse_save_path, res_save_path)

message("\nNext step: Run Annotation.qmd for RT correction and annotation.")
```

---

## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```

---

# Reference

1. Metabonaut tutorial: <https://rformassspectrometry.github.io/Metabonaut/articles/a-end-to-end-untargeted-metabolomics.html#retention-time-alignment>
