---
title: >-
  METFORMIN-HIIE dataset — Stage 3: Cross-Ionization Mode Data Integration
subtitle: >-
  Positive and negative ionization mode integration for coverage reporting
  and downstream analysis
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "METFORMIN-HIIE"
format:
  html:
    output-file: "METFORMIN-HIIE_Cross_mode_integration.html"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

```{r}
library("tictoc")
tic("Quarto compilation")
```

# Overview

::: {.callout-tip}
## Study-specific configuration
This is a configured copy of `generic_workflow/POS_NEG_merge.qmd` for the
**METFORMIN-HIIE study**. Run after completing `Annotation_pos.qmd` and
`Annotation_neg.qmd`.
:::

This document merges positive and negative ionization mode annotation results
into a single `SummarizedExperiment`. It handles:

1. Adding ionization mode metadata
2. Aligning samples between ionization modes
3. Resolving duplicate lipids across modes by retaining the feature with
   higher mean normalized abundance
4. Visualizing lipid distribution
5. Performing PCA on of the combined dataset

# Configuration

```{r config}
#| label: configuration

# ============================================================================
# METFORMIN-HIIE STUDY CONFIGURATION
# ============================================================================

PROJECT_ROOT <- "../.."

# Study identifier: must match what was used in the annotation workflow
STUDY_ID <- "METFORMIN-HIIE"

# Paths to the final result objects (from Annotation.qmd)
RES_POS_PATH <- file.path("positive", "objects",
                          paste0(STUDY_ID, "_final_res_pos"))
RES_NEG_PATH <- file.path("negative", "objects",
                          paste0(STUDY_ID, "_final_res_neg"))

# Output path for combined result
RES_COMBINED_PATH <- file.path("objects", paste0(STUDY_ID, "_res_combined"))

# Whether IS participate in downstream plots, metrics, and PCA.
# When FALSE, IS are dropped from res_combined *after* the Save section,
# so the saved binary still has them (with an origin column) for reuse.
INCLUDE_IS <- FALSE
```

# Setup

```{r}
#| label: load-packages
library(alabaster.se)
library(SummarizedExperiment)
library(UpSetR)
library(ggfortify)
library(ggplot2)
library(RColorBrewer)
library(writexl)
source(file.path(PROJECT_ROOT, "R/lipid_helpers.R"))
```

# Load Results

```{r}
#| label: load-data
res_pos <- readObject(RES_POS_PATH)
res_neg <- readObject(RES_NEG_PATH)

cat("Positive ionization mode:", nrow(res_pos), "features x", ncol(res_pos),
    "samples\n")
cat("Negative ionization mode:", nrow(res_neg), "features x", ncol(res_neg),
    "samples\n")
```

# Add Ionization Mode

```{r}
#| label: add-mode
rowData(res_pos)$ionization_mode <- "positive"
rowData(res_neg)$ionization_mode <- "negative"
```

# Align Samples

```{r}
#| label: align-samples
# Match samples on the polarity-independent `sample_name` column from colData,
# so this works regardless of mzML filename conventions.
pos_samples <- colData(res_pos)$sample_name
neg_samples <- colData(res_neg)$sample_name

stopifnot(
    !anyDuplicated(pos_samples),
    !anyDuplicated(neg_samples),
    setequal(pos_samples, neg_samples)
)

res_neg <- res_neg[, match(pos_samples, neg_samples)]

stopifnot(all(colData(res_pos)$sample_name == colData(res_neg)$sample_name))
```

Within-mode duplicates are resolved in the lipid-ambiguity step of
`Annotation_pos.qmd` and `Annotation_neg.qmd`.

# Prefer Resolved sn-1/sn-2 Annotations Across Modes

An unresolved lysoglycerophospholipid sn-1/sn-2 pair (e.g.
`LPC(0:0/18:1)_1; LPC(18:1/0:0)_1`) is removed when either isomer is
annotated in the other mode. Other merged annotations are kept. Removals are
computed in both directions before changing either dataset.

```{r}
#| label: prefer-resolved-sn-position
find_unresolved_rows <- function(ambiguous, other_mode) {
    ambiguous_name <- as.character(rowData(ambiguous)$target_lipid_name_unique)
    other_name <- as.character(rowData(other_mode)$target_lipid_name_unique)
    which(vapply(ambiguous_name, function(nm)
        any(other_name %in% sn_pair_options(nm)), logical(1),
        USE.NAMES = FALSE))
}

pos_unresolved <- find_unresolved_rows(res_pos, res_neg)
neg_unresolved <- find_unresolved_rows(res_neg, res_pos)

cat("Removing", length(pos_unresolved), "unresolved positive and",
    length(neg_unresolved), "unresolved negative annotations\n")
res_pos <- res_pos[
    base::setdiff(seq_len(nrow(res_pos)), pos_unresolved), , drop = FALSE
]
res_neg <- res_neg[
    base::setdiff(seq_len(nrow(res_neg)), neg_unresolved), , drop = FALSE
]
```

# Resolve Duplicate Annotations

For exact lipid names shared between modes, repeatedly compare a
positive/negative pair regardless of RT. Pairs are ordered by RT difference,
but RT does not determine eligibility. Remove the feature with lower mean
normalized abundance and repeat until no cross-mode pair remains.

```{r}
#| label: find-duplicates
pos_name <- as.character(rowData(res_pos)$target_lipid_name_unique)
neg_name <- as.character(rowData(res_neg)$target_lipid_name_unique)
pos_rt <- as.numeric(rowData(res_pos)$rtmed)
neg_rt <- as.numeric(rowData(res_neg)$rtmed)
pos_mean <- rowMeans(as.matrix(assay(res_pos, "ISnorm_filled_imputed")),
                     na.rm = TRUE)
neg_mean <- rowMeans(as.matrix(assay(res_neg, "ISnorm_filled_imputed")),
                     na.rm = TRUE)
pos_mean[!is.finite(pos_mean)] <- -Inf
neg_mean[!is.finite(neg_mean)] <- -Inf

common_lipids <- intersect(
    unique(pos_name[!is.na(pos_name) & nzchar(pos_name)]),
    unique(neg_name[!is.na(neg_name) & nzchar(neg_name)])
)

cat("Number of common lipids:", length(common_lipids), "\n")

comparison_rows <- list()
pos_remove_idx <- integer()
neg_remove_idx <- integer()

for (lipid in common_lipids) {
    pos_idx <- which(pos_name == lipid)
    neg_idx <- which(neg_name == lipid)

    while (length(pos_idx) && length(neg_idx)) {
        pairs <- expand.grid(pos_idx = pos_idx, neg_idx = neg_idx,
                             KEEP.OUT.ATTRS = FALSE)
        pairs$rt_diff <- abs(pos_rt[pairs$pos_idx] - neg_rt[pairs$neg_idx])

        pair <- pairs[order(pairs$rt_diff, pairs$pos_idx,
                            pairs$neg_idx, na.last = TRUE)[1], , drop = FALSE]
        p <- pair$pos_idx
        n <- pair$neg_idx
        keep_mode <- if (pos_mean[p] >= neg_mean[n]) "positive" else "negative"

        comparison_rows[[length(comparison_rows) + 1L]] <- data.frame(
            lipid_name_unique = lipid,
            pos_idx = p,
            neg_idx = n,
            pos_feature_id = rownames(res_pos)[p],
            neg_feature_id = rownames(res_neg)[n],
            pos_mean_abundance = pos_mean[p],
            neg_mean_abundance = neg_mean[n],
            pos_rt = pos_rt[p],
            neg_rt = neg_rt[n],
            rt_diff = pair$rt_diff,
            keep_mode = keep_mode
        )

        if (keep_mode == "positive") {
            neg_remove_idx <- c(neg_remove_idx, n)
            neg_idx <- neg_idx[neg_idx != n]
        } else {
            pos_remove_idx <- c(pos_remove_idx, p)
            pos_idx <- pos_idx[pos_idx != p]
        }
    }
}

comparison_df <- if (length(comparison_rows)) {
    do.call(rbind, comparison_rows)
} else {
    data.frame(lipid_name_unique = character(), pos_idx = integer(),
               neg_idx = integer(), pos_feature_id = character(),
               neg_feature_id = character(), pos_mean_abundance = numeric(),
               neg_mean_abundance = numeric(), pos_rt = numeric(),
               neg_rt = numeric(), rt_diff = numeric(), keep_mode = character())
}

cat("Cross-mode comparisons resolved:", nrow(comparison_df), "\n")
cat("Retention time differences of resolved comparisons:\n")
print(summary(comparison_df$rt_diff))
```

```{r}
#| label: remove-duplicates
cat("Removing", length(pos_remove_idx),
    "duplicates from positive ionization mode\n")
cat("Removing", length(neg_remove_idx),
    "duplicates from negative ionization mode\n")

res_pos_filtered <- res_pos[
    base::setdiff(seq_len(nrow(res_pos)), pos_remove_idx), , drop = FALSE
]
res_neg_filtered <- res_neg[
    base::setdiff(seq_len(nrow(res_neg)), neg_remove_idx), , drop = FALSE
]

remaining_common <- intersect(
    as.character(rowData(res_pos_filtered)$target_lipid_name_unique),
    as.character(rowData(res_neg_filtered)$target_lipid_name_unique)
)
cat("Shared lipid names retained in both modes:",
    length(remaining_common), "\n")
```

# Combine Lipid Annotations

```{r}
#| label: combine-results
colnames(res_pos_filtered) <- colData(res_pos_filtered)$sample_name
colnames(res_neg_filtered) <- colData(res_neg_filtered)$sample_name

# Drop polarity-specific colData; `file_name` differs between modes and
# can't survive the rbind below.
colData(res_pos_filtered)$file_name <- NULL
colData(res_neg_filtered)$file_name <- NULL
colData(res_pos_filtered)$spectraOrigin <- NULL
colData(res_neg_filtered)$spectraOrigin <- NULL
colData(res_pos_filtered)$polarity <- NULL
colData(res_neg_filtered)$polarity <- NULL

if (!"ionization_mode" %in% colnames(rowData(res_pos_filtered)))
    rowData(res_pos_filtered)$ionization_mode <- "positive"
if (!"ionization_mode" %in% colnames(rowData(res_neg_filtered)))
    rowData(res_neg_filtered)$ionization_mode <- "negative"

res_combined <- rbind(res_pos_filtered, res_neg_filtered)

cat("Combined dataset:", nrow(res_combined), "features x",
    ncol(res_combined), "samples\n")
cat("Features from positive ionization mode:",
    sum(rowData(res_combined)$ionization_mode == "positive"), "\n")
cat("Features from negative ionization mode:",
    sum(rowData(res_combined)$ionization_mode == "negative"), "\n")
```

```{r}
#| label: mark-origin

# Use IS_norm to differentiate between Internal Standards and Endogenous lipids
is_IS <- !is.na(rowData(res_combined)$target_IS_norm) &
    rowData(res_combined)$target_lipid.name ==
        rowData(res_combined)$target_IS_norm
rowData(res_combined)$origin <- ifelse(is_IS, "IS", "endogenous")

cat("Origin marked:",
    sum(rowData(res_combined)$origin == "IS"), "IS,",
    sum(rowData(res_combined)$origin == "endogenous"), "endogenous\n")
```

# Save Results

```{r}
#| label: save-combined
dir.create(dirname(RES_COMBINED_PATH), recursive = TRUE, showWarnings = FALSE)
replace_saved_directory(RES_COMBINED_PATH, function(path) saveObject(res_combined, path))
```

```{r}
#| label: filter-IS-for-analysis

# Keep an unfiltered snapshot for the xlsx export at the end of this file
res_combined_all <- res_combined
if (!INCLUDE_IS) {
    n_before <- nrow(res_combined)
    res_combined <- res_combined[rowData(res_combined)$origin == "endogenous", ]
    cat("INCLUDE_IS = FALSE: dropped",
        n_before - nrow(res_combined),
        "IS from downstream analysis (",
        nrow(res_combined), "features remain).\n")
} else {
    cat("INCLUDE_IS = TRUE: keeping",
        sum(rowData(res_combined)$origin == "IS"),
        "IS in downstream analysis.\n")
}
```

# Lipid Distribution

```{r}
#| label: lipid-coverage
#| fig-cap: "Annotated lipid category distribution in the combined dataset"

counts_cat <- table(rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`)
bar_colors_cat <- rep(c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854"),
                      length.out = length(counts_cat))

par(mar = c(1, 1, 2, 1))
pie(counts_cat, col = bar_colors_cat, labels = counts_cat,
    border = "white",
    main = paste0("Lipid Category Distribution (n = ", sum(counts_cat), ")"),
    cex = 0.9)
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")
legend("topright", legend = names(counts_cat), fill = bar_colors_cat,
       border = NA, bty = "n", title = "Lipid Category", cex = 0.8)
```

```{r}
#| label: lipid-coverage-subclass-by-category
#| fig-cap: >-
#|   Annotated lipid subclass distribution, grouped and colored by category in
#|   the combined dataset

counts_df <- prepare_subclass_counts(res_combined)
cat_palette <- c(FA = "#66c2a5", GL = "#fc8d62", GP = "#8da0cb",
                 SP = "#e78ac3", ST = "#a6d854")

ggplot(counts_df, aes(x = subclass, y = Freq, fill = category)) +
    geom_col() +
    facet_grid(~ category, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = cat_palette) +
    labs(title = "Lipid Subclass Distribution",
         x = NULL, y = "Count", fill = "Category") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.placement = "outside",
          strip.background = element_rect(fill = "grey95"),
          panel.spacing = unit(0.2, "lines"))
```

```{r}
#| label: lipid-coverage-by-mode
#| fig-cap: "Lipid category distribution by ionization mode"

coverage_df <- as.data.frame(
  table(
    category = rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`,
    mode = rowData(res_combined)$ionization_mode
  )
)

ggplot(coverage_df, aes(x = category, y = Freq, fill = mode)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("positive" = "#e41a1c",
                               "negative" = "#377eb8")) +
  labs(title = "Lipid Category Distribution by Ionization Mode",
       x = "Lipid Category", y = "Count", fill = "Mode") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Duplicate Resolution Summary

```{r}
#| label: duplicate-summary
dup_summary <- comparison_df
dup_summary$decision <- ifelse(
    dup_summary$keep_mode == "positive",
    "Removed negative (lower mean abundance)",
    "Removed positive (lower mean abundance)"
)

cat("Sequential cross-mode duplicate resolution summary:\n")
table(dup_summary$decision)

dup_summary[, c("lipid_name_unique", "pos_feature_id", "neg_feature_id",
                "rt_diff", "decision", "pos_mean_abundance",
                "neg_mean_abundance")]
```

# PCA Analysis

```{r}
#| label: pca-setup
res_combined$sample_type <- ifelse(
    grepl("^QC", colnames(res_combined)), "QC", "sample"
)
```

```{r}
#| label: pca-all
#| fig-cap: "PCA of all samples (POS + NEG ionization mode combined)"

col_sample_type <- c("QC" = "#1b9e77", "sample" = "#d95f02")

vals <- assay(res_combined, "ISnorm_filled_imputed") |>
    as.matrix() |> t() |> log2() |> scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- as.data.frame(vals)
vals_st$sample_type <- res_combined$sample_type

autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
    scale_color_manual(values = col_sample_type) +
    labs(
        title = "PCA - All samples (from positive and negative ionization mode)"
    )
```

```{r}
#| label: pca-by-mode
#| fig-cap: "PCA of features colored by ionization mode"

is_sample <- res_combined$sample_type == "sample"
res_samples <- res_combined[, is_sample]

vals_features <- assay(res_samples, "ISnorm_filled_imputed") |>
    as.matrix() |> log2() |> t() |> scale(center = TRUE, scale = TRUE) |> t()
pca_features <- prcomp(vals_features, scale = FALSE, center = FALSE)

feature_df <- data.frame(
    PC1 = pca_features$x[, 1],
    PC2 = pca_features$x[, 2],
    ionization_mode = rowData(res_samples)$ionization_mode
)

ggplot(feature_df, aes(x = PC1, y = PC2, color = ionization_mode)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("positive" = "#e41a1c",
                                  "negative" = "#377eb8")) +
    labs(title = "PCA of features colored by ionization mode") +
    theme_bw()
```

# Export Annotated Compounds

```{r}
#| label: export-annotated-xlsx

annot_cols <- intersect(
    c("target_lipid.name", "target_lipid_name_unique",
      "target_LIPID.CATEGORY..ABBREV.", "target_LIPID.SUBCLASS..ABBREV.",
      "target_Adduct", "target_IS_norm", "origin", "ionization_mode",
      "mzmed", "rtmed"),
    colnames(rowData(res_combined_all))
)

annotation_df <- data.frame(
    feature_id = rownames(res_combined_all),
    as.data.frame(rowData(res_combined_all)[, annot_cols, drop = FALSE]),
    check.names = FALSE,
    stringsAsFactors = FALSE
)

abundance_df <- as.data.frame(
    assay(res_combined_all, "ISnorm_filled_imputed"),
    check.names = FALSE
)

export_df <- cbind(annotation_df, abundance_df)

xlsx_path <- file.path("objects",
                       paste0(STUDY_ID, "_annotated_compounds_abundances.xlsx"))
dir.create(dirname(xlsx_path), recursive = TRUE, showWarnings = FALSE)
writexl::write_xlsx(export_df, path = xlsx_path)
cat("Saved annotated compound abundances to:", xlsx_path, "\n")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```

```{r}
toc()
```
