---
title: "Exercise Study — POS & NEG Mode Merge"
subtitle: "Combine positive and negative ionization mode results"
author: "CEMBIO-EURAC Team"
date: today
params:
  study_id: "exercise"
format:
  html:
    output-file: "exercise_POS_NEG_merge.html"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

```{r}
library("tictoc")
tic("Quarto compilation")
```

# Overview

::: {.callout-tip}
## Study-specific instance
This is a configured copy of `generic_workflow/POS_NEG_merge.qmd` for the
**exercise study**. Run after completing `Annotation_pos.qmd` and `Annotation_neg.qmd`.
:::

This document merges positive and negative mode annotation results into a
single `SummarizedExperiment`. It handles:

1. Adding ionization mode metadata
2. Aligning samples between modes
3. Resolving duplicate lipids (keeps higher abundance when RT is similar)
4. Lipid coverage visualization
5. PCA analysis of the combined dataset

# Configuration

```{r config}
#| label: configuration

# ============================================================================
# EXERCISE STUDY CONFIGURATION
# ============================================================================

# Study identifier: must match what was used in the annotation workflow
STUDY_ID <- "exercise"

# Paths to the final result objects (from Annotation.qmd)
RES_POS_PATH <- file.path("positive", "objects", paste0(STUDY_ID, "_final_res_pos"))
RES_NEG_PATH <- file.path("negative", "objects", paste0(STUDY_ID, "_final_res_neg"))

# Output path for combined result
RES_COMBINED_PATH <- file.path("objects", paste0(STUDY_ID, "_res_combined"))

# RT difference threshold for duplicate resolution (seconds)
RT_DIFF_THRESHOLD <- 30
```

# Setup

```{r}
#| label: load-packages
library(alabaster.se)
library(SummarizedExperiment)
library(UpSetR)
library(ggfortify)
library(ggplot2)
library(RColorBrewer)
library(writexl)
```

# Load Results

```{r}
#| label: load-data
res_pos <- readObject(RES_POS_PATH)
res_neg <- readObject(RES_NEG_PATH)

cat("Positive mode:", nrow(res_pos), "features x", ncol(res_pos), "samples\n")
cat("Negative mode:", nrow(res_neg), "features x", ncol(res_neg), "samples\n")
```

# Add Ionization Mode

```{r}
#| label: add-mode
rowData(res_pos)$ionization_mode <- "positive"
rowData(res_neg)$ionization_mode <- "negative"
```

# Align Samples

```{r}
#| label: align-samples
# Match samples on the polarity-independent `sample_name` column from colData,
# so this works regardless of mzML filename conventions.
pos_samples <- colData(res_pos)$sample_name
neg_samples <- colData(res_neg)$sample_name

stopifnot(
    !anyDuplicated(pos_samples),
    !anyDuplicated(neg_samples),
    setequal(pos_samples, neg_samples)
)

res_neg <- res_neg[, match(pos_samples, neg_samples)]

stopifnot(all(colData(res_pos)$sample_name == colData(res_neg)$sample_name))
```

# Resolve Duplicate Lipids

For duplicated lipid names between modes, keep the entry with higher mean
abundance when retention times are similar (within `RT_DIFF_THRESHOLD` seconds).

```{r}
#| label: find-duplicates
common_lipids <- intersect(
    rowData(res_pos)$target_lipid_name_unique,
    rowData(res_neg)$target_lipid_name_unique
)

cat("Number of common lipids:", length(common_lipids), "\n")

comparison_df <- data.frame(
    lipid_name_unique = common_lipids,
    pos_idx = match(common_lipids, rowData(res_pos)$target_lipid_name_unique),
    neg_idx = match(common_lipids, rowData(res_neg)$target_lipid_name_unique)
)

comparison_df$pos_mean_abundance <- sapply(comparison_df$pos_idx, function(i) {
    mean(assay(res_pos, "ISnorm_filled_imputed")[i, ], na.rm = TRUE)
})
comparison_df$neg_mean_abundance <- sapply(comparison_df$neg_idx, function(i) {
    mean(assay(res_neg, "ISnorm_filled_imputed")[i, ], na.rm = TRUE)
})

comparison_df$pos_rt <- rowData(res_pos)$rtmed[comparison_df$pos_idx]
comparison_df$neg_rt <- rowData(res_neg)$rtmed[comparison_df$neg_idx]
comparison_df$rt_diff <- abs(comparison_df$pos_rt - comparison_df$neg_rt)
comparison_df$rt_similar <- comparison_df$rt_diff < RT_DIFF_THRESHOLD

comparison_df$keep_mode <- ifelse(
    comparison_df$pos_mean_abundance >= comparison_df$neg_mean_abundance,
    "positive", "negative"
)

cat("\nRetention time differences summary:\n")
summary(comparison_df$rt_diff)

cat("\nLipids with dissimilar RT (kept in both modes):\n")
comparison_df[!comparison_df$rt_similar,
              c("lipid_name_unique", "pos_rt", "neg_rt", "rt_diff")]
```

```{r}
#| label: remove-duplicates
pos_remove_idx <- comparison_df$pos_idx[
    comparison_df$keep_mode == "negative" & comparison_df$rt_similar
]
neg_remove_idx <- comparison_df$neg_idx[
    comparison_df$keep_mode == "positive" & comparison_df$rt_similar
]

cat("Removing", length(pos_remove_idx), "duplicates from positive mode\n")
cat("Removing", length(neg_remove_idx), "duplicates from negative mode\n")

res_pos_filtered <- res_pos[-pos_remove_idx, ]
res_neg_filtered <- res_neg[-neg_remove_idx, ]
```

# Combine

```{r}
#| label: combine-results
colnames(res_pos_filtered) <- colData(res_pos_filtered)$sample_name
colnames(res_neg_filtered) <- colData(res_neg_filtered)$sample_name

# Drop polarity-specific colData; `file_name` differs between modes and
# can't survive the rbind below.
colData(res_pos_filtered)$file_name <- NULL
colData(res_neg_filtered)$file_name <- NULL
colData(res_pos_filtered)$spectraOrigin <- NULL
colData(res_neg_filtered)$spectraOrigin <- NULL
colData(res_pos_filtered)$polarity <- NULL
colData(res_neg_filtered)$polarity <- NULL

if (!"ionization_mode" %in% colnames(rowData(res_pos_filtered)))
    rowData(res_pos_filtered)$ionization_mode <- "positive"
if (!"ionization_mode" %in% colnames(rowData(res_neg_filtered)))
    rowData(res_neg_filtered)$ionization_mode <- "negative"

res_combined <- rbind(res_pos_filtered, res_neg_filtered)

cat("Combined dataset:", nrow(res_combined), "features x", ncol(res_combined), "samples\n")
cat("Features from positive mode:", sum(rowData(res_combined)$ionization_mode == "positive"), "\n")
cat("Features from negative mode:", sum(rowData(res_combined)$ionization_mode == "negative"), "\n")
```

# Save

```{r}
#| label: save-combined
dir.create(dirname(RES_COMBINED_PATH), recursive = TRUE, showWarnings = FALSE)
saveObject(res_combined, RES_COMBINED_PATH)
```

# Lipid Coverage

```{r}
#| label: lipid-coverage
#| fig-cap: "Lipid class distribution in the combined POS+NEG dataset"

counts_cat <- table(rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`)
bar_colors_cat <- rep(c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854"),
                      length.out = length(counts_cat))

par(mar = c(1, 1, 2, 1))
pie(counts_cat, col = bar_colors_cat, labels = counts_cat,
    border = "white",
    main = paste0("Combined Lipid Distribution (n = ", sum(counts_cat), ")"),
    cex = 0.9)
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")
legend("topright", legend = names(counts_cat), fill = bar_colors_cat,
       border = NA, bty = "n", title = "Lipid Category", cex = 0.8)
```

```{r}
#| label: lipid-coverage-subclass
#| fig-cap: "Lipid subclass distribution in the combined dataset"

counts_sub <- table(rowData(res_combined)$`target_LIPID.SUBCLASS..ABBREV.`)
sub_colors <- rep(brewer.pal(9, "Set1"), length.out = length(counts_sub))

par(mar = c(1, 1, 2, 1))
pie(counts_sub, col = sub_colors, labels = counts_sub,
    border = "white",
    main = paste0("Subclass Distribution (n = ", sum(counts_sub), ")"),
    cex = 0.9)
symbols(0, 0, circles = 0.5, inches = FALSE, add = TRUE, bg = "white")
legend("topright", legend = names(counts_sub), fill = sub_colors,
       border = NA, bty = "n", title = "Lipid Subclass", cex = 0.5)
```

```{r}
#| label: lipid-coverage-by-mode
#| fig-cap: "Lipid categories by ionization mode"

coverage_df <- as.data.frame(
  table(
    category = rowData(res_combined)$`target_LIPID.CATEGORY..ABBREV.`,
    mode = rowData(res_combined)$ionization_mode
  )
)

ggplot(coverage_df, aes(x = category, y = Freq, fill = mode)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("positive" = "#e41a1c", "negative" = "#377eb8")) +
  labs(title = "Lipid Categories by Ionization Mode",
       x = "Lipid Category", y = "Count", fill = "Mode") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Duplicate Resolution Summary

```{r}
#| label: duplicate-summary
dup_summary <- comparison_df |>
    dplyr::mutate(
        decision = dplyr::case_when(
            !rt_similar ~ "Kept both (different RT)",
            keep_mode == "positive" ~ "Kept positive (higher abundance)",
            keep_mode == "negative" ~ "Kept negative (higher abundance)"
        )
    )

cat("Duplicate resolution summary:\n")
table(dup_summary$decision)

dup_summary[, c("lipid_name_unique", "decision",
                "pos_mean_abundance", "neg_mean_abundance")]
```

# PCA Analysis

```{r}
#| label: pca-setup
res_combined$sample_type <- ifelse(
    grepl("^QC", colnames(res_combined)), "QC", "sample"
)
```

```{r}
#| label: pca-all
#| fig-cap: "PCA of all samples (POS + NEG combined)"

col_sample_type <- c("QC" = "#1b9e77", "sample" = "#d95f02")

vals <- assay(res_combined, "ISnorm_filled_imputed") |>
    as.matrix() |> t() |> log2() |> scale(center = TRUE, scale = TRUE)
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st <- as.data.frame(vals)
vals_st$sample_type <- res_combined$sample_type

autoplot(pca_res, data = vals_st, colour = "sample_type", scale = 0) +
    scale_color_manual(values = col_sample_type) +
    labs(title = "PCA - All samples (Positive + Negative combined)")
```

```{r}
#| label: pca-by-mode
#| fig-cap: "PCA of features colored by ionization mode"

is_sample <- res_combined$sample_type == "sample"
res_samples <- res_combined[, is_sample]

vals_features <- assay(res_samples, "ISnorm_filled_imputed") |>
    as.matrix() |> log2() |> t() |> scale(center = TRUE, scale = TRUE) |> t()
pca_features <- prcomp(vals_features, scale = FALSE, center = FALSE)

feature_df <- data.frame(
    PC1 = pca_features$x[, 1],
    PC2 = pca_features$x[, 2],
    ionization_mode = rowData(res_samples)$ionization_mode
)

ggplot(feature_df, aes(x = PC1, y = PC2, color = ionization_mode)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("positive" = "#e41a1c", "negative" = "#377eb8")) +
    labs(title = "PCA of features colored by ionization mode") +
    theme_bw()
```

# Export Annotated Compounds

```{r}
#| label: export-annotated-xlsx

annot_cols <- intersect(
    c("target_lipid.name", "target_lipid_name_unique",
      "target_LIPID.CATEGORY..ABBREV.", "target_LIPID.SUBCLASS..ABBREV.",
      "target_Adduct", "target_IS_norm", "ionization_mode",
      "mzmed", "rtmed"),
    colnames(rowData(res_combined))
)

annotation_df <- data.frame(
    feature_id = rownames(res_combined),
    as.data.frame(rowData(res_combined)[, annot_cols, drop = FALSE]),
    check.names = FALSE,
    stringsAsFactors = FALSE
)

abundance_df <- as.data.frame(
    assay(res_combined, "ISnorm_filled_imputed"),
    check.names = FALSE
)

export_df <- cbind(annotation_df, abundance_df)

xlsx_path <- file.path("objects",
                       paste0(STUDY_ID, "_annotated_compounds_abundances.xlsx"))
dir.create(dirname(xlsx_path), recursive = TRUE, showWarnings = FALSE)
writexl::write_xlsx(export_df, path = xlsx_path)
cat("Saved annotated compound abundances to:", xlsx_path, "\n")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```

```{r}
toc()
```
